### 2010-01-03 papers/bioinf_apps_2009/bio-app.tex
Adding papers directory, with bio. apps paper and NAR proposal

\documentclass{bioinfo}
\copyrightyear{2009}
\pubyear{2009}

\bibliographystyle{natbib}
\usepackage{subfigure}

% \usepackage{flafter,epsf}       
% \usepackage{amsmath}
% \usepackage{amsfonts,amsbsy,amssymb,bbm}
% \usepackage{amsthm}  %% proof enviroment
% \usepackage{natbib}
% %\usepackage{showkeys}

% \sloppy 
% \usepackage{graphicx}

% \bibliographystyle{mbe_bibliography}

% \input{epsf}

% \newcommand {\rel} {\mathbb{R}}
% \newcommand {\com} {\mathbb{C}}
% \newcommand {\nat} {\mathbb{N}}
% \newcommand {\rat} {\mathbb{Q}}
% \newcommand {\ganz} {\mathbb{Z}}
% \newcommand {\indicator} {\mathbbm{1}}

% \newtheorem{defi}{Definition}
% \newtheorem{theo}{Theorem}
% \newtheorem{corol}{Corollary}
% \newtheorem{prop}{Proposition}
% \newtheorem{example}{Example}
% \newtheorem{lemma}{Lemma}
% \newtheorem{problem}{Problem}[section]

% \textheight 21cm
% \textwidth 16cm
% \oddsidemargin -0.1cm
% \evensidemargin -0.1cm

\newlabel{d2_section}{A}
\newlabel{mst_section}{B}
\newlabel{sim_results}{C}
\newlabel{real_results}{D}
\newlabel{ESTsim_sup}{C.1}
\newlabel{seq_time}{S4}

\begin{document}

\firstpage{1}

\title[PEACE cluster]{PEACE: {\underline P}arallel {\underline E}ST {\underline A}nalysis
  and {\underline C}lustering {\underline E}ngine}
\author{D.M. Rao\,$^{1}$, J.C. Moler\,$^{1}$, M. Ozden\,$^1$, Y. Zhang\,$^{1}$,
  C. Liang$^{1,2*}$\, and J.E. Karro\,$^{1,3}$\footnote{to whom
    correspondence should be addressed}}
\address{$^1$ Department of Computer Science and Software Engineering, \\
  $^2$ Department of Botany, \\
  $^3$ and Department of Microbiology, Miami University, Oxford, Ohio,
  USA}

\history{Received on XXXXX}

\editor{Associate Editor: XXXXX}

\maketitle

\begin{abstract} 
  \noindent {\bf Summary: } \textsc{peace} is a software tool for the
  clustering of large sets of Expressed Sequence Tags by gene
  association, executable as a single processor tool with the
  infrastructure for distributing its computation over multiple
  processors on any platform supporting {\bf MPI}.  \textsc{peace}
  demonstrates considerably higher sensitivity, than other tools
  addressing the EST clustering problem, while exhibiting slightly
  faster performance.\\
  {\bf Availability: } Source Code and manual are available at
  www.conifergdb\\.org/PEACE.php.  Code is implemented in C++, and ready
  to run as a command line tool under Windows, Linux and OS X. \\
  {\bf Contact: } karroje@muohio.edu \\
  {\bf Supplementary Information:} A supplementary materials section
  has been submitted with this paper that contains additional details on the
  algorithm and additional results.
\end{abstract}

\section {Introduction}

Expressed Sequence Tags, or ESTs, are short DNA sequences obtained
from cDNA that represent transcripts of expressed genes.  They have
proved a powerful tool in the study of genes and gene organization, helping us to
identify genes and providing us access to the
transcriptome.  They are an inexpensive and easy way to obtain DNA-level
knowledge about any genome, but the computational cost of the EST analysis
is high [\cite{Nagaraj07}].

A critical step in the analysis of any EST data set is that of
clustering: partitioning the data into disjoint sets based on gene
association.  Clustering is generally done through overlap
identification, but the accuracy of alignments can be disrupted by EST
sequencing error.  Single base read errors are reported to occur in as
much as 5\% of bases [\cite{Nagaraj07}], while insertions, deletions and stuttering can
also be an issue; the genetic variations inherent in a set of transcriptomes 
can also disrupt overlap-based detection.  Further, the large size of
EST data sets poses its own problems; computing alignments of all
pairs is infeasible in terms of both runtime and memory usage.

Here we present \textsc{peace}, a tool for the clustering of EST sets
by gene, applicable to ESTs of length 150 bp or greater using the
default parameters.  Using a {\it minimum spanning tree} (MST) based
clustering strategy [\cite{Jain99,Wan08}] and the $d^2$ sequence
distance function [\cite{Hide94}], \textsc{peace} generates
the MSTs using a tailored version of Prim's algorithm [\cite{Prim57}].
The result is a clustering tool that is considerably more sensitive to
sequencing errors than the competing \textsc{wcd} tool
[\cite{Hazelhurst08a}] and others in the literature without any
sacrifice of performance (e.g. [\cite{Burke99}, \cite{Slater00},
\cite{Huang99}, \cite{Parkinson02}, \cite{Kalyanaraman03},
\cite{Malde03}, \cite{Ptitsyn05}, \cite{Picardi09}]).

\section {Methods}

The clustering performed by \textsc{peace} is based on the use of minimum
spanning trees (MSTs), known to be an effective approach for narrow
band single linkage clustering [\cite{Jain99,Wan08}]. We use the $d^2$
distance measure [\cite{Hide94}] to represent edge weights.  Prim's
algorithm [\cite{Prim57}] then allows us to efficiently calculate an
MST from which we can infer a high-quality clustering solution.

The $d^2$ distance measure used to assign edge weights is an
alignment-free measurement of sequence distance calculated
significantly faster than a alignment \cite{Hide94,Foret09}.  $d^2$
works by comparing the frequency of words (strings of a fixed length)
appearing in a limited region of each string.  ESTs overlapping by a
sufficient length will share neighborhoods of enough similarity to
ensure a small distance even in the presence of a moderate number of
base errors.  Our method of calculating $d^2$, the {\it two-pass $d^2$
  heuristic}, is a refinement of that used in the \textsc{wcd}
clustering tool [\cite{Hazelhurst08a}].  We augment the computation
with a preliminary pass that narrows down the two regions of maximum
similarity, and then explore those local regions in more detail.

We can model the EST input as a weighted, undirected graph: the ESTs
are represented as nodes, with $d^2$ sequence distances assigned to
the connecting edges as weights.  Conceptually, we want to remove each
edge exceeding a threshold score from the complete graph, and define
our partitions by the remaining connected components of the graph.  An
edge with a large weight connects ESTs which are likely unrelated;
once such edges are removed, the components define a series of
overlaps.  Hence those ESTs that can still be connected by some path
correspond to the same gene.  However, such an approach requires the
calculation of all edge weights, a task infeasible both in terms of
runtime and memory usage for the data set sizes we expect to process.

We approach the problem by generating a minimum spanning tree of the
described graph, then removing edges exceeding our threshold.  By
using Prim's algorithm we are able to calculate edge weights
on-the-fly (reducing memory requirements), and we can skip the
calculation of a majority of edge distances using the $u/v$ and $t/v$
filtering heuristics employed in \textsc{wcd} [\cite{Hazelhurst08a}].
These heuristics allow us to quickly dismiss many of the edges as too
large without the need to apply the full $d^2$ algorithm (see
Sections~\ref{d2_section} and \ref{mst_section} of the Supplementary
Materials for more details).

\section{Results}

We have tested \textsc{peace} on both simulated and real data,
comparing results against those produced by the \textsc{wcd}
clustering tool [\cite{Hazelhurst08a}] and the \textsc{cap3} assembly
tool [\cite{Huang99}] (which calculates a cluster as an intermediate
step).  For our simulation tests we used the {\bf ESTSim} tool
[\cite{Hazelhurst03}] to generate simulated ESTs under varying models
of error (Supplementary Materials, Section~\ref{ESTsim_sup}),
generating the fragments from the list of 100 zebra fish genes used in
the \textsc{wcd} testing [\cite{Hazelhurst08a}].  Tool parameters were
taken to match, as closely as possible, those used in the \textsc{wcd}
study (see supplementary materials).  Our primary methods of quality
assessment were {\it sensitivity} (the fraction of EST pairs from the
same gene that were correctly clustered together) and {\it Type 1
  error} (the fraction of genes that were divided between clusters)
[\cite{Wang04,Hazelhurst08a}]; other measurements are also discussed
in the supplementary materials.  In Figure~1 (left) we see a
significant improvement of \textsc{peace} (blue) over \textsc{wcd}
(green) and \textsc{cap3} (black) in its sensitivity to sequencing
errors, while in Figure~1 (right) we see a comparable
improvement in Type 1 error.  In Figure~2 we look at
the number of singleton clusters (ESTs not joined to any cluster),
which will not occur in our simulated sets; we again see significant
improvement in \textsc{peace}.  The sequential runtime of
\textsc{peace} is slightly improved over that of \textsc{wcd} (see
Supplementary Materials, Figure~S4).

\begin{figure*}[b]
\centerline{
\includegraphics[scale=0.5]{SeT1.eps} 
}
\label{SeT1}
\caption{Comparisons of Sensitivity and Type 1 error, based on the
  average over 30 Simulated EST sets derived from 100 zebra fish genes
  (see Supplementary Materials, Section~\ref{sim_results}, for more
  details).   Blue/Solid = \textsc{peace}, Green/Dash =
  \textsc{wcd}, Black/Dot-Dash = \textsc{cap3}; vertical tics = 95\% confidence
  intervals on estimates.  Intervals are not presented for Type 1
  error due to the extremely small variance.}
\end{figure*}

\begin{figure*}[b]
\centerline{
\includegraphics[scale=0.35]{singletons.eps}
}
\label{singletons}
\caption{Average number of ESTs flagged as singletons by each tool
  when run on the simulated sequences; correct answer in all cases is
  zero.  Blue/Solid = \textsc{peace}, Green/Dash = \textsc{wcd},
  Black/Dot-Dash = \textsc{cap3}.}
\end{figure*}

In applying the tools to real data, we started with a set of
approximately $190,000$ ESTs derived from the {\it Chlamydomonas
  reinhardtii} genome.  To compute sensitivity we used the {\it gmap}
tool [\cite{Wu05}] to map the EST set to the genome, taking this as
our reference clustering.  We see slight improvements in
\textsc{peace} over \textsc{wcd} in both sensitivity and Type 1 error
(both significantly out preforming \textsc{cap3}).  Using the mouse EST
set used in Hazelhurst {\it et al.}  [\cite{Hazelhurst08a}]
\textsc{peace} again shows slight improvements in sensitivity and Type
1 error rates, with an 18\% speed up for \textsc{peace} (see
Supplementary Materials, Section~\ref{real_results}, for more
details).

\section{Acknowledgements}

Dr. Karro was funded under a PhRMA Foundation Informatics Research
Starters Grant while conducting this research.  We would also like to
acknowledge Iddo Friedberg, David Woods, Jens Mueller and David
Scoville at Miami University for their help with this project.

\vspace{3mm}
%Bib TeX

\bibliography{peace.bib}

\end{document}

% LocalWords:  arallel nalysis lustering ngine Rao Moler Ozden Zhang Liang ESTs
% LocalWords:  Karro XXXXX transcriptome pre isoforms SNPs unsequenced runtime
% LocalWords:  chlamydomondomonas ESTate PaCE xsact CLU WCD Hazelhurst et al th
% LocalWords:  CHUN nalyzer Prim's MSTs subgraph MPI SourceCode ESTSim runtimes
% LocalWords:  ests Chlamydomonas Reihartdii gmap CDNA transcriptomes cDNA www
% LocalWords:  reinhartdii conifergdb DNAc bp wcd reinhardtii PhRMA Iddo
% LocalWords:  Friedberg Scoville

### 2010-01-03 peace/papers/bioinf_apps_2009/bio-app.sup.tex
Adding papers directory, with bio. apps paper and NAR proposal

\documentclass[a4paper,12pt]{article}
\usepackage{flafter,epsf}       
\usepackage{amsmath}
\usepackage{amsfonts,amsbsy,amssymb,bbm}
\usepackage{amsthm}  %% proof enviroment
\usepackage{natbibOrig}

\sloppy 
\usepackage{graphicx}
\bibliographystyle{natbib}

% Using the following to mark changes for review
\usepackage{color}
\definecolor{blue}{rgb}{0,0,1}
\newcommand{\mc}[1]{\textcolor{blue}{#1}}
%\newcommand{\mc}[1]{#1}   % Replace above line with this to eliminate all coloring

\input{epsf}

\newcommand {\rel} {\mathbb{R}}
\newcommand {\com} {\mathbb{C}}
\newcommand {\nat} {\mathbb{N}}
\newcommand {\rat} {\mathbb{Q}}
\newcommand {\ganz} {\mathbb{Z}}
\newcommand {\indicator} {\mathbbm{1}}

\newtheorem{defi}{Definition}
\newtheorem{theo}{Theorem}
\newtheorem{corol}{Corollary}
\newtheorem{prop}{Proposition}
\newtheorem{example}{Example}
\newtheorem{lemma}{Lemma}
\newtheorem{problem}{Problem}[section]

\textheight 21cm
\textwidth 16cm
\oddsidemargin -0.1cm
\evensidemargin -0.1cm

\newlabel{sensitivityFig}{1}

\begin{document}

\title{PEACE: {\underline P}arallel {\underline E}ST {\underline A}nalysis
  and {\underline C}lustering {\underline E}ngine \\Supplementary
  Materials}

\date{}

\maketitle

\begin{appendix}

\renewcommand{\theequation}{S\arabic{equation}}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{equation}{0}

\section{Distance metrics, heuristics and filters}

\subsection{The $d^2$ distance metric}
$d^2$, as outlined by {\it Hide et al.} [\cite{Hide94}], is an {\it
  alignment free} distance pseudo-metric which quantifies local
similarity between sequences based on a simple word count.  Let
$c_x(w)$ denote the number of times word $w$ occurs in string $x$.  We
search for similarity between strings $x$ and $y$ by looking at the
difference between $c_x(w)$ and $c_y(w)$ for different words $w$.
Specifically,  for all words of a fixed length $k$, we calculate:
\begin{equation*}
d^2_k(x,y) = \sum_{|w|=k}(c_x(w) - c_y(w))^2
\end{equation*}
However, applying such a definition to two sequences as a whole leads
to a measure of global similarity, while we want to measure local
similarity (thus, for example, assigning two sequences with
sufficiently long overlapping ends to be at a distance of zero -- or a
small distance if errors are present).  For this, instead of comparing
the entire two strings, we compare sliding windows from each string of a fixed
size $r$.  Formally, for sequences $x$ and $y$ ($|x| \geq r$, $|y|
\geq r$), we define:
\begin{equation}
\label{d2def}
d^2(x,y) = 
\min\left\{d^2(u,v) \; : \; u \sqsubseteq x, v \sqsubseteq y, |u|=|v|=r\right\}
\end{equation}
(where $u \sqsubseteq x$ denotes that $u$ is a substring of $x$).
Defined as such, $d^2$ is, in a mathematical sense, a {\it
  pseudo-metric}: $d^2(x,y) = 0$ does not imply $x=y$.

The \textsc{peace} implementation of $d^2$ was initially based on the description from
{\it Hazelhurst} [\cite{Hazelhurst04}].  For parameters
we adapted those used by the \textsc{wcd} clustering tool [\cite{Hazelhurst08a}]:
a word size $k=6$ and a window size $r=100$.

\subsection{Two-pass $d^2$}
Our ``two-pass $d^2$'' algorithm works by sampling a subset of window
pairs evenly distributed across the sequences, narrowing down a
smaller region in which to search for the best scoring window-pair.
In the first pass, the algorithm look at every length $r$ window on one sequence,
but on the other sequence we skip the window by $s$ bases between
every sampling.  It then take the best such window pair, extend
each of these two windows by $s$ bases in each direction, and apply
(\ref{d2def}) to this limited region.

In \textsc{peace}, we use $k=6$, $r=100$, and $s=50$.

\subsection{Filtering heuristics}
\textsc{peace} uses the $u/v$ and $t/v$ heuristics, roughly as described by {\it
  Hazelhurst et al.} [\cite{Hazelhurst08a}] as filters that allows us to avoid
over 99\% of the potential $d^2$ calculations.  Each heuristic takes two
sequences and estimates whether it is worth proceeding to the $d^2$
computation by sampling and comparing word frequency across the two
sequences.  The $u/v$ heuristic looks at every $h$-th word of size $v$
on one sequence, and rejects if it does not find at least $u$ occurrences
of these words on the other sequence.  The $t/v$ heuristic demands
that there be at least $t$ size $v$ words on one sequence that occur
within a $l$ base range within the second sequence.  If a sequence
pair meets the requirements of both these filters, then we compute the
$d^2$ distance.

In \textsc{peace}, we set $h=16$, $v=8$, $u=4$, $t=65$, and $l=100$.

\section{MST-Based  Calculations}

While $d^2$ serves to quantify sequence distance, the basis for the
clustering algorithm is the minimum spanning tree (MST).  Viewing the
ESTs as nodes and the data set as an $d^2$-weighted graph, the
derivation of a minimum spanning tree results in the placement of
nodes of a given tree into a restricted neighborhood of the graph.  By
then removing larger edges, we are left with connected components
corresponding to the gene-based clusters.  The MST-based clustering
method has been used effectively in other applications, but to our
knowledge this is the first such use for the  EST
clustering problem [\cite{Jain99,Wan08}].

\subsection{Calculation of the MST}

To calculate this MST we use Prim's algorithm [\cite{Prim57}].  For a
graph of $n$ nodes and $e$ edges, Prim's can be implemented such that
the algorithm has an $O(e + n \log n)$ worst-case runtime.  However,
the {\it narrow-band} nature of the model allows us to reduce this
bound.  If we were to model only the edges connecting adjacent nodes,
we would find the node degrees to be very small relative to $n$: any
EST overlaps only a few others, and has no connection to a vast
majority of the data set members.  Since we can quickly eliminate most
of these excess edges with the $u/v$ and $t/v$ heuristics (removing
more than 99\% of all edges before applying Prim's), we find in
practice that $e$ is a very small fraction of $n^2$.  In the
runtime results (see Section~\ref{rt_section} and
Figure~\ref{seq_runtime}), we find that when holding the EST size
distribution constant, the tool as a whole has a runtime of $O(n^2)$
-- appearing to dominated by the time required to apply
the filtering heuristics to every EST pair.

\subsection{Removing edges}

Once the MST is has been calculated, our last step is to remove all
edges exceeding a threshold weight $T$, taking the resulting
components as our clusters.  \textsc{wcd}, when faced with a similar
challenge, sets the threshold at $T=40$, hypothesizing that EST pairs
with a $d^2$ distance of greater than 40 are unlikely to overlap.  We
have set out threshold to $T=130$.

\section{Simulated Test Results}

In {\it Hazelhurst et al.} [\cite{Hazelhurst08a}] the authors conduct
an investigation of \textsc{wcd} against a number of clustering
tools.  As they make a convincing argument that \textsc{wcd} is returning
better results then the tools against which they compare, we limit our
analysis to a comparison of \textsc{peace} against \textsc{wcd} and the \textsc{cap3} assembly
tool [\cite{Huang99}] -- a tool which implicitly clusters while
performing assembly.  We do so by applying all three tools to a number
of data sets, both simulated and real.

\subsection{Simulation Tool and Parameters}

For simulated data sets we use the {\bf ESTsim} tool to generate
simulated EST data sets [\cite{Hazelhurst03}], using the collection of
zebra fish genes that served as a basis for the \textsc{wcd} simulations
[\cite{Hazelhurst08a}].  In generating the ESTs, ESTsim models three
types of error (general base read errors, errors due to
polymerase decay, and primer interference), and allows those errors to
take the form of substitutions, deletions, and insertions (of bases
and Ns).  See the paper for a discussion of the probability
distributions and default parameters.  For generating our simulated
data, we use the default values for all parameters that are not
explicitly being subjected to variation in our experiments --
paralleling the testing of the \textsc{wcd} tool.

\subsection{Methodology}

Each estimate given in the main paper, or in the following, is
averaged over 30 trials.  Each trial consists of the application of
all three tools to a simulated data set, the set having been derived
from the application of the ESTsim tool to a set of 100 zebra fish
gene sequences [\cite{Hazelhurst03}].  All confidence intervals are
calculated at a 95\% level of significance.

\subsection{Result Quality}

While our primary measurement of tool result is sensitivity
(Figure~\ref{sensitivityFig}), a number of alternative measurements
were investigated as well. The {\it Jaccard} index measures both
sensitivity and specificity, calculated as the ratio of true positives
to the sum of true positives, true negatives, and false positives ($tp
/ (tp+fn+fp)$) [\cite{Hazelhurst08a}].  Notably, when we removed duplicated genes before
generating the simulated EST sets, we found all three tools to have
100\% specificity: no false positives.  In such a case the Jaccard
Index is identical to the Sensitivity Index, hence \textsc{peace} exhibits the
same improvement on both indexes relative to \textsc{wcd} and \textsc{cap3}.  Applying
the tools to the original genes (in which 21\% of the genes were
highly similar to genes in the remaining 79\%), we find only a slight
difference in the Jaccard Index, and the same relative performance of
the three tools (Figure~\ref{JiSe}).  The sensitivity of each tool to
duplications is discussed below.

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tbp]
\centerline{
\includegraphics[scale=0.5]{4JiSe.eps}
\label{JiSe}
}
\caption{Comparisons of Sensitivity and the Jaccard Index as a
  function of base read error rate, based on 30 simulated EST sets
  derived from a set of 79 genes with no significant similarity (left
  column), and an additional 21 genes showing significant similarity
  to the base set (right column).  Blue/Solid = \textsc{peace}, Green/Dash =
  \textsc{wcd}, Black/Dot-Dash = \textsc{cap3}; vertical tics = 95\% confidence
  intervals on estimates.}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%

Type 1 and Type 2 errors, as defined in {\it Wang et al.}
[\cite{Wang04}], measure quality at the level of the genes from which
the ESTs were derived.  For Type 1 error, we look at the fraction of genes
that were  broken into two or more partitions, while in
Type 2 error we look at the fraction of clusters that contain two or
more genes.   In Figure~\ref{t1t2} we compare the Type 1 and Type 2 error rates of
the three tools for varying levels of simulated base read error.  For
Type 1 error we find that while all tools do equally poorly when the
EST is subjected to more than a $5\%$ error rate, \textsc{peace} does
significantly better than \textsc{wcd} in the lower ranges (while
\textsc{wcd} does correspondingly better than \textsc{cap3}).  When
looking at Type 2 error rate, we find that \textsc{peace} does a
generally better job than \textsc{cap3} in separating genes into
different clusters.  We note, however, that Type 1 error is the more
important measurement: genes incorrectly split between clusters cannot
be later recovered (without re-examining the cluster as a whole), but
there is the potential for an assembly program to later identify
incorrectly merged clusters based only on local cluster information.

%%%%%%%%%%%%%%
\begin{figure}[tbp]
\centerline{\includegraphics[scale=0.5]{t1t2.eps}
\label{t1t2}
}
\caption{Type 1 error rate (fraction of genes partitioned) and Type 2
  error rate (fraction of clusters containing multiple genes) as a
  function of base read error rate.  Estimates averaged over
  application to 30 simulated ESTs sets derived from the full 100
  zebra fish gene set. Blue/Solid = \textsc{peace}, Green/Dash =
  \textsc{wcd}, Black/Dot-Dash = \textsc{cap3}; vertical tics = 95\%
  confidence intervals on estimates (not shown for Type 1 error due to
  the extremely small variance in estimates).}
\end{figure}
%%%%%%%%%%%%%%

One of the difficulties in clustering EST data is dealing with
highly similar genes.  Genes with a high degree of similarity will
produce ESTs that reflect that similarity, hence appear to overlap --
resulting in the incorrect clustering of ESTs from separate genes.  Assembly tools
such as \textsc{cap3} may have more ability to discriminate between clusters
given their more intensive investigation of overlaps, but highly
similar sequences are going to cause a problem for any tool.

In Figure~\ref{dups} we look at the ability of each tool to separate
duplicates as a function of the \% divergence between the
duplications.  Unsurprisingly, \textsc{cap3} (the assembler) does the best
here, able to effectively separate duplicates at $~92\%$ similarity
or less.  \textsc{peace} and \textsc{wcd} are roughly comparable, both clearly able to
separate duplicates out at a similarity level of $83\%$ -- but completely 
unable to distinguish sharing a similarity of $~88\%$ or more.

\begin{figure}[tbp]
\centerline{
\label{dups}
\includegraphics[scale=0.35]{duplicates.eps}
}
\caption{Ability to distinguish duplicates as a function of
  divergence.  Estimates averaged over 30 trials; one trial consists
  of taking a random gene, copying it and stochastically changing bases
  at the specified rate, then using the two genes as the bases for
  generating a simulated set. Blue/Solid = \textsc{peace}, Green/Dashed = \textsc{wcd}, Black/Dot-Dashed = \textsc{cap3};
  variance was too small for visible plotting of
  confidence intervals.}
\end{figure}

\subsection{Runtime}
\label{rt_section}

In Figure~\ref{seq_runtime} we see give a runtime comparison of \textsc{peace} and \textsc{wcd},
seeing a slight improvement in \textsc{peace}.  (As \textsc{cap3} performs both
clustering and assembly, runtime comparisons are not meaningful.)  As
stated, we see an almost perfectly quadratic curve ($r > 0.99$),
indicating that the runtime is dominated by the time required to look
at every EST pair (either filtering or computing $d^2$).
As both \textsc{peace} and \textsc{wcd} can be run in a parallel mode, we also
examined runtime as a function of the ratio of EST set sizes to number
of processors used.  In Figure~\ref{par_runtime} we show two such ratios,
again seeing an improvement in \textsc{peace} over \textsc{wcd}.

%%%%%%%%%%%%%
\begin{figure}[tbp]
\centerline{
\includegraphics[scale=0.35]{seq_time.eps}
\label{seq_runtime}
}
\caption{Single Processor Runtime as a function of input size,
  averaged over 50 simulated EST sets; Blue/Solid = \textsc{peace},
  Green/Dashed = \textsc{wcd}; vertical tics represent= 95\% confidence
  intervals on estimates.}
\end{figure}
\begin{figure}[tbp]
\centerline{
\includegraphics[scale=0.5]{par_time.eps}
\label{par_runtime}
}
\caption{Runtime of a function of input size when holding constant the
  ratio of set size to number of processors used; averaged over 30
  simulated EST sets.  Blue/Solid = \textsc{peace}, Green/Dashed = \textsc{wcd};
  vertical tics represent= 95\% confidence intervals on estimates.}
\end{figure}
%%%%%%%%%%%%%

\section{Applications to Real Data Sets}

In Table~\ref{chlamy} we lay out more comprehensive results for the
application of our tool to an EST set collected from the {\it
  Chlamydomonas reinhartdii} genome, while in Table~\ref{mouse} 
we see the results for the mouse set used for the \textsc{wcd} analysis
[\cite{Hazelhurst08a}].  We note that for the Chlamydomonas genome the
correct clustering is not known; our estimates of sensitivity are
based on as assignment of ESTs to the sequenced genomes by the Gmap tool
[\cite{Wu05}], leaving some potential for error in the  reference.
Further, we found a number of ligations in the data (ESTs derived by
incorrectly combining multiple transcripts),  which were removed before testing.

Interestingly, where in the simulated results we saw significant
improvement in \textsc{peace} over \textsc{wcd} in terms of result quality and only a
light improvement in runtime, here we see the reverse. The two tools are
roughly comparable  in terms of sensitivity, Type 1 error and Type 2
error, while \textsc{peace} shows a 10\% and 20\% improvement in runtime (both
showing significant improvements over \textsc{cap3}).  Predictably , the full
assembly tool does somewhat better in terms of specificity (the Jaccard
Index and Type 2 error), though scores here are worse for all three
tools due to the increased number of homologous gene pairs as compared
to our simulated sets.

\begin{table}[tbp]
\begin{center}
\begin{tabular} {| c || c | c | c | c | c | c | c | c |}
\hline
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} & \begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & Singletons & \begin{tabular}{c} Single
      \\ processor \\ runtime (s) \end{tabular}\\
\hline \hline \textsc{peace} & 0.958 & 0.386 & 0.623 & 0.026 & 19649 & 10493 & 8799 \\
\hline \textsc{wcd} & 0.94 & 0.500 & 0.654 & 0.021 & 22433 & 13000 & 9913 \\
\hline \textsc{cap3} & 0.77 & 0.740 & 0.737 & 0.015 & 33771 & 22169 &  \\
\hline
\end{tabular}
\label{chlamy}
\end{center}
\caption{{\bf Chlamydomonas reinhartdii}: 189975 ESTs, average length
  553 bp, estimated 9886 actual genes represented.}
\end{table}

\begin{table}[tbp]
\begin{center}
\begin{tabular} {| c || c | c | c | c | c | c | c | c |}
\hline
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} & \begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & Singletons & \begin{tabular}{c} Single
      \\ processor \\ runtime (s) \end{tabular}\\
\hline \hline \textsc{peace} & 0.936 & 0.348 & 0.340 & 0.034 & 18338 & 8206 & 820\\
\hline \textsc{wcd} & 0.932 & 0.476 & 0.350 & 0.0275 & 18787 & 8554 & 1000\\
\hline \textsc{cap3} & 0.830 & 0.805 & 0.486 & 0.0135 & 25040 & 14925 & \\
\hline
\end{tabular}
\caption{{\bf A076951 (Mouse EST set)}: 76914 ESTs, average length 427,
  13240 actual genes represented.}
\label{mouse}
\end{center}
\end{table}

%Bib TeX
\bibliography{peace.bib}

\end{appendix}

\end{document}

% LocalWords:  overline TpG li lj lk ijk i'j'k jj kk trinucleotides rb Arndt et
% LocalWords:  transversion Arndt's dinucleotides transversions RepBase arallel
% LocalWords:  nalysis lustering ngine al substring Hazelhurst \textsc{wcd}  ESTsim ESTs
% LocalWords:  polymearse Methodoloy Jaccard tp fn fp  wcd th Prim's Ns
% LocalWords:  Chlamydomonas reinhartdii

### 2010-02-09 papers/NAR_web_server_2009/paper.tex
Source for NAR paper

\documentclass[a4,center,fleqn]{NAR}
%\documentclass{article}
\bibliographystyle{natbib}
%\usepackage{graphicx}

\copyrightyear{2010}
\pubdate{X}
\pubyear{X}
\jvolume{X}
\jissue{X}

\newlabel{d2_section}{A}
\newlabel{mst_section}{B}
\newlabel{sim_results}{C}
\newlabel{real_results}{D}
\newlabel{ESTsim_sup}{C.1}
\newlabel{seq_time}{S4}

\newcommand{\wcd}{\textsc{wcd }}
\newcommand{\capthree}{\textsc{cap3 }}

\begin{document}

\firstpage{1}

\title[PEACE cluster]{PEACE: {\underline P}arallel {\underline E}ST {\underline A}nalysis
  and {\underline C}lustering {\underline E}ngine}
\title{PEACE: {\underline P}arallel {\underline E}ST {\underline A}nalysis
  and {\underline C}lustering {\underline E}ngine}
\author{D.M. Rao\,$^{1}$, J.C. Moler\,$^{1}$, M. Ozden\,$^1$, Y. Zhang\,$^{1}$,
  C. Liang$^{1,2*}$\, and J.E. Karro\,$^{1,3}$\footnote{to whom
    correspondence should be addressed}}
\address{$^1$ Department of Computer Science and Software Engineering, \\
  $^2$ Department of Botany, \\
  $^3$ and Department of Microbiology, Miami University, Oxford, Ohio,
  USA}

%\history{Received on XXXXX}

\maketitle

\begin{abstract} 
  We present {\sc Peace}, a tool for the high-throughput clustering of
  transcript fragment sequences produced Next Generation or Sanger
  sequence technologies.  Installed and easily run through an easily
  downloaded GUI, {\sc Peace} can process large data sets, group the
  fragments by gene association, faster and with greater sensitivity
  than competing clustering tools such as {\sc WCD}, and with
  considerably better sensitivity then the implicit clustering
  preformed by assembly tools such as {\sc Cap3}.  Through the GUI the
  user can collect cluster statistics and easily examine specific
  clusters for more comprehensive study.  And by using {\sc Peace} as
  a ``pre-assembly'' tool to allow the application of an assembly tool
  to the individual clusters, the user gains both the significantly
  increase sensitivity of the clustering method and a large decrease
  in overall runtime for the clustering and assembly as compared to
  the application of an assembly tool to the data as a whole.
\end{abstract}

\section {Introduction}

Understanding an organism's transcriptome, the set of (spliced)
transcripts expressed by genes of the organism, is a vital step in
understanding the full functional and organization roll of the genome
in the life cycle of any organism.  Studying the transcriptome has led
to gene discovery, provided information of protein isoforms, and
generally help shed new light on the biological processes both
controlling and dictated by the genome \cite{Nagaraj07}.  But in order
to get access those transcripts, we must first deal with the
fragmented data produced by both Next Generation and traditional
Sanger sequencing technology.

Given a set of transcript fragments sampled from multiple transcripts
across the genome, a necessary first step is that of clustering:
separating the fragments according to the transcript from which they
were derived.  While clustering is frequently preformes implicity by
assembly tools such as {\sc Cap3} \cite{Huang99}, clustering the data as a
``pre-assembly'' step has a number of advantages.  Most significant
among these: preforming this step will allow the application of the
assembly tool to individucal clusters -- saving significant amounts of
time due to the reduced input size \cite{Hazelhurst08a}.  But clustering is
a computationally challenging problem, especially when done {\it ab
  initio} instead of relying on access to a sequenced genome.  Even
with the number of fragments produced using Sanger sequencing, the
runtime and memory requirements to cluster based on pair-wise sequence
alignments make such an approach infeasible.  The much larger data-set
sizes produced by the short-read technologies exacerbates this
problem.

Here we present {\sc Peace}, a tool for the {\it ab initio} clustering
of transcript fragments by gene, applicable to both short-read and traditional
Sanger Sequencing technologies.  Available through the
www.peace-tools.org website, the PEACE GUI allows the user
to both easily install (locally or remotely) and run the computational
tool, as well as providing various tools for result analysis. {\sc
  Peace} can be run in single or multi-processor mode, and requires no
Internet connectivity past the initial download.  Using a {\it minimum
  spanning tree} (MST) based clustering strategy \cite{Jain99,Wan08}
and the $d^2$ sequence distance function \cite{Hide94}, {\sc Peace}
generates the MSTs using a tailored version of Prim's algorithm
\cite{Prim57}.  The result is an easy-to-use clustering tool that is considerably
more sensitive to sequencing errors than the competing \wcd tool
\cite{Hazelhurst08a} and others in the literature without any
sacrifice of performance (CITE) (e.g. \cite{Burke99}, \cite{Slater00},
\cite{Huang99}, \cite{Parkinson02}, \cite{Kalyanaraman03},
\cite{Malde03}, \cite{Ptitsyn05}, \cite{Picardi09}).

\section{PEACE: Instillation and Use}

The {\sc Peace} GUI, available from the {\sc Peace} website
(www.peace-tools.org), handles the details of both instillation and
job management and provides tools for the initial analysis of the
results.  By downloading and executing the {\tt peace.jar} GUI Java
program on a Linux, Windows or OS-X machine running Java 1.6+, the
user is presented with the {\sc Peace} management tool.  Through this
the user can install the computational tool and preform a clustering
of a data file, view an initial analysis of the clusters and produce
files containing subsets of the clusters for assembly by a tool such
as {\sc Cap3} \cite{Huang99}.  A typical (first) use of the tool would
generally require the the following steps.

\noindent {\bf Tool Instillation:} Installing the GUI requires only
that the Java file be downloaded from the website.  To install the {\sc
  Peace} tool onto a local or remote machine, the user will select
from within the GUI the {\tt Server > Add New Server} tab from the GUI
main menu (Figure~1(a)), which then starts the install
wizard that will prompt for the appropriate information.
Figure~1(b) illustrates the request for server information;
in this example, the user has chosen to install the {\sc Peace}
computational tool on a remote machine and is providing the necessary
connection information.  Server information is persistent between GUI
sessions, giving the user access to {\sc Peace} on the target machine
as needed.

\noindent {\bf Job Processing:} After importing the target file into
the GUI ({\tt File > New Data Set}), the user can start a new job with
the {\tt Job > Job to Complete Clusters} tab and follow the wizard
menus.  Figure~1(c) illustrates the process of specifying
the number of processors available (if running on a machine supporting
the OpenMPI protocol -- which will be determined during job
instillation).  Once executed, the GUI will manage the job thread,
alert the user when the job is completed (or when the user next
runs the GUI after completion) and 
copy the final results back to the local machine if necessary.

\noindent {\bf Result Analysis:} Once the resulting clusters have been
computed and (if necessary) copied back, the user has several options
for analysis:
\begin{itemize}
\item {\bf Export:} The user can export the contents of one or more
  clusters into a fasta file, obtaining a subset of the original file
  containing the sequences corresponding to the selected clusters
  ready for processing by an assembly tool (e.g. {\sc Cap3} \cite{Huang99}).
\item {\bf View Clustering}: The user may view a list of all clusters,
  expanding selected clusters to a list of all fragments (illustrated in
  Figure~1(d)).  
\item {\bf Classified Summary Graph:} The user may view a distribution
  of cluster sizes in a linear or $\log$ scale.  Further, the user may
  set up a {\it classifier}, associating certain patterns with
  specific colors.  These patterns will be matched against the fragment
  header information from the original fasta file, allowing the overlay of
  a colored cluster size distribution does clusters containing matching elements in a
  specified colors (e.g. if the user wished to restrict their view to
  clusters containing transcript fragments from a specific source).  The method of
  setting up a these classifiers, and the resulting histogram, is
  illustrated in Figure~1(e).
\end{itemize}

Extensive documentation for the tool has been posted on the {\sc
  Peace} website, as well as links to several videos demonstrating
{\sc Peace} use and capabilities.

\section{Methods}

The clustering performed by {\sc Peace} is based on the use of minimum
spanning trees (MSTs), known to be an effective approach for narrow
band single linkage clustering \cite{Jain99,Wan08}. We use the $d^2$
distance measure \cite{Hide94} to represent edge weights.  Prim's
algorithm \cite{Prim57} then allows us to efficiently calculate an
MST from which we can infer a high-quality clustering solution.

The $d^2$ distance measure used to assign edge weights is an
alignment-free measurement of sequence distance that can be calculated
significantly faster than a Smith-Waterman alignment
\cite{Hide94,Foret09}.  $d^2$ works by comparing the frequency of
words (strings of a fixed length) appearing in a limited region of
each string.  Fragments overlapping by a sufficient length will share
neighborhoods of enough similarity to ensure a small distance even in
the presence of a moderate number of base errors.  Our method of
calculating $d^2$, the {\it two-pass $d^2$ heuristic}, is a refinement
of that used in the \wcd clustering tool \cite{Hazelhurst08a}.  We
augment the computation with a preliminary pass that narrows down the
two regions of maximum similarity, and then explore those local
regions in more detail.

We can model the fragment input as a weighted, undirected graph: the fragments
are represented as nodes, with $d^2$ sequence distances assigned to
the connecting edges as weights.  Conceptually, we want to remove each
edge exceeding a threshold score from the complete graph, and define
our partitions by the remaining connected components of the graph.  An
edge with a large weight connects fragments which are likely unrelated;
once such edges are removed, the components define a series of
overlaps.  Hence those fragments that can still be connected by some path
correspond to the same gene.  However, such an approach requires the
calculation of all edge weights.  Such a task is infeasible both in terms of
runtime and memory usage for the data set sizes we expect to process.

We approach the problem by generating a minimum spanning tree of the
described graph, then removing edges exceeding our threshold.  By
using Prim's algorithm we are able to calculate edge weights
on-the-fly (reducing memory requirements), and we can skip the
calculation of a majority of edge distances using the $u/v$ and $t/v$
filtering heuristics employed in \wcd \cite{Hazelhurst08a}.
These heuristics allow us to quickly dismiss many of the edges as too
large without the need to apply the full $d^2$ algorithm (see
Sections~\ref{d2_section} and \ref{mst_section} of the Supplementary
Materials for more details).

\section{Results}

We have tested {\sc Peace} on both simulated and real data,
comparing results against those produced by the \wcd
clustering tool \cite{Hazelhurst08a} and the \capthree assembly
tool \cite{Huang99} (which calculates a cluster as an intermediate
step).  For our simulation tests we used the {\bf ESTSim} tool
\cite{Hazelhurst03} to generate simulated transcript fragments under varying models
of error (Supplementary Materials, Section~\ref{ESTsim_sup}),
generating the fragments from the list of 100 zebra fish genes used in
the \wcd testing \cite{Hazelhurst08a}.  Tool parameters were
taken to match, as closely as possible, those used in the \wcd
study (see supplementary materials).  Our primary methods of quality
assessment were {\it sensitivity} (the fraction of fragment pairs from the
same gene that were correctly clustered together) and {\it Type 1
  error} (the fraction of genes that were divided between clusters)
\cite{Wang04,Hazelhurst08a}; other measurements are also discussed
in the supplementary materials.  In Figure~2(left) we see a
significant improvement of {\sc Peace} (blue) over \wcd
(green) and \capthree (black) in its sensitivity to sequencing
errors, while in Figure~2(right) we see a comparable
improvement in Type 1 error.  In Figure~3 we look at
the number of singleton clusters (fragments not joined to any cluster),
which will not occur in our simulated sets; we again see significant
improvement in {\sc Peace}.  The sequential runtime of
{\sc Peace} is slightly improved over that of \wcd (see
Supplementary Materials, Figure~S4).

\begin{figure*}[b]
\centerline{
\includegraphics[trim=0cm 3cm 0cm 1cm, clip, scale=0.9]{screen.d/screen.pdf} 
}
\label{screen}
\caption{Screenshots of the PEACE GUI during execution, including (a)
GUI Welcome and server instillation menu; (b) setup wizard for
installing the computational tool on a remote server; (c) execution
wizard for starting a selected job to be executed in parallel mode;
(d) basic cluster output; and (e) histogram view of cluster results
and classifier editor.}
\end{figure*}

\begin{figure*}[b]
\centerline{
\includegraphics[scale=0.5]{pics.d/SeT1.pdf} 
}
\label{SeT1}
\caption{Comparisons of Sensitivity and Type 1 error, based on the
  average over 30 Simulated fragment sets derived from 100 zebra fish genes
  (see Supplementary Materials, Section~\ref{sim_results}, for more
  details).   Blue/Solid = {\sc Peace}, Green/Dash =
  \wcd, Black/Dot-Dash = \capthree; vertical tics = 95\% confidence
  intervals on estimates.  Intervals are not presented for Type 1
  error due to the extremely small variance.}
\end{figure*}

\begin{figure*}[b]
\centerline{
\includegraphics[scale=0.35]{pics.d/singletons.pdf}
}
\label{singletons}
\caption{Average number of fragments flagged as singletons by each tool
  when run on the simulated sequences; correct answer in all cases is
  zero.  Blue/Solid = {\sc Peace}, Green/Dash = \wcd,
  Black/Dot-Dash = \capthree.}
\end{figure*}

In applying the tools to real data, we started with a set of
approximately $190,000$ fragments derived from the {\it Chlamydomonas
  reinhardtii} genome.  To compute sensitivity we used the {\it gmap}
tool \cite{Wu05} to map the fragment set to the genome, taking this as
our reference clustering.  We see slight improvements in
{\sc Peace} over \wcd in both sensitivity and Type 1 error
(both significantly out preforming \capthree).  Using the mouse fragment
set used in Hazelhurst {\it et al.}  \cite{Hazelhurst08a}
{\sc Peace} again shows slight improvements in sensitivity and Type
1 error rates, with an 18\% speed up for {\sc Peace} (see
Supplementary Materials, Section~\ref{real_results}, for more
details).

\section{Acknowledgements}

Dr. Karro was funded under a PhRMA Foundation Informatics Research
Starters Grant while conducting this research.  We would also like to
acknowledge Iddo Friedberg, David Woods, Jens Mueller and David
Scoville at Miami University for their help with this project.

\vspace{3mm}
%Bib TeX

\bibliography{peace.bib}

\end{document}

% LocalWords:  arallel nalysis lustering ngine Rao Moler Ozden Zhang Liang ESTs
% LocalWords:  Karro XXXXX transcriptome pre isoforms SNPs unsequenced runtime
% LocalWords:  chlamydomondomonas ESTate PaCE xsact CLU WCD Hazelhurst et al th
% LocalWords:  CHUN nalyzer Prim's MSTs subgraph MPI SourceCode ESTSim runtimes
% LocalWords:  ests Chlamydomonas Reihartdii gmap CDNA transcriptomes cDNA www
% LocalWords:  reinhartdii conifergdb DNAc bp wcd reinhardtii PhRMA Iddo fasta
% LocalWords:  Friedberg Scoville OpenMPI initio multi Screenshots

### 2010-02-10 src/ClusterMaker.h
This check-in provides a significantly enhanced infrastructure for filtering out FASTA entries with low complexity segments. This infrastructure has been implemented in the form of Filters.
Filters are a new concept that have been added to PEACE. The
filters have been implemented in a similar manner as heuristics.
Accordingly, I have added the following classes:

Filter: The base class for all filters. It provides common
interface methods to interact with all filters.

FilterFactory: Centralized factory class for instantiating
various filters.

FilterChain: A chain of filters through which an

@@ -100,9 +100,79 @@ public:
         EST analysis method.  This method is a pure-virtual method.
         Therefore all cluster maker classes must override this method
         to perform all the necessary operations.
+
+		\note This method must be invoked only after the initialize()
+		method is invoked.
     */
     virtual int makeClusters() = 0;
 
+    /** Obtain the EST analyzer set for this cluster maker.
+
+        This method must be used to obtain the EST analyzer set for
+        this cluster maker.
+
+        \return The EST analyzer set for this cluster maker. If a
+        valid EST analyzer has not been set then this method returns
+        NULL.
+    */
+    inline ESTAnalyzer *getAnalyzer() const { return analyzer; }
+
+    /** Add a dummy cluster to the cluster maker.
+
+        This method can be used to add a dummy cluster to the cluster
+        maker. The dummy clusters are added as direct descendants of
+        the root cluster with the given name.
+
+        \note This method is currently used by the Filter hierarchy to
+        add ESTs that are logically filtered out and must not be part
+        of the core clustering process.
+        
+        \param[in] name A human readable name to be set for this
+        cluster.
+        
+        \return If a cluster was successfully added, then this method
+        returns a unique integer that identifies the newly added
+        cluster. This value must be used to add entries to this
+        cluster via the ClusterMaker::addEST method.
+    */
+    virtual int addDummyCluster(const std::string name) = 0;
+
+    /** Add a EST directly to a given cluster.
+
+        This method can be used to add an EST directly to a
+        cluster. This bypasses any traditional mechanism and directly
+        adds the EST to the specified cluster.
+
+        \note The EST is added with an invalid metric value. ESTs
+        added to a cluster are not included in the standard clustering
+        process. Adding an EST that has already been added to the
+        same/another cluster results in undefined behaviors.
+
+        \param[in] clusterID The unique ID of the cluster to which the
+        EST is to be added. This value must have been obtained from an
+        earlier (successful) call to the ClusterMaker::addDummyCluster
+        method.
+        
+        \param[in] estIdx The EST to be added to the given
+        cluster. Once the EST has been added to this cluster it will
+        not be included in the clustering process performed by this
+        cluster maker.
+    */
+    virtual void addEST(const int clusterID, const int estIdx) = 0;
+
+	/** A method to handle initialization tasks for the ClusterMaker.
+		
+		This method is called after the cluster maker has been created
[…]

### 2010-02-10 src/EST.cpp

@@ -58,6 +58,7 @@ EST::EST(const int idValue, const char *information, const char* seq,
     info       = EST::duplicate(information);
     sequence   = EST::duplicate(seq);
     similarity = 0;
+    processed  = false;
 }
 
 EST::~EST() {
@@ -114,7 +115,7 @@ EST::create(const int id, const char *info,
 }
 
 EST*
-EST::create(FILE* fastaFile, int& lineNum, const int minLength) {
+EST::create(FILE* fastaFile, int& lineNum) {
     // Some basic validation on the fastFile first.
     if (feof(fastaFile) || ferror(fastaFile)) {
         // Can't read information from the fasta file.
@@ -161,20 +162,15 @@ EST::create(FILE* fastaFile, int& lineNum, const int minLength) {
                        toupper);
         const char* const seqBP = sequence.c_str();
         size_t seqLen = strlen(seqBP);
-        if (seqLen >= minLength) { // Only accept ESTs of sufficient length
-            // Compute new max EST length
-            maxESTlen = std::max(maxESTlen, strlen(seqBP));
-            // Create a new est with all the information.
-            EST *est = new EST((int) estList.size(), headerLine.c_str(), seqBP,
-                               offset);
-            // Add it to the est list.
-            estList.push_back(est);
-            // Return newly created est back to the caller.
-            return est;
-        } else {
-            // Return a dummy EST to the caller.
-            return dummy;
-        }
+        // Compute new max EST length
+        maxESTlen = std::max(maxESTlen, strlen(seqBP));
+        // Create a new est with all the information.
+        EST *est = new EST((int) estList.size(), headerLine.c_str(), seqBP,
+                           offset);
+        // Add it to the est list.
+        estList.push_back(est);
+        // Return newly created est back to the caller.
+        return est;
     }
 
     // Can't create a valid EST
@@ -197,6 +193,19 @@ EST::dumpESTList(std::ostream& os) {
     for(int id = 0; (id < EstCount); id++) {
         // Dump the EST information out.
         estList[id]->dumpEST(os);
+    }
+}
+
+
+void
+EST::dumpESTList(std::ostream& os, const bool processed) {
+    const int EstCount = (int) estList.size();
+    for(int id = 0; (id < EstCount); id++) {
+        // Dump the EST information out based on processed status.
+        if (estList[id]->processed == processed) {
[…]

### 2010-02-10 src/FWAnalyzer.cpp

@@ -178,7 +178,7 @@ FWAnalyzer::initialize() {
                       << std::endl;
         }
     }
-    // Initializawtion successful.
+    // Initialization successful.
     return 0;
 }
 

### 2010-02-10 src/FWAnalyzer.h

@@ -136,9 +136,9 @@ public:
 
     /** Method to begin EST analysis.
 
-        This method is invoked just before commencement of EST
+        <p>This method is invoked just before commencement of EST
         analysis.  This method loads the list of ESTs from a given
-        input multi-FASTA file and pouplates the list of ESTs.
+        input multi-FASTA file and pouplates the list of ESTs.</p>
 
         \return If the ESTs were successfully loaded from the FATA
         file then this method returns 0.  Otherwise this method
@@ -204,6 +204,21 @@ public:
         non-zero value.
     */
     virtual int analyze();
+
+    /** Determine preferred dummy EST lengths to be used with this
+        analyzer.
+
+        <b>Note: For more detailed description of the motivation for
+        dummy ESTs please refer to the documentation for the
+        corresponding method in the base class --
+        getPreferredDummyESTLength().
+        
+        \return This method overrides the default implementation in
+        the base class to return twice the length of the frame (aka
+        window) size.
+    */
+    virtual int getPreferredDummyESTLength() const
+    { return frameSize * 2; }
     
 protected:    
     /** Analyze and obtain a similarity metric.
@@ -247,20 +262,20 @@ protected:
         to perform the necessary operations.
     */
     virtual float analyzeFrame(const std::string& refFrame,
-			       const std::string& otherFrame,
-			       const int wordSize);
+                               const std::string& otherFrame,
+                               const int wordSize);
     
     /** Helper method to dump result log header.
 
-	This is a helper method that is invoked from the analyze()
-	method to dump a result log header.  This method was
-	introduced to keep the code clustter in the analyze method to
-	a minimum.
+        This is a helper method that is invoked from the analyze()
+        method to dump a result log header.  This method was
+        introduced to keep the code clustter in the analyze method to
+        a minimum.
 
-	This method dumps some of the analysis parameters to the
-	supplied log.
+        This method dumps some of the analysis parameters to the
+        supplied log.
 
-	\param[out] log The log to which the header is to be dumped.
+        \param[out] log The log to which the header is to be dumped.
 
         \param[in] mean The overall mean similarity for this set of
         ESTs.
@@ -291,7 +306,7 @@ protected:
                              const EST* refEST,
                              ResultLog& log);
 
-    /** Dumps a given EST in 3-column format using R
+    /** Dumps a given EST in 3-column format using R.
 
         This method is a helper method that dumps a given EST out to
[…]

### 2010-02-10 src/Filter.cpp

@@ -1,0 +1,113 @@
+#ifndef FILTER_CPP
+#define FILTER_CPP
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include "Filter.h"
+#include "Utilities.h"
+#include "ClusterMaker.h"
+
+Filter::Filter(const std::string& name, ClusterMaker* clsMaker)
+    : filterName(name), clusterMaker(clsMaker) {
+    runCount    = 0;
+    filterCount = 0;
+    // Nothing else to be done for now.
+}
+
+Filter::~Filter() {
+    // Empty constructor begets an empty destructor
+}
+
+int
+Filter::applyFilter(const int estIndex) {
+    int clusterIndex;
+    if ((clusterIndex = runFilter(estIndex)) != -1) {
+        // This EST was filtered out.
+        filterCount++;
+        // Track the filtered EST information for distribution to
+        // other processes later on.
+        filteredESTList[clusterIndex].push_back(estIndex);
+        // Add this EST to the clsuter
+        clusterMaker->addEST(clusterIndex, estIndex);
+    }
+    // Track number of times this filter was run.
+    runCount++;
[…]

### 2010-02-10 src/FilterChain.cpp

@@ -1,0 +1,270 @@
+#ifndef FILTER_CHAIN_CPP
+#define FILTER_CHAIN_CPP
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include "FilterChain.h"
+#include "FilterFactory.h"
+#include "Utilities.h"
+#include "MPIHelper.h"
+#include "MPIStats.h"
+#include "EST.h"
+
+// Get rid of magic numbers
+#define NO_ERROR 0
+
+// The static pointer to the singleton filter chain instance.
+FilterChain* FilterChain::ptrInstance = NULL;
+
+void
+FilterChain::showArguments(std::ostream& os) {
+    for (size_t i = 0; (i < chain.size()); i++) {
+        chain[i]->showArguments(os);
+    }
+}
+
+bool
+FilterChain::parseArguments(int& argc, char **argv) {
+    bool flag = true;
+    for (size_t i = 0; (i < chain.size()); i++) {
+        if (!chain[i]->parseArguments(argc, argv)) {
+            flag = false;
+        }
+    }
+    return flag;
+}
+
+bool
[…]

### 2010-02-10 src/FilterChain.h

@@ -1,0 +1,373 @@
+#ifndef FILTER_CHAIN_H
+#define FILTER_CHAIN_H
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include <vector>
+#include "Filter.h"
+#include "HashMap.h"
+
+/** Class that manages a list of filters.
+
+    <p>This class represents a list of filters that are used to try
+    and elimintate entries that could potentially deteriorate the
+    quality of clustering generated by PEACE.  The filters are run
+    prior to commencement of the core clustering operation.  For
+    convenient post-processing analysis the filtered ESTs are gathered
+    in suitable "dummy" clusters to be ignored.</p>
+
+    <p>Each filter object in the chain implements a specific type of
+    filteration operation and returns an integer indicating the
+    cluster to which an EST is to be assigned. If the cluster ID is
+    -1, then that indicates that the EST must be subjected to regular
+    clustering operations.</p>
+
+	<p>Note that filter processing proceeds in a given order. The
+	first filter in the chain is invoked. If that filter returns a
+	positive cluster ID then rest of the filters are not invoked. On
[…]

### 2010-02-10 src/FilterFactory.h

@@ -1,0 +1,115 @@
+#ifndef FILTER_FACTORY_H
+#define FILTER_FACTORY_H
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include <iostream>
+
+// For declaration to make compiler happy.
+class Filter;
+class ClusterMaker;
+
+/** Centralized factory for creating filters.
+
+	This is a centralized helper class that can be used to create new
+	instances of various filters. This class provides a convenient
+	FilterFactory::create() method for this purpose. In addition, it
+	also provides a mechanism to list all the known filters.
+
+    <p>A set of filters (stored in the FilterChain class) are run on
+    all entries in a FASTA file prior to commencement of the core
+    clustering operation. The filters perform various validation
+    operations to ensure ESTs are good prior to clustering. Filtering
+    ensures that the overall quality of clustering provided by PEACE
+    is good.</p>
+
+    <p>Each filter object in the chain implements a specific type of
+    filteration operation and ultimately returns an integer indicating
+    the cluster to which an EST is to be assigned. If the cluster ID
+    is -1, then that indicates that the EST must be subjected to
+    regular clustering operations.</p>
+
+    \see FilterChain
+*/
+class FilterFactory {
+public:
[…]

### 2010-02-10 src/LengthFilter.cpp

@@ -1,0 +1,104 @@
+#ifndef LENGTH_FILTER_CPP
+#define LENGTH_FILTER_CPP
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include "LengthFilter.h"
+#include "ClusterMaker.h"
+#include "EST.h"
+
+// Define the static parameters
+int LengthFilter::minESTLen = 50;
+
+// The set of arguments for this class.
+arg_parser::arg_record LengthFilter::argsList[] = {
+    {"--minESTLen", "Minimum EST length to be permitted by this filter",
+     &LengthFilter::minESTLen, arg_parser::INTEGER},    
+    {NULL, NULL, NULL, arg_parser::BOOLEAN}
+};
+
+LengthFilter::LengthFilter(ClusterMaker *clusterMaker) :
+    Filter("lengthFilter", clusterMaker) {
+    // Initialize instance variables.
+    clusterID = -1;
+}
+
+
+void
+LengthFilter::showArguments(std::ostream& os) {
+    // Use a arg parser object to conveniently display common options.
[…]

### 2010-02-10 src/MSTCluster.h

@@ -48,7 +48,6 @@ class MSTCluster;
 */
 typedef std::vector<MSTCluster*> ClusterList;
 
-
 /** Class to encapsulate cluster information and aid in building
     clusters using Minimum Spanning Tree (MST) data.
 
@@ -59,12 +58,39 @@ class MSTCluster {
     // Insertion operator to make dumping MST for debugging easier.
     friend std::ostream& operator<<(std::ostream&, const MSTCluster&);  
 public:
-    MSTCluster(MSTCluster* owner = NULL);
+    MSTCluster(MSTCluster* owner = NULL, const std::string& name = "");
     ~MSTCluster();
 
     double makeClusters(NodeList& nodeList, const ESTAnalyzer* analyzer);
     void add(const MSTNode& node);
 
+    /** Add a child cluster to this cluster.
+
+        This method must be used to add a child cluster this
+        cluster. Note that the parent cluster must be appropriately
+        set in the child to refer to this MSTCluster.
+
+        \note This cluster node takes ownership of the child node.
+
+        \param[in] child The child node to be added to the this
+        cluster.
+    */
+    void add(MSTCluster *child);
+    
+    /** Method to directly add a given EST to a dummy clusterID.
+
+        This method is typically used to add a EST that has been
+        filtered out by a filter to a specific cluster.  The cluster
+        maker typically creates a dummy MSTNode with the necessary
+        information to be added.
+        
+        \param clusterID The ID/index of the cluster to which the
+        MST node is to be added.
+        
+        \param[in] node The MSTNode entry to be added to this cluster.
+    */
+    void add(const int clusterID, const MSTNode& node);
+    
     void printClusterTree(std::ostream& os = std::cout,
                           const std::string& prefix = "") const;
 
@@ -74,16 +100,36 @@ public:
     void makeMergedClusters(const int size, int* parent, bool* root);
                
     inline int getClusterID() const { return clusterID; }
-    
+
+
 protected:
 	void guiPrintTree(std::ostream& os) const;
 	
 private:
+    /** List of sub-clusters for this cluster.
+
+        This instance variable is used to maintain the list of
+        sub-clusters for this cluster. Entries are added to this list
+        whenever sub-clusters are created in the
+        MSTCluster::makeClusters() method.
+    */
     ClusterList clusterList;
+    
     const MSTCluster* parent;
     NodeList members;
     const int clusterID;
-    
+
[…]

### 2010-02-10 src/PMSTClusterMaker.cpp

@@ -483,7 +483,8 @@ PMSTClusterMaker::populateCache(const int estIdx, SMList* metricList) {
     bool needInvalidMetric = true;
     
     for(int otherIdx = startESTidx; (otherIdx < endESTidx); otherIdx++) {
-        if ((otherIdx == estIdx) || (cache->isESTinMST(otherIdx))) {
+        if ((otherIdx == estIdx) || (cache->isESTinMST(otherIdx)) ||
+            (EST::getEST(otherIdx)->hasBeenProcessed())) {
             // This EST entry can be ignored as this similarity metric
             // is not needed for MST construction.
             continue;

### 2010-02-10 src/UVSampleHeuristic.h

@@ -171,7 +171,7 @@ protected:
 					  const std::string& outputFileName);
     
     /** Determine whether the analyzer should analyze, according to
-	this heuristic.
+        this heuristic.
         
         This method can be used to compare a given EST with the
         reference EST (set via the call to the setReferenceEST())
@@ -181,7 +181,7 @@ protected:
         which the reference EST is to be compared.
         
         \return This method returns true if the heuristic says the
-	EST pair should be analyzed, and false if it should not.
+        EST pair should be analyzed, and false if it should not.
     */
     virtual bool runHeuristic(const int otherEST);
 
@@ -201,31 +201,31 @@ protected:
     void computeHash(const int estIdx);
     
     /** Instance variable to track if a given word (of length \c v)
-	appears in reference EST.
+        appears in reference EST.
 
-	This instance variable is created in the initialize() method
-	to point to an array of 4<sup>v</sup>
+        This instance variable is created in the initialize() method
+        to point to an array of 4<sup>v</sup>
 
     */
     char* s1WordMap;
 
     /** Instance variable to track if a given word (of length \c v)
-	appears in reference EST.
+        appears in reference EST.
 
-	This instance variable is created in the initialize() method
-	to point to an array of 4<sup>v</sup> 
+        This instance variable is created in the initialize() method
+        to point to an array of 4<sup>v</sup> 
     */
     char* s1RCWordMap;
 
     /** Flag to indicate if normal or reverse-complement version
-	provided best match.
+        provided best match.
 
-	This flag is set at the end of the runHeuristic method in this
-	class to indicate if the normal or the reverse-complement
-	check yielded the best possible match. If this flag is \c
-	false, then the normal check yielded the best match. If this
-	value is \c true, then the reverse-complement check yielded
-	the best match.
+        This flag is set at the end of the runHeuristic method in this
+        class to indicate if the normal or the reverse-complement
+        check yielded the best possible match. If this flag is \c
+        false, then the normal check yielded the best match. If this
+        value is \c true, then the reverse-complement check yielded
+        the best match.
     */
     bool bestMatchIsRC;
 
@@ -297,11 +297,11 @@ private:
     UVHashTable uvCache;
 
     /** The hint key that is used to add hint for normal or
-	reverse-complement D2 computation.
-
-	This hint key is used to set a hint in the \c hints hash
-	map. This string is defined as a constant to save compute time
-	in the core \c runHeuristics method.
[…]

### 2010-02-10 peace/src/EST.h
This check-in provides a significantly enhanced infrastructure for filtering out FASTA entries with low complexity segments. This infrastructure has been implemented in the form of Filters.
Filters are a new concept that have been added to PEACE. The
filters have been implemented in a similar manner as heuristics.
Accordingly, I have added the following classes:

Filter: The base class for all filters. It provides common
interface methods to interact with all filters.

FilterFactory: Centralized factory class for instantiating
various filters.

FilterChain: A chain of filters through which an

@@ -95,10 +95,10 @@ public:
         as the parameter.  The EST names are expected to be unique in
         a given file.
 
-	\note If the new EST is successfully instantiated, then this
-	method adds the newly created EST to the end of the list of
-	ESTs maintianed by this class.  Consequenlty, the parameter \c
-	id must be equal to estList.size().
+        \note If the new EST is successfully instantiated, then this
+        method adds the newly created EST to the end of the list of
+        ESTs maintianed by this class.  Consequenlty, the parameter \c
+        id must be equal to estList.size().
 	
         \param[in] id The unqiue ID value to be set for this EST.
         
@@ -137,79 +137,106 @@ public:
         \param[in,out] lineNum A line number counter to be updated to
         provide the user with a more meaningful error message.
 
-	\param[in] minLength The minimum length of EST data that
-	will be accepted and used for clustering.  Sequences shorter
-	than this length will not be loaded into memory.
-        
         \note At the end of this method the fastaFile's file pointer
         will point at the beginning of the next EST (if any) in the
         file.
     */
-    static EST* create(FILE* fastaFile, int& lineNum, const int minLength);
+    static EST* create(FILE* fastaFile, int& lineNum);
 
     /** Obtain the list of ESTs.
 
-	This method may be used to obtain a reference to the list of
-	ESTs currently defined.
+        This method may be used to obtain a reference to the list of
+        ESTs currently defined.
 
-	\return The list of ESTs currently defined.
+        \return The list of ESTs currently defined.
     */
     static std::vector<EST*>& getESTList() { return estList; }
 
+	/** Obtain count of ESTs that have been flagged as being processed.
+
+		This method can be used to determine the number of ESTs that
+		have been flagged as being processed. Subtracting this number
+		from the total number of ESTs indicates the number of ESTs to
+		be processed.
+
+		\note This method iterates over the list of ESTs to determine
+		the current number of processed ESTs. So use this method
+		sparingly.
+
+		\return The number of ESTs that have been flagged as having
+		been processed.
+	*/
+	static int getProcessedESTCount();
+	
     /** Obtain the number of ESTs in this list.
 
-	This method may be used to determine the number of ESTs that
-	have been defined and added to this list.
+        This method may be used to determine the number of ESTs that
+        have been defined and added to this list.
 
-	\return The number of ESTs currently defined.
+        \return The number of ESTs currently defined.
     */
     static int getESTCount() { return (int) estList.size(); }
 
     /** Helper method to determine the longest EST.
 
[…]

### 2010-02-10 peace/src/ESTAnalyzer.cpp

@@ -45,7 +45,6 @@
 bool  ESTAnalyzer::readAhead      = false;
 char* ESTAnalyzer::estFileName    = NULL;
 bool  ESTAnalyzer::htmlLog        = false;
-int   ESTAnalyzer::estMinLength   = 50;
 
 // The common set of arguments for all EST analyzers
 arg_parser::arg_record ESTAnalyzer::commonArgsList[] = {
@@ -53,10 +52,8 @@ arg_parser::arg_record ESTAnalyzer::commonArgsList[] = {
      &ESTAnalyzer::readAhead, arg_parser::BOOLEAN},
     {"--estFile", "Name of EST file (in FASTA format) to be processed",
      &ESTAnalyzer::estFileName, arg_parser::STRING},
-    {"--minLength", "Minimum length of sequences to be clustered",
-     &ESTAnalyzer::estMinLength, arg_parser::INTEGER},
     {"--html", "Generate analysis report in HTML format",
-     &ESTAnalyzer::htmlLog, arg_parser::BOOLEAN},   
+     &ESTAnalyzer::htmlLog, arg_parser::BOOLEAN},
     {NULL, NULL, NULL, arg_parser::BOOLEAN}
 };
 
@@ -65,7 +62,7 @@ ESTAnalyzer::ESTAnalyzer(const std::string& name, const int estIdx,
     : refESTidx(estIdx), chain(NULL), outputFileName(outputFile),
       analyzerName(name) {
     // Nothing else to be done for now.
-}
+      }
 
 ESTAnalyzer::~ESTAnalyzer() {
     // Empty constructor begets an empty destructor
@@ -122,9 +119,9 @@ bool
 ESTAnalyzer::loadFASTAFile(const char *fileName, const bool unpopulate) {
     static const std::string IgnoreFileName = "<none>";
     if (IgnoreFileName == fileName) {
-	// The user does not want to use a file name. This possibly
-	// happens when using peace libary from custom applications.
-	return true;
+        // The user does not want to use a file name. This possibly
+        // happens when using peace libary from custom applications.
+        return true;
     }
     FILE *fastaFile = NULL;
 #ifndef _WINDOWS
@@ -145,7 +142,7 @@ ESTAnalyzer::loadFASTAFile(const char *fileName, const bool unpopulate) {
     int filteredCount = 0;
     // Repeatedly read EST's from the file.
     while (!feof(fastaFile)) {
[…]

### 2010-02-10 peace/src/ESTAnalyzer.h

@@ -99,7 +99,8 @@ public:
         This method is invoked just before commencement of EST
         analysis.  This method typically loads the list of ESTs from a
         given input file.  In addition, it may perform any
-        pre-processing as the case may be.
+        pre-processing as the case may be. Some EST analyzers may also
+        add dummy entries to aid in various operations.
 
         \note Derived classes must override this method.
 
@@ -139,28 +140,28 @@ public:
     virtual int setReferenceEST(const int estIdx) = 0;
 
     /** Analyze and obtain a similarity metric using the attached
-	heuristic chain (if one exists) followed by the appropriate
-	heavy weight distance/similarity measure associated with
-	this ESTAnalyzer.
-
+		heuristic chain (if one exists) followed by the appropriate
+		heavy weight distance/similarity measure associated with
+		this ESTAnalyzer.
+		
         This method can be used to compare a given EST with the
         reference EST (set via the call to the setReferenceEST())
         method.
-
-	\note This method may return -1, if the otherEST is
+		
+		\note This method may return -1, if the otherEST is
         significantly different from the reference EST (possibly
         warranting no further analysis) that a meanigful metric cannot
         be generated.
-	
+		
         \param[in] otherEST The index (zero based) of the EST with
         which the reference EST is to be compared.
-
-	\param[in] useHeuristics A directive instructing the ESTAnalyzer
-	on whether or not to use its heuristis chain.  Defaults to true.
-
-	\param[in] useHeavyWeight A directive instructing the ESTAnalyzer
-	on whether or not to use the heavy weight metric.  Defaults to true.
-
+		
+		\param[in] useHeuristics A directive instructing the ESTAnalyzer
+		on whether or not to use its heuristis chain.  Defaults to true.
+		
+		\param[in] useHeavyWeight A directive instructing the ESTAnalyzer
+		on whether or not to use the heavy weight metric.  Defaults to true.
+		
         \return This method returns a similarity/distance metric by
         comparing the ESTs. This method may return -1, if the otherEST
         is significantly different from the reference EST (possibly
@@ -168,7 +169,7 @@ public:
         be generated.
     */
     float analyze(const int otherEST, const bool useHeuristics = true,
-		  const bool useHeavyWeight = true);
+				  const bool useHeavyWeight = true);
 
     /** Method to perform EST analysis.
 
@@ -237,13 +238,13 @@ public:
 
     /** Obtain the input file name.
 
-	This method returns the input file from where the EST data was
-	read.
[…]

### 2010-02-10 peace/src/Filter.h

@@ -1,0 +1,397 @@
+#ifndef FILTER_H
+#define FILTER_H
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include "arg_parser.h"
+#include "HashMap.h"
+
+#include <vector>
+
+// Forward declarations to keep compiler fast and happy
+class ClusterMaker;
+
+/** \def FilteredESTList
+
+    \brief Typedef for HashMap<int, std::vector<int> > FilteredESTList;
+
+    This typedef is a convenience definition to define a hash map
+    tracks the set of entries that were filetered out by this
+    filter. The key to the hash map is the cluster ID to which
+    filtered ESTs were added. The value in the hash map is a vector
+    that contains the index of the ESTs that were added to the
+    cluster.
+*/
+typedef HashMap<int, std::vector<int> > FilteredESTList;
+
+/** The base class of all filters.
+    
+    <p>This class must be the base class of all filters in the
+    system. This class provides some default functionality that can be
+    readily used by the filters. This class enables the FilterChain to
+    manage a list of filters and dispatch method calls to various
+    filters.</p>
+
+    <p>A set of filters (stored in the FilterChain class) are run on
[…]

### 2010-02-10 peace/src/FilterFactory.cpp

@@ -1,0 +1,78 @@
+#ifndef FILTER_FACTORY_CPP
+#define FILTER_FACTORY_CPP
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include "FilterFactory.h"
+#include "LengthFilter.h"
+#include "arg_parser.h"
+
+void
+FilterFactory::displayList(std::ostream &os) {
+    // Create dummy command-line args to make display prettier and
+    // easier.
+    arg_parser::arg_record dummy_args[] = {
+        {"lengthFilter", "Filter out short ESTs in the FASTA file",
+         NULL, arg_parser::STRING},
+        {"lcFilter", "Filter out entries with Low Complexity regions",
+         NULL, arg_parser::STRING},
+        {NULL, NULL, NULL, arg_parser::BOOLEAN}
+    };
+    arg_parser dummyParser(dummy_args);
+    os << dummyParser;
+}
+
+Filter*
+FilterFactory::create(const char* name, ClusterMaker *clusterMaker) {
+    if (name == NULL) {
+        return NULL;
+    }
+    if (clusterMaker == NULL) {
[…]

### 2010-02-10 peace/src/LengthFilter.h

@@ -1,0 +1,224 @@
+#ifndef LENGTH_FILTER_H
+#define LENGTH_FILTER_H
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include "arg_parser.h"
+#include "Filter.h"
+
+// Forward declarations to keep compiler fast and happy
+class ClusterMaker;
+
+/** A simple filter to filter out reads shorter than a given length.
+    
+	<p>This class provides a simple filter that can be used to filter
+	out ESTs that are shorter than a given number of nucleotides.
+	This filter can be enabled by specifying its name, namely \c
+	lengthFilter, in the filter chain (that is specified as the
+	command line argument to PEACE). If this filter is specified then
+	the lengthFilterSize command line argument is used to determine
+	the threshold length value (ESTs shorter than this value will be
+	filtered out).</p>
+
+    <p>This filter creates an dummy cluster with the meta name "Short
+    ESTs (filtered by LengthFilter)" and adds any ESTs filtered out by
+    this EST to that cluster.</p>
+
+    <p>This class has been developed by extending the Filter base
+    class and implementing the necessary API methods specified by the
[…]

### 2010-02-10 peace/src/MSTCluster.cpp

@@ -46,9 +46,14 @@ int MSTCluster::clusterIDSequence = 0;
 // A type-def for HashMap to make code cleaner.
 typedef HashMap<int, MSTCluster*> ClusterMap;
 
-MSTCluster::MSTCluster(MSTCluster* owner) :
-    parent(owner), clusterID(clusterIDSequence++) {
-    // Nothing else to be done for now.
+// Global list to maintain reference to all the MSTClusters created.
+ClusterList MSTCluster::globalClusterList;
+
+MSTCluster::MSTCluster(MSTCluster* owner, const std::string& clsName) :
+    parent(owner), clusterID(clusterIDSequence++), name(clsName) {
+    // Save reference to this cluster entry in the global list
+    ASSERT ( clusterID == (int) globalClusterList.size() );
+    globalClusterList.push_back(this);
 }
 
 MSTCluster::~MSTCluster() {
@@ -60,11 +65,30 @@ MSTCluster::~MSTCluster() {
     clusterList.clear();
     // Remove all nodes in the members list.
     members.clear();
+    // Clear out entry in the global EST list
+    globalClusterList[clusterID] = NULL;
 }
 
 void
 MSTCluster::add(const MSTNode& node) {
     members.push_back(node);
+}
+
+void
+MSTCluster::add(const int clusterID, const MSTNode& node) {
+    // Validate context and parameters
+    ASSERT ((clusterID >= 0) && (clusterID < (int) globalClusterList.size()));
+    ASSERT (globalClusterList[clusterID] != NULL);
+    ASSERT (globalClusterList[clusterID]->clusterID == clusterID);
+    // Add node to the specified cluster.
+    globalClusterList[clusterID]->members.push_back(node);
+}
+
+void
+MSTCluster::add(MSTCluster *child) {
+    ASSERT ( child->parent == this );
+    // Add newly created cluster to list of sub-clusters
+    clusterList.push_back(child);
 }
 
 double
@@ -89,8 +113,8 @@ MSTCluster::makeClusters(NodeList& nodeList, const ESTAnalyzer* analyzer) {
             subCluster = new MSTCluster(this);
             // Add cluster to temporary look up hash_map. 
             clusterMap[(*node).parentIdx] = subCluster;
-            // Add newly created cluster to permanent list of sub-clusters
-            clusterList.push_back(subCluster);
+            // Add to list of sub-clusters
+            add(subCluster);
             // Add the parent node as well to the cluster as it would
[…]

### 2010-02-10 peace/src/MSTClusterMaker.cpp

@@ -36,7 +36,6 @@
 
 #include "MSTClusterMaker.h"
 #include "ESTAnalyzer.h"
-#include "MSTCluster.h"
 #include "MSTMultiListCache.h"
 #include "MSTHeapCache.h"
 #include "EST.h"
@@ -300,7 +299,8 @@ MSTClusterMaker::updateProgress(const int estsAnalyzed,
 int
 MSTClusterMaker::manager() {
     // The number of pending nodes to be added to the MST.
-    const int TotalESTcount = EST::getESTList().size();
+    const int TotalESTcount = EST::getESTList().size() -
+        EST::getProcessedESTCount();;
     int pendingESTs         = TotalESTcount;
     // The minimum spanning tree that is built by this manager.
     int dummy;
@@ -438,6 +438,13 @@ MSTClusterMaker::analyze(const int otherEST) {
 
 int
 MSTClusterMaker::initialize() {
+    // All the parallel MPI processes load the data for processing
+    // into memory.
+    int result;
+    if ((result = analyzer->initialize()) != NO_ERROR) {
+        // Error occured during initialization. Bail out.
+        return result;
+    }
     // Hack to fix the baton maxUse problem until a better solution is found
     if (!analyzer->getName().compare("baton") && maxUse > -1) {
         maxUse = 100;
@@ -462,9 +469,10 @@ MSTClusterMaker::populateCache(const int estIdx, SMList* metricList) {
     // Flag to ensure only one invalid metric gets added
     bool needInvalidMetric = true;
     for(int otherIdx = startESTidx; (otherIdx < endESTidx); otherIdx++) {
-        if (cache->isESTinMST(otherIdx) || (otherIdx == estIdx)) {
-            // This EST entry can be ignored as this similarity metric
-            // is not needed for MST construction.
+        if (cache->isESTinMST(otherIdx) || (otherIdx == estIdx) ||
+            (EST::getEST(otherIdx)->hasBeenProcessed())) {
+            // This EST entry can be ignored as this metric is not
+            // needed (or must not be used) for MST construction.
             continue;
         }
         // Get similarity/distance metric.
@@ -650,8 +658,25 @@ MSTClusterMaker::populateMST() {
 }
 
 int
+MSTClusterMaker::addDummyCluster(const std::string name) {
+    MSTCluster *dummy = new MSTCluster(&root, name);
+    root.add(dummy);
+    // Return id of newly added cluster
[…]

### 2010-02-10 peace/src/MSTClusterMaker.h

@@ -36,7 +36,9 @@
 
 #include "ClusterMaker.h"
 #include "MSTCache.h"
+#include "MSTCluster.h"
 #include "MST.h"
+
 #include <fstream>
 
 /** A Minimum Spanning Tree (MST) based parallel cluster maker.
@@ -63,9 +65,9 @@ public:
         parallel/distributed manner.
     */
     enum MessageTags{REPOPULATE_REQUEST, COMPUTE_SIMILARITY_REQUEST,
-		     SIMILARITY_LIST, SIMILARITY_COMPUTATION_DONE,
-		     COMPUTE_MAX_SIMILARITY_REQUEST, MAX_SIMILARITY_RESPONSE,
-		     ADD_EST, TRANSITIVITY_LIST, COMPUTE_TOTAL_ANALYSIS_COUNT};
+                     SIMILARITY_LIST, SIMILARITY_COMPUTATION_DONE,
+                     COMPUTE_MAX_SIMILARITY_REQUEST, MAX_SIMILARITY_RESPONSE,
+                     ADD_EST, TRANSITIVITY_LIST, COMPUTE_TOTAL_ANALYSIS_COUNT};
     
     /** The destructor.
         
@@ -130,7 +132,49 @@ public:
         be written.        
     */
     virtual void displayStats(std::ostream& os);
-    
+
+    /** Add a dummy cluster to the cluster maker.
+        
+        This method can be used to add a dummy cluster to the cluster
+        maker. The dummy clusters are added as direct descendants of
+        the \c root cluster with the given name.
+
+        \note This method is currently used by the Filter hierarchy to
+        add ESTs that are logically filtered out and must not be part
+        of the core clustering process.
+        
+        \param[in] name A human readable name to be set for this
+        cluster. No special checks are made on the contents of the
+        string.
+        
+        \return This method returns the ID value set for the newly
+        added dummy index.
+    */
+    virtual int addDummyCluster(const std::string name);
+
+    /** Add a EST directly to a given cluster.
+
+        This method can be used to add an EST directly to a
+        cluster. This bypasses any traditional mechanism and directly
+        adds the EST to the specified cluster.
+
+        \note The EST is added with an invalid metric value. ESTs
+        added to a cluster are not included in the standard clustering
+        process. Adding an EST that has already been added to the
+        same/another cluster results in undefined behaviors.
+
+        \param[in] clusterID The unique ID of the cluster to which the
+        EST is to be added. This value must have been obtained from an
[…]

### 2010-02-10 peace/src/main.cpp

@@ -41,13 +41,17 @@
 #include "ESTAnalyzer.h"
 #include "HeuristicFactory.h"
 #include "HeuristicChain.h"
+#include "FilterFactory.h"
+#include "FilterChain.h"
 #include "MPIHelper.h"
 #include "InteractiveConsole.h"
+#include "EST.h"
 
 #include <string>
 #include <sstream>
+#include <fstream>
 
-/** \fn void showUsage(arg_parser& ap, ESTAnalyzer *analyzer, ClusterMaker *cMaker, HeuristicChain *hChain)
+/** \fn void showUsage(arg_parser& ap, ESTAnalyzer *analyzer, ClusterMaker *cMaker, HeuristicChain *hChain, FilterChain fChain)
 
     A simple helper method to show usage and clean up the temporarily
     created EST analyzer and cluster Maker objects.
@@ -68,10 +72,14 @@
 
     \param[in] hChain The chain of heuristics that are currently
     defined for this run.
+
+    \param[in] fChain The chain of filters that are currently defined
+    for this run.
 */
 static void showUsage(arg_parser& ap,
                       ESTAnalyzer *analyzer, ClusterMaker *cMaker,
-                      HeuristicChain *hChain) {
+                      HeuristicChain *hChain,
+                      FilterChain *fChain) {
     std::cout << "Usage: PEACE [options]\n"
               << "where options are:\n";
     std::cout << ap;
@@ -81,6 +89,8 @@ static void showUsage(arg_parser& ap,
     ClusterMakerFactory::displayList(std::cerr);
     std::cout << "Names of heuristics available are:\n";
     HeuristicFactory::displayList(std::cerr);
+    std::cout << "Names of filters available are:\n";
+    FilterFactory::displayList(std::cerr);
     // Display any analyzer specific options.
     if (analyzer != NULL) {
         analyzer->showArguments(std::cout);
@@ -95,6 +105,66 @@ static void showUsage(arg_parser& ap,
     if (hChain != NULL) {
         hChain->showArguments(std::cout);
         delete hChain;
+    }
+    // Display any filter specific options.
+    if (fChain != NULL) {
+        fChain->showArguments(std::cout);
+        delete fChain;
+    }
+}
+
+/** \fn int applyFilters(ClusterMaker *, FilterChain *, const char *, const char *)
+
+    \brief Helper method to apply filters (if any).
+
+    This method is a helper method that is invoked from the main
+    method to apply any filters that the user has sepecified to the
+    list of ESTs (before they are actually clustered).
+
[…]

### 2010-02-10 peace/papers/NAR_web_server_2009/paper.tex
Update of manuscript; addition of all files necessary for a pdflatex compile

--- 2010-02-09 papers/NAR_web_server_2009/paper.tex
@@
 \documentclass[a4,center,fleqn]{NAR}
-%\documentclass{article}
-\bibliographystyle{natbib}
-%\usepackage{graphicx}
+\usepackage[pdftex,
+colorlinks=true, %makes the links show up in color, rather than in a box
+citecolor=black, %color of the in-text citation numbers
+urlcolor=blue %color of the url links
+]{hyperref}
 
-\copyrightyear{2010}
-\pubdate{X}
-\pubyear{X}
-\jvolume{X}
-\jissue{X}
+\bibliographystyle{unsrt}
 
-\newlabel{d2_section}{A}
-\newlabel{mst_section}{B}
-\newlabel{sim_results}{C}
-\newlabel{real_results}{D}
-\newlabel{ESTsim_sup}{C.1}
-\newlabel{seq_time}{S4}
+% Enter dates of publication
+\copyrightyear{}
+\pubdate{}
+\pubyear{}
+\jvolume{}
+\jissue{}
 
-\newcommand{\wcd}{\textsc{wcd }}
-\newcommand{\capthree}{\textsc{cap3 }}
+%\articlesubtype{This is the article type (optional)}
 
@@
 
-\firstpage{1}
+\title{PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly
+  and {\underline C}lustering of Gene {\underline E}xpression}
 
-\title[PEACE cluster]{PEACE: {\underline P}arallel {\underline E}ST {\underline A}nalysis
-  and {\underline C}lustering {\underline E}ngine}
-\title{PEACE: {\underline P}arallel {\underline E}ST {\underline A}nalysis
-  and {\underline C}lustering {\underline E}ngine}
 \author{D.M. Rao\,$^{1}$, J.C. Moler\,$^{1}$, M. Ozden\,$^1$, Y. Zhang\,$^{1}$,
@@
     correspondence should be addressed}}
+
 \address{$^1$ Department of Computer Science and Software Engineering, \\
@@
 
-%\history{Received on XXXXX}
+\history{Received Feb. 12, 2010}
 
@@
   transcript fragment sequences produced Next Generation or Sanger
-  sequence technologies.  Installed and easily run through an easily
-  downloaded GUI, {\sc Peace} can process large data sets, group the
+  Sequencing technologies.  Installed and easily run through an easily
+  downloaded GUI, {\sc Peace} can process large data sets, grouping the
   fragments by gene association, faster and with greater sensitivity
   than competing clustering tools such as {\sc WCD}, and with
-  considerably better sensitivity then the implicit clustering
+  considerably greater sensitivity than the implicit clustering
   preformed by assembly tools such as {\sc Cap3}.  Through the GUI the
-  user can collect cluster statistics and easily examine specific
+  user can collect cluster statistics and examine specific
   clusters for more comprehensive study.  And by using {\sc Peace} as
   a ``pre-assembly'' tool to allow the application of an assembly tool
-  to the individual clusters, the user gains both the significantly
+  to the resultant individual clusters, the user gains both a significant
   increase sensitivity of the clustering method and a large decrease
-  in overall runtime for the clustering and assembly as compared to
-  the application of an assembly tool to the data as a whole.
+  in overall runtime for the clustering and assembly together.
 \end{abstract}
 
-\section {Introduction}
+\section{Introduction}
 
@@
 to gene discovery, provided information of protein isoforms, and
-generally help shed new light on the biological processes both
+generally helped shed light on the biological processes both
 controlling and dictated by the genome \cite{Nagaraj07}.  But in order
@@
 
-Given a set of transcript fragments sampled from multiple transcripts
-across the genome, a necessary first step is that of clustering:
-separating the fragments according to the transcript from which they
-were derived.  While clustering is frequently preformes implicity by
-assembly tools such as {\sc Cap3} \cite{Huang99}, clustering the data as a
+Until recently, a primary method of gaining access to a transcriptome
+sequence was through the use of Express Sequence Tags (ESTs), which
+were primarily single-pass cDNA sequences derived from transcribed
+mRNAs and traditionally sequenced by Sanger Sequencing technology.
+More recently, Next-Generation Sequence (NGS) technologies have
+replaced the Sanger Sequencing techniques, allowing for considerably
+greater depth and coverage of the transcriptome.  For example, ESTs in
+GenBank dbEST are increasingly the product of NGS technologies such as
+454 pyrosequencing, which enables the sequencing of novel and rate
+transcripts at a considerably higher rate of coverage than did Sanger
+Sequencing \cite{Cheung2006,Emrich2007}.  But from a computational
+perspective, this is a mixed blessing: while NGS provides immense
+quantities of new information, it does provides immensely larger data sets
+for analysis, requiring even more efficient algorithms than were
+already needed.
+
+Given a set of transcript fragments, produced by either sequencing
+technology, sampled from multiple transcripts across the genome, a
+necessary first step is that of clustering: separating the fragments
+according to the transcript from which they were derived.  While
+clustering is frequently preformed implicitly by assembly tools such
+as {\sc Cap3} \cite{Huang99}, clustering the data as a
 ``pre-assembly'' step has a number of advantages.  Most significant
 among these: preforming this step will allow the application of the
-assembly tool to individucal clusters -- saving significant amounts of
-time due to the reduced input size \cite{Hazelhurst08a}.  But clustering is
-a computationally challenging problem, especially when done {\it ab
-  initio} instead of relying on access to a sequenced genome.  Even
-with the number of fragments produced using Sanger sequencing, the
-runtime and memory requirements to cluster based on pair-wise sequence
-alignments make such an approach infeasible.  The much larger data-set
-sizes produced by the short-read technologies exacerbates this
-problem.
+assembly tool to individual clusters -- saving significant amounts of
+time due to the reduced input size \cite{Hazelhurst08a}.  But
+clustering is a computationally challenging problem, especially when
+done {\it ab initio} instead of relying on access to a sequenced
+genome.  Even with the number of ESTs produced using Sanger
+sequencing, the runtime and memory requirements to cluster based on
+pair-wise sequence alignments make such an approach infeasible.  The
+much larger data-set sizes produced by the next-generation sequencing
+technologies exacerbates this problem.
 
@@
 Sanger Sequencing technologies.  Available through the
-www.peace-tools.org website, the PEACE GUI allows the user
+\href{http://www.peace-tools.org}{www.peace-tools.org} website, the PEACE GUI allows the user
 to both easily install (locally or remotely) and run the computational
@@
 \cite{Prim57}.  The result is an easy-to-use clustering tool that is considerably
-more sensitive to sequencing errors than the competing \wcd tool
+more sensitive to sequencing errors than the competing {\sc wcd} tool
 \cite{Hazelhurst08a} and others in the literature without any
-sacrifice of performance (CITE) (e.g. \cite{Burke99}, \cite{Slater00},
+sacrifice of performance (e.g. \cite{Burke99}, \cite{Slater00},
 \cite{Huang99}, \cite{Parkinson02}, \cite{Kalyanaraman03},
 \cite{Malde03}, \cite{Ptitsyn05}, \cite{Picardi09}).
+
+\enlargethispage{-65.1pt}
 
@@
 The {\sc Peace} GUI, available from the {\sc Peace} website
-(www.peace-tools.org), handles the details of both instillation and
-job management and provides tools for the initial analysis of the
-results.  By downloading and executing the {\tt peace.jar} GUI Java
-program on a Linux, Windows or OS-X machine running Java 1.6+, the
-user is presented with the {\sc Peace} management tool.  Through this
-the user can install the computational tool and preform a clustering
-of a data file, view an initial analysis of the clusters and produce
-files containing subsets of the clusters for assembly by a tool such
-as {\sc Cap3} \cite{Huang99}.  A typical (first) use of the tool would
-generally require the the following steps.
+(\href{http://www.peace-tools.org}{www.peace-tools.org}), handles the
+details of both instillation and job management and provides tools for
+the initial analysis of the results.  By downloading and executing the
+{\tt peace.jar} GUI Java program on a Linux, Windows or OS-X machine
+running Java 1.6+, the user is presented with the {\sc Peace}
+management tool.  Through this the user can install the computational
+tool and preform a clustering of a data file, view an initial analysis
+of the clusters and produce files containing subsets of the clusters
+for assembly by a tool such as {\sc Cap3} \cite{Huang99}.  A typical
+(first) use of the tool would generally require the the following
+steps.
 
@@
   set up a {\it classifier}, associating certain patterns with
-  specific colors.  These patterns will be matched against the fragment
-  header information from the original fasta file, allowing the overlay of
-  a colored cluster size distribution does clusters containing matching elements in a
-  specified colors (e.g. if the user wished to restrict their view to
-  clusters containing transcript fragments from a specific source).  The method of
-  setting up a these classifiers, and the resulting histogram, is
-  illustrated in Figure~1(e).
+  specific colors.  These patterns will be matched against the
+  fragment header information from the original fasta file, allowing
+  the overlay of a colored cluster size distribution does clusters
+  containing matching elements in a specified colors.  For example, if
+  the sequence names contain unique string patterns denoting different
+  cDNA libraries, the classifier can help the user to determine the
+  differential expression profiles of different libraries for a given
+  cluster by color coding the strains.  The method of setting up a
+  these classifiers, and the resulting histogram, is illustrated in
+  Figure~1(e).
 \end{itemize}
@@
 calculating $d^2$, the {\it two-pass $d^2$ heuristic}, is a refinement
-of that used in the \wcd clustering tool \cite{Hazelhurst08a}.  We
+of that used in the {\sc wcd} clustering tool \cite{Hazelhurst08a}.  We
 augment the computation with a preliminary pass that narrows down the
@@
 calculation of a majority of edge distances using the $u/v$ and $t/v$
-filtering heuristics employed in \wcd \cite{Hazelhurst08a}.
+filtering heuristics employed in {\sc wcd} \cite{Hazelhurst08a}.
 These heuristics allow us to quickly dismiss many of the edges as too
 large without the need to apply the full $d^2$ algorithm (see
-Sections~\ref{d2_section} and \ref{mst_section} of the Supplementary
-Materials for more details).
+Sections~A and B of the Supplementary Materials for more details).
 
@@
 We have tested {\sc Peace} on both simulated and real data,
-comparing results against those produced by the \wcd
-clustering tool \cite{Hazelhurst08a} and the \capthree assembly
+comparing results against those produced by the {\sc wcd}
+clustering tool \cite{Hazelhurst08a} and the {\sc Cap3} assembly
 tool \cite{Huang99} (which calculates a cluster as an intermediate
@@
 \cite{Hazelhurst03} to generate simulated transcript fragments under varying models
-of error (Supplementary Materials, Section~\ref{ESTsim_sup}),
+of error (Supplementary Materials, Section~\href{ESTsim_sup}),
 generating the fragments from the list of 100 zebra fish genes used in
-the \wcd testing \cite{Hazelhurst08a}.  Tool parameters were
-taken to match, as closely as possible, those used in the \wcd
+the {\sc wcd} testing \cite{Hazelhurst08a}.  Tool parameters were
+taken to match, as closely as possible, those used in the {\sc wcd}
 study (see supplementary materials).  Our primary methods of quality
@@
 in the supplementary materials.  In Figure~2(left) we see a
-significant improvement of {\sc Peace} (blue) over \wcd
-(green) and \capthree (black) in its sensitivity to sequencing
+significant improvement of {\sc Peace} (blue) over {\sc wcd}
+(green) and {\sc Cap3} (black) in its sensitivity to sequencing
 errors, while in Figure~2(right) we see a comparable
@@
 improvement in {\sc Peace}.  The sequential runtime of
-{\sc Peace} is slightly improved over that of \wcd (see
+{\sc Peace} is slightly improved over that of {\sc wcd} (see
 Supplementary Materials, Figure~S4).
@@
 (d) basic cluster output; and (e) histogram view of cluster results
-and classifier editor.}
+and classifier editor for setting up differential expression profiles.}
 \end{figure*}
@@
   average over 30 Simulated fragment sets derived from 100 zebra fish genes
-  (see Supplementary Materials, Section~\ref{sim_results}, for more
+  (see Supplementary Materials, Section~\href{sim_results}, for more
   details).   Blue/Solid = {\sc Peace}, Green/Dash =
-  \wcd, Black/Dot-Dash = \capthree; vertical tics = 95\% confidence
+  {\sc wcd}, Black/Dot-Dash = {\sc Cap3}; vertical tics = 95\% confidence
   intervals on estimates.  Intervals are not presented for Type 1
@@
   when run on the simulated sequences; correct answer in all cases is
-  zero.  Blue/Solid = {\sc Peace}, Green/Dash = \wcd,
-  Black/Dot-Dash = \capthree.}
+  zero.  Blue/Solid = {\sc Peace}, Green/Dash = {\sc wcd},
+  Black/Dot-Dash = {\sc Cap3}.}
 \end{figure*}
@@
 In applying the tools to real data, we started with a set of
-approximately $190,000$ fragments derived from the {\it Chlamydomonas
-  reinhardtii} genome.  To compute sensitivity we used the {\it gmap}
-tool \cite{Wu05} to map the fragment set to the genome, taking this as
-our reference clustering.  We see slight improvements in
-{\sc Peace} over \wcd in both sensitivity and Type 1 error
-(both significantly out preforming \capthree).  Using the mouse fragment
-set used in Hazelhurst {\it et al.}  \cite{Hazelhurst08a}
-{\sc Peace} again shows slight improvements in sensitivity and Type
-1 error rates, with an 18\% speed up for {\sc Peace} (see
-Supplementary Materials, Section~\ref{real_results}, for more
-details).
+approximately $190,000$ clean Sanger ESTs derived from the {\it
+  Chlamydomonas reinhardtii} genome \cite{Liang2008}.  To compute
+sensitivity we used the {\it gmap} tool \cite{Wu05} to map the
+fragment set to the genome, taking this as our reference clustering.
+We see slight improvements in {\sc Peace} over {\sc wcd} in both
+sensitivity and Type 1 error (both significantly out preforming {\sc
+  Cap3}).  Using the mouse fragment set used in Hazelhurst {\it et
+  al.}  \cite{Hazelhurst08a} {\sc Peace} again shows slight
+improvements in sensitivity and Type 1 error rates, with an 18\% speed
+up for {\sc Peace} (see Supplementary Materials,
+Section~\href{real_results}, for more details).
 
@@
 % LocalWords:  reinhartdii conifergdb DNAc bp wcd reinhardtii PhRMA Iddo fasta
-% LocalWords:  Friedberg Scoville OpenMPI initio multi Screenshots
+% LocalWords:  Friedberg Scoville OpenMPI initio multi Screenshots nvironment
+% LocalWords:  ssembly xpression GenBank dbEST pyrosequencing ESTsim sim

### 2010-02-10 peace/src/main.cpp
Added missing return statement at end of applyFilters method

@@ -166,6 +166,7 @@ int applyFilters(ClusterMaker *clusterMaker, FilterChain *chain,
         EST::dumpESTList(outFile, true);
         outFile.close();
     }
+    return 0;
 }
 
 /** \fn int main(int argc, char *argv[])

### 2010-02-10 peace/src/TwoPassD2.h
Just some minor indentation and formatting changes.

@@ -144,7 +144,7 @@ public:
         Interactive Console).
 
         \return This method returns the string "layeredD2" identifiying 
-	this analyzer.
+        this analyzer.
     */    
     virtual std::string getName() const { return "layeredD2"; }
     
@@ -178,11 +178,11 @@ public:
         merely calls the corresponding base class implementation that
         performs all the necessary operations.
 
-	\note This method was introduced to avoid the unnecessary
-	warning about partial overloading of virtual methods (warning
-	#654) in ICC.
+        \note This method was introduced to avoid the unnecessary
+        warning about partial overloading of virtual methods (warning
+        #654) in ICC.
 	
-	\return This method returns zero if all the processing
+        \return This method returns zero if all the processing
         proceeded successfully. On errors this method returns a
         non-zero value.
     */
@@ -306,28 +306,28 @@ protected:
     }
 
     /** Instance variable that maps an index in the reference sequence
-	(sequence s1) to the hash of a word.  This hash can then be used
-	as an index in the fdHashMap to get the frequency differential
-	for that word.
+        (sequence s1) to the hash of a word.  This hash can then be used
+        as an index in the fdHashMap to get the frequency differential
+        for that word.
 
-	The word table is created in the initialize() method and
-	filled in using the buildWordTable() method.  For the reference
-	EST, buildWordTable() is called in the setReferenceEST() method,
-	meaning it does not need to be rebuilt every time we analyze
-	a new comparison sequence.
+        The word table is created in the initialize() method and
+        filled in using the buildWordTable() method.  For the reference
+        EST, buildWordTable() is called in the setReferenceEST() method,
+        meaning it does not need to be rebuilt every time we analyze
+        a new comparison sequence.
     */
     int* s1WordTable;
 
     /** Instance variable that maps an index in the comparison sequence
-	(sequence s2) to the hash of a word.
+        (sequence s2) to the hash of a word.
 
         This hash is used as an index in the fdHashMap to get the
-	frequency differential for that word.
+        frequency differential for that word.
 
-	The word table is created in the initialize() method and
-	filled in using the buildWordTable() method.  For the comparison
-	EST, buildWordTable() must be called in the analyze() method because
-	a new comparison sequence is given every time analyze() is called.
+        The word table is created in the initialize() method and
+        filled in using the buildWordTable() method.  For the comparison
[…]

### 2010-02-10 src/TwoPassD2.cpp
Just some minor indentation and formatting changes.

@@ -271,12 +271,11 @@ TwoPassD2::runD2Asymmetric(const int otherEST, int* s1MinScoreIdx,
         // Break out of this loop if we have found a potential match
         if (minScore <= threshold) {
             break;
-	}
+        }
         
         // Check every window in EST #2 against current window in EST
         // #1 by sliding EST #2 window to left.
-        for(s2Win = sq2End - wordSize; (s2Win > FirstWindowInSq2);
-            s2Win--) {
+        for(s2Win = sq2End - wordSize; (s2Win > FirstWindowInSq2); s2Win--) {
             // The word at s2Win - numWordsInWindow is moving in while
             // the word at s2Win is moving out as we move window
             // associated with EST #2 from right-to-left.

### 2010-02-11 papers/NAR_web_server_2009/paper.tex
Fixed typographical errors and misspellings, and made minor grammatical revisions to the paper.

@@
   We present {\sc Peace}, a tool for the high-throughput clustering of
-  transcript fragment sequences produced Next Generation or Sanger
-  sequence technologies.  Installed and easily run through an easily
-  downloaded GUI, {\sc Peace} can process large data sets, group the
+  transcript fragment sequences produced by Next Generation or Sanger
+  Sequencing technologies.  Installed and run through an easily
+  downloaded GUI, {\sc Peace} can process large data sets, grouping the
   fragments by gene association, faster and with greater sensitivity
   than competing clustering tools such as {\sc WCD}, and with
-  considerably better sensitivity then the implicit clustering
-  preformed by assembly tools such as {\sc Cap3}.  Through the GUI the
-  user can collect cluster statistics and easily examine specific
+  considerably greater sensitivity than the implicit clustering
+  performed by assembly tools such as {\sc Cap3}.  Through the GUI the
+  user can collect cluster statistics and examine specific
   clusters for more comprehensive study.  And by using {\sc Peace} as
   a ``pre-assembly'' tool to allow the application of an assembly tool
-  to the individual clusters, the user gains both the significantly
-  increase sensitivity of the clustering method and a large decrease
-  in overall runtime for the clustering and assembly as compared to
-  the application of an assembly tool to the data as a whole.
+  to the resultant individual clusters, the user gains both a significant
+  increase in sensitivity of the clustering method and a large decrease
+  in overall runtime for the clustering and assembly together.
 \end{abstract}
 
-\section {Introduction}
+\section{Introduction}
 
@@
 transcripts expressed by genes of the organism, is a vital step in
-understanding the full functional and organization roll of the genome
+understanding the full functional and organizational role of the genome
 in the life cycle of any organism.  Studying the transcriptome has led
-to gene discovery, provided information of protein isoforms, and
-generally help shed new light on the biological processes both
+to gene discovery, provided information on protein isoforms, and
+generally helped shed light on the biological processes both
 controlling and dictated by the genome \cite{Nagaraj07}.  But in order
-to get access those transcripts, we must first deal with the
+to get access to those transcripts, we must first deal with the
 fragmented data produced by both Next Generation and traditional
@@
 
-Given a set of transcript fragments sampled from multiple transcripts
-across the genome, a necessary first step is that of clustering:
-separating the fragments according to the transcript from which they
-were derived.  While clustering is frequently preformes implicity by
-assembly tools such as {\sc Cap3} \cite{Huang99}, clustering the data as a
+Until recently, a primary method of gaining access to a transcriptome
+sequence was through the use of Expressed Sequence Tags (ESTs), which
+were primarily single-pass cDNA sequences derived from transcribed
+mRNAs and traditionally sequenced by Sanger Sequencing technology.
+More recently, Next-Generation Sequencing (NGS) technologies have
+replaced the Sanger Sequencing techniques, allowing for considerably
+greater depth and coverage of the transcriptome.  For example, ESTs in
+GenBank dbEST are increasingly the product of NGS technologies such as
+454 pyrosequencing, which enables the sequencing of novel and rate
+transcripts at a considerably higher rate of coverage than did Sanger
+Sequencing \cite{Cheung2006,Emrich2007}.  But from a computational
+perspective, this is a mixed blessing: while NGS provides immense
+quantities of new information, it also provides immensely larger data sets,
+requiring even more efficient algorithms than were already needed
+for analysis.
+
+Given a set of transcript fragments, produced by either sequencing
+technology, sampled from multiple transcripts across the genome, a
+necessary first step is that of clustering: separating the fragments
+according to the transcript from which they were derived.  While
+clustering is frequently performed implicitly by assembly tools such
+as {\sc Cap3} \cite{Huang99}, clustering the data as a
 ``pre-assembly'' step has a number of advantages.  Most significant
-among these: preforming this step will allow the application of the
-assembly tool to individucal clusters -- saving significant amounts of
-time due to the reduced input size \cite{Hazelhurst08a}.  But clustering is
-a computationally challenging problem, especially when done {\it ab
-  initio} instead of relying on access to a sequenced genome.  Even
-with the number of fragments produced using Sanger sequencing, the
-runtime and memory requirements to cluster based on pair-wise sequence
-alignments make such an approach infeasible.  The much larger data-set
-sizes produced by the short-read technologies exacerbates this
-problem.
+among these: performing this step will allow the application of the
+assembly tool to individual clusters -- saving significant amounts of
+time due to the reduced input size \cite{Hazelhurst08a}.  But
+clustering is a computationally challenging problem, especially when
+done {\it ab initio} instead of relying on access to a sequenced
+genome.  Even with the smaller number of ESTs produced using Sanger
+sequencing, the runtime and memory requirements to cluster based on
+pair-wise sequence alignments make such an approach infeasible.  The
+much larger data-set size produced by next-generation sequencing
+technologies exacerbates this problem.
 
@@
 \cite{Prim57}.  The result is an easy-to-use clustering tool that is considerably
-more sensitive to sequencing errors than the competing \wcd tool
+more sensitive to sequencing errors than the competing {\sc wcd} tool
 \cite{Hazelhurst08a} and others in the literature without any
-sacrifice of performance (CITE) (e.g. \cite{Burke99}, \cite{Slater00},
+sacrifice of performance (e.g. \cite{Burke99}, \cite{Slater00},
 \cite{Huang99}, \cite{Parkinson02}, \cite{Kalyanaraman03},
@@
 
-\section{PEACE: Instillation and Use}
+\enlargethispage{-65.1pt}
+
+\section{PEACE: Installation and Use}
 
 The {\sc Peace} GUI, available from the {\sc Peace} website
-(www.peace-tools.org), handles the details of both instillation and
-job management and provides tools for the initial analysis of the
-results.  By downloading and executing the {\tt peace.jar} GUI Java
-program on a Linux, Windows or OS-X machine running Java 1.6+, the
-user is presented with the {\sc Peace} management tool.  Through this
-the user can install the computational tool and preform a clustering
-of a data file, view an initial analysis of the clusters and produce
-files containing subsets of the clusters for assembly by a tool such
-as {\sc Cap3} \cite{Huang99}.  A typical (first) use of the tool would
-generally require the the following steps.
+(\href{http://www.peace-tools.org}{www.peace-tools.org}), handles the
+details of both installation and job management and provides tools for
+the initial analysis of the results.  By downloading and executing the
+{\tt peace.jar} GUI Java program on a Linux, Windows or OS-X machine
+running Java 1.6+, the user is presented with the {\sc Peace}
+management tool.  Through this the user can install the computational
+tool and perform a clustering of a data file, view an initial analysis
+of the clusters and produce files containing subsets of the clusters
+for assembly by a tool such as {\sc Cap3} \cite{Huang99}.  A typical
+(first) use of the tool would generally require the the following
+steps.
 
-\noindent {\bf Tool Instillation:} Installing the GUI requires only
+\noindent {\bf Tool Installation:} Installing the GUI requires only
 that the Java file be downloaded from the website.  To install the {\sc
@@
 \item {\bf Export:} The user can export the contents of one or more
-  clusters into a fasta file, obtaining a subset of the original file
+  clusters into a FASTA format file, obtaining a subset of the original file
   containing the sequences corresponding to the selected clusters
@@
   set up a {\it classifier}, associating certain patterns with
-  specific colors.  These patterns will be matched against the fragment
-  header information from the original fasta file, allowing the overlay of
-  a colored cluster size distribution does clusters containing matching elements in a
-  specified colors (e.g. if the user wished to restrict their view to
-  clusters containing transcript fragments from a specific source).  The method of
-  setting up a these classifiers, and the resulting histogram, is
-  illustrated in Figure~1(e).
+  specific colors.  These patterns will be matched against the
+  fragment header information from the original FASTA file, allowing
+  the overlay of a colored cluster size distribution does clusters
+  containing matching elements in a specified colors (comment by molerjc: please reword the preceding sentence fragment).  For example, if
+  the sequence names contain unique string patterns denoting different
+  cDNA libraries, the classifier can help the user to determine the
+  differential expression profiles of different libraries for a given
+  cluster by color coding the strains.  The method of setting up a
+  these classifiers, and the resulting histogram, is illustrated in
+  Figure~1(e).
 \end{itemize}
@@
 \caption{Screenshots of the PEACE GUI during execution, including (a)
-GUI Welcome and server instillation menu; (b) setup wizard for
+GUI Welcome and server installation menu; (b) setup wizard for
 installing the computational tool on a remote server; (c) execution
@@
 \caption{Comparisons of Sensitivity and Type 1 error, based on the
-  average over 30 Simulated fragment sets derived from 100 zebra fish genes
-  (see Supplementary Materials, Section~\ref{sim_results}, for more
+  average over 30 simulated fragment sets derived from 100 zebra fish genes
+  (see Supplementary Materials, Section~\href{sim_results}, for more
   details).   Blue/Solid = {\sc Peace}, Green/Dash =
-  \wcd, Black/Dot-Dash = \capthree; vertical tics = 95\% confidence
+  {\sc wcd}, Black/Dot-Dash = {\sc Cap3}; vertical tics = 95\% confidence
   intervals on estimates.  Intervals are not presented for Type 1
@@
 In applying the tools to real data, we started with a set of
-approximately $190,000$ fragments derived from the {\it Chlamydomonas
-  reinhardtii} genome.  To compute sensitivity we used the {\it gmap}
-tool \cite{Wu05} to map the fragment set to the genome, taking this as
-our reference clustering.  We see slight improvements in
-{\sc Peace} over \wcd in both sensitivity and Type 1 error
-(both significantly out preforming \capthree).  Using the mouse fragment
-set used in Hazelhurst {\it et al.}  \cite{Hazelhurst08a}
-{\sc Peace} again shows slight improvements in sensitivity and Type
-1 error rates, with an 18\% speed up for {\sc Peace} (see
-Supplementary Materials, Section~\ref{real_results}, for more
-details).
+approximately $190,000$ clean Sanger ESTs derived from the {\it
+  Chlamydomonas reinhardtii} genome \cite{Liang2008}.  To compute
+sensitivity we used the {\it gmap} tool \cite{Wu05} to map the
+fragment set to the genome, taking this as our reference clustering.
+We see slight improvements in {\sc Peace} over {\sc wcd} in both
+sensitivity and Type 1 error (both significantly outperforming {\sc
+  Cap3}).  Using the mouse fragment set used in Hazelhurst {\it et
+  al.}  \cite{Hazelhurst08a} {\sc Peace} again shows slight
+improvements in sensitivity and Type 1 error rates, with an 18\% speedup
+for {\sc Peace} (see Supplementary Materials,
+Section~\href{real_results}, for more details).
 

### 2010-02-12 peace/papers/NAR_web_server_2009/paper.tex
Modifications to manuscript: Conclusion added; corrections made

@@
 \begin{abstract}
-  We present {\sc Peace}, a tool for the high-throughput clustering of
-  transcript fragment sequences produced Next Generation or Sanger
-  Sequencing technologies.  Installed and easily run through an easily
-  downloaded GUI, {\sc Peace} can process large data sets, grouping the
-  fragments by gene association, faster and with greater sensitivity
-  than competing clustering tools such as {\sc WCD}, and with
-  considerably greater sensitivity than the implicit clustering
-  preformed by assembly tools such as {\sc Cap3}.  Through the GUI the
-  user can collect cluster statistics and examine specific
-  clusters for more comprehensive study.  And by using {\sc Peace} as
-  a ``pre-assembly'' tool to allow the application of an assembly tool
-  to the resultant individual clusters, the user gains both a significant
-  increase sensitivity of the clustering method and a large decrease
-  in overall runtime for the clustering and assembly together.
+  We present {\sc Peace}, a standalone tool for the high-throughput
+  {\it ab initio} clustering of transcript fragment sequences produced by Next
+  Generation or Sanger Sequencing technologies.  Installed and run
+  through an easily downloaded GUI, {\sc Peace} can process large data
+  sets and group the fragments by gene association, achieving a
+  greater sensitivity (without sacrificing runtime) than the competing
+  clustering tool {\sc WCD}.  Through the GUI the user can collect
+  cluster statistics and examine specific clusters for more
+  comprehensive study.  By using {\sc Peace} as a ``pre-assembly''
+  tool to allow the application of an assembly tool to the resultant
+  individual clusters, the user gains both a significant increase in
+  sensitivity and a large decrease in overall runtime for the
+  clustering and assembly together.
 \end{abstract}
@@
 transcripts expressed by genes of the organism, is a vital step in
-understanding the full functional and organization roll of the genome
-in the life cycle of any organism.  Studying the transcriptome has led
-to gene discovery, provided information of protein isoforms, and
-generally helped shed light on the biological processes both
-controlling and dictated by the genome \cite{Nagaraj07}.  But in order
-to get access those transcripts, we must first deal with the
-fragmented data produced by both Next Generation and traditional
-Sanger sequencing technology.
+understanding the full functional and organizational role of the
+genome in the life cycle of any organism.  Studying the transcriptome
+has led to gene discovery, provided information on splice variants,
+and helped shed light on the biological processes both controlling and
+dictated by the genome \cite{Nagaraj07}.  But to access to those
+transcripts, we must deal with the fragmented data produced by
+both Next Generation and traditional Sanger sequencing technology.
 
-Until recently, a primary method of gaining access to a transcriptome
-sequence was through the use of Express Sequence Tags (ESTs), which
-were primarily single-pass cDNA sequences derived from transcribed
-mRNAs and traditionally sequenced by Sanger Sequencing technology.
-More recently, Next-Generation Sequence (NGS) technologies have
-replaced the Sanger Sequencing techniques, allowing for considerably
-greater depth and coverage of the transcriptome.  For example, ESTs in
-GenBank dbEST are increasingly the product of NGS technologies such as
-454 pyrosequencing, which enables the sequencing of novel and rate
-transcripts at a considerably higher rate of coverage than did Sanger
-Sequencing \cite{Cheung2006,Emrich2007}.  But from a computational
+Until recently, access to a transcriptome sequence was primarily
+through the use of Expressed Sequence Tags (ESTs), single-pass cDNA
+sequences derived from transcribed mRNAs and sequenced by Sanger
+Sequencing technology.  More recently, Next Generation Sequencing
+(NGS) technologies have replaced the Sanger Sequencing techniques,
+allowing for considerably coverage of the transcriptome.  For example,
+ESTs in GenBank dbEST are increasingly the product of NGS technologies
+such as 454 pyrosequencing, which enables the sequencing of novel and
+rare transcripts at a considerably higher rate of coverage than did
+Sanger Sequencing \cite{Cheung2006,Emrich2007}.  From a computational
 perspective, this is a mixed blessing: while NGS provides immense
-quantities of new information, it does provides immensely larger data sets
-for analysis, requiring even more efficient algorithms than were
-already needed.
+quantities of new information, it also provides immensely larger data
+sets, requiring fast computational techniques if we are to process
+them.
 
-Given a set of transcript fragments, produced by either sequencing
-technology, sampled from multiple transcripts across the genome, a
-necessary first step is that of clustering: separating the fragments
-according to the transcript from which they were derived.  While
-clustering is frequently preformed implicitly by assembly tools such
-as {\sc Cap3} \cite{Huang99}, clustering the data as a
-``pre-assembly'' step has a number of advantages.  Most significant
-among these: preforming this step will allow the application of the
-assembly tool to individual clusters -- saving significant amounts of
-time due to the reduced input size \cite{Hazelhurst08a}.  But
-clustering is a computationally challenging problem, especially when
-done {\it ab initio} instead of relying on access to a sequenced
-genome.  Even with the number of ESTs produced using Sanger
+Given a set of transcript fragments sampled from across the genome, a
+necessary first step in the analysis is that of clustering: separating
+the fragments according to the transcript from which they were
+derived.  Frequently performed implicitly by assembly tools,
+clustering the data as a ``pre-assembly'' step has a number of
+advantages.  Most significant among these: performing this step will
+allow the application of the assembly tool to individual clusters --
+saving significant amounts of time due to the reduced input size
+\cite{Hazelhurst08a}.  But clustering is a computationally challenging
+problem.  Even with the smaller number of ESTs produced using Sanger
 sequencing, the runtime and memory requirements to cluster based on
 pair-wise sequence alignments make such an approach infeasible.  The
-much larger data-set sizes produced by the next-generation sequencing
+much larger data-set size produced by Next-Generation sequencing
 technologies exacerbates this problem.
@@
 Here we present {\sc Peace}, a tool for the {\it ab initio} clustering
-of transcript fragments by gene, applicable to both short-read and traditional
-Sanger Sequencing technologies.  Available through the
-\href{http://www.peace-tools.org}{www.peace-tools.org} website, the PEACE GUI allows the user
-to both easily install (locally or remotely) and run the computational
-tool, as well as providing various tools for result analysis. {\sc
-  Peace} can be run in single or multi-processor mode, and requires no
-Internet connectivity past the initial download.  Using a {\it minimum
-  spanning tree} (MST) based clustering strategy \cite{Jain99,Wan08}
-and the $d^2$ sequence distance function \cite{Hide94}, {\sc Peace}
-generates the MSTs using a tailored version of Prim's algorithm
-\cite{Prim57}.  The result is an easy-to-use clustering tool that is considerably
-more sensitive to sequencing errors than the competing {\sc wcd} tool
-\cite{Hazelhurst08a} and others in the literature without any
-sacrifice of performance (e.g. \cite{Burke99}, \cite{Slater00},
-\cite{Huang99}, \cite{Parkinson02}, \cite{Kalyanaraman03},
-\cite{Malde03}, \cite{Ptitsyn05}, \cite{Picardi09}).
+of transcript fragments by gene, applicable to both short-read and
+traditional Sanger Sequencing technologies.  Available through the
+\href{http://www.peace-tools.org}{www.peace-tools.org} website, the
+PEACE GUI allows the user to both easily install (locally or remotely)
+and run the computational tool, as well as enabling transparent
+parallel processing and providing various tools for result analysis.
+Using a {\it minimum spanning tree} (MST) based clustering strategy
+\cite{Jain99,Wan08} and the $d^2$ sequence distance function
+\cite{Hide94}, {\sc Peace} generates the MSTs using a tailored version
+of Prim's algorithm \cite{Prim57}.  The result is an easy-to-use and
+clustering tool that , without any sacrifice in runtime, is
+considerably more sensitive and robust to sequencing errors than other
+clusters tools in the literature \cite{Burke99,Slater00,
+  Huang99,Parkinson02,Kalyanaraman03,Malde03,Ptitsyn05,Hazelhurst08a, Picardi09}.
 
@@
 
-\section{PEACE: Instillation and Use}
+\section{PEACE: Installation and Use}
 
@@
 (\href{http://www.peace-tools.org}{www.peace-tools.org}), handles the
-details of both instillation and job management and provides tools for
+details of both installation and job management and provides tools for
 the initial analysis of the results.  By downloading and executing the
@@
 management tool.  Through this the user can install the computational
-tool and preform a clustering of a data file, view an initial analysis
+tool and perform a clustering of a data file, view an initial analysis
 of the clusters and produce files containing subsets of the clusters
@@
 
-\noindent {\bf Tool Instillation:} Installing the GUI requires only
-that the Java file be downloaded from the website.  To install the {\sc
-  Peace} tool onto a local or remote machine, the user will select
-from within the GUI the {\tt Server > Add New Server} tab from the GUI
-main menu (Figure~1(a)), which then starts the install
-wizard that will prompt for the appropriate information.
-Figure~1(b) illustrates the request for server information;
-in this example, the user has chosen to install the {\sc Peace}
+\noindent {\bf Tool Installation:} Installing the GUI requires only
+that the Java file be downloaded from the website.  To install the
+{\sc Peace} tool onto a local or remote machine, the user will select
+from within the GUI the appropriate menu tab (Figure~1(a)), which
+then starts the install wizard that will prompt for the appropriate
+information.  Figure~1(b) illustrates the request for server
+information; the user has chosen to install the {\sc Peace}
 computational tool on a remote machine and is providing the necessary
@@
 \noindent {\bf Job Processing:} After importing the target file into
-the GUI ({\tt File > New Data Set}), the user can start a new job with
-the {\tt Job > Job to Complete Clusters} tab and follow the wizard
-menus.  Figure~1(c) illustrates the process of specifying
-the number of processors available (if running on a machine supporting
-the OpenMPI protocol -- which will be determined during job
-instillation).  Once executed, the GUI will manage the job thread,
-alert the user when the job is completed (or when the user next
-runs the GUI after completion) and 
-copy the final results back to the local machine if necessary.
+the GUI, the user can start a new job and follow the wizard menus.
+Figure~1(c) illustrates the process of specifying the number of
+processors available (if running on a machine supporting the OpenMPI
+protocol -- which will be determined during job installation).  Once
+executed, the GUI will manage the job thread, alert the user when the
+job is completed (or when the user next runs the GUI after completion)
+and copy the final results back to the local machine if necessary.
 
 \noindent {\bf Result Analysis:} Once the resulting clusters have been
-computed and (if necessary) copied back, the user has several options
+computed, the user has several options
 for analysis:
@@
 \item {\bf Export:} The user can export the contents of one or more
-  clusters into a fasta file, obtaining a subset of the original file
+  clusters into a FASTA format file, obtaining a subset of the original file
   containing the sequences corresponding to the selected clusters
   ready for processing by an assembly tool (e.g. {\sc Cap3} \cite{Huang99}).
-\item {\bf View Clustering}: The user may view a list of all clusters,
-  expanding selected clusters to a list of all fragments (illustrated in
-  Figure~1(d)).  
+\item {\bf View Clustering}: The user may view a list of clusters,
+  expanding selected clusters to a list of all fragments (illustrated
+  in Figure~1(d)).
 \item {\bf Classified Summary Graph:} The user may view a distribution
-  of cluster sizes in a linear or $\log$ scale.  Further, the user may
-  set up a {\it classifier}, associating certain patterns with
-  specific colors.  These patterns will be matched against the
-  fragment header information from the original fasta file, allowing
-  the overlay of a colored cluster size distribution does clusters
-  containing matching elements in a specified colors.  For example, if
-  the sequence names contain unique string patterns denoting different
-  cDNA libraries, the classifier can help the user to determine the
-  differential expression profiles of different libraries for a given
-  cluster by color coding the strains.  The method of setting up a
-  these classifiers, and the resulting histogram, is illustrated in
-  Figure~1(e).
+  of cluster sizes.  Further, the user may set up a {\it classifier},
+  associating certain patterns with specific colors.  These patterns
+  will be matched against the fragment header information from the
+  original FASTA file, allowing the overlay of a colored cluster size
+  distributions.  For example, if the sequence names contain unique
+  string patterns denoting different cDNA libraries, the classifier
+  can help the user to determine the differential expression profiles
+  of different libraries for a given cluster by color coding the
+  strains.  The method of setting up a these classifiers, and the
+  resulting histogram, is illustrated in Figure~1(e).
 \end{itemize}
-
 Extensive documentation for the tool has been posted on the {\sc
@@
 alignment-free measurement of sequence distance that can be calculated
-significantly faster than a Smith-Waterman alignment
-\cite{Hide94,Foret09}.  $d^2$ works by comparing the frequency of
-words (strings of a fixed length) appearing in a limited region of
-each string.  Fragments overlapping by a sufficient length will share
-neighborhoods of enough similarity to ensure a small distance even in
-the presence of a moderate number of base errors.  Our method of
-calculating $d^2$, the {\it two-pass $d^2$ heuristic}, is a refinement
-of that used in the {\sc wcd} clustering tool \cite{Hazelhurst08a}.  We
-augment the computation with a preliminary pass that narrows down the
-two regions of maximum similarity, and then explore those local
-regions in more detail.
+significantly faster than a Smith-Waterman alignment \cite{Hide94}.
+$d^2$ works by comparing the frequency of words (strings of a fixed
+length) appearing in a limited region of each string.  Fragments
+overlapping by a sufficient length will share neighborhoods of enough
+similarity to ensure a small distance even in the presence of a
+moderate number of base errors.
 
@@
 step).  For our simulation tests we used the {\bf ESTSim} tool
-\cite{Hazelhurst03} to generate simulated transcript fragments under varying models
-of error (Supplementary Materials, Section~\href{ESTsim_sup}),
+\cite{Hazelhurst03} to generate simulated transcript of varying length
+fragments under different models
+of error (Supplementary Materials, Section~C.1),
 generating the fragments from the list of 100 zebra fish genes used in
@@
 \caption{Comparisons of Sensitivity and Type 1 error, based on the
-  average over 30 Simulated fragment sets derived from 100 zebra fish genes
+  average over 30 simulated fragment sets derived from 100 zebra fish genes
   (see Supplementary Materials, Section~\href{sim_results}, for more
@@
 We see slight improvements in {\sc Peace} over {\sc wcd} in both
-sensitivity and Type 1 error (both significantly out preforming {\sc
+sensitivity and Type 1 error (both significantly outperforming {\sc
   Cap3}).  Using the mouse fragment set used in Hazelhurst {\it et
   al.}  \cite{Hazelhurst08a} {\sc Peace} again shows slight
-improvements in sensitivity and Type 1 error rates, with an 18\% speed
-up for {\sc Peace} (see Supplementary Materials,
+improvements in sensitivity and Type 1 error rates, with an 18\% speedup
+for {\sc Peace} (see Supplementary Materials,
 Section~\href{real_results}, for more details).
+
+\section{Conclusions}
+
+Here we have presented {\sc Peace}, a stand-alone tool for the
+high-throughput clustering of transcript fragments, managed through a
+user friendly GUI that enables both local and remote instillation and
+execution in sequential or parallel mode.  Based on a novel algorithm
+for the clustering of the fragments by gene association, {\sc Peace}
+shows significant improvement in sensitivity while increasing runtime
+over the competing {\sc WCD} tool \cite{Hazelhurst08a}, and shows an
+order of magnitude in improvement over the clustering preformed in the
+course of assembly by the {\sc Cap3} tool \cite{Huang99}.
+
+As a clustering tool based on sequence distance, {\sc Peace} faces
+certain inherent limitations. For example, {\sc Peace} cannot handle
+duplicate genes; like {\sc WCD}, it is unable to separate clusters
+corresponding to genes with a greater than 88\% similarity.
+Similarly, naturally biological effects such as the trans-splicing of
+transcripts, or effects from poorly cleaned transcript data (e.g. the
+failure to remove polyA tailed introduced in the closing process) can
+cause similar effects in these clustering tools.  The problems can be
+handled through the application of the assembler, and the ability to
+apply any assembler to small cluster (as opposed to the data set as a
+while) results in a significant reduction in overall assembly time.
+
+In short: {\sc Peace} can out-preform existing fragment cluster tools
+on fragment data.  By using the {\sc Peace} GUI, the user can easily
+set up a first analysis of transcript distributions and set the user
+up for a significant reduction in the overall time for assembly a
+transcript from fragment transcript sequences.
 
@@
 % LocalWords:  Friedberg Scoville OpenMPI initio multi Screenshots nvironment
-% LocalWords:  ssembly xpression GenBank dbEST pyrosequencing ESTsim sim
+% LocalWords:  ssembly xpression GenBank dbEST pyrosequencing ESTsim sim mRNAs
+% LocalWords:  NGS polyA

### 2010-02-12 peace/papers/NAR_web_server_2009/paper.tex
Corrections to manuscript, as found by Jenna

@@
 high-throughput clustering of transcript fragments, managed through a
-user friendly GUI that enables both local and remote instillation and
+user friendly GUI that enables both local and remote installation and
 execution in sequential or parallel mode.  Based on a novel algorithm
 for the clustering of the fragments by gene association, {\sc Peace}
-shows significant improvement in sensitivity while increasing runtime
+shows significant improvement in sensitivity, without sacrificing  runtime,
 over the competing {\sc WCD} tool \cite{Hazelhurst08a}, and shows an
@@
 while) results in a significant reduction in overall assembly time.
-
-In short: {\sc Peace} can out-preform existing fragment cluster tools
-on fragment data.  By using the {\sc Peace} GUI, the user can easily
-set up a first analysis of transcript distributions and set the user
-up for a significant reduction in the overall time for assembly a
-transcript from fragment transcript sequences.
 

### 2010-02-12 papers/NAR_web_server_2009/paper.tex
Incorperation of comments on manuscript from Mufit and Chun

@@
 \begin{abstract}
-  We present {\sc Peace}, a tool for the high-throughput clustering of
-  transcript fragment sequences produced by Next Generation or Sanger
-  Sequencing technologies.  Installed and run through an easily
-  downloaded GUI, {\sc Peace} can process large data sets, grouping the
-  fragments by gene association, faster and with greater sensitivity
-  than competing clustering tools such as {\sc WCD}, and with
-  considerably greater sensitivity than the implicit clustering
-  performed by assembly tools such as {\sc Cap3}.  Through the GUI the
-  user can collect cluster statistics and examine specific
-  clusters for more comprehensive study.  And by using {\sc Peace} as
-  a ``pre-assembly'' tool to allow the application of an assembly tool
-  to the resultant individual clusters, the user gains both a significant
-  increase in sensitivity of the clustering method and a large decrease
-  in overall runtime for the clustering and assembly together.
+  We present {\sc Peace}, a standalone tool for the high-throughput
+  {\it ab initio} clustering of transcript fragment sequences produced
+  by Next Generation or Sanger Sequencing technologies.  Installed and
+  run through an easily downloaded GUI, {\sc Peace} can process large
+  data sets and group the fragments by gene association, achieving a
+  greater sensitivity (without sacrificing runtime) than the competing
+  clustering tool {\sc WCD}.  Through the GUI the user can collect
+  cluster statistics and examine specific clusters for more
+  comprehensive study.  {\sc Peace}, used as a ``pre-assembly'' tool,
+  enables any assembly tool to work with individual small clusters of
+  closely related sequences, resulting in a signficant: increases in
+  sensitivity, robustness to sequenceing error, and reduction in
+  overall runtime.
 \end{abstract}
@@
 transcripts expressed by genes of the organism, is a vital step in
-understanding the full functional and organizational role of the genome
-in the life cycle of any organism.  Studying the transcriptome has led
-to gene discovery, provided information on protein isoforms, and
-generally helped shed light on the biological processes both
-controlling and dictated by the genome \cite{Nagaraj07}.  But in order
-to get access to those transcripts, we must first deal with the
-fragmented data produced by both Next Generation and traditional
-Sanger sequencing technology.
+understanding the full functional and organizational role of the
+genome in the life cycle of any eukaryote.  Studying the transcriptome
+has led to gene discovery, provided information on splice variants,  
+and helped shed light on the biological processes both controlling and
+dictated by the genome \cite{Nagaraj07}.  But to access to those
+transcripts, we must deal with the fragmented data produced by
+both Next Generation and traditional Sanger sequencing technology.
 
-Until recently, a primary method of gaining access to a transcriptome
-sequence was through the use of Expressed Sequence Tags (ESTs), which
-were primarily single-pass cDNA sequences derived from transcribed
-mRNAs and traditionally sequenced by Sanger Sequencing technology.
-More recently, Next-Generation Sequencing (NGS) technologies have
-replaced the Sanger Sequencing techniques, allowing for considerably
-greater depth and coverage of the transcriptome.  For example, ESTs in
-GenBank dbEST are increasingly the product of NGS technologies such as
-454 pyrosequencing, which enables the sequencing of novel and rate
-transcripts at a considerably higher rate of coverage than did Sanger
-Sequencing \cite{Cheung2006,Emrich2007}.  But from a computational
+Until recently, access to a transcriptome sequence was primarily
+through the use of Expressed Sequence Tags (ESTs), single-pass cDNA
+sequences derived from transcribed mRNAs and sequenced by Sanger
+Sequencing technology.  More recently, Next Generation Sequencing
+technology (NGS) is rapidly replacing Sanger Sequencing, allowing for
+much more comprehensive coverage of the transcriptome.  For example,
+ESTs in GenBank dbEST are increasingly the product of NGS technologies
+such as 454 pyrosequencing, which enables the sequencing of novel and
+rare transcripts at a considerably higher rate of coverage than did
+Sanger Sequencing \cite{Cheung2006,Emrich2007}.  From a computational
 perspective, this is a mixed blessing: while NGS provides immense
-quantities of new information, it also provides immensely larger data sets,
-requiring even more efficient algorithms than were already needed
-for analysis.
+quantities of new information, it also provides immensely larger data
+sets, requiring fast computational techniques if we are to process
+them.
 
-Given a set of transcript fragments, produced by either sequencing
-technology, sampled from multiple transcripts across the genome, a
-necessary first step is that of clustering: separating the fragments
-according to the transcript from which they were derived.  While
-clustering is frequently performed implicitly by assembly tools such
-as {\sc Cap3} \cite{Huang99}, clustering the data as a
-``pre-assembly'' step has a number of advantages.  Most significant
-among these: performing this step will allow the application of the
-assembly tool to individual clusters -- saving significant amounts of
-time due to the reduced input size \cite{Hazelhurst08a}.  But
-clustering is a computationally challenging problem, especially when
-done {\it ab initio} instead of relying on access to a sequenced
-genome.  Even with the smaller number of ESTs produced using Sanger
+Given a set of transcript fragments sampled from across the genome, a
+necessary first step in the analysis is that of clustering: separating
+the fragments according to the transcript from which they were
+derived.  Frequently performed implicitly by assembly tools,
+clustering the data as a ``pre-assembly'' step has a number of
+advantages.  Most significant among these: performing this step will
+allow the application of the assembly tool to individual clusters --
+saving significant amounts of time due to the reduced input size
+\cite{Hazelhurst08a}.  But clustering is a computationally challenging
+problem.  Even with the smaller number of ESTs produced using Sanger
 sequencing, the runtime and memory requirements to cluster based on
 pair-wise sequence alignments make such an approach infeasible.  The
-much larger data-set size produced by next-generation sequencing
+much larger data-set size produced by NGS 
 technologies exacerbates this problem.
@@
 Here we present {\sc Peace}, a tool for the {\it ab initio} clustering
-of transcript fragments by gene, applicable to both short-read and traditional
-Sanger Sequencing technologies.  Available through the
-\href{http://www.peace-tools.org}{www.peace-tools.org} website, the PEACE GUI allows the user
-to both easily install (locally or remotely) and run the computational
-tool, as well as providing various tools for result analysis. {\sc
-  Peace} can be run in single or multi-processor mode, and requires no
-Internet connectivity past the initial download.  Using a {\it minimum
-  spanning tree} (MST) based clustering strategy \cite{Jain99,Wan08}
-and the $d^2$ sequence distance function \cite{Hide94}, {\sc Peace}
-generates the MSTs using a tailored version of Prim's algorithm
-\cite{Prim57}.  The result is an easy-to-use clustering tool that is considerably
-more sensitive to sequencing errors than the competing {\sc wcd} tool
-\cite{Hazelhurst08a} and others in the literature without any
-sacrifice of performance (e.g. \cite{Burke99}, \cite{Slater00},
-\cite{Huang99}, \cite{Parkinson02}, \cite{Kalyanaraman03},
-\cite{Malde03}, \cite{Ptitsyn05}, \cite{Picardi09}).
+of transcript fragments by gene, applicable to both NGS and
+traditional Sanger Sequencing technologies.  Available through the
+\href{http://www.peace-tools.org}{www.peace-tools.org} website, the
+{\sc Peace} GUI allows the user to both easily install (locally or remotely)
+and run the computational tool, as well as enabling transparent
+parallel processing and providing various tools for result analysis.
+Using a {\it minimum spanning tree} (MST) based clustering strategy
+\cite{Jain99,Wan08} and the $d^2$ sequence distance function
+\cite{Hide94}, {\sc Peace} generates the MSTs using a tailored version
+of Prim's algorithm \cite{Prim57}.  The result is an easy-to-use and
+clustering tool that , without any sacrifice in runtime, is
+considerably more sensitive and robust to sequencing errors than other
+clusters tools in the literature \cite{Burke99,Slater00,
+  Huang99,Parkinson02,Kalyanaraman03,Malde03,Ptitsyn05,Hazelhurst08a, Picardi09}.
 
@@
 
-\section{PEACE: Installation and Use}
+\section{{\sc Peace}: Installation and Use}
 
@@
 management tool.  Through this the user can install the computational
-tool and perform a clustering of a data file, view an initial analysis
+tool and perform a clustering of a data file in the FASTA format, view an initial analysis
 of the clusters and produce files containing subsets of the clusters
 for assembly by a tool such as {\sc Cap3} \cite{Huang99}.  A typical
-(first) use of the tool would generally require the the following
+(first) use of {\sc Peace} would generally require the the following
 steps.
@@
 \noindent {\bf Tool Installation:} Installing the GUI requires only
-that the Java file be downloaded from the website.  To install the {\sc
-  Peace} tool onto a local or remote machine, the user will select
-from within the GUI the {\tt Server > Add New Server} tab from the GUI
-main menu (Figure~1(a)), which then starts the install
-wizard that will prompt for the appropriate information.
-Figure~1(b) illustrates the request for server information;
-in this example, the user has chosen to install the {\sc Peace}
+that the Java file be downloaded from the website.  To install the
+{\sc Peace} tool onto a local or remote machine, the user will select
+from within the GUI the appropriate menu tab (Figure~1(a)), which
+then starts the install wizard that will prompt for the appropriate
+information.  Figure~1(b) illustrates the request for server
+information; the user has chosen to install the {\sc Peace}
 computational tool on a remote machine and is providing the necessary
@@
 
-\noindent {\bf Job Processing:} After importing the target file into
-the GUI ({\tt File > New Data Set}), the user can start a new job with
-the {\tt Job > Job to Complete Clusters} tab and follow the wizard
-menus.  Figure~1(c) illustrates the process of specifying
-the number of processors available (if running on a machine supporting
-the OpenMPI protocol -- which will be determined during job
-instillation).  Once executed, the GUI will manage the job thread,
-alert the user when the job is completed (or when the user next
-runs the GUI after completion) and 
-copy the final results back to the local machine if necessary.
+\noindent {\bf Job Processing:} After importing the target sequence file into
+the GUI, the user can start a new job and follow the wizard menus.
+Figure~1(c) illustrates the process of specifying the number of
+processors available (if running on a machine supporting the OpenMPI
+protocol -- which will be determined during job installation).  Once
+executed, the GUI will manage the job thread, alert the user when the
+job is completed (or when the user next runs the GUI after completion)
+and copy the final results back to the local machine if necessary.
 
 \noindent {\bf Result Analysis:} Once the resulting clusters have been
-computed and (if necessary) copied back, the user has several options
+computed, the user has several options
 for analysis:
@@
 \item {\bf Export:} The user can export the contents of one or more
-  clusters into a FASTA format file, obtaining a subset of the original file
+  clusters into a FASTA format file, obtaining a subset of the original target file
   containing the sequences corresponding to the selected clusters
   ready for processing by an assembly tool (e.g. {\sc Cap3} \cite{Huang99}).
-\item {\bf View Clustering}: The user may view a list of all clusters,
-  expanding selected clusters to a list of all fragments (illustrated in
-  Figure~1(d)).  
+\item {\bf View Clustering}: The user may view a list of clusters,
+  expanding selected clusters to a list of all individual sequences (illustrated
+  in Figure~1(d)).
 \item {\bf Classified Summary Graph:} The user may view a distribution
-  of cluster sizes in a linear or $\log$ scale.  Further, the user may
-  set up a {\it classifier}, associating certain patterns with
-  specific colors.  These patterns will be matched against the
-  fragment header information from the original FASTA file, allowing
-  the overlay of a colored cluster size distribution does clusters
-  containing matching elements in a specified colors (comment by molerjc: please reword the preceding sentence fragment).  For example, if
-  the sequence names contain unique string patterns denoting different
-  cDNA libraries, the classifier can help the user to determine the
-  differential expression profiles of different libraries for a given
-  cluster by color coding the strains.  The method of setting up a
-  these classifiers, and the resulting histogram, is illustrated in
-  Figure~1(e).
+  of cluster sizes.  Further, the user may set up a {\it classifier},
+  associating certain patterns with specific colors.  These patterns
+  will be matched against the fragment header information from the
+  original FASTA file, allowing the overlay of a colored cluster size
+  distributions.  For example, if the sequence names contain unique
+  string patterns denoting different cDNA libraries, the classifier
+  can help the user to determine and visualize the differential expression profiles
+  of different libraries for a given cluster.  The method of setting up a these classifiers, and the
+  resulting histogram, is illustrated in Figure~1(e).
 \end{itemize}
-
 Extensive documentation for the tool has been posted on the {\sc
-  Peace} website, as well as links to several videos demonstrating
+  Peace} website, as well as links to several tutorial videos demonstrating
 {\sc Peace} use and capabilities.
@@
 clustering tool \cite{Hazelhurst08a} and the {\sc Cap3} assembly
-tool \cite{Huang99} (which calculates a cluster as an intermediate
+tool \cite{Huang99} (which calculates a cluster as an implicit, intermediate
 step).  For our simulation tests we used the {\bf ESTSim} tool
-\cite{Hazelhurst03} to generate simulated transcript fragments under varying models
-of error (Supplementary Materials, Section~\href{ESTsim_sup}),
+\cite{Hazelhurst03} to generate simulated transcript of varying length
+fragments under different models
+of error (Supplementary Materials, Section~C.1),
 generating the fragments from the list of 100 zebra fish genes used in
@@
 \label{screen}
-\caption{Screenshots of the PEACE GUI during execution, including (a)
+\caption{Screenshots of the {\sc Peace} GUI during execution, including (a)
 GUI Welcome and server installation menu; (b) setup wizard for
@@
 sensitivity and Type 1 error (both significantly outperforming {\sc
-  Cap3}).  Using the mouse fragment set used in Hazelhurst {\it et
+  Cap3}).  Using the mouse dataset used in Hazelhurst {\it et
   al.}  \cite{Hazelhurst08a} {\sc Peace} again shows slight
@@
 Section~\href{real_results}, for more details).
+
+\section{Conclusions}
+
+Here we have presented {\sc Peace}, a stand-alone tool for the
+high-throughput clustering of transcript fragments, managed through a
+user friendly GUI that enables both local and remote installation and
+execution in sequential or parallel mode.  Based on a novel algorithm
+for the clustering of the fragments by gene association, {\sc Peace}
+shows significant improvement in sensitivity, without sacrificing runtime,
+over the competing {\sc WCD} tool \cite{Hazelhurst08a}, and shows an
+order of magnitude in improvement over the clustering preformed in the
+course of assembly by the {\sc Cap3} tool \cite{Huang99}.
+
+As a clustering tool based on sequence distance, {\sc Peace} faces
+certain inherent limitations. For example, {\sc Peace} cannot handle
+duplicate genes; like {\sc WCD}, it is unable to separate clusters
+corresponding to genes with a greater than 88\% similarity.
+Similarly, naturally biological effects such as the trans-splicing of
+transcripts, or effects from poorly cleaned transcript data (e.g. the
+failure to remove sequencing adapters), post-transcriptional
+poly(A)/(T) tails, and low-complexity repeats can cause similar
+effects in these clustering tools.  The problems can be handled
+through the application of the assembler, and the ability to apply any
+assembler to small cluster (as opposed to the data set as a while)
+results in a significant reduction in overall assembly time.
 

### 2010-02-12 src/TVHeuristic.cpp
Changes to TVHeuristic, so that parameters now scale with EST size.  This should cut down on the supercluster problem and improve the specificity of clustering without sacrificing much in the way of sensitivity.  Also slightly tweaked UV Heuristic parameters and fixed a compile warning in EST.cpp.

@@ -38,26 +38,24 @@
 #include "ESTCodec.h"
 #include "EST.h"
 
-// Define the static parameters
-int TVHeuristic::t = 18;
-
-int TVHeuristic::windowLen = 50;
-
 // The set of arguments for this class.  Note that some of the base
 // class static instance variables are reused here so that the values
 // are consistently set.
 arg_parser::arg_record TVHeuristic::argsList[] = {
-    {"--tv_t", "t (number of v-word matches) (default=65)",
-     &TVHeuristic::t, arg_parser::INTEGER},
-    {"--tv_win", "Window size for t/v heuristics (default=100)",
-     &TVHeuristic::windowLen, arg_parser::INTEGER},    
+    //    {"--tv_t", "t (number of v-word matches) (default=65)",
+    //     &TVHeuristic::t, arg_parser::INTEGER},
+    //    {"--tv_win", "Window size for t/v heuristics (default=100)",
+    //     &TVHeuristic::windowLen, arg_parser::INTEGER},    
     {NULL, NULL, NULL, arg_parser::BOOLEAN}
-};
+    };
 
 TVHeuristic::TVHeuristic(const std::string& outputFileName)
     : NewUVHeuristic("tv", outputFileName) {
     matchTable     = NULL;
     uvSuccessCount = 0;
+    t = 18;
+    windowLen = 50;
+    refESTLen = 0;
 }
 
 TVHeuristic::~TVHeuristic() {
@@ -104,7 +102,8 @@ TVHeuristic::initialize() {
     size_t maxESTlen = EST::getMaxESTLen();
     
     // Add extra 100 characters to ease processing.
-    matchTable = new char[maxESTlen + windowLen + v];
+    //matchTable = new char[maxESTlen + windowLen + v];
+    matchTable = new char[maxESTlen + 150 + v];
     // Everything went well
     return 0;
 }
@@ -115,6 +114,7 @@ TVHeuristic::setReferenceEST(const int estIdx) {
         // The reference EST is the same. Nothing to be done.
         return 0;
     }
+    refESTLen = (int)strlen(EST::getEST(estIdx)->getSequence());
     // Simply let the base class to its job
     return NewUVHeuristic::setReferenceEST(estIdx);
 }
@@ -126,11 +126,15 @@ TVHeuristic::runHeuristic(const int otherEST) {
     if (!NewUVHeuristic::runHeuristic(otherEST)) {
         // This pair need not be analyzed further.
         return false;
-    }
+    }    
[…]

### 2010-02-12 peace/src/EST.cpp
Changes to TVHeuristic, so that parameters now scale with EST size.  This should cut down on the supercluster problem and improve the specificity of clustering without sacrificing much in the way of sensitivity.  Also slightly tweaked UV Heuristic parameters and fixed a compile warning in EST.cpp.

@@ -161,7 +161,6 @@ EST::create(FILE* fastaFile, int& lineNum) {
         std::transform(sequence.begin(), sequence.end(), sequence.begin(),
                        toupper);
         const char* const seqBP = sequence.c_str();
-        size_t seqLen = strlen(seqBP);
         // Compute new max EST length
         maxESTlen = std::max(maxESTlen, strlen(seqBP));
         // Create a new est with all the information.

### 2010-02-12 peace/src/FWAnalyzer.cpp

@@ -167,12 +167,12 @@ FWAnalyzer::initialize() {
         // heuristic are consistently set. This feels a bit hoaky, but
         // should be OK for now.
         Heuristic* tv = chain->getHeuristic("tv");
-        if ((tv != NULL) && (TVHeuristic::getWindowLen() != frameSize)) {
+        if ((tv != NULL) && (((TVHeuristic*)tv)->getWindowLen() != frameSize)) {
             std::cerr << "Warning: The frame/window size for "
                       << getName() << " analyzer (which is set to "
                       << frameSize << ") is different from that for "
                       << "the t/v heuristic (which is set to "
-                      << TVHeuristic::getWindowLen() << ").\n"
+                      << ((TVHeuristic*)tv)->getWindowLen() << ").\n"
                       << "       : This will most likely cause problems. "
                       << "To avoid it use both --frame and --tv_win options."
                       << std::endl;

### 2010-02-12 peace/src/NewUVHeuristic.cpp

@@ -41,7 +41,7 @@
 
 // default params
 int NewUVHeuristic::u           = 4;
-int NewUVHeuristic::v           = 6;
+int NewUVHeuristic::v           = 8;
 int NewUVHeuristic::wordShift   = 4;
 int NewUVHeuristic::BitMask     = 0;
 int NewUVHeuristic::passes      = 2;

### 2010-02-12 peace/src/TVHeuristic.h

@@ -155,7 +155,7 @@ public:
         \return The current window (or frame) size set for <i>t/v</i>
         heuristic.
     */
-    static int getWindowLen() { return windowLen; }
+    int getWindowLen() { return windowLen; }
     
 protected:
     /** The default constructor.
@@ -200,6 +200,8 @@ protected:
 	EST pair should be analyzed, and \c false if it should not.
     */
     virtual bool runHeuristic(const int otherEST);
+
+    virtual void adjustParameters(const int otherESTLen);
 
     template <typename Encoder>
     int countCommonWords(const int otherEST, Encoder encoder,
@@ -265,7 +267,7 @@ private:
         in pairs of ESTs.  The default is 65. However, this value can
         be overridden by a command line argument.
     */
-    static int t;
+    int t;
 
     /** The window length to be used for <i>t/v</i> heuristic.
 
@@ -276,7 +278,9 @@ private:
         default value is 100.  This value can be overridden by the
         user via suitable command line arguments.
     */
-    static int windowLen;
+    int windowLen;
+
+    int refESTLen;
     
     /** A large table to track matches.
 

### 2010-02-12 papers/NAR_web_server_2009/paper.tex
Update to manuscript

@@
   greater sensitivity (without sacrificing runtime) than the competing
-  clustering tool {\sc WCD}.  Through the GUI the user can collect
+  clustering tools.  Through the GUI the user can collect
   cluster statistics and examine specific clusters for more
@@
   enables any assembly tool to work with individual small clusters of
-  closely related sequences, resulting in a signficant: increases in
-  sensitivity, robustness to sequenceing error, and reduction in
+  closely related sequences, resulting in signficant increases in
+  sensitivit and robustness to sequenceing error, as well as a reduction in
   overall runtime.
@@
 and helped shed light on the biological processes both controlling and
-dictated by the genome \cite{Nagaraj07}.  But to access to those
+dictated by the genome \cite{Nagaraj07}.  However, to access to those
 transcripts, we must deal with the fragmented data produced by
@@
 Sequencing technology.  More recently, Next Generation Sequencing
-technology (NGS) is rapidly replacing Sanger Sequencing, allowing for
-much more comprehensive coverage of the transcriptome.  For example,
-ESTs in GenBank dbEST are increasingly the product of NGS technologies
-such as 454 pyrosequencing, which enables the sequencing of novel and
-rare transcripts at a considerably higher rate of coverage than did
-Sanger Sequencing \cite{Cheung2006,Emrich2007}.  From a computational
-perspective, this is a mixed blessing: while NGS provides immense
-quantities of new information, it also provides immensely larger data
-sets, requiring fast computational techniques if we are to process
-them.
+(NGS) technology has begun to rapidly replace Sanger Sequencing,
+allowing for more comprehensive coverage of the transcriptome.  For
+example, ESTs now being added to the GenBank dbEST are increasingly
+the product of NGS technologies such as 454 pyrosequencing, which
+enables the sequencing of novel and rare transcripts at a considerably
+higher rate of coverage than Sanger Sequencing
+\cite{Cheung2006,Emrich2007}.  From a computational perspective, this
+is a mixed blessing: while NGS provides immense quantities of new
+information, it also provides immensely larger data sets, requiring
+faster computational techniques if we are to process them.
 
@@
 clustering the data as a ``pre-assembly'' step has a number of
-advantages.  Most significant among these: performing this step will
+advantages.  Most significantly: performing this step will
 allow the application of the assembly tool to individual clusters --
@@
 problem.  Even with the smaller number of ESTs produced using Sanger
-sequencing, the runtime and memory requirements to cluster based on
-pair-wise sequence alignments make such an approach infeasible.  The
+Sequencing, the runtime and memory requirements to cluster based on
+pair-wise sequence alignments make such an approach infeasible in practice.  The
 much larger data-set size produced by NGS 
@@
 of Prim's algorithm \cite{Prim57}.  The result is an easy-to-use and
-clustering tool that , without any sacrifice in runtime, is
+clustering tool that, without any sacrifice in runtime, is
 considerably more sensitive and robust to sequencing errors than other
@@
 
-The {\sc Peace} GUI, available from the {\sc Peace} website
-(\href{http://www.peace-tools.org}{www.peace-tools.org}), handles the
-details of both installation and job management and provides tools for
-the initial analysis of the results.  By downloading and executing the
-{\tt peace.jar} GUI Java program on a Linux, Windows or OS-X machine
-running Java 1.6+, the user is presented with the {\sc Peace}
-management tool.  Through this the user can install the computational
-tool and perform a clustering of a data file in the FASTA format, view an initial analysis
-of the clusters and produce files containing subsets of the clusters
-for assembly by a tool such as {\sc Cap3} \cite{Huang99}.  A typical
-(first) use of {\sc Peace} would generally require the the following
-steps.
+The {\sc Peace} GUI can be launched by downloading and executing the
+Jar file available on the {\sc Peace} wesite
+(\href{http://www.peace-tools.org}{www.peace-tools.org}) to any
+machine running a standard Java Virtual Machine (JVM).  Once running,
+the user can employ the GUI to install the computational tool and perform
+a clustering of a data file in the FASTA format, view an initial
+analysis of the clusters, and produce files containing subsets of the
+clusters for assembly by a tool such as {\sc Cap3} \cite{Huang99}.
+A typical (first) use of {\sc Peace} generally
+requires the the following steps.
 
-\noindent {\bf Tool Installation:} Installing the GUI requires only
-that the Java file be downloaded from the website.  To install the
-{\sc Peace} tool onto a local or remote machine, the user will select
-from within the GUI the appropriate menu tab (Figure~1(a)), which
-then starts the install wizard that will prompt for the appropriate
-information.  Figure~1(b) illustrates the request for server
-information; the user has chosen to install the {\sc Peace}
+\noindent {\bf Tool Installation:} (First use only.)  To install the
+{\sc Peace} clustering onto a local or remote machine, the
+user selects from within the GUI the appropriate menu tab
+(Figure~1(a)), which then starts an install wizard that will prompt
+for the appropriate information.  Figure~1(b) illustrates the request
+for server information; the user has chosen to install the {\sc Peace}
 computational tool on a remote machine and is providing the necessary
@@
 \noindent {\bf Job Processing:} After importing the target sequence file into
-the GUI, the user can start a new job and follow the wizard menus.
+the GUI, the user starts a new job by following the wizard menus.
 Figure~1(c) illustrates the process of specifying the number of
@@
 executed, the GUI will manage the job thread, alert the user when the
-job is completed (or when the user next runs the GUI after completion)
+job is completed (or when the user next runs the GUI after completion),
 and copy the final results back to the local machine if necessary.
@@
 \noindent {\bf Result Analysis:} Once the resulting clusters have been
-computed, the user has several options
-for analysis:
+computed, the user has several options for analysis:
 \begin{itemize}
@@
   can help the user to determine and visualize the differential expression profiles
-  of different libraries for a given cluster.  The method of setting up a these classifiers, and the
+  of different libraries for a given cluster.  The method of setting up these classifiers, and the
   resulting histogram, is illustrated in Figure~1(e).
@@
 the connecting edges as weights.  Conceptually, we want to remove each
-edge exceeding a threshold score from the complete graph, and define
+edge exceeding a threshold score from the complete graph and define
 our partitions by the remaining connected components of the graph.  An
 edge with a large weight connects fragments which are likely unrelated;
-once such edges are removed, the components define a series of
-overlaps.  Hence those fragments that can still be connected by some path
+once such edges are removed the components define a series of
+overlaps.  Those fragments that can still be connected by some path
 correspond to the same gene.  However, such an approach requires the
@@
 using Prim's algorithm we are able to calculate edge weights
-on-the-fly (reducing memory requirements), and we can skip the
+on-the-fly (reducing memory requirements) and we can skip the
 calculation of a majority of edge distances using the $u/v$ and $t/v$
@@
 
-\begin{figure*}[b]
+\begin{figure*}[h]
 \centerline{
+\begin{minipage}[b]{2in}
 \includegraphics[scale=0.5]{pics.d/SeT1.pdf} 
+\end{minipage}{2in}
 }
@@
 \begin{figure*}[b]
+\begin{minipage}{2in}
 \centerline{
@@
 }
+\end{minipage}
 \label{singletons}
@@
 corresponding to genes with a greater than 88\% similarity.
-Similarly, naturally biological effects such as the trans-splicing of
+Similarly, natural biological effects such as the trans-splicing of
 transcripts, or effects from poorly cleaned transcript data (e.g. the
@@
 through the application of the assembler, and the ability to apply any
-assembler to small cluster (as opposed to the data set as a while)
+assembler to small cluster (as opposed to the data set as a whole)
 results in a significant reduction in overall assembly time.
@@
 Starters Grant while conducting this research.  We would also like to
-acknowledge Iddo Friedberg, David Woods, Jens Mueller and David
+acknowledge Iddo Friedberg, David Woods, Elizabeth Bikun, Jens Mueller and David
 Scoville at Miami University for their help with this project.

### 2010-02-13 peace/papers/NAR_web_server_2009/paper.tex
Further esits to manuscript

@@
 \begin{abstract}
-  We present {\sc Peace}, a standalone tool for the high-throughput
-  {\it ab initio} clustering of transcript fragment sequences produced by Next
-  Generation or Sanger Sequencing technologies.  Installed and run
-  through an easily downloaded GUI, {\sc Peace} can process large data
-  sets and group the fragments by gene association, achieving a
+  We present {\sc Peace}, a stand alone tool for the high-throughput
+  {\it ab initio} clustering of transcript fragment sequences produced
+  by Next Generation or Sanger Sequencing technologies.  Installed and
+  run through an easily downloaded GUI, {\sc Peace} can process large
+  data sets and group the fragments by gene association, achieving a
   greater sensitivity (without sacrificing runtime) than the competing
-  clustering tool {\sc WCD}.  Through the GUI the user can collect
+  clustering tools.  Through the GUI the user can collect
   cluster statistics and examine specific clusters for more
-  comprehensive study.  By using {\sc Peace} as a ``pre-assembly''
-  tool to allow the application of an assembly tool to the resultant
-  individual clusters, the user gains both a significant increase in
-  sensitivity and a large decrease in overall runtime for the
-  clustering and assembly together.
+  comprehensive study.  {\sc Peace}, used as a ``pre-assembly'' tool,
+  enables any assembly tool to work with individual small clusters of
+  closely related sequences, resulting in significant increases in
+  sensitive and robustness to sequencing error, as well as a reduction in
+  overall runtime.
 \end{abstract}
@@
 understanding the full functional and organizational role of the
-genome in the life cycle of any organism.  Studying the transcriptome
+genome in the life cycle of any eukaryote.  Studying the transcriptome
 has led to gene discovery, provided information on splice variants,  
 and helped shed light on the biological processes both controlling and
-dictated by the genome \cite{Nagaraj07}.  But to access to those
+dictated by the genome \cite{Nagaraj07}.  However, to access those
 transcripts, we must deal with the fragmented data produced by
@@
 Sequencing technology.  More recently, Next Generation Sequencing
-(NGS) technologies have replaced the Sanger Sequencing techniques,
-allowing for considerably coverage of the transcriptome.  For example,
-ESTs in GenBank dbEST are increasingly the product of NGS technologies
-such as 454 pyrosequencing, which enables the sequencing of novel and
-rare transcripts at a considerably higher rate of coverage than did
-Sanger Sequencing \cite{Cheung2006,Emrich2007}.  From a computational
-perspective, this is a mixed blessing: while NGS provides immense
-quantities of new information, it also provides immensely larger data
-sets, requiring fast computational techniques if we are to process
-them.
+(NGS) technology has begun to rapidly replace Sanger Sequencing,
+allowing for more comprehensive coverage of the transcriptome.  For
+example, ESTs now being added to the GenBank dbEST are increasingly
+the product of NGS technologies such as 454 pyrosequencing, which
+enables the sequencing of novel and rare transcripts at a considerably
+higher rate of coverage than Sanger Sequencing
+\cite{Cheung2006,Emrich2007}.  From a computational perspective, this
+is a mixed blessing: while NGS provides immense quantities of new
+information, it also provides immensely larger data sets -- and thus a
+need for faster, more efficient analysis algorithms.
 
 Given a set of transcript fragments sampled from across the genome, a
-necessary first step in the analysis is that of clustering: separating
+necessary first step of their analysis is that of clustering: separating
 the fragments according to the transcript from which they were
@@
 clustering the data as a ``pre-assembly'' step has a number of
-advantages.  Most significant among these: performing this step will
+advantages.  Most significantly: performing this step will
 allow the application of the assembly tool to individual clusters --
 saving significant amounts of time due to the reduced input size
-\cite{Hazelhurst08a}.  But clustering is a computationally challenging
-problem.  Even with the smaller number of ESTs produced using Sanger
-sequencing, the runtime and memory requirements to cluster based on
-pair-wise sequence alignments make such an approach infeasible.  The
-much larger data-set size produced by Next-Generation sequencing
-technologies exacerbates this problem.
+\cite{Hazelhurst08a}.
 
-Here we present {\sc Peace}, a tool for the {\it ab initio} clustering
-of transcript fragments by gene, applicable to both short-read and
+But clustering is a computationally challenging problem.  Even with
+the smaller number of ESTs produced using Sanger Sequencing, the
+runtime and memory requirements to cluster based on pair-wise sequence
+alignments make such an approach infeasible in practice.  The much
+larger data-set size produced by NGS technologies exacerbates this
+problem.  To deal with this, {\sc Peace} combines the $d^2$
+alignment-free sequence distance function \cite{Hide94} and the
+concept of a {\it minimum spanning tree} \cite{Prim57} to quickly and
+accurately find clusters of ESTs expressed from the same gene without
+reference to a sequenced genome.  Compared against competing tools
+for the same purpose \cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
+  Ptitsyn05,Hazelhurst08a,Picardi09}, {\it Peace} proves to be both more sensitive and
+more robust to sequencing error without sacrificing runtime.
+
+In short, {\sc Peace} is computational tool for the {\it ab initio}
+clustering of transcript fragments by gene association, applicable to both NGS and
 traditional Sanger Sequencing technologies.  Available through the
 \href{http://www.peace-tools.org}{www.peace-tools.org} website, the
-PEACE GUI allows the user to both easily install (locally or remotely)
-and run the computational tool, as well as enabling transparent
-parallel processing and providing various tools for result analysis.
-Using a {\it minimum spanning tree} (MST) based clustering strategy
-\cite{Jain99,Wan08} and the $d^2$ sequence distance function
-\cite{Hide94}, {\sc Peace} generates the MSTs using a tailored version
-of Prim's algorithm \cite{Prim57}.  The result is an easy-to-use and
-clustering tool that , without any sacrifice in runtime, is
-considerably more sensitive and robust to sequencing errors than other
-clusters tools in the literature \cite{Burke99,Slater00,
-  Huang99,Parkinson02,Kalyanaraman03,Malde03,Ptitsyn05,Hazelhurst08a, Picardi09}.
+{\sc Peace} GUI allows the user to both easily install (locally or
+remotely) and run the clustering engine, as well as enabling
+transparent parallel processing and providing various tools for result
+analysis.
 
@@
 
-\section{PEACE: Installation and Use}
+\section{{\sc Peace}: Installation and Use}
 
-The {\sc Peace} GUI, available from the {\sc Peace} website
-(\href{http://www.peace-tools.org}{www.peace-tools.org}), handles the
-details of both installation and job management and provides tools for
-the initial analysis of the results.  By downloading and executing the
-{\tt peace.jar} GUI Java program on a Linux, Windows or OS-X machine
-running Java 1.6+, the user is presented with the {\sc Peace}
-management tool.  Through this the user can install the computational
-tool and perform a clustering of a data file, view an initial analysis
-of the clusters and produce files containing subsets of the clusters
-for assembly by a tool such as {\sc Cap3} \cite{Huang99}.  A typical
-(first) use of the tool would generally require the the following
-steps.
+The {\sc Peace} GUI can be launched by downloading and executing the
+Jar file available on the {\sc Peace} website
+(\href{http://www.peace-tools.org}{www.peace-tools.org}) to any
+machine running the standard Java Virtual Machine (JVM).  Once running,
+the user can employ the GUI to install the computational tool and perform
+a clustering of a data file in the FASTA format, view an initial
+analysis of the clusters, and produce files containing subsets of the
+clusters as input to assembly tools such as {\sc Cap3} \cite{Huang99}.
+A typical (first) use of {\sc Peace} generally
+requires the following steps.
 
-\noindent {\bf Tool Installation:} Installing the GUI requires only
-that the Java file be downloaded from the website.  To install the
-{\sc Peace} tool onto a local or remote machine, the user will select
-from within the GUI the appropriate menu tab (Figure~1(a)), which
-then starts the install wizard that will prompt for the appropriate
-information.  Figure~1(b) illustrates the request for server
-information; the user has chosen to install the {\sc Peace}
+\noindent {\bf Tool Installation:} (First use only.)  To install the
+{\sc Peace} clustering onto a local or remote machine, the
+user selects from within the GUI the appropriate menu tab
+(Figure~1(a)), which then starts an install wizard that will prompt
+for the appropriate information.  Figure~1(b) illustrates the request
+for server information; the user has chosen to install the {\sc Peace}
 computational tool on a remote machine and is providing the necessary
@@
 
-\noindent {\bf Job Processing:} After importing the target file into
-the GUI, the user can start a new job and follow the wizard menus.
+\noindent {\bf Job Processing:} After importing the target sequence file into
+the GUI, the user starts a new job by following the wizard menus.
 Figure~1(c) illustrates the process of specifying the number of
@@
 \noindent {\bf Result Analysis:} Once the resulting clusters have been
-computed, the user has several options
-for analysis:
+computed, the user has several options for analysis:
 \begin{itemize}
 \item {\bf Export:} The user can export the contents of one or more
-  clusters into a FASTA format file, obtaining a subset of the original file
+  clusters into a FASTA format file, obtaining a subset of the original target file
   containing the sequences corresponding to the selected clusters
@@
 \item {\bf View Clustering}: The user may view a list of clusters,
-  expanding selected clusters to a list of all fragments (illustrated
+  expanding selected clusters to a list of all individual sequences (illustrated
   in Figure~1(d)).
 \item {\bf Classified Summary Graph:} The user may view a distribution
-  of cluster sizes.  Further, the user may set up a {\it classifier},
+  of cluster sizes.  Further, the user can set up a {\it classifier},
   associating certain patterns with specific colors.  These patterns
-  will be matched against the fragment header information from the
+  ware matched against the fragment header information from the
   original FASTA file, allowing the overlay of a colored cluster size
@@
   string patterns denoting different cDNA libraries, the classifier
-  can help the user to determine the differential expression profiles
-  of different libraries for a given cluster by color coding the
-  strains.  The method of setting up a these classifiers, and the
+  can help the user to determine and visualize the differential expression profiles
+  of different libraries for a given cluster.  The method of setting up these classifiers, and the
   resulting histogram, is illustrated in Figure~1(e).
@@
 Extensive documentation for the tool has been posted on the {\sc
-  Peace} website, as well as links to several videos demonstrating
+  Peace} website, as well as links to several tutorial videos demonstrating
 {\sc Peace} use and capabilities.
@@
 spanning trees (MSTs), known to be an effective approach for narrow
-band single linkage clustering \cite{Jain99,Wan08}. We use the $d^2$
-distance measure \cite{Hide94} to represent edge weights.  Prim's
-algorithm \cite{Prim57} then allows us to efficiently calculate an
-MST from which we can infer a high-quality clustering solution.
+band single linkage clustering \cite{Jain99,Wan08}. Edge weights are
+assigned using the $d^2$ distance measure \cite{Hide94} to represent
+edge weights.  Prim's algorithm \cite{Prim57} then allows us to
+efficiently calculate an MST from which we can infer a high-quality
+clustering solution.
 
@@
 
-We can model the fragment input as a weighted, undirected graph: the fragments
+Fragment input is modeled as a weighted, undirected graph: the fragments
 are represented as nodes, with $d^2$ sequence distances assigned to
 the connecting edges as weights.  Conceptually, we want to remove each
-edge exceeding a threshold score from the complete graph, and define
+edge exceeding a threshold score from the complete graph and define
 our partitions by the remaining connected components of the graph.  An
 edge with a large weight connects fragments which are likely unrelated;
-once such edges are removed, the components define a series of
-overlaps.  Hence those fragments that can still be connected by some path
+once such edges are removed the components define a series of
+overlaps.  Those fragments that can still be connected by some path
 correspond to the same gene.  However, such an approach requires the
@@
 
-We approach the problem by generating a minimum spanning tree of the
+{\sc Peace} approaches the problem by generating a minimum spanning tree of the
 described graph, then removing edges exceeding our threshold.  By
 using Prim's algorithm we are able to calculate edge weights
-on-the-fly (reducing memory requirements), and we can skip the
+on-the-fly (reducing memory requirements) and can skip the
 calculation of a majority of edge distances using the $u/v$ and $t/v$
@@
 
-We have tested {\sc Peace} on both simulated and real data,
-comparing results against those produced by the {\sc wcd}
-clustering tool \cite{Hazelhurst08a} and the {\sc Cap3} assembly
-tool \cite{Huang99} (which calculates a cluster as an intermediate
-step).  For our simulation tests we used the {\bf ESTSim} tool
-\cite{Hazelhurst03} to generate simulated transcript of varying length
-fragments under different models
-of error (Supplementary Materials, Section~C.1),
-generating the fragments from the list of 100 zebra fish genes used in
-the {\sc wcd} testing \cite{Hazelhurst08a}.  Tool parameters were
-taken to match, as closely as possible, those used in the {\sc wcd}
-study (see supplementary materials).  Our primary methods of quality
-assessment were {\it sensitivity} (the fraction of fragment pairs from the
-same gene that were correctly clustered together) and {\it Type 1
-  error} (the fraction of genes that were divided between clusters)
-\cite{Wang04,Hazelhurst08a}; other measurements are also discussed
-in the supplementary materials.  In Figure~2(left) we see a
-significant improvement of {\sc Peace} (blue) over {\sc wcd}
-(green) and {\sc Cap3} (black) in its sensitivity to sequencing
-errors, while in Figure~2(right) we see a comparable
-improvement in Type 1 error.  In Figure~3 we look at
-the number of singleton clusters (fragments not joined to any cluster),
-which will not occur in our simulated sets; we again see significant
-improvement in {\sc Peace}.  The sequential runtime of
-{\sc Peace} is slightly improved over that of {\sc wcd} (see
-Supplementary Materials, Figure~S4).
+{\sc Peace} has been test on both simulated and real data from NGS and
+Sanger Sequencing technologies, comparing results against those
+produced by the {\sc wcd} clustering tool \cite{Hazelhurst08a} and the
+{\sc Cap3} assembly tool \cite{Huang99} (which calculates a cluster as
+an implicit, intermediate step).  For our simulation tests we used the
+{\bf ESTSim} tool \cite{Hazelhurst03} to generate simulated transcript
+of varying length fragments under different models of error
+(Supplementary Materials, Section~C.1), generating the fragments from
+the list of 100 zebra fish genes used in the {\sc wcd} testing
+\cite{Hazelhurst08a}.  Tool parameters were taken to match, as closely
+as possible, those used in the {\sc wcd} study (see supplementary
+materials).  Our primary methods of quality assessment were {\it
+  sensitivity} (the fraction of fragment pairs from the same gene that
+were correctly clustered together) and {\it Type 1 error} (the
+fraction of genes that were divided between clusters)
+\cite{Wang04,Hazelhurst08a}; other measurements are also discussed in
+the supplementary materials.  In Figure~2(left) we see a significant
+improvement of {\sc Peace} (blue) over {\sc wcd} (green) and {\sc
+  Cap3} (black) in its sensitivity to sequencing errors, while in
+Figure~2(right) we see a comparable improvement in Type 1 error.  In
+Figure~3 we look at the number of singleton clusters (fragments not
+joined to any cluster), which will not occur in our simulated sets; we
+again see significant improvement in {\sc Peace}.  The sequential
+runtime of {\sc Peace} is slightly improved over that of {\sc wcd}
+(see Supplementary Materials, Figure~S4).
 
@@
 sensitivity and Type 1 error (both significantly outperforming {\sc
-  Cap3}).  Using the mouse fragment set used in Hazelhurst {\it et
-  al.}  \cite{Hazelhurst08a} {\sc Peace} again shows slight
-improvements in sensitivity and Type 1 error rates, with an 18\% speedup
-for {\sc Peace} (see Supplementary Materials,
-Section~\href{real_results}, for more details).
+  Cap3}).  Using the mouse data set used in Hazelhurst {\it et al.}
+\cite{Hazelhurst08a} {\sc Peace} again shows slight improvements in
+sensitivity and Type 1 error rates, with an 18\% speedup for {\sc
+  Peace} (see Supplementary Materials, Section~D, for more details).
 
@@
 
-Here we have presented {\sc Peace}, a stand-alone tool for the
+Here we have presented {\sc Peace}, a stand alone tool for the
 high-throughput clustering of transcript fragments, managed through a
@@
 corresponding to genes with a greater than 88\% similarity.
-Similarly, naturally biological effects such as the trans-splicing of
+Similarly, natural biological effects such as the trans-splicing of
 transcripts, or effects from poorly cleaned transcript data (e.g. the
-failure to remove polyA tailed introduced in the closing process) can
-cause similar effects in these clustering tools.  The problems can be
-handled through the application of the assembler, and the ability to
-apply any assembler to small cluster (as opposed to the data set as a
-while) results in a significant reduction in overall assembly time.
+failure to remove sequencing adapters or post-transcriptional
+poly(A)/(T) tails), and low-complexity repeats can cause similar
+effects in these clustering tools.  The problems can be handled
+through the application of the assembler, and the ability to apply any
+assembler to small cluster (as opposed to the data set as a whole)
+results in a significant reduction in overall assembly time.
 

### 2010-02-13 papers/NAR_web_server_2009/paper.tex
Added workflow figure

@@
 
-\history{Received Feb. 12, 2010}
+\history{Received Feb. 13, 2010}
 
@@
 \begin{abstract}
-  We present {\sc Peace}, a standalone tool for the high-throughput
+  We present {\sc Peace}, a stand alone tool for the high-throughput
   {\it ab initio} clustering of transcript fragment sequences produced
@@
   enables any assembly tool to work with individual small clusters of
-  closely related sequences, resulting in signficant increases in
-  sensitivit and robustness to sequenceing error, as well as a reduction in
+  closely related sequences, resulting in significant increases in
+  sensitive and robustness to sequencing error, as well as a reduction in
   overall runtime.
@@
 and helped shed light on the biological processes both controlling and
-dictated by the genome \cite{Nagaraj07}.  However, to access to those
+dictated by the genome \cite{Nagaraj07}.  However, to access those
 transcripts, we must deal with the fragmented data produced by
@@
 is a mixed blessing: while NGS provides immense quantities of new
-information, it also provides immensely larger data sets, requiring
-faster computational techniques if we are to process them.
+information, it also provides immensely larger data sets -- and thus a
+need for faster, more efficient analysis algorithms.
 
 Given a set of transcript fragments sampled from across the genome, a
-necessary first step in the analysis is that of clustering: separating
+necessary first step of their analysis is that of clustering: separating
 the fragments according to the transcript from which they were
@@
 saving significant amounts of time due to the reduced input size
-\cite{Hazelhurst08a}.  But clustering is a computationally challenging
-problem.  Even with the smaller number of ESTs produced using Sanger
-Sequencing, the runtime and memory requirements to cluster based on
-pair-wise sequence alignments make such an approach infeasible in practice.  The
-much larger data-set size produced by NGS 
-technologies exacerbates this problem.
+\cite{Hazelhurst08a}.
 
-Here we present {\sc Peace}, a tool for the {\it ab initio} clustering
-of transcript fragments by gene, applicable to both NGS and
+But clustering is a computationally challenging problem.  Even with
+the smaller number of ESTs produced using Sanger Sequencing, the
+runtime and memory requirements to cluster based on pair-wise sequence
+alignments make such an approach infeasible in practice.  The much
+larger data-set size produced by NGS technologies exacerbates this
+problem.  To deal with this, {\sc Peace} combines the $d^2$
+alignment-free sequence distance function \cite{Hide94} and the
+concept of a {\it minimum spanning tree} \cite{Prim57} to quickly and
+accurately find clusters of ESTs expressed from the same gene without
+reference to a sequenced genome.  Compared against competing tools
+for the same purpose \cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
+  Ptitsyn05,Hazelhurst08a,Picardi09}, {\it Peace} proves to be both more sensitive and
+more robust to sequencing error without sacrificing runtime.
+
+In short, {\sc Peace} is computational tool for the {\it ab initio}
+clustering of transcript fragments by gene association, applicable to both NGS and
 traditional Sanger Sequencing technologies.  Available through the
 \href{http://www.peace-tools.org}{www.peace-tools.org} website, the
-{\sc Peace} GUI allows the user to both easily install (locally or remotely)
-and run the computational tool, as well as enabling transparent
-parallel processing and providing various tools for result analysis.
-Using a {\it minimum spanning tree} (MST) based clustering strategy
-\cite{Jain99,Wan08} and the $d^2$ sequence distance function
-\cite{Hide94}, {\sc Peace} generates the MSTs using a tailored version
-of Prim's algorithm \cite{Prim57}.  The result is an easy-to-use and
-clustering tool that, without any sacrifice in runtime, is
-considerably more sensitive and robust to sequencing errors than other
-clusters tools in the literature \cite{Burke99,Slater00,
-  Huang99,Parkinson02,Kalyanaraman03,Malde03,Ptitsyn05,Hazelhurst08a, Picardi09}.
+{\sc Peace} GUI allows the user to both easily install (locally or
+remotely) and run the clustering engine, as well as enabling
+transparent parallel processing and providing various tools for result
+analysis.
 
@@
 The {\sc Peace} GUI can be launched by downloading and executing the
-Jar file available on the {\sc Peace} wesite
+Jar file available on the {\sc Peace} website
 (\href{http://www.peace-tools.org}{www.peace-tools.org}) to any
-machine running a standard Java Virtual Machine (JVM).  Once running,
+machine running the standard Java Virtual Machine (JVM).  Once running,
 the user can employ the GUI to install the computational tool and perform
@@
 analysis of the clusters, and produce files containing subsets of the
-clusters for assembly by a tool such as {\sc Cap3} \cite{Huang99}.
+clusters as input to assembly tools such as {\sc Cap3} \cite{Huang99}.
 A typical (first) use of {\sc Peace} generally
-requires the the following steps.
+requires the following steps.
 
@@
 \item {\bf Classified Summary Graph:} The user may view a distribution
-  of cluster sizes.  Further, the user may set up a {\it classifier},
+  of cluster sizes.  Further, the user can set up a {\it classifier},
   associating certain patterns with specific colors.  These patterns
-  will be matched against the fragment header information from the
+  ware matched against the fragment header information from the
   original FASTA file, allowing the overlay of a colored cluster size
@@
 
-We can model the fragment input as a weighted, undirected graph: the fragments
+Fragment input is modeled as a weighted, undirected graph: the fragments
 are represented as nodes, with $d^2$ sequence distances assigned to
@@
 
-We approach the problem by generating a minimum spanning tree of the
+{\sc Peace} approaches the problem by generating a minimum spanning tree of the
 described graph, then removing edges exceeding our threshold.  By
 using Prim's algorithm we are able to calculate edge weights
-on-the-fly (reducing memory requirements) and we can skip the
+on-the-fly (reducing memory requirements) and can skip the
 calculation of a majority of edge distances using the $u/v$ and $t/v$
@@
 
-We have tested {\sc Peace} on both simulated and real data,
-comparing results against those produced by the {\sc wcd}
-clustering tool \cite{Hazelhurst08a} and the {\sc Cap3} assembly
-tool \cite{Huang99} (which calculates a cluster as an implicit, intermediate
-step).  For our simulation tests we used the {\bf ESTSim} tool
-\cite{Hazelhurst03} to generate simulated transcript of varying length
-fragments under different models
-of error (Supplementary Materials, Section~C.1),
-generating the fragments from the list of 100 zebra fish genes used in
-the {\sc wcd} testing \cite{Hazelhurst08a}.  Tool parameters were
-taken to match, as closely as possible, those used in the {\sc wcd}
-study (see supplementary materials).  Our primary methods of quality
-assessment were {\it sensitivity} (the fraction of fragment pairs from the
-same gene that were correctly clustered together) and {\it Type 1
-  error} (the fraction of genes that were divided between clusters)
-\cite{Wang04,Hazelhurst08a}; other measurements are also discussed
-in the supplementary materials.  In Figure~2(left) we see a
-significant improvement of {\sc Peace} (blue) over {\sc wcd}
-(green) and {\sc Cap3} (black) in its sensitivity to sequencing
-errors, while in Figure~2(right) we see a comparable
-improvement in Type 1 error.  In Figure~3 we look at
-the number of singleton clusters (fragments not joined to any cluster),
-which will not occur in our simulated sets; we again see significant
-improvement in {\sc Peace}.  The sequential runtime of
-{\sc Peace} is slightly improved over that of {\sc wcd} (see
-Supplementary Materials, Figure~S4).
+{\sc Peace} has been test on both simulated and real data from NGS and
+Sanger Sequencing technologies, comparing results against those
+produced by the {\sc wcd} clustering tool \cite{Hazelhurst08a} and the
+{\sc Cap3} assembly tool \cite{Huang99} (which calculates a cluster as
+an implicit, intermediate step).  For our simulation tests we used the
+{\bf ESTSim} tool \cite{Hazelhurst03} to generate simulated transcript
+of varying length fragments under different models of error
+(Supplementary Materials, Section~C.1), generating the fragments from
+the list of 100 zebra fish genes used in the {\sc wcd} testing
+\cite{Hazelhurst08a}.  Tool parameters were taken to match, as closely
+as possible, those used in the {\sc wcd} study (see supplementary
+materials).  Our primary methods of quality assessment were {\it
+  sensitivity} (the fraction of fragment pairs from the same gene that
+were correctly clustered together) and {\it Type 1 error} (the
+fraction of genes that were divided between clusters)
+\cite{Wang04,Hazelhurst08a}; other measurements are also discussed in
+the supplementary materials.  In Figure~2(left) we see a significant
+improvement of {\sc Peace} (blue) over {\sc wcd} (green) and {\sc
+  Cap3} (black) in its sensitivity to sequencing errors, while in
+Figure~2(right) we see a comparable improvement in Type 1 error.  In
+Figure~3 we look at the number of singleton clusters (fragments not
+joined to any cluster), which will not occur in our simulated sets; we
+again see significant improvement in {\sc Peace}.  The sequential
+runtime of {\sc Peace} is slightly improved over that of {\sc wcd}
+(see Supplementary Materials, Figure~S4).
 
@@
 sensitivity and Type 1 error (both significantly outperforming {\sc
-  Cap3}).  Using the mouse dataset used in Hazelhurst {\it et
-  al.}  \cite{Hazelhurst08a} {\sc Peace} again shows slight
-improvements in sensitivity and Type 1 error rates, with an 18\% speedup
-for {\sc Peace} (see Supplementary Materials,
-Section~\href{real_results}, for more details).
+  Cap3}).  Using the mouse data set used in Hazelhurst {\it et al.}
+\cite{Hazelhurst08a} {\sc Peace} again shows slight improvements in
+sensitivity and Type 1 error rates, with an 18\% speedup for {\sc
+  Peace} (see Supplementary Materials, Section~D, for more details).
 
@@
 transcripts, or effects from poorly cleaned transcript data (e.g. the
-failure to remove sequencing adapters), post-transcriptional
-poly(A)/(T) tails, and low-complexity repeats can cause similar
+failure to remove sequencing adapters or post-transcriptional
+poly(A)/(T) tails), and low-complexity repeats can cause similar
 effects in these clustering tools.  The problems can be handled

### 2010-02-13 papers/NAR_web_server_2009/paper.tex
Several changes to the paper to actually make it more LaTex friendly. One of the important modifications is to change Figure~1 to Figure~\ref{fig:workflow}. Figures need to be referred. Otherwise links and layout will be broken.
I have done these references only to a few spots and we need to scan the 
paper for MS-Word like use of Figure~2(c) etc. and replace them with LaTex
references.

I have also replaced all uses of {\sc peace} with custom macro \peace which
will permit us to conveniently refer and change the formatting by touching
just 1 line.

I cropped several pdf files with pdfcrop to

@@
 ]{hyperref}
+\usepackage{wrapfig}
 
@@
 
-%\articlesubtype{This is the article type (optional)}
+% \articlesubtype{This is the article type (optional)}
+
+% Some shortcuts to make life easier when writing latex
+\newcommand{\peace} {{\small PEACE}}
 
@@
 \begin{abstract}
-  We present {\sc Peace}, a stand alone tool for the high-throughput
-  {\it ab initio} clustering of transcript fragment sequences produced
-  by Next Generation or Sanger Sequencing technologies.  Installed and
-  run through an easily downloaded GUI, {\sc Peace} can process large
+  We present \peace, a stand alone tool for the high-throughput {\it
+    ab initio} clustering of transcript fragment sequences produced by
+  Next Generation or Sanger Sequencing technologies.  Installed and
+  run through an easily downloaded GUI, \peace\/ can process large
   data sets and group the fragments by gene association, achieving a
   greater sensitivity (without sacrificing runtime) than the competing
-  clustering tools.  Through the GUI the user can collect
-  cluster statistics and examine specific clusters for more
-  comprehensive study.  {\sc Peace}, used as a ``pre-assembly'' tool,
-  enables any assembly tool to work with individual small clusters of
-  closely related sequences, resulting in significant increases in
-  sensitive and robustness to sequencing error, as well as a reduction in
+  clustering tools.  Through the GUI the user can collect cluster
+  statistics and examine specific clusters for more comprehensive
+  study.  \peace\/, used as a ``pre-assembly'' tool, enables any
+  assembly tool to work with individual small clusters of closely
+  related sequences, resulting in significant increases in sensitive
+  and robustness to sequencing error, as well as a reduction in
   overall runtime.
@@
 larger data-set size produced by NGS technologies exacerbates this
-problem.  To deal with this, {\sc Peace} combines the $d^2$
+problem.  To deal with this, \peace\/ combines the $d^2$
 alignment-free sequence distance function \cite{Hide94} and the
@@
 accurately find clusters of ESTs expressed from the same gene without
-reference to a sequenced genome.  Compared against competing tools
-for the same purpose \cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
-  Ptitsyn05,Hazelhurst08a,Picardi09}, {\it Peace} proves to be both more sensitive and
-more robust to sequencing error without sacrificing runtime.
+reference to a sequenced genome.  Compared against competing tools for
+the same purpose
+\cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
+  Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be both more
+sensitive and more robust to sequencing error without sacrificing
+runtime.
 
-In short, {\sc Peace} is computational tool for the {\it ab initio}
-clustering of transcript fragments by gene association, applicable to both NGS and
-traditional Sanger Sequencing technologies.  Available through the
-\href{http://www.peace-tools.org}{www.peace-tools.org} website, the
-{\sc Peace} GUI allows the user to both easily install (locally or
-remotely) and run the clustering engine, as well as enabling
-transparent parallel processing and providing various tools for result
-analysis.
+In short, \peace\/ is computational tool for the {\it ab initio}
+clustering of transcript fragments by gene association, applicable to
+both NGS and traditional Sanger Sequencing technologies.  Available
+through the \href{http://www.peace-tools.org}{www.peace-tools.org}
+website, the \peace\/ GUI allows the user to both easily install
+(locally or remotely) and run the clustering engine, as well as
+enabling transparent parallel processing and providing various tools
+for result analysis.
 
@@
 
-\section{{\sc Peace}: Installation and Use}
+\section{\peace\/: Installation and Use}
 
-The {\sc Peace} GUI can be launched by downloading and executing the
-Jar file available on the {\sc Peace} website
+The \peace\/ GUI can be launched by downloading and executing the
+Jar file available on the \peace\/ website
 (\href{http://www.peace-tools.org}{www.peace-tools.org}) to any
-machine running the standard Java Virtual Machine (JVM).  Once running,
-the user can employ the GUI to install the computational tool and perform
-a clustering of a data file in the FASTA format, view an initial
-analysis of the clusters, and produce files containing subsets of the
-clusters as input to assembly tools such as {\sc Cap3} \cite{Huang99}.
-A typical (first) use of {\sc Peace} generally
-requires the following steps.
+machine running the standard Java Virtual Machine (JVM).  Once
+running, the user can employ the GUI to install the computational tool
+and perform a clustering of a data file in the FASTA format, view an
+initial analysis of the clusters, and produce files containing subsets
+of the clusters as input to assembly tools such as {\sc Cap3}
+\cite{Huang99}.  A typical (first) use of \peace\/ must be
+performed in the following manner (see Figure~\ref{fig:workflow}):
 
-\noindent {\bf Tool Installation:} (First use only.)  To install the
-{\sc Peace} clustering onto a local or remote machine, the
-user selects from within the GUI the appropriate menu tab
-(Figure~1(a)), which then starts an install wizard that will prompt
-for the appropriate information.  Figure~1(b) illustrates the request
-for server information; the user has chosen to install the {\sc Peace}
-computational tool on a remote machine and is providing the necessary
-connection information.  Server information is persistent between GUI
-sessions, giving the user access to {\sc Peace} on the target machine
-as needed.
+\begin{figure}
+  \centerline{\includegraphics[width=3in]{screen.d/workflow-crop.pdf}}
+  \caption{Overview of the procedure for clustering and analysis using
+    PEACE.}\label{fig:workflow}
+\end{figure}
 
-\noindent {\bf Job Processing:} After importing the target sequence file into
-the GUI, the user starts a new job by following the wizard menus.
-Figure~1(c) illustrates the process of specifying the number of
-processors available (if running on a machine supporting the OpenMPI
-protocol -- which will be determined during job installation).  Once
-executed, the GUI will manage the job thread, alert the user when the
-job is completed (or when the user next runs the GUI after completion),
-and copy the final results back to the local machine if necessary.
+\noindent {\bf Tool Installation:} (First use) To install the {\sc
+  Peace} clustering onto a local or remote machine, the user selects
+from within the GUI the appropriate menu tab (Figure~\ref{screen}(a)),
+which then starts an install wizard that will prompt for the
+appropriate information.  Figure~\ref{screen}(b) illustrates the
+request for server information; the user has chosen to install the
+\peace\/ computational tool on a remote machine and is providing
+the necessary connection information.  Server information is
+persistent between GUI sessions, giving the user access to \peace\/
+on the target machine as needed.
+
+\begin{figure*}
+  \centerline{ 
+    \includegraphics[trim=0cm 3cm 0cm 1cm, clip, scale=0.9]{screen.d/screen.pdf}
+  }
+  \caption{Screenshots of the \peace\/ GUI during execution, including (a)
+    GUI Welcome and server installation menu; (b) setup wizard for
+    installing the computational tool on a remote server; (c) execution
+    wizard for starting a selected job to be executed in parallel mode;
+    (d) basic cluster output; and (e) histogram view of cluster results
+    and classifier editor for setting up differential expression profiles.}\label{screen}
+\end{figure*}
+
+\noindent {\bf Job Processing:} After importing the target sequence
+file into the GUI, the user starts a new job by following the wizard
+menus.  Figure~\ref{screen}(c) illustrates the process of specifying
+the number of processors available (if running on a machine supporting
+the OpenMPI protocol -- which will be determined during job
+installation).  Once executed, the GUI will manage the job thread,
+alert the user when the job is completed (or when the user next runs
+the GUI after completion), and copy the final results back to the
+local machine if necessary.
 
@@
 \item {\bf Export:} The user can export the contents of one or more
-  clusters into a FASTA format file, obtaining a subset of the original target file
-  containing the sequences corresponding to the selected clusters
-  ready for processing by an assembly tool (e.g. {\sc Cap3} \cite{Huang99}).
+  clusters into a FASTA format file, obtaining a subset of the
+  original target file containing the sequences corresponding to the
+  selected clusters ready for processing by an assembly tool
+  (e.g. {\sc Cap3} \cite{Huang99}).
 \item {\bf View Clustering}: The user may view a list of clusters,
-  expanding selected clusters to a list of all individual sequences (illustrated
-  in Figure~1(d)).
+  expanding selected clusters to a list of all individual sequences
+  (illustrated in Figure~\ref{screen}(d)).
 \item {\bf Classified Summary Graph:} The user may view a distribution
@@
   string patterns denoting different cDNA libraries, the classifier
-  can help the user to determine and visualize the differential expression profiles
-  of different libraries for a given cluster.  The method of setting up these classifiers, and the
-  resulting histogram, is illustrated in Figure~1(e).
+  can help the user to determine and visualize the differential
+  expression profiles of different libraries for a given cluster.  The
+  method of setting up these classifiers, and the resulting histogram,
+  is illustrated in Figure~\ref{screen}(e).
 \end{itemize}
 Extensive documentation for the tool has been posted on the {\sc
-  Peace} website, as well as links to several tutorial videos demonstrating
-{\sc Peace} use and capabilities.
+  Peace} website, as well as links to several tutorial videos
+demonstrating \peace\/ use and capabilities.
 
@@
 
-The clustering performed by {\sc Peace} is based on the use of minimum
+The clustering performed by \peace\/ is based on the use of minimum
 spanning trees (MSTs), known to be an effective approach for narrow
@@
 
-{\sc Peace} approaches the problem by generating a minimum spanning tree of the
+\peace\/ approaches the problem by generating a minimum spanning tree of the
 described graph, then removing edges exceeding our threshold.  By
@@
 
-{\sc Peace} has been test on both simulated and real data from NGS and
+\peace\/ has been test on both simulated and real data from NGS and
 Sanger Sequencing technologies, comparing results against those
@@
 the supplementary materials.  In Figure~2(left) we see a significant
-improvement of {\sc Peace} (blue) over {\sc wcd} (green) and {\sc
+improvement of \peace\/ (blue) over {\sc wcd} (green) and {\sc
   Cap3} (black) in its sensitivity to sequencing errors, while in
@@
 joined to any cluster), which will not occur in our simulated sets; we
-again see significant improvement in {\sc Peace}.  The sequential
-runtime of {\sc Peace} is slightly improved over that of {\sc wcd}
+again see significant improvement in \peace.  The sequential
+runtime of \peace\/ is slightly improved over that of {\sc wcd}
 (see Supplementary Materials, Figure~S4).
 
-\begin{figure*}[b]
-\centerline{
-\includegraphics[trim=0cm 3cm 0cm 1cm, clip, scale=0.9]{screen.d/screen.pdf} 
-}
-\label{screen}
-\caption{Screenshots of the {\sc Peace} GUI during execution, including (a)
-GUI Welcome and server installation menu; (b) setup wizard for
-installing the computational tool on a remote server; (c) execution
-wizard for starting a selected job to be executed in parallel mode;
-(d) basic cluster output; and (e) histogram view of cluster results
-and classifier editor for setting up differential expression profiles.}
-\end{figure*}
-
-\begin{figure*}[h]
-\centerline{
-\begin{minipage}[b]{2in}
-\includegraphics[scale=0.5]{pics.d/SeT1.pdf} 
-\end{minipage}{2in}
-}
+\begin{figure}
+  \centerline{\includegraphics[width=3.35in]{pics.d/SeT1.pdf}}
   \label{SeT1}
@@
     (see Supplementary Materials, Section~\href{sim_results}, for more
-  details).   Blue/Solid = {\sc Peace}, Green/Dash =
+    details).   Blue/Solid = \peace, Green/Dash =
     {\sc wcd}, Black/Dot-Dash = {\sc Cap3}; vertical tics = 95\% confidence
@@
     error due to the extremely small variance.}
-\end{figure*}
+\end{figure}
 
-\begin{figure*}[b]
-\begin{minipage}{2in}
-\centerline{
-\includegraphics[scale=0.35]{pics.d/singletons.pdf}
-}
-\end{minipage}
+\begin{figure}
+  \centerline{\includegraphics[width=3in]{pics.d/singletons.pdf}}
   \label{singletons}
@@
     when run on the simulated sequences; correct answer in all cases is
-  zero.  Blue/Solid = {\sc Peace}, Green/Dash = {\sc wcd},
+    zero.  Blue/Solid = \peace, Green/Dash = {\sc wcd},
     Black/Dot-Dash = {\sc Cap3}.}
-\end{figure*}
+\end{figure}
 
@@
 fragment set to the genome, taking this as our reference clustering.
-We see slight improvements in {\sc Peace} over {\sc wcd} in both
+We see slight improvements in \peace\/ over {\sc wcd} in both
 sensitivity and Type 1 error (both significantly outperforming {\sc
   Cap3}).  Using the mouse data set used in Hazelhurst {\it et al.}
-\cite{Hazelhurst08a} {\sc Peace} again shows slight improvements in
+\cite{Hazelhurst08a} \peace\/ again shows slight improvements in
 sensitivity and Type 1 error rates, with an 18\% speedup for {\sc
@@
 
-Here we have presented {\sc Peace}, a stand alone tool for the
+Here we have presented \peace, a stand alone tool for the
 high-throughput clustering of transcript fragments, managed through a
@@
 execution in sequential or parallel mode.  Based on a novel algorithm
-for the clustering of the fragments by gene association, {\sc Peace}
+for the clustering of the fragments by gene association, \peace\/
 shows significant improvement in sensitivity, without sacrificing runtime,
@@
 
-As a clustering tool based on sequence distance, {\sc Peace} faces
-certain inherent limitations. For example, {\sc Peace} cannot handle
+As a clustering tool based on sequence distance, \peace\/ faces
+certain inherent limitations. For example, \peace\/ cannot handle
 duplicate genes; like {\sc WCD}, it is unable to separate clusters
@@
 \vspace{3mm}
-%Bib TeX
+% Bib TeX
 

### 2010-02-13 peace/src/TVHeuristic.cpp
Fixed a compile warning and added a missing check to the adjustParameters function in TVHeuristic class.

@@ -162,7 +162,10 @@ TVHeuristic::adjustParameters(const int otherESTLen) {
     } else {
         windowLen = 50;
     }
-    t = 6 + 0.25*minLen;
+    t = 6 + (int) (0.25*minLen);
+    if (t > 125) {
+        t = 125;
+    }
 }
 
 void

### 2010-02-13 peace/src/FWAnalyzer.cpp
Redesign of two-pass D2 to allow for the dynamic calculation of parameters including threshold and frame size.  Modifications to FWAnalyzer and MSTCluster to facilitate this change.

@@ -43,13 +43,13 @@
 #include <time.h>
 
 // The static instance variables for command line arguments.
-int FWAnalyzer::frameSize = 50; // 100 is default from WCD implementation
+//int FWAnalyzer::frameSize = 50; // 100 is default from WCD implementation
 int FWAnalyzer::wordSize  = 6;   // 6   is default from CLU implementation.
 
 // The common set of arguments for all FW EST analyzers
 arg_parser::arg_record FWAnalyzer::commonArgsList[] = {
-    {"--frame", "Frame size (in base pairs, default=100)",
-     &FWAnalyzer::frameSize, arg_parser::INTEGER},
+    //{"--frame", "Frame size (in base pairs, default=100)",
+    // &FWAnalyzer::frameSize, arg_parser::INTEGER},
     {"--word", "Word size (in base pairs, default=6)",
      &FWAnalyzer::wordSize, arg_parser::INTEGER},    
     {NULL, NULL, NULL, arg_parser::BOOLEAN}
@@ -58,6 +58,7 @@ arg_parser::arg_record FWAnalyzer::commonArgsList[] = {
 FWAnalyzer::FWAnalyzer(const std::string& analyzerName, const int refESTidx,
                        const std::string& outputFile)
     : ESTAnalyzer(analyzerName, refESTidx, outputFile) {
+    frameSize = 50;
     // Nothing else to be done here for now.
 }
 

### 2010-02-13 peace/src/FWAnalyzer.h

@@ -329,7 +329,7 @@ protected:
         the parseArguments method depending on the actual value
         specified by the user.
     */
-    static int frameSize;
+    int frameSize;
 
     /** The word size to be used by this analyzer.
 

### 2010-02-13 peace/src/TwoPassD2.cpp

@@ -40,7 +40,7 @@
 #include <algorithm>
 
 // Skip parameter for d2 asymmetric.
-int TwoPassD2::frameShift = 50;
+//int TwoPassD2::frameShift = 50;
 
 // The bitmak to be used when build hash values.
 int TwoPassD2::BitMask   = 0;
@@ -51,21 +51,21 @@ int TwoPassD2::bitShift  = 0;
 
 // The threshold score below which two ESTs are considered
 // sufficiently similar to be clustered.
-int TwoPassD2::threshold = 0;
+int TwoPassD2::minThreshold = 0;
 
 // The threshold score above which two ESTs are considered
 // sufficiently dissimilar that symmetric bounded D2 does not
 // need to be run.
-int TwoPassD2::maxThreshold = 130;
+//int TwoPassD2::maxThreshold = 130;
 
 // The set of arguments for this class.
 arg_parser::arg_record TwoPassD2::argsList[] = {
-    {"--frameShift", "Frame Shift (default=50)",
-     &TwoPassD2::frameShift, arg_parser::INTEGER},
+    //{"--frameShift", "Frame Shift (default=50)",
+    // &TwoPassD2::frameShift, arg_parser::INTEGER},
     {"--threshold", "Threshold score to break out of D2 (default=0)",
-     &TwoPassD2::threshold, arg_parser::INTEGER},    
-    {"--maxThreshold", "Threshold score to run bounded symmetric D2 (default=130)",
-     &TwoPassD2::maxThreshold, arg_parser::INTEGER},    
+     &TwoPassD2::minThreshold, arg_parser::INTEGER},    
+    //{"--maxThreshold", "Threshold score to run bounded symmetric D2 (default=130)",
+    // &TwoPassD2::maxThreshold, arg_parser::INTEGER},    
     {NULL, NULL, NULL, arg_parser::BOOLEAN}
 };
 
@@ -75,6 +75,12 @@ TwoPassD2::TwoPassD2(const int refESTidx, const std::string& outputFileName)
     s2WordTable     = NULL;
     delta           = NULL;
     alignmentMetric = 0;
+    refESTLen = 0;
+    
+    // Below are formerly static parameters that are now dynamic.
+    frameShift = 50;
+    maxThreshold = 130;
+    threshold = 40;
 }
 
 TwoPassD2::~TwoPassD2() {
@@ -152,6 +158,7 @@ TwoPassD2::setReferenceEST(const int estIdx) {
         // init ref-est word table
[…]

### 2010-02-13 src/MSTCluster.cpp
Redesign of two-pass D2 to allow for the dynamic calculation of parameters including threshold and frame size.  Modifications to FWAnalyzer and MSTCluster to facilitate this change.

@@ -93,7 +93,7 @@ MSTCluster::add(MSTCluster *child) {
 
 double
 MSTCluster::makeClusters(NodeList& nodeList, const ESTAnalyzer* analyzer) {
-    const float threshold = 52;
+    const float threshold = 1;
     // Now create a hash map to track cluster for a given node
     ClusterMap clusterMap;
     // Now extract nodes from the nodeList and add it to appropriate

### 2010-02-13 src/TwoPassD2.h

@@ -42,6 +42,14 @@
 class EST;
 class ResultLog;
 
+// Define some constants for the dynamic frame/threshold adjustment
+// functionality, in order to avoid magic numbers in the code
+
+#define MIN_FRAME_SIZE     50
+#define MAX_FRAME_SIZE    500
+#define COVERAGE_FRACTION 0.3
+#define FRAME_SIZE_ADJUST 1.2
+
 /** \brief EST Analyzer that uses the D2 algorithm in both its
 	symmetric and asymmetric variations to compute distance between
 	two ESTs.
@@ -349,10 +357,8 @@ protected:
         shifting the frame on the reference sequence, resulting in
         fewer computations but a possible loss of accuracy from not
         comparing every frame in both sequences.
-
-        \note Currently this value is not used.
     */
-    static int frameShift;
+    int frameShift;
     
     /** Instance variable to store the number of bits to be shifted to
         create hash values.
@@ -379,10 +385,10 @@ protected:
         to break out of the core D2 loop.  This value essentially
         represents the D2 score below which two ESTs are considered
         sufficiently similar. Currently, the default threshold value
-        is set to 40.  This value can be overridden by the user with
+        is set to 0.  This value can be overridden by the user with
         the "--threshold" command line argument.
     */
-    static int threshold;
+    static int minThreshold;
 	
     /** The threshold score above which two ESTs are considered
         sufficiently dissimilar that a more accurate symmetric
@@ -395,7 +401,9 @@ protected:
         130.  This value can be overridden by the user with the
         "--maxThreshold" command line argument.
     */
-    static int maxThreshold;
+    int maxThreshold;
+
+    int threshold;
     
 private:
     /** The set of arguments specific to the D2 algorithm
@@ -465,6 +473,11 @@ private:
     */
     int numWordsInWindow;
 
+    /** Length of the reference EST.  Stored to avoid making multiple
+	unnecessary strlen() calls.
+     */
+    int refESTLen;
+
     /** Helper method to update the delta table as well as the
         minimum score (if it has changed) when the window is shifted.
 		
@@ -516,6 +529,13 @@ private:
             alignmentMetric = windowDistance;
         }
     }
+
+    /** Method to dynamically compute parameters such as frame size
+	and threshold.  Sets these parameters to their (hopefully)
+	optimum values for better clustering of data sets with wide
+	variation in length, such as data generated by 454 pyrosequencing.
+    */
[…]

### 2010-02-13 peace/papers/NAR_web_server_2009/paper.tex
Some more Rao-magic applied to the formatting of the paper. Specifically, I performed the following changes:
1. Increased the DPI of all images to 300 DPI via gimp. gimp did a wonderful
   job on them.

2. Modified the paper to directly import the png (which LaTex shines at) so
   the complete 300 DPI images are exposed directly in the PDF.

3. This makes the images look pretty decent at 100% and even at 400% zooms.

4. The images print pretty decently on a color Laser. Did not check on a 
   B&W laser tho'

5. Set the layout to be controlled by latex rather than include the images
   as a sin

@@
 
-But clustering is a computationally challenging problem.  Even with
-the smaller number of ESTs produced using Sanger Sequencing, the
+However clustering is a computationally challenging problem.  Even
+with the smaller number of ESTs produced using Sanger Sequencing, the
 runtime and memory requirements to cluster based on pair-wise sequence
@@
 
-\section{{\sc Peace}: Installation and Use}
+\section{\peace\/: Installation and Use}
 
-The {\sc Peace} GUI can be launched by downloading and executing the
-Jar file available on the {\sc Peace} website
+The \peace\/ GUI can be launched by downloading and executing the
+Jar file available on the \peace\/ website
 (\href{http://www.peace-tools.org}{www.peace-tools.org}) to any
-machine running the standard Java Virtual Machine (JVM).  Once running,
-the user can employ the GUI to install the computational tool and perform
-a clustering of a data file in the FASTA format, view an initial
-analysis of the clusters, and produce files containing subsets of the
-clusters as input to assembly tools such as {\sc Cap3} \cite{Huang99}.
-A typical (first) use of {\sc Peace} generally
-requires the following steps.
+machine running the standard Java Virtual Machine (JVM).  Once
+running, the user can employ the GUI to install the computational tool
+and perform a clustering of a data file in the FASTA format, view an
+initial analysis of the clusters, and produce files containing subsets
+of the clusters as input to assembly tools such as {\sc Cap3}
+\cite{Huang99}.  A typical (first) use of \peace\/ must be
+performed in the following manner (see Figure~\ref{fig:workflow}):
 
-\noindent {\bf Tool Installation:} (First use only.)  To install the
-{\sc Peace} clustering onto a local or remote machine, the
-user selects from within the GUI the appropriate menu tab
-(Figure~1(a)), which then starts an install wizard that will prompt
-for the appropriate information.  Figure~1(b) illustrates the request
-for server information; the user has chosen to install the {\sc Peace}
-computational tool on a remote machine and is providing the necessary
-connection information.  Server information is persistent between GUI
-sessions, giving the user access to {\sc Peace} on the target machine
-as needed.
+\begin{figure}
+  \centerline{\includegraphics[width=3in]{screen.d/workflow.pdf}}
+  \caption{Overview of the procedure for clustering and analysis using
+    PEACE.}\label{fig:workflow}
+\end{figure}
 
-\noindent {\bf Job Processing:} After importing the target sequence file into
-the GUI, the user starts a new job by following the wizard menus.
-Figure~1(c) illustrates the process of specifying the number of
-processors available (if running on a machine supporting the OpenMPI
-protocol -- which will be determined during job installation).  Once
-executed, the GUI will manage the job thread, alert the user when the
-job is completed (or when the user next runs the GUI after completion),
-and copy the final results back to the local machine if necessary.
+\noindent {\bf Tool Installation:} (First use) To install the {\sc
+  Peace} clustering onto a local or remote machine, the user selects
+from within the GUI the appropriate menu tab (Figure~\ref{screen}(a)),
+which then starts an install wizard that will prompt for the
+appropriate information.  Figure~\ref{screen}(b) illustrates the
+request for server information; the user has chosen to install the
+\peace\/ computational tool on a remote machine and is providing
+the necessary connection information.  Server information is
+persistent between GUI sessions, giving the user access to \peace\/
+on the target machine as needed.
+
+\begin{figure*}
+  \begin{minipage}{3in}
+    \includegraphics[width=3in]{screen.d/install_page_big.png}
+    \centerline{\small{(a)}}
+  \end{minipage}
+  \begin{minipage}{2in}
+    \vspace{0.47in}
+    \includegraphics[width=2in]{screen.d/add_server_big.png}
+    \centerline{\small{(b)}}
+  \end{minipage}
+  \begin{minipage}{2in}
+    \vspace{0.47in}
+    \includegraphics[width=2in]{screen.d/add_server_big.png}
+    \centerline{\small{(c)}}
+  \end{minipage}
+
+  \begin{minipage}{3in}
+    \includegraphics[width=3in]{screen.d/cluster_list_big.png}
+    \centerline{\small{(d)}}
+  \end{minipage}
+  \begin{minipage}{1in}
+    \hspace{1in}
+  \end{minipage}
+  \begin{minipage}{3in}
+    \includegraphics[width=3in]{screen.d/classifier_big.png}
+    \centerline{\small{(e)}}
+  \end{minipage}
+
+  \caption{Screenshots of the \peace\/ GUI during execution, including (a)
+    GUI Welcome and server installation menu; (b) setup wizard for
+    installing the computational tool on a remote server; (c) execution
+    wizard for starting a selected job to be executed in parallel mode;
+    (d) basic cluster output; and (e) histogram view of cluster results
+    and classifier editor for setting up differential expression profiles.}\label{screen}
+\end{figure*}
+
+\noindent {\bf Job Processing:} After importing the target sequence
+file into the GUI, the user starts a new job by following the wizard
+menus.  Figure~\ref{screen}(c) illustrates the process of specifying
+the number of processors available (if running on a machine supporting
+the OpenMPI protocol -- which will be determined during job
+installation).  Once executed, the GUI will manage the job thread,
+alert the user when the job is completed (or when the user next runs
+the GUI after completion), and copy the final results back to the
+local machine if necessary.
 
@@
 \cite{Wang04,Hazelhurst08a}; other measurements are also discussed in
-the supplementary materials.  In Figure~2(left) we see a significant
-improvement of {\sc Peace} (blue) over {\sc wcd} (green) and {\sc
-  Cap3} (black) in its sensitivity to sequencing errors, while in
-Figure~2(right) we see a comparable improvement in Type 1 error.  In
-Figure~3 we look at the number of singleton clusters (fragments not
-joined to any cluster), which will not occur in our simulated sets; we
-again see significant improvement in {\sc Peace}.  The sequential
-runtime of {\sc Peace} is slightly improved over that of {\sc wcd}
-(see Supplementary Materials, Figure~S4).
+the supplementary materials.  In Figure~\ref{SeT1} (left) we see a
+significant improvement of \peace\/ (blue) over {\sc wcd} (green) and
+{\sc Cap3} (black) in its sensitivity to sequencing errors, while in
+Figure~\ref{SeT1} (right) we see a comparable improvement in Type 1
+error.  In Figure~\ref{singletons} we look at the number of singleton
+clusters (fragments not joined to any cluster), which will not occur
+in our simulated sets; we again see significant improvement in \peace.
+The sequential runtime of \peace\/ is slightly improved over that of
+{\sc wcd} (see Supplementary Materials, Figure~S4).
 
-\begin{figure*}[b]
-\centerline{
-\includegraphics[trim=0cm 3cm 0cm 1cm, clip, scale=0.9]{screen.d/screen.pdf} 
-}
-\label{screen}
-\caption{Screenshots of the {\sc Peace} GUI during execution, including (a)
-GUI Welcome and server installation menu; (b) setup wizard for
-installing the computational tool on a remote server; (c) execution
-wizard for starting a selected job to be executed in parallel mode;
-(d) basic cluster output; and (e) histogram view of cluster results
-and classifier editor for setting up differential expression profiles.}
-\end{figure*}
+\begin{figure}
+  \centerline{\includegraphics[width=3.35in]{pics.d/SeT1.pdf}}
+  \caption{Comparisons of Sensitivity and Type 1 error, based on the
+    average over 30 simulated fragment sets derived from 100 zebra
+    fish genes (see Supplementary Materials, Section C, for more
+    details).  Blue/Solid = \peace, Green/Dash = {\sc wcd},
+    Black/Dot-Dash = {\sc Cap3}; vertical tics = 95\% confidence
+    intervals on estimates.  Intervals are not presented for Type 1
+    error due to the extremely small variance.}\label{SeT1}
+\end{figure}
 
-\begin{figure*}[h]
-\centerline{
-\begin{minipage}[b]{2in}
-\includegraphics[scale=0.5]{pics.d/SeT1.pdf} 
-\end{minipage}{2in}
-}
-\label{SeT1}
-\caption{Comparisons of Sensitivity and Type 1 error, based on the
-  average over 30 simulated fragment sets derived from 100 zebra fish genes
-  (see Supplementary Materials, Section~\href{sim_results}, for more
-  details).   Blue/Solid = {\sc Peace}, Green/Dash =
-  {\sc wcd}, Black/Dot-Dash = {\sc Cap3}; vertical tics = 95\% confidence
-  intervals on estimates.  Intervals are not presented for Type 1
-  error due to the extremely small variance.}
-\end{figure*}
-
-\begin{figure*}[b]
-\begin{minipage}{2in}
-\centerline{
-\includegraphics[scale=0.35]{pics.d/singletons.pdf}
-}
-\end{minipage}
-\label{singletons}
+\begin{figure}
+  \centerline{\includegraphics[width=3in]{pics.d/singletons.pdf}}
   \caption{Average number of fragments flagged as singletons by each tool
     when run on the simulated sequences; correct answer in all cases is
-  zero.  Blue/Solid = {\sc Peace}, Green/Dash = {\sc wcd},
-  Black/Dot-Dash = {\sc Cap3}.}
-\end{figure*}
+    zero.  Blue/Solid = \peace, Green/Dash = {\sc wcd},
+    Black/Dot-Dash = {\sc Cap3}.}\label{singletons}
+\end{figure}
 

### 2010-02-13 src/EST.cpp
Fixed a bug with the new code.

@@ -48,9 +48,6 @@ std::vector<EST*> EST::estList;
 // The length of the longest EST we have
 size_t EST::maxESTlen = 0;
 
-// A "dummy" EST for returning in place of filtered-out ESTs
-EST* dummy = new EST(-1, "", "", 0);
-        
 EST::EST(const int idValue, const char *information, const char* seq,
          const int fileOffset) : id(idValue), offset(fileOffset),
                                  customData(NULL) {

### 2010-02-13 src/NewUVHeuristic.cpp

@@ -40,11 +40,11 @@
 #include "EST.h"
 
 // default params
-int NewUVHeuristic::u           = 4;
+int NewUVHeuristic::u           = 6;
 int NewUVHeuristic::v           = 8;
-int NewUVHeuristic::wordShift   = 4;
+int NewUVHeuristic::wordShift   = 8;
 int NewUVHeuristic::BitMask     = 0;
-int NewUVHeuristic::passes      = 2;
+int NewUVHeuristic::passes      = 3;
 
 // Instance variable to track number of bits to shift for building
 // hash values.  This is set to 2*(v-1) in initialize method.

### 2010-02-13 src/TVHeuristic.cpp

@@ -101,9 +101,8 @@ TVHeuristic::initialize() {
     // First compute the longest EST we have.
     size_t maxESTlen = EST::getMaxESTLen();
     
-    // Add extra 100 characters to ease processing.
-    //matchTable = new char[maxESTlen + windowLen + v];
-    matchTable = new char[maxESTlen + 150 + v];
+    // Add extra windowLen characters to ease processing.
+    matchTable = new char[maxESTlen + MAX_WINDOW_LEN + v];
     // Everything went well
     return 0;
 }
@@ -155,7 +154,9 @@ TVHeuristic::runHeuristic(const int otherEST) {
 void
 TVHeuristic::adjustParameters(const int otherESTLen) {
     int minLen = std::min(refESTLen, otherESTLen);
-    if (minLen >= 300) {
+    if (minLen >= 500) {
+        windowLen = MAX_WINDOW_LEN;
+    } else if (minLen >= 300) {
         windowLen = 150;
     } else if (minLen >= 100) {
         windowLen = 100;
@@ -163,8 +164,8 @@ TVHeuristic::adjustParameters(const int otherESTLen) {
         windowLen = 50;
     }
     t = 6 + (int) (0.25*minLen);
-    if (t > 125) {
-        t = 125;
+    if (t > 150) {
+        t = 150;
     }
 }
 

### 2010-02-13 src/TVHeuristic.h

@@ -38,6 +38,8 @@
 #include "EST.h"
 #include <utility>
 
+#define MAX_WINDOW_LEN 200
+
 /** Heuristic based upon the T/V heuristic used in WCD, a type of
     common word heuristic.
 
@@ -215,7 +217,7 @@ protected:
             hash = encoder(hash, otherSeq[i]);
         }
         // Set first window length entries to zero.
-        memset(matchTable, 0, sizeof(char) *( windowLen + v));
+        memset(matchTable, 0, sizeof(char) *( windowLen + NewUVHeuristic::v));
         // Skip first windowLen entries to simplify logic in loop below.
         char *matchTicker = matchTable + windowLen;
         // Now see how many common words exist in the two ESTs

### 2010-02-13 src/TwoPassD2.cpp

@@ -213,7 +213,7 @@ TwoPassD2::updateParameters(const int otherESTLen) {
     if (greaterLen >= 200) {
         int smallerLen = std::min(refESTLen, otherESTLen);
         coverageNum = (float)std::min(COVERAGE_FRACTION*greaterLen,
-                                      (double)smallerLen);
+                                      ((double)smallerLen)/FRAME_SIZE_ADJUST);
     } else if (greaterLen >= 100) {
         coverageNum = 60;
     } else {
@@ -241,7 +241,7 @@ TwoPassD2::runD2Asymmetric(const int otherEST, int* s1MinScoreIdx,
     const EST * estS2  = EST::getEST(otherEST);
     const char* sq2    = estS2->getSequence();
     const int   sq2Len = strlen(sq2);
-    
+
     // Build the word table for otherEST depending on normal or
     // reverse complement suggestion using hint UVSampleHeuristic.
     int bestMatchIsRC = 0;

### 2010-02-13 papers/NAR_web_server_2009/paper.tex
Edits to manuscript; additon of some short-read results

@@
 \newcommand{\peace} {{\small PEACE}}
+\newcommand{\wcd} {{\small WCD}}
+\newcommand{\capthree} {{\small Cap3}}
 
@@
   greater sensitivity (without sacrificing runtime) than the competing
-  clustering tools.  Through the GUI the user can collect cluster
-  statistics and examine specific clusters for more comprehensive
-  study.  \peace\/, used as a ``pre-assembly'' tool, enables any
-  assembly tool to work with individual small clusters of closely
-  related sequences, resulting in significant increases in sensitive
-  and robustness to sequencing error, as well as a reduction in
-  overall runtime.
+  clustering tools and applicable to bases as small as 75 bases.
+  Through the GUI the user can collect cluster statistics and examine
+  specific clusters for more comprehensive study.  \peace\/, used as a
+  ``pre-assembly'' tool, enables any assembly tool to work with
+  individual small clusters of closely related sequences, resulting in
+  significant increases in sensitive and robustness to sequencing
+  error (e.g. with a 52\% improvement in sensitivity over the cluster
+  tool, \wcd\/, and a 184\% improvement in sensitivity over a leading
+  assembly tool, \capthree, given a a 5\% rate of sequencing errors), as
+  well as a reduction in runtime (with a 40\% speedup, as compared to
+  \wcd, when applied to a 361 Mb set of Arabidopsis transcriptome fragments).
 \end{abstract}
@@
 and helped shed light on the biological processes both controlling and
-dictated by the genome \cite{Nagaraj07}.  However, to access those
+controlled by the genome \cite{Nagaraj07}.  However, to access those
 transcripts, we must deal with the fragmented data produced by
@@
 
-Until recently, access to a transcriptome sequence was primarily
+, access to a transcriptome sequence was primarily
 through the use of Expressed Sequence Tags (ESTs), single-pass cDNA
@@
 
+\enlargethispage{-65.1pt}
+
 Given a set of transcript fragments sampled from across the genome, a
@@
 allow the application of the assembly tool to individual clusters --
-saving significant amounts of time due to the reduced input size
-\cite{Hazelhurst08a}.
+saving significant amounts of time \cite{Hazelhurst08a}.
 
-But clustering is a computationally challenging problem.  Even with
-the smaller number of ESTs produced using Sanger Sequencing, the
+However clustering is a computationally challenging problem.  Even
+with the smaller number of ESTs produced using Sanger Sequencing, the
 runtime and memory requirements to cluster based on pair-wise sequence
@@
 larger data-set size produced by NGS technologies exacerbates this
-problem.  To deal with this, \peace\/ combines the $d^2$
+problem.  To deal with this, \peace\/ combines our own version of the $d^2$
 alignment-free sequence distance function \cite{Hide94} and the
@@
 
-\enlargethispage{-65.1pt}
-
 \section{\peace\/: Installation and Use}
@@
 machine running the standard Java Virtual Machine (JVM).  Once
-running, the user can employ the GUI to install the computational tool
-and perform a clustering of a data file in the FASTA format, view an
+running, the user can employ the GUI to install the clustering engine
+and perform a clustering of a data file in {\sc fasta} format, view an
 initial analysis of the clusters, and produce files containing subsets
-of the clusters as input to assembly tools such as {\sc Cap3}
+of the clusters as input to assembly tools such as \capthree\/ 
 \cite{Huang99}.  A typical (first) use of \peace\/ must be
@@
 \begin{figure}
-  \centerline{\includegraphics[width=3in]{screen.d/workflow-crop.pdf}}
+  \centerline{\includegraphics[width=3in]{screen.d/workflow.pdf}}
   \caption{Overview of the procedure for clustering and analysis using
@@
 
-\noindent {\bf Tool Installation:} (First use) To install the {\sc
-  Peace} clustering onto a local or remote machine, the user selects
+\noindent {\bf Tool Installation:} (First use only.) To install the
+\peace\/ clustering engine onto a local or remote machine, the user selects
 from within the GUI the appropriate menu tab (Figure~\ref{screen}(a)),
@@
 \begin{figure*}
-  \centerline{ 
-    \includegraphics[trim=0cm 3cm 0cm 1cm, clip, scale=0.9]{screen.d/screen.pdf}
-  }
+  \begin{minipage}{3in}
+    \includegraphics[width=3in]{screen.d/install_page_big.png}
+    \centerline{\small{(a)}}
+  \end{minipage}
+  \begin{minipage}{2in}
+    \vspace{0.47in}
+    \includegraphics[width=2in]{screen.d/add_server_big.png}
+    \centerline{\small{(b)}}
+  \end{minipage}
+  \begin{minipage}{2in}
+    \vspace{0.47in}
+    \includegraphics[width=2in]{screen.d/add_server_big.png}
+    \centerline{\small{(c)}}
+  \end{minipage}
+
+  \begin{minipage}{3in}
+    \includegraphics[width=3in]{screen.d/cluster_list_big.png}
+    \centerline{\small{(d)}}
+  \end{minipage}
+  \begin{minipage}{1in}
+    \hspace{1in}
+  \end{minipage}
+  \begin{minipage}{3in}
+    \includegraphics[width=3in]{screen.d/classifier_big.png}
+    \centerline{\small{(e)}}
+  \end{minipage}
+
   \caption{Screenshots of the \peace\/ GUI during execution, including (a)
@@
   selected clusters ready for processing by an assembly tool
-  (e.g. {\sc Cap3} \cite{Huang99}).
+  (e.g. \capthree\/ \cite{Huang99}).
 \item {\bf View Clustering}: The user may view a list of clusters,
@@
 \end{itemize}
-Extensive documentation for the tool has been posted on the {\sc
-  Peace} website, as well as links to several tutorial videos
+Extensive documentation for the tool has been posted on the \peace\/ website, as well as links to several tutorial videos
 demonstrating \peace\/ use and capabilities.
@@
 spanning trees (MSTs), known to be an effective approach for narrow
-band single linkage clustering \cite{Jain99,Wan08}. Edge weights are
-assigned using the $d^2$ distance measure \cite{Hide94} to represent
-edge weights.  Prim's algorithm \cite{Prim57} then allows us to
-efficiently calculate an MST from which we can infer a high-quality
-clustering solution.
+band single linkage clustering \cite{Jain99,Wan08}.  Using a graph
+structure to model the fragment relationships and the $d^2$ distance
+measure to assign edge weights \cite{Hide94}, we can employ Prim's
+algorithm \cite{Prim57} to efficiently calculate an MST from which we
+can infer a high-quality clustering solution.
 
@@
 similarity to ensure a small distance even in the presence of a
-moderate number of base errors.
+moderate number of base errors.  In practice we employ our own
+variation of $d^2$, the {\it two-pass $d^2$ algorithm}, which
+heuristically searches for a neighborhood of maximum similarity and
+then finding the $d^2$ score within that neighborhood (see
+Supplementary Materials for details).
 
-Fragment input is modeled as a weighted, undirected graph: the fragments
-are represented as nodes, with $d^2$ sequence distances assigned to
-the connecting edges as weights.  Conceptually, we want to remove each
-edge exceeding a threshold score from the complete graph and define
-our partitions by the remaining connected components of the graph.  An
-edge with a large weight connects fragments which are likely unrelated;
-once such edges are removed the components define a series of
-overlaps.  Those fragments that can still be connected by some path
+Fragment input is modeled as a weighted, undirected graph: the
+fragments are represented as nodes, with $d^2$ sequence distances
+assigned to the connecting edges as weights.  Conceptually, we want to
+remove each edge exceeding a threshold score from the complete graph
+and define our partitions by the remaining connected components.  An
+edge with a large weight connects fragments which are likely
+unrelated; once such edges are removed the components define a series
+of overlaps.  Those fragments that can still be connected by some path
 correspond to the same gene.  However, such an approach requires the
-calculation of all edge weights.  Such a task is infeasible both in terms of
-runtime and memory usage for the data set sizes we expect to process.
+calculation of all edge weights.  Given the data set sizes, that a
+task is infeasible both in terms of runtime and memory usage for the
+data set sizes we expect to process.
 
@@
 calculation of a majority of edge distances using the $u/v$ and $t/v$
-filtering heuristics employed in {\sc wcd} \cite{Hazelhurst08a}.
+filtering heuristics employed in \wcd\/ \cite{Hazelhurst08a}.
 These heuristics allow us to quickly dismiss many of the edges as too
@@
 Sanger Sequencing technologies, comparing results against those
-produced by the {\sc wcd} clustering tool \cite{Hazelhurst08a} and the
-{\sc Cap3} assembly tool \cite{Huang99} (which calculates a cluster as
-an implicit, intermediate step).  For our simulation tests we used the
-{\bf ESTSim} tool \cite{Hazelhurst03} to generate simulated transcript
-of varying length fragments under different models of error
-(Supplementary Materials, Section~C.1), generating the fragments from
-the list of 100 zebra fish genes used in the {\sc wcd} testing
-\cite{Hazelhurst08a}.  Tool parameters were taken to match, as closely
-as possible, those used in the {\sc wcd} study (see supplementary
-materials).  Our primary methods of quality assessment were {\it
-  sensitivity} (the fraction of fragment pairs from the same gene that
-were correctly clustered together) and {\it Type 1 error} (the
-fraction of genes that were divided between clusters)
-\cite{Wang04,Hazelhurst08a}; other measurements are also discussed in
-the supplementary materials.  In Figure~2(left) we see a significant
-improvement of \peace\/ (blue) over {\sc wcd} (green) and {\sc
-  Cap3} (black) in its sensitivity to sequencing errors, while in
-Figure~2(right) we see a comparable improvement in Type 1 error.  In
-Figure~3 we look at the number of singleton clusters (fragments not
-joined to any cluster), which will not occur in our simulated sets; we
-again see significant improvement in \peace.  The sequential
-runtime of \peace\/ is slightly improved over that of {\sc wcd}
-(see Supplementary Materials, Figure~S4).
+produced by the \wcd\/ clustering tool \cite{Hazelhurst08a} and the
+\capthree\/ assembly tool \cite{Huang99} (the latter of which
+implicitly calculates a cluster in the process of assembly).  For our
+simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03} to
+generate simulated transcript of varying length fragments under
+different models of error (Supplementary Materials, Section~C.1),
+generating the fragments from the list of 100 zebra fish genes used in
+the \wcd\/ testing \cite{Hazelhurst08a}.  Tool parameters were taken
+to match, as closely as possible, those used in the \wcd\/ study (see
+supplementary materials).  Our primary methods of quality assessment
+were {\it sensitivity} (the fraction of fragment pairs from the same
+gene that were correctly clustered together) and {\it Type 1 error}
+(the fraction of genes that were divided between clusters)
+\cite{Wang04,Hazelhurst08a}.  Specificity was not an issue in
+simulated tests, as both \peace and \wcd were completely successful
+in separating unrelated ESTs (save when dealing with recently
+duplicated genes).  Other measurements are also discussed in the
+supplementary materials.  In Figure~\ref{SeT1} (left), reflecting
+tests done on simulated Sanger EST set, we see a
+significant improvement of \peace\/ (blue) over \wcd\/ (green) and
+\capthree\/ (black) in its sensitivity to sequencing errors, while in
+Figure~\ref{SeT1} (right) we see a comparable improvement in Type 1
+error.  In Figure~\ref{singletons} we look at the number of singleton
+clusters (fragments not joined to any cluster), which should not occur
+in our simulated sets.  We again see significant improvement in \peace.
+The sequential runtime of \peace\/ is slightly improved over that of
+\wcd\/ (see Supplementary Materials, Figure~S4).
 
@@
   \centerline{\includegraphics[width=3.35in]{pics.d/SeT1.pdf}}
-  \label{SeT1}
   \caption{Comparisons of Sensitivity and Type 1 error, based on the
-    average over 30 simulated fragment sets derived from 100 zebra fish genes
-    (see Supplementary Materials, Section~\href{sim_results}, for more
-    details).   Blue/Solid = \peace, Green/Dash =
-    {\sc wcd}, Black/Dot-Dash = {\sc Cap3}; vertical tics = 95\% confidence
-    intervals on estimates.  Intervals are not presented for Type 1
-    error due to the extremely small variance.}
+    average over 30 simulated Sanger Sequence ESTs sets derived from 100 zebra
+    fish genes  (see
+    Supplementary Materials, Section C, for more details).  Blue/Solid
+    = \peace, Green/Dash = \wcd, Black/Dot-Dash = \capthree; vertical
+    tics = 95\% confidence intervals on estimates.  Intervals are not
+    presented for Type 1 error due to the extremely small
+    variance.}\label{SeT1}
 \end{figure}
@@
   \centerline{\includegraphics[width=3in]{pics.d/singletons.pdf}}
-  \label{singletons}
   \caption{Average number of fragments flagged as singletons by each tool
     when run on the simulated sequences; correct answer in all cases is
-    zero.  Blue/Solid = \peace, Green/Dash = {\sc wcd},
-    Black/Dot-Dash = {\sc Cap3}.}
+    zero.  Blue/Solid = \peace, Green/Dash = \wcd,
+    Black/Dot-Dash = \capthree.}\label{singletons}
 \end{figure}
@@
 fragment set to the genome, taking this as our reference clustering.
-We see slight improvements in \peace\/ over {\sc wcd} in both
-sensitivity and Type 1 error (both significantly outperforming {\sc
-  Cap3}).  Using the mouse data set used in Hazelhurst {\it et al.}
+We see slight improvements in \peace\/ over \wcd\/ in both sensitivity
+and Type 1 error (both significantly outperforming \capthree).  Using
+the mouse data set used in Hazelhurst {\it et al.}
 \cite{Hazelhurst08a} \peace\/ again shows slight improvements in
-sensitivity and Type 1 error rates, with an 18\% speedup for {\sc
-  Peace} (see Supplementary Materials, Section~D, for more details).
+sensitivity and Type 1 error rates, with an 18\% speedup for \peace
+(see Supplementary Materials, Section~D, for more details).
+
+A comparative test of \peace against \wcd on NGS short-read sequences
+was not feasible due to the implementation of \wcd.  In theory that
+tool should be able to handle short-read sequences with the
+appropriate parameter settings.  However, in the distributed code those
+parameters seemed to be hard-coded, thus rendering testing impossible.
+\peace was run on both simulated and real NGS fragment sets.  For the
+former we found \peace to do well, achieving a sensitivity of 0.75
+when simulating a sequencing error rate of 2\% (with comparable {\it
+  Jaccard Index} value, which reflects both sensitivity {\it and}
+specificity \cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was
+approxemently 17.6 minutes to process the 7.9 Mb of data when run in
+sequential mode on a 3.2 GHz Intel Xeon CPU.
 
@@
 Here we have presented \peace, a stand alone tool for the
-high-throughput clustering of transcript fragments, managed through a
-user friendly GUI that enables both local and remote installation and
-execution in sequential or parallel mode.  Based on a novel algorithm
-for the clustering of the fragments by gene association, \peace\/
-shows significant improvement in sensitivity, without sacrificing runtime,
-over the competing {\sc WCD} tool \cite{Hazelhurst08a}, and shows an
-order of magnitude in improvement over the clustering preformed in the
-course of assembly by the {\sc Cap3} tool \cite{Huang99}.
+high-throughput clustering of transcript fragments capable of dealing
+with sequences as short as 75 bases.  \peace\/, available at
+\href{http://www.peace-tools.org}{www.peace-tools.org}, is open-source
+and managed through a user friendly GUI that enables both local and
+remote installation and execution in sequential or parallel mode.
+Based on a novel algorithm for the clustering of the fragments by gene
+association, \peace\/ shows significant improvement in sensitivity,
+without sacrificing runtime, over the competing \wcd\/ tool
+\cite{Hazelhurst08a}, and shows an order of magnitude in improvement
+over the clustering preformed in the course of assembly by the
+\capthree\/ tool \cite{Huang99}.
 
@@
 certain inherent limitations. For example, \peace\/ cannot handle
-duplicate genes; like {\sc WCD}, it is unable to separate clusters
+duplicate genes; like \wcd\/, it is unable to separate clusters
 corresponding to genes with a greater than 88\% similarity.
@@
 failure to remove sequencing adapters or post-transcriptional
-poly(A)/(T) tails), and low-complexity repeats can cause similar
-effects in these clustering tools.  The problems can be handled
-through the application of the assembler, and the ability to apply any
-assembler to small cluster (as opposed to the data set as a whole)
-results in a significant reduction in overall assembly time.
+poly(A)/(T) tails), and the presence of low-complexity repeats can
+cause similar effects in these clustering tools.  The problems can be
+handled through the application of the assembler, and the ability to
+apply any assembler to small cluster (as opposed to the data set as a
+whole) results in a significant reduction in overall assembly time.
 

### 2010-02-14 peace/papers/NAR_web_server_2009/paper.tex
Corrections to manuscript

@@
     ab initio} clustering of transcript fragment sequences produced by
-  Next Generation or Sanger Sequencing technologies.  Installed and
-  run through an easily downloaded GUI, \peace\/ can process large
+  Next Generation or Sanger Sequencing technologies, available from
+  \href{http://www.peace-tools.org}{www.peace-tools.org} .  Installed
+  and run through an easily downloaded GUI, \peace\/ can process large
   data sets and group the fragments by gene association, achieving a
   greater sensitivity (without sacrificing runtime) than the competing
-  clustering tools.  Through the GUI the user can collect cluster
-  statistics and examine specific clusters for more comprehensive
-  study.  \peace\/, used as a ``pre-assembly'' tool, enables any
-  assembly tool to work with individual small clusters of closely
-  related sequences, resulting in significant increases in sensitive
-  and robustness to sequencing error, as well as a reduction in
-  overall runtime.
+  clustering tools and applicable to bases as small as 75 bases.
+  Through the GUI the user can collect cluster statistics and examine
+  specific clusters for more comprehensive study.  \peace\/, used as a
+  ``pre-assembly'' tool, enables any assembly tool to work with
+  individual small clusters of closely related sequences, resulting in
+  significant increases in sensitive and robustness to sequencing
+  error (e.g. with a 52\% improvement in sensitivity over the cluster
+  tool, \wcd\/, and a 184\% improvement in sensitivity over a leading
+  assembly tool, \capthree, given a a 5\% rate of sequencing errors),
+  as well as a reduction in runtime (with a 40\% speedup, as compared
+  to \wcd, when applied to a 361 Mb set of Arabidopsis transcriptome
+  fragments).
 \end{abstract}
@@
 
-Until recently, access to a transcriptome sequence was primarily
+In the past, access to a transcriptome sequence was primarily
 through the use of Expressed Sequence Tags (ESTs), single-pass cDNA
@@
 
+\enlargethispage{-65.1pt}
+
 Given a set of transcript fragments sampled from across the genome, a
-necessary first step of their analysis is that of clustering: separating
+necessary first step of the set's analysis is that of clustering: separating
 the fragments according to the transcript from which they were
@@
 allow the application of the assembly tool to individual clusters --
-saving significant amounts of time due to the reduced input size
-\cite{Hazelhurst08a}.
+saving significant amounts of time \cite{Hazelhurst08a}.
 
-However clustering is a computationally challenging problem.  Even
+However, clustering is a computationally challenging problem.  Even
 with the smaller number of ESTs produced using Sanger Sequencing, the
-runtime and memory requirements to cluster based on pair-wise sequence
-alignments make such an approach infeasible in practice.  The much
-larger data-set size produced by NGS technologies exacerbates this
-problem.  To deal with this, \peace\/ combines the $d^2$
-alignment-free sequence distance function \cite{Hide94} and the
-concept of a {\it minimum spanning tree} \cite{Prim57} to quickly and
-accurately find clusters of ESTs expressed from the same gene without
-reference to a sequenced genome.  Compared against competing tools for
-the same purpose
+runtime and memory requirements to cluster on the basis of pair-wise
+sequence alignments make such an approach infeasible in practice.  The
+much larger data-set size produced by NGS technologies exacerbates
+this problem.  To deal with this, \peace\/ combines our own version of
+the $d^2$ alignment-free sequence distance function \cite{Hide94} and
+the concept of a {\it minimum spanning tree} \cite{Prim57} to quickly
+and accurately find clusters of ESTs expressed from the same gene
+without reference to a sequenced genome.  Compared against \wcd, the
+leading cluster tool in the literature \cite{Hazelhurst08a}, as well as
+other tools designed for the same purpose 
 \cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
@@
 sensitive and more robust to sequencing error without sacrificing
-runtime.
+runtime.  Nor are of these tools designed for ease of installation or
+use to the extent of \peace.
 
@@
 similarity to ensure a small distance even in the presence of a
-moderate number of base errors.
+moderate number of base errors.  In practice we employ our own
+variation of $d^2$, the {\it two-pass $d^2$ algorithm}, which
+heuristically searches for a neighborhood of maximum similarity and
+then finds the $d^2$ score based on that neighborhood (see
+Supplementary Materials for details).
 
-Fragment input is modeled as a weighted, undirected graph: the fragments
-are represented as nodes, with $d^2$ sequence distances assigned to
-the connecting edges as weights.  Conceptually, we want to remove each
-edge exceeding a threshold score from the complete graph and define
-our partitions by the remaining connected components of the graph.  An
-edge with a large weight connects fragments which are likely unrelated;
-once such edges are removed the components define a series of
-overlaps.  Those fragments that can still be connected by some path
+Fragment input is modeled as a weighted, undirected graph: the
+fragments are represented as nodes, with $d^2$ sequence distances
+assigned to the connecting edges as weights.  Conceptually, we want to
+remove each edge exceeding a threshold score from the complete graph
+and define our partitions by the remaining connected components.  An
+edge with a large weight connects fragments which are likely
+unrelated; once such edges are removed the components define a series
+of overlaps.  Those fragments that can still be connected by some path
 correspond to the same gene.  However, such an approach requires the
-calculation of all edge weights.  Such a task is infeasible both in terms of
-runtime and memory usage for the data set sizes we expect to process.
+calculation of all edge weights.  Given the data set sizes, that a
+task is infeasible both in terms of runtime and memory usage for the
+data set sizes we expect to process.
 
@@
 Sanger Sequencing technologies, comparing results against those
-produced by the {\sc wcd} clustering tool \cite{Hazelhurst08a} and the
-{\sc Cap3} assembly tool \cite{Huang99} (which calculates a cluster as
-an implicit, intermediate step).  For our simulation tests we used the
-{\bf ESTSim} tool \cite{Hazelhurst03} to generate simulated transcript
-of varying length fragments under different models of error
-(Supplementary Materials, Section~C.1), generating the fragments from
-the list of 100 zebra fish genes used in the {\sc wcd} testing
-\cite{Hazelhurst08a}.  Tool parameters were taken to match, as closely
-as possible, those used in the {\sc wcd} study (see supplementary
-materials).  Our primary methods of quality assessment were {\it
-  sensitivity} (the fraction of fragment pairs from the same gene that
-were correctly clustered together) and {\it Type 1 error} (the
-fraction of genes that were divided between clusters)
-\cite{Wang04,Hazelhurst08a}; other measurements are also discussed in
-the supplementary materials.  In Figure~\ref{SeT1} (left) we see a
-significant improvement of \peace\/ (blue) over {\sc wcd} (green) and
-{\sc Cap3} (black) in its sensitivity to sequencing errors, while in
+produced by the \wcd\/ clustering tool \cite{Hazelhurst08a} and the
+\capthree\/ assembly tool \cite{Huang99} (the latter of which
+implicitly calculates a cluster in the process of assembly).  For our
+simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03} to
+generate simulated transcript of varying length fragments under
+different models of error (Supplementary Materials, Section~C.1),
+generating the fragments from the list of 100 zebra fish genes used in
+the \wcd\/ testing \cite{Hazelhurst08a}.  Tool parameters were taken
+to match, as closely as possible, those used in the \wcd\/ study (see
+supplementary materials).  Our primary methods of quality assessment
+were {\it sensitivity} (the fraction of fragment pairs from the same
+gene that were correctly clustered together) and {\it Type 1 error}
+(the fraction of genes that were divided between clusters)
+\cite{Wang04,Hazelhurst08a}.  Specificity was not an issue in
+simulated tests, as both \peace\/ and \wcd\/ were completely successful
+in separating unrelated ESTs (save when dealing with recently
+duplicated genes).  Other measurements are also discussed in the
+supplementary materials.  In Figure~\ref{SeT1} (left), reflecting
+tests done on simulated Sanger EST set, we see a
+significant improvement of \peace\/ (blue) over \wcd\/ (green) and
+\capthree\/ (black) in its sensitivity to sequencing errors, while in
 Figure~\ref{SeT1} (right) we see a comparable improvement in Type 1
 error.  In Figure~\ref{singletons} we look at the number of singleton
-clusters (fragments not joined to any cluster), which will not occur
-in our simulated sets; we again see significant improvement in \peace.
+clusters (fragments not joined to any cluster), which should not occur
+in our simulated sets.  We again see significant improvement in \peace.
 The sequential runtime of \peace\/ is slightly improved over that of
-{\sc wcd} (see Supplementary Materials, Figure~S4).
+\wcd\/ (see Supplementary Materials, Figure~S4).
 
@@
   \caption{Comparisons of Sensitivity and Type 1 error, based on the
-    average over 30 simulated fragment sets derived from 100 zebra
-    fish genes (see Supplementary Materials, Section C, for more
-    details).  Blue/Solid = \peace, Green/Dash = {\sc wcd},
-    Black/Dot-Dash = {\sc Cap3}; vertical tics = 95\% confidence
-    intervals on estimates.  Intervals are not presented for Type 1
-    error due to the extremely small variance.}\label{SeT1}
+    average over 30 simulated Sanger Sequence ESTs sets derived from 100 zebra
+    fish genes  (see
+    Supplementary Materials, Section C, for more details).  Blue/Solid
+    = \peace, Green/Dash = \wcd, Black/Dot-Dash = \capthree; vertical
+    tics = 95\% confidence intervals on estimates.  Intervals are not
+    presented for Type 1 error due to the extremely small
+    variance.}\label{SeT1}
 \end{figure}
@@
     when run on the simulated sequences; correct answer in all cases is
-    zero.  Blue/Solid = \peace, Green/Dash = {\sc wcd},
-    Black/Dot-Dash = {\sc Cap3}.}\label{singletons}
+    zero.  Blue/Solid = \peace, Green/Dash = \wcd,
+    Black/Dot-Dash = \capthree.}\label{singletons}
 \end{figure}
@@
 fragment set to the genome, taking this as our reference clustering.
-We see slight improvements in \peace\/ over {\sc wcd} in both
-sensitivity and Type 1 error (both significantly outperforming {\sc
-  Cap3}).  Using the mouse data set used in Hazelhurst {\it et al.}
+We see slight improvements in \peace\/ over \wcd\/ in both sensitivity
+and Type 1 error (both significantly outperforming \capthree).  Using
+the mouse data set used in Hazelhurst {\it et al.}
 \cite{Hazelhurst08a} \peace\/ again shows slight improvements in
-sensitivity and Type 1 error rates, with an 18\% speedup for {\sc
-  Peace} (see Supplementary Materials, Section~D, for more details).
+sensitivity and Type 1 error rates, with an 18\% speedup for \peace\/ 
+(see Supplementary Materials, Section~D, for more details).
+
+A comparative test of \peace\/ against \wcd\/ on NGS short-read sequences
+was not feasible due to the implementation of \wcd.  In theory that
+tool should be able to handle short-read sequences with the
+appropriate parameter settings.  However, in the distributed versions those
+parameters seemed to be hard-coded, thus rendering testing impossible.
+\peace\/ was run on both simulated and real NGS fragment sets.  For the
+former we found \peace\/ to do well, achieving a sensitivity of 0.75
+when simulating a sequencing error rate of 2\% (with comparable {\it
+  Jaccard Index} value, which reflects both sensitivity {\it and}
+specificity \cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was
+approximately 17.6 minutes to process the 7.9 Mb of data when run in
+sequential mode on a 3.2 GHz Intel Xeon CPU.
 
@@
 certain inherent limitations. For example, \peace\/ cannot handle
-duplicate genes; like {\sc WCD}, it is unable to separate clusters
+duplicate genes; like \wcd\/, it is unable to separate clusters
 corresponding to genes with a greater than 88\% similarity.
-Similarly, natural biological effects such as the trans-splicing of
-transcripts, or effects from poorly cleaned transcript data (e.g. the
+Similarly, other natural biological effects (e.g. the trans-splicing of
+transcripts), effects from poorly cleaned transcript data (e.g. the
 failure to remove sequencing adapters or post-transcriptional
-poly(A)/(T) tails), and low-complexity repeats can cause similar
-effects in these clustering tools.  The problems can be handled
-through the application of the assembler, and the ability to apply any
-assembler to small cluster (as opposed to the data set as a whole)
-results in a significant reduction in overall assembly time.
+poly(A)/(T) tails), and the presence of low-complexity repeats can
+cause similar effects in these clustering tools.  The problems can be
+handled through the application of the assembler, and the ability to
+apply any assembler to small cluster (as opposed to the data set as a
+whole) results in a significant reduction in overall assembly time.
 
@@
 % LocalWords:  ssembly xpression GenBank dbEST pyrosequencing ESTsim sim mRNAs
-% LocalWords:  NGS polyA
+% LocalWords:  NGS polyA set's GHz Xeon Bikun

### 2010-02-14 papers/NAR_web_server_2009/paper.tex
Modificatons to abstract.

@@
     ab initio} clustering of transcript fragment sequences produced by
-  Next Generation or Sanger Sequencing technologies.  Installed and
-  run through an easily downloaded GUI, \peace\/ can process large
-  data sets and group the fragments by gene association, achieving a
-  greater sensitivity (without sacrificing runtime) than the competing
-  clustering tools and applicable to bases as small as 75 bases.
-  Through the GUI the user can collect cluster statistics and examine
-  specific clusters for more comprehensive study.  \peace\/, used as a
-  ``pre-assembly'' tool, enables any assembly tool to work with
-  individual small clusters of closely related sequences, resulting in
-  significant increases in sensitive and robustness to sequencing
-  error (e.g. with a 52\% improvement in sensitivity over the cluster
-  tool, \wcd\/, and a 184\% improvement in sensitivity over a leading
-  assembly tool, \capthree, given a a 5\% rate of sequencing errors), as
-  well as a reduction in runtime (with a 40\% speedup, as compared to
-  \wcd, when applied to a 361 Mb set of Arabidopsis transcriptome fragments).
+  Next Generation or Sanger Sequencing technologies, freely available
+  from \href{http://www.peace-tools.org}{www.peace-tools.org} .
+  Installed and run through an easily downloaded GUI, \peace\/ can
+  process large data sets of transcripts fragments of length 50 bases
+  or greater, grouping the fragments by gene association and achieving
+  a greater sensitivity (without sacrificing runtime) than the
+  competing clustering tools.  Once clusered, the user can use the GUI
+  to to collect statistics and single out specific clusters for more
+  comprehensive study.  \peace\/, when used as a ``pre-assembly''
+  tool, enables any assembly tool to work with individual small
+  clusters of closely related sequences.  The result is a significant
+  increase in sensitivity, as well as improved robustness to
+  sequencing error as well as a reduction in runtime.  For a 5\%
+  sequencing error rate, we see a 52\% improvement in sensitivity over
+  the leading cluster tool, \wcd\/, a 184\% improvement in sensitivity
+  over the assembly tool \capthree.  When applied to a 361 Mb set of
+  Arabidopsis transcriptome fragments, we see a 40\% speedup over
+  \wcd.
 \end{abstract}
@@
 
-, access to a transcriptome sequence was primarily
+In the past, access to a transcriptome sequence was primarily
 through the use of Expressed Sequence Tags (ESTs), single-pass cDNA
@@
 Given a set of transcript fragments sampled from across the genome, a
-necessary first step of their analysis is that of clustering: separating
+necessary first step of the set's analysis is that of clustering: separating
 the fragments according to the transcript from which they were
@@
 
-However clustering is a computationally challenging problem.  Even
+However, clustering is a computationally challenging problem.  Even
 with the smaller number of ESTs produced using Sanger Sequencing, the
-runtime and memory requirements to cluster based on pair-wise sequence
-alignments make such an approach infeasible in practice.  The much
-larger data-set size produced by NGS technologies exacerbates this
-problem.  To deal with this, \peace\/ combines our own version of the $d^2$
-alignment-free sequence distance function \cite{Hide94} and the
-concept of a {\it minimum spanning tree} \cite{Prim57} to quickly and
-accurately find clusters of ESTs expressed from the same gene without
-reference to a sequenced genome.  Compared against competing tools for
-the same purpose
+runtime and memory requirements to cluster on the basis of pair-wise
+sequence alignments make such an approach infeasible in practice.  The
+much larger data-set size produced by NGS technologies exacerbates
+this problem.  To deal with this, \peace\/ combines our own version of
+the $d^2$ alignment-free sequence distance function \cite{Hide94} and
+the concept of a {\it minimum spanning tree} \cite{Prim57} to quickly
+and accurately find clusters of ESTs expressed from the same gene
+without reference to a sequenced genome.  Compared against \wcd, the
+leading cluster tool in the literature \cite{Hazelhurst08a}, as well as
+other tools designed for the same purpose 
 \cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
@@
 heuristically searches for a neighborhood of maximum similarity and
-then finding the $d^2$ score within that neighborhood (see
+then finds the $d^2$ score based on that neighborhood (see
 Supplementary Materials for details).
@@
 \cite{Wang04,Hazelhurst08a}.  Specificity was not an issue in
-simulated tests, as both \peace and \wcd were completely successful
+simulated tests, as both \peace\/ and \wcd\/ were completely successful
 in separating unrelated ESTs (save when dealing with recently
@@
 \cite{Hazelhurst08a} \peace\/ again shows slight improvements in
-sensitivity and Type 1 error rates, with an 18\% speedup for \peace
+sensitivity and Type 1 error rates, with an 18\% speedup for \peace\/ 
 (see Supplementary Materials, Section~D, for more details).
 
-A comparative test of \peace against \wcd on NGS short-read sequences
+A comparative test of \peace\/ against \wcd\/ on NGS short-read sequences
 was not feasible due to the implementation of \wcd.  In theory that
 tool should be able to handle short-read sequences with the
-appropriate parameter settings.  However, in the distributed code those
+appropriate parameter settings.  However, in the distributed versions those
 parameters seemed to be hard-coded, thus rendering testing impossible.
-\peace was run on both simulated and real NGS fragment sets.  For the
-former we found \peace to do well, achieving a sensitivity of 0.75
+\peace\/ was run on both simulated and real NGS fragment sets.  For the
+former we found \peace\/ to do well, achieving a sensitivity of 0.75
 when simulating a sequencing error rate of 2\% (with comparable {\it
@@
 specificity \cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was
-approxemently 17.6 minutes to process the 7.9 Mb of data when run in
+approximately 17.6 minutes to process the 7.9 Mb of data when run in
 sequential mode on a 3.2 GHz Intel Xeon CPU.
@@
 high-throughput clustering of transcript fragments capable of dealing
-with sequences as short as 75 bases.  \peace\/, available at
+with sequences as short as 50 bases.  \peace\/, available at
 \href{http://www.peace-tools.org}{www.peace-tools.org}, is open-source
@@
 corresponding to genes with a greater than 88\% similarity.
-Similarly, natural biological effects such as the trans-splicing of
-transcripts, or effects from poorly cleaned transcript data (e.g. the
+Similarly, other natural biological effects (e.g. the trans-splicing of
+transcripts), effects from poorly cleaned transcript data (e.g. the
 failure to remove sequencing adapters or post-transcriptional

### 2010-02-14 papers/NAR_web_server_2009/paper.tex
Cleaning up of typos and grammatical errors in the paper.

@@
   Installed and run through an easily downloaded GUI, \peace\/ can
-  process large data sets of transcripts fragments of length 50 bases
+  process large data sets of transcript fragments of length 50 bases
   or greater, grouping the fragments by gene association and achieving
   a greater sensitivity (without sacrificing runtime) than the
-  competing clustering tools.  Once clusered, the user can use the GUI
-  to to collect statistics and single out specific clusters for more
+  competing clustering tools.  Once clustered, the user can use the GUI
+  to collect statistics and single out specific clusters for more
   comprehensive study.  \peace\/, when used as a ``pre-assembly''
@@
   clusters of closely related sequences.  The result is a significant
-  increase in sensitivity, as well as improved robustness to
-  sequencing error as well as a reduction in runtime.  For a 5\%
-  sequencing error rate, we see a 52\% improvement in sensitivity over
-  the leading cluster tool, \wcd\/, a 184\% improvement in sensitivity
-  over the assembly tool \capthree.  When applied to a 361 Mb set of
-  Arabidopsis transcriptome fragments, we see a 40\% speedup over
-  \wcd.
+  increase in sensitivity, improved robustness to sequencing error,
+  and a reduction in runtime.  For a 5\% sequencing error rate, we see
+  a 52\% improvement in sensitivity over the leading cluster tool,
+  \wcd\/, and 184\% improvement in sensitivity over the assembly tool
+  \capthree.  When applied to a 361 Mb set of Arabidopsis transcriptome
+  fragments, we see a 40\% speedup over \wcd.
 \end{abstract}
@@
 without reference to a sequenced genome.  Compared against \wcd, the
-leading cluster tool in the literature \cite{Hazelhurst08a}, as well as
+leading clustering tool in the literature \cite{Hazelhurst08a}, as well as
 other tools designed for the same purpose 
@@
 sensitive and more robust to sequencing error without sacrificing
-runtime.  Nor are of these tools designed for ease of installation or
-use to the extent of \peace.
+runtime.  Nor are any of these tools designed for the ease of installation
+and use that \peace provides.
 
-In short, \peace\/ is computational tool for the {\it ab initio}
+In short, \peace\/ is a computational tool for the {\it ab initio}
 clustering of transcript fragments by gene association, applicable to
@@
 correspond to the same gene.  However, such an approach requires the
-calculation of all edge weights.  Given the data set sizes, that a
-task is infeasible both in terms of runtime and memory usage for the
-data set sizes we expect to process.
+calculation of all edge weights.  That task is infeasible both in terms
+of runtime and memory usage for the data set sizes we expect to process.
 
@@
 
-\peace\/ has been test on both simulated and real data from NGS and
+\peace\/ has been tested on both simulated and real data from NGS and
 Sanger Sequencing technologies, comparing results against those
@@
 \capthree\/ assembly tool \cite{Huang99} (the latter of which
-implicitly calculates a cluster in the process of assembly).  For our
+implicitly calculates a clustering in the process of assembly).  For our
 simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03} to
-generate simulated transcript of varying length fragments under
+generate simulated transcript fragments of varying length under
 different models of error (Supplementary Materials, Section~C.1),
@@
 supplementary materials.  In Figure~\ref{SeT1} (left), reflecting
-tests done on simulated Sanger EST set, we see a
+tests done on simulated Sanger EST sets, we see a
 significant improvement of \peace\/ (blue) over \wcd\/ (green) and
@@
 A comparative test of \peace\/ against \wcd\/ on NGS short-read sequences
-was not feasible due to the implementation of \wcd.  In theory that
+was not feasible due to the implementation of \wcd.  In theory, that
 tool should be able to handle short-read sequences with the
@@
 \cite{Hazelhurst08a}, and shows an order of magnitude in improvement
-over the clustering preformed in the course of assembly by the
+over the clustering performed in the course of assembly by the
 \capthree\/ tool \cite{Huang99}.

### 2010-02-14 peace/papers/NAR_web_server_2009/paper.tex
Final drafts of cover leter and paper

@@
     ab initio} clustering of transcript fragment sequences produced by
-  Next Generation or Sanger Sequencing technologies, available from
-  \href{http://www.peace-tools.org}{www.peace-tools.org} .  Installed
-  and run through an easily downloaded GUI, \peace\/ can process large
-  data sets and group the fragments by gene association, achieving a
-  greater sensitivity (without sacrificing runtime) than the competing
-  clustering tools and applicable to bases as small as 75 bases.
-  Through the GUI the user can collect cluster statistics and examine
-  specific clusters for more comprehensive study.  \peace\/, used as a
-  ``pre-assembly'' tool, enables any assembly tool to work with
-  individual small clusters of closely related sequences, resulting in
-  significant increases in sensitive and robustness to sequencing
-  error (e.g. with a 52\% improvement in sensitivity over the cluster
-  tool, \wcd\/, and a 184\% improvement in sensitivity over a leading
-  assembly tool, \capthree, given a a 5\% rate of sequencing errors),
-  as well as a reduction in runtime (with a 40\% speedup, as compared
-  to \wcd, when applied to a 361 Mb set of Arabidopsis transcriptome
-  fragments).
+  Next Generation or Sanger Sequencing technologies, freely available
+  from \href{http://www.peace-tools.org}{www.peace-tools.org}.
+  Installed and run through an easily downloaded GUI, \peace\/ can
+  process large data sets of transcript fragments of length 50 bases
+  or greater, grouping the fragments by gene association and achieving
+  a greater sensitivity (without sacrificing runtime) than the
+  competing clustering tools.  Once clustered, the user can employ the
+  GUI to collect statistics and single out specific clusters for more
+  comprehensive study.  \peace\/, when used as a ``pre-assembly''
+  tool, enables any assembly tool to work with individual small
+  clusters of closely related sequences.  For a 5\% sequencing error
+  rate, we see a 52\% improvement in sensitivity over the leading
+  cluster tool, \wcd\/, and 184\% improvement in sensitivity over the
+  assembly tool \capthree.  When applied to a 361 Mb set of
+  Arabidopsis transcriptome fragments, we see a 40\% speedup over
+  \wcd.
 \end{abstract}
@@
 to match, as closely as possible, those used in the \wcd\/ study (see
-supplementary materials).  Our primary methods of quality assessment
+Supplementary Materials).  Our primary methods of quality assessment
 were {\it sensitivity} (the fraction of fragment pairs from the same
@@
 
-A comparative test of \peace\/ against \wcd\/ on NGS short-read sequences
-was not feasible due to the implementation of \wcd.  In theory that
-tool should be able to handle short-read sequences with the
-appropriate parameter settings.  However, in the distributed versions those
-parameters seemed to be hard-coded, thus rendering testing impossible.
-\peace\/ was run on both simulated and real NGS fragment sets.  For the
-former we found \peace\/ to do well, achieving a sensitivity of 0.75
-when simulating a sequencing error rate of 2\% (with comparable {\it
-  Jaccard Index} value, which reflects both sensitivity {\it and}
-specificity \cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was
-approximately 17.6 minutes to process the 7.9 Mb of data when run in
-sequential mode on a 3.2 GHz Intel Xeon CPU.
+We were unable to run \wcd\/ on short-read sequences, as the
+distributed implementation is unable to handle such data without an
+adjustment to the code.  Hence, while \peace\/ was run on both
+simulated and real NGS fragment sets, there was no cluster tool to
+which the results could be compared.  \peace\/ did well, achieving a
+sensitivity of 0.75 when applied to a simulated short-read set subjected to
+a sequencing error rate of 2\% (with comparable {\it Jaccard Index}
+values reflecting both sensitivity {\it and} specificity
+\cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was approximately 17.6
+minutes to process the 7.9 Mb of data when run in sequential mode on a
+3.2 GHz Intel Xeon CPU.  When applied to a 64 Mb set of 165,490 {\it
+  Chlamydomonas reinhardtii} transcript fragments sequenced by 454
+pyrosequencing technology (ranging in size from 50 bases to several
+hundred), \peace\/ completed the job in 49 minutes using 30 of the
+Xeon processors.
 

### 2010-02-14 papers/NAR_web_server_2009/paper.tex
Added a missing image that I had missed out earlier.

@@
   Next Generation or Sanger Sequencing technologies, freely available
-  from \href{http://www.peace-tools.org}{www.peace-tools.org} .
+  from \href{http://www.peace-tools.org}{www.peace-tools.org}.
   Installed and run through an easily downloaded GUI, \peace\/ can
@@
   a greater sensitivity (without sacrificing runtime) than the
-  competing clustering tools.  Once clustered, the user can use the GUI
-  to collect statistics and single out specific clusters for more
+  competing clustering tools.  Once clustered, the user can employ the
+  GUI to collect statistics and single out specific clusters for more
   comprehensive study.  \peace\/, when used as a ``pre-assembly''
   tool, enables any assembly tool to work with individual small
-  clusters of closely related sequences.  The result is a significant
-  increase in sensitivity, improved robustness to sequencing error,
-  and a reduction in runtime.  For a 5\% sequencing error rate, we see
-  a 52\% improvement in sensitivity over the leading cluster tool,
-  \wcd\/, and 184\% improvement in sensitivity over the assembly tool
-  \capthree.  When applied to a 361 Mb set of Arabidopsis transcriptome
-  fragments, we see a 40\% speedup over \wcd.
+  clusters of closely related sequences.  For a 5\% sequencing error
+  rate, we see a 52\% improvement in sensitivity over the leading
+  cluster tool, \wcd\/, and 184\% improvement in sensitivity over the
+  assembly tool \capthree.  When applied to a 361 Mb set of
+  Arabidopsis transcriptome fragments, we see a 40\% speedup over
+  \wcd.
 \end{abstract}
@@
   \begin{minipage}{2in}
-    \vspace{0.47in}
-    \includegraphics[width=2in]{screen.d/add_server_big.png}
+    \vspace{0.66in}
+    \includegraphics[width=2in]{screen.d/processor_assignment_big.png}
     \centerline{\small{(c)}}
@@
 
-A comparative test of \peace\/ against \wcd\/ on NGS short-read sequences
-was not feasible due to the implementation of \wcd.  In theory, that
-tool should be able to handle short-read sequences with the
-appropriate parameter settings.  However, in the distributed versions those
-parameters seemed to be hard-coded, thus rendering testing impossible.
-\peace\/ was run on both simulated and real NGS fragment sets.  For the
-former we found \peace\/ to do well, achieving a sensitivity of 0.75
-when simulating a sequencing error rate of 2\% (with comparable {\it
-  Jaccard Index} value, which reflects both sensitivity {\it and}
-specificity \cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was
-approximately 17.6 minutes to process the 7.9 Mb of data when run in
-sequential mode on a 3.2 GHz Intel Xeon CPU.
+We were unable to run \wcd\/ on short-read sequences, as the
+distributed implementation is unable to handle such data without an
+adjustment to the code.  Hence, while \peace\/ was run on both
+simulated and real NGS fragment sets, there was no cluster tool to
+which the results could be compared.  \peace\/ did well, achieving a
+sensitivity of 0.75 when applied to a simulated short-read set subjected to
+a sequencing error rate of 2\% (with comparable {\it Jaccard Index}
+values reflecting both sensitivity {\it and} specificity
+\cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was approximately 17.6
+minutes to process the 7.9 Mb of data when run in sequential mode on a
+3.2 GHz Intel Xeon CPU.  When applied to a 64 Mb set of 165,490 {\it
+  Chlamydomonas reinhardtii} transcript fragments sequenced by 454
+pyrosequencing technology (ranging in size from 50 bases to several
+hundred), \peace\/ completed the job in 49 minutes using 30 of the
+Xeon processors.
 

### 2010-04-16 peace/papers/NAR_web_server_2009/paper.tex
Small modifications to NAR paper; added the (partial draft) of the response to the reviewers.  Response is still very much a work in progress.

@@
 \usepackage{wrapfig}
-
 \bibliographystyle{unsrt}
@@
 \jissue{}
+
+\usepackage[usenames,dvipsnames]{color}
+\definecolor{red}{rgb}{1,0,0}
+\newcommand{\mc}[1]{\textcolor{red}{#1}}
 
@@
 \begin{figure}
-  \centerline{\includegraphics[width=3in]{screen.d/workflow.pdf}}
+  \centerline{\includegraphics[width=3in]{workflow.eps}}
   \caption{Overview of the procedure for clustering and analysis using
@@
   \begin{minipage}{3in}
-    \includegraphics[width=3in]{screen.d/install_page_big.png}
+    \includegraphics[width=3in]{install_page_big.eps}
     \centerline{\small{(a)}}
@@
     \vspace{0.47in}
-    \includegraphics[width=2in]{screen.d/add_server_big.png}
+    \includegraphics[width=2in]{add_server_big.eps}
     \centerline{\small{(b)}}
@@
     \vspace{0.47in}
-    \includegraphics[width=2in]{screen.d/add_server_big.png}
+    \includegraphics[width=2in]{processor_assignment_big.eps}
     \centerline{\small{(c)}}
@@
   \begin{minipage}{3in}
-    \includegraphics[width=3in]{screen.d/cluster_list_big.png}
+    \includegraphics[width=3in]{cluster_list_big.eps}
     \centerline{\small{(d)}}
@@
   \begin{minipage}{3in}
-    \includegraphics[width=3in]{screen.d/classifier_big.png}
+    \includegraphics[width=3in]{classifier_big.eps}
     \centerline{\small{(e)}}
@@
 Sanger Sequencing technologies, comparing results against those
-produced by the \wcd\/ clustering tool \cite{Hazelhurst08a} and the
+produced by the \wcd\/ clustering tool \wc{(version 0.5.1)}\cite{Hazelhurst08a} and the
 \capthree\/ assembly tool \cite{Huang99} (the latter of which
@@
 \begin{figure}
-  \centerline{\includegraphics[width=3.35in]{pics.d/SeT1.pdf}}
+  \centerline{\includegraphics[width=3.35in]{SeT1.eps}}
   \caption{Comparisons of Sensitivity and Type 1 error, based on the
@@
 \begin{figure}
-  \centerline{\includegraphics[width=3in]{pics.d/singletons.pdf}}
+  \centerline{\includegraphics[width=3in]{singletons.eps}}
   \caption{Average number of fragments flagged as singletons by each tool
@@
 and Type 1 error (both significantly outperforming \capthree).  Using
-the mouse data set used in Hazelhurst {\it et al.}
+the \mc{Arabidopsis} data set used in Hazelhurst {\it et al.}
 \cite{Hazelhurst08a} \peace\/ again shows slight improvements in
@@
 which the results could be compared.  \peace\/ did well, achieving a
-sensitivity of 0.75 when applied to a simulated short-read set subjected to
-a sequencing error rate of 2\% (with comparable {\it Jaccard Index}
-values reflecting both sensitivity {\it and} specificity
-\cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was approximately 17.6
-minutes to process the 7.9 Mb of data when run in sequential mode on a
-3.2 GHz Intel Xeon CPU.  When applied to a 64 Mb set of 165,490 {\it
-  Chlamydomonas reinhardtii} transcript fragments sequenced by 454
-pyrosequencing technology (ranging in size from 50 bases to several
-hundred), \peace\/ completed the job in 49 minutes using 30 of the
-Xeon processors.
+sensitivity of 0.75 when applied to a simulated short-read set
+subjected to a sequencing error rate of 2\% (with comparable {\it
+  Jaccard Index} values reflecting both sensitivity {\it and}
+specificity \cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was
+approximately 17.6 minutes to process the 7.9 Mb of data when run in
+sequential mode on a \mc{3.2 GHz Intel Xeon EM64T CPU with a 2 MB
+  cache and 800 MHz font side bus}.  When applied to a 64 Mb set of
+165,490 {\it
+ Chlamydomonas reinhardtii} transcript fragments
+sequenced by 454 pyrosequencing technology (ranging in size from 50
+bases to several hundred), \peace\/ completed the job in 49 minutes
+using 30 of the Xeon processors.
 

### 2010-04-17 src/main.cpp
Updated just the version number and date to reflect the upcoming release of the software for NAR resubmission.

@@ -80,7 +80,8 @@ static void showUsage(arg_parser& ap,
                       ESTAnalyzer *analyzer, ClusterMaker *cMaker,
                       HeuristicChain *hChain,
                       FilterChain *fChain) {
-    std::cout << "Usage: PEACE [options]\n"
+    std::cout << "PEACE Version 0.95 (Released April 19 2010)\n"
+              << "Usage: PEACE [options]\n"
               << "where options are:\n";
     std::cout << ap;
     std::cout << "Names of EST analyzers available are:\n";

### 2010-04-18 peace/papers/NAR_web_server_2009/paper.tex
Upated to paper

@@
 \begin{figure}
-  \centerline{\includegraphics[width=3in]{workflow.eps}}
+  \centerline{\includegraphics[width=3in]{pics.d/workflow.eps}}
   \caption{Overview of the procedure for clustering and analysis using
@@
 Sanger Sequencing technologies, comparing results against those
-produced by the \wcd\/ clustering tool \wc{(version 0.5.1)}\cite{Hazelhurst08a} and the
-\capthree\/ assembly tool \cite{Huang99} (the latter of which
-implicitly calculates a clustering in the process of assembly).  For our
-simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03} to
-generate simulated transcript fragments of varying length under
-different models of error (Supplementary Materials, Section~C.1),
-generating the fragments from the list of 100 zebra fish genes used in
-the \wcd\/ testing \cite{Hazelhurst08a}.  Tool parameters were taken
-to match, as closely as possible, those used in the \wcd\/ study (see
-Supplementary Materials).  Our primary methods of quality assessment
-were {\it sensitivity} (the fraction of fragment pairs from the same
-gene that were correctly clustered together) and {\it Type 1 error}
-(the fraction of genes that were divided between clusters)
-\cite{Wang04,Hazelhurst08a}.  Specificity was not an issue in
-simulated tests, as both \peace\/ and \wcd\/ were completely successful
-in separating unrelated ESTs (save when dealing with recently
-duplicated genes).  Other measurements are also discussed in the
-supplementary materials.  In Figure~\ref{SeT1} (left), reflecting
-tests done on simulated Sanger EST sets, we see a
-significant improvement of \peace\/ (blue) over \wcd\/ (green) and
-\capthree\/ (black) in its sensitivity to sequencing errors, while in
-Figure~\ref{SeT1} (right) we see a comparable improvement in Type 1
-error.  In Figure~\ref{singletons} we look at the number of singleton
-clusters (fragments not joined to any cluster), which should not occur
-in our simulated sets.  We again see significant improvement in \peace.
-The sequential runtime of \peace\/ is slightly improved over that of
-\wcd\/ (see Supplementary Materials, Figure~S4).
+produced by the \wcd\/ clustering tool \mc{(version
+  0.5.1)}\cite{Hazelhurst08a} and the \capthree\/ assembly tool
+\cite{Huang99} (the latter of which implicitly calculates a clustering
+in the process of assembly).  For our simulation tests we used the
+{\bf ESTSim} tool \cite{Hazelhurst03} to generate simulated transcript
+fragments of varying length under different models of error
+(Supplementary Materials, Section~C.1), generating the fragments from
+the list of 100 zebra fish genes used in the \wcd\/ testing
+\cite{Hazelhurst08a}.  Tool parameters were taken to match, as closely
+as possible, those used in the \wcd\/ study (see Supplementary
+Materials).  \mc{The most important method of quality assessment is
+  {\it sensitivity} (the fraction of fragment pairs from the same gene
+  that were correctly clustered together).  But we also look at the
+  {\it Jaccard Index} (which balances sensitivity with the number of
+  false positives by looking at the ratio of true positives to the sum
+  of true positives, false positives, and true negitives), {\it Type 1
+    error} (the fraction of genes that were divided between clusters),
+  and {\it Type 2 error} (the fraction of clusters containing two or
+  more genes).  Specificity was not an issue in simulated tests, as
+  both \peace\/ and \wcd\/ were completely successful in separating
+  unrelated ESTs (save when dealing with recently duplicated genes).
+  Other measurements are also discussed in the supplementary
+  materials.  In Figure~\ref{SeJiT1T2} (left) we see these four tests,...}
+
+% , reflecting tests done on
+%   simulated Sanger EST sets, we see a significant improvement of
+%   \peace\/ (blue) over \wcd\/ (green) and \capthree\/ (black) in its
+%   sensitivity to sequencing errors, while in Figure~\ref{SeT1} (right)
+%   we see a comparable improvement in Type 1 error.  In
+%   Figure~\ref{singletons} we look at the number of singleton clusters
+%   (fragments not joined to any cluster), which should not occur in our
+%   simulated sets.  We again see significant improvement in \peace.
+%   The sequential runtime of \peace\/ is slightly improved over that of
+%   \wcd\/ (see Supplementary Materials, Figure~S4).
 
 \begin{figure}
-  \centerline{\includegraphics[width=3.35in]{SeT1.eps}}
+  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2.40.eps}}
   \caption{Comparisons of Sensitivity and Type 1 error, based on the
@@
     presented for Type 1 error due to the extremely small
-    variance.}\label{SeT1}
-\end{figure}
-
-\begin{figure}
-  \centerline{\includegraphics[width=3in]{singletons.eps}}
-  \caption{Average number of fragments flagged as singletons by each tool
-    when run on the simulated sequences; correct answer in all cases is
-    zero.  Blue/Solid = \peace, Green/Dash = \wcd,
-    Black/Dot-Dash = \capthree.}\label{singletons}
+    variance.}\label{SeJiT1T2}
 \end{figure}

### 2010-04-18 papers/NAR_web_server_2009/paper.tex
Modifications to introduction of response to reviewers

@@
 
+\usepackage[usenames,dvipsnames]{color}
+\definecolor{red}{rgb}{1,0,0}
+\newcommand{\mc}[1]{\textcolor{red}{#1}}
+
 % \articlesubtype{This is the article type (optional)}
@@
 \newcommand{\capthree} {{\small Cap3}}
+\newcommand{\easycluster} {{\small EasyCluster}}
 
@@
 \begin{figure}
-  \centerline{\includegraphics[width=3in]{screen.d/workflow.pdf}}
+  \centerline{\includegraphics[width=3in]{pics.d/workflow.eps}}
   \caption{Overview of the procedure for clustering and analysis using
@@
   \begin{minipage}{2in}
-    \vspace{0.66in}
-    \includegraphics[width=2in]{screen.d/processor_assignment_big.png}
+    \vspace{0.47in}
+    \includegraphics[width=2in]{processor_assignment_big.eps}
     \centerline{\small{(c)}}
@@
 Sanger Sequencing technologies, comparing results against those
-produced by the \wcd\/ clustering tool \cite{Hazelhurst08a} and the
-\capthree\/ assembly tool \cite{Huang99} (the latter of which
-implicitly calculates a clustering in the process of assembly).  For our
-simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03} to
-generate simulated transcript fragments of varying length under
-different models of error (Supplementary Materials, Section~C.1),
-generating the fragments from the list of 100 zebra fish genes used in
-the \wcd\/ testing \cite{Hazelhurst08a}.  Tool parameters were taken
-to match, as closely as possible, those used in the \wcd\/ study (see
-Supplementary Materials).  Our primary methods of quality assessment
-were {\it sensitivity} (the fraction of fragment pairs from the same
-gene that were correctly clustered together) and {\it Type 1 error}
-(the fraction of genes that were divided between clusters)
-\cite{Wang04,Hazelhurst08a}.  Specificity was not an issue in
-simulated tests, as both \peace\/ and \wcd\/ were completely successful
-in separating unrelated ESTs (save when dealing with recently
-duplicated genes).  Other measurements are also discussed in the
-supplementary materials.  In Figure~\ref{SeT1} (left), reflecting
-tests done on simulated Sanger EST sets, we see a
-significant improvement of \peace\/ (blue) over \wcd\/ (green) and
-\capthree\/ (black) in its sensitivity to sequencing errors, while in
-Figure~\ref{SeT1} (right) we see a comparable improvement in Type 1
-error.  In Figure~\ref{singletons} we look at the number of singleton
-clusters (fragments not joined to any cluster), which should not occur
-in our simulated sets.  We again see significant improvement in \peace.
-The sequential runtime of \peace\/ is slightly improved over that of
-\wcd\/ (see Supplementary Materials, Figure~S4).
+produced by the \wcd\/ clustering tool \mc{(version
+  0.5.1)}\cite{Hazelhurst08a} and the \capthree\/ assembly tool
+\cite{Huang99} (the latter of which implicitly calculates a clustering
+in the process of assembly).  For our simulation tests we used the
+{\bf ESTSim} tool \cite{Hazelhurst03} to generate simulated transcript
+fragments of varying length under different models of error
+(Supplementary Materials, Section~C.1), generating the fragments from
+the list of 100 zebra fish genes used in the \wcd\/ testing
+\cite{Hazelhurst08a}.  Tool parameters were taken to match, as closely
+as possible, those used in the \wcd\/ study (see Supplementary
+Materials).  \mc{The most important method of quality assessment is
+  {\it sensitivity} (the fraction of fragment pairs from the same gene
+  that were correctly clustered together).  But we also look at the
+  {\it Jaccard Index} (which balances sensitivity with the number of
+  false positives by looking at the ratio of true positives to the sum
+  of true positives, false positives, and true negitives), {\it Type 1
+    error} (the fraction of genes that were divided between clusters),
+  and {\it Type 2 error} (the fraction of clusters containing two or
+  more genes).  Specificity was not an issue in simulated tests, as
+  both \peace\/ and \wcd\/ were completely successful in separating
+  unrelated ESTs (save when dealing with recently duplicated genes).
+  Other measurements are also discussed in the supplementary
+  materials.  In Figure~\ref{SeJiT1T2} (left) we see these four tests,...}
+
+% , reflecting tests done on
+%   simulated Sanger EST sets, we see a significant improvement of
+%   \peace\/ (blue) over \wcd\/ (green) and \capthree\/ (black) in its
+%   sensitivity to sequencing errors, while in Figure~\ref{SeT1} (right)
+%   we see a comparable improvement in Type 1 error.  In
+%   Figure~\ref{singletons} we look at the number of singleton clusters
+%   (fragments not joined to any cluster), which should not occur in our
+%   simulated sets.  We again see significant improvement in \peace.
+%   The sequential runtime of \peace\/ is slightly improved over that of
+%   \wcd\/ (see Supplementary Materials, Figure~S4).
 
 \begin{figure}
-  \centerline{\includegraphics[width=3.35in]{pics.d/SeT1.pdf}}
+  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2.40.eps}}
   \caption{Comparisons of Sensitivity and Type 1 error, based on the
@@
     presented for Type 1 error due to the extremely small
-    variance.}\label{SeT1}
+    variance.}\label{SeJiT1T2}
 \end{figure}
 
-\begin{figure}
-  \centerline{\includegraphics[width=3in]{pics.d/singletons.pdf}}
-  \caption{Average number of fragments flagged as singletons by each tool
-    when run on the simulated sequences; correct answer in all cases is
-    zero.  Blue/Solid = \peace, Green/Dash = \wcd,
-    Black/Dot-Dash = \capthree.}\label{singletons}
-\end{figure}
+\mc{In applying the tools to real data Sanger data, we used the Human
+Benchmark Dataset used to test \easycluster\/ \cite{Picardi09}, and
+the A076941.fa {\it Arabidopsis thaliana} dataset used to test WCD
+\cite{Hazelhurst08,Hazelhurst08a}.  For the first, we find an almost
+exact match in results in quality between PEACE and WCD (both
+out-preforming Cap3), but a significant gain in runtime for PEACE over
+WCD, as shown in Tabe\ref{XXX}}.  For the A076941.fa...}
 
-In applying the tools to real data, we started with a set of
-approximately $190,000$ clean Sanger ESTs derived from the {\it
-  Chlamydomonas reinhardtii} genome \cite{Liang2008}.  To compute
-sensitivity we used the {\it gmap} tool \cite{Wu05} to map the
-fragment set to the genome, taking this as our reference clustering.
-We see slight improvements in \peace\/ over \wcd\/ in both sensitivity
-and Type 1 error (both significantly outperforming \capthree).  Using
-the mouse data set used in Hazelhurst {\it et al.}
-\cite{Hazelhurst08a} \peace\/ again shows slight improvements in
-sensitivity and Type 1 error rates, with an 18\% speedup for \peace\/ 
-(see Supplementary Materials, Section~D, for more details).
+% we started with a set of
+% approximately $190,000$ clean Sanger ESTs derived from the {\it
+%   Chlamydomonas reinhardtii} genome \cite{Liang2008}.  To compute
+% sensitivity we used the {\it gmap} tool \cite{Wu05} to map the
+% fragment set to the genome, taking this as our reference clustering.
+% We see slight improvements in \peace\/ over \wcd\/ in both sensitivity
+% and Type 1 error (both significantly outperforming \capthree).  Using
+% the \mc{Arabidopsis} data set used in Hazelhurst {\it et al.}
+% \cite{Hazelhurst08a} \peace\/ again shows slight improvements in
+% sensitivity and Type 1 error rates, with an 18\% speedup for \peace\/ 
+% (see Supplementary Materials, Section~D, for more details).
 

### 2010-04-19 peace/papers/NAR_web_server_2009/paper.tex
Updates to NAR paper, reviewer response

@@
   Other measurements are also discussed in the supplementary
-  materials.  In Figure~\ref{SeJiT1T2} (left) we see these four tests,...}
-
-% , reflecting tests done on
-%   simulated Sanger EST sets, we see a significant improvement of
-%   \peace\/ (blue) over \wcd\/ (green) and \capthree\/ (black) in its
-%   sensitivity to sequencing errors, while in Figure~\ref{SeT1} (right)
-%   we see a comparable improvement in Type 1 error.  In
-%   Figure~\ref{singletons} we look at the number of singleton clusters
-%   (fragments not joined to any cluster), which should not occur in our
-%   simulated sets.  We again see significant improvement in \peace.
-%   The sequential runtime of \peace\/ is slightly improved over that of
-%   \wcd\/ (see Supplementary Materials, Figure~S4).
+  materials.  In Figure~\ref{SeJiT1T2} we plot these four tests as a
+  function of erro rate, and observe the almost identical results
+  between \peace\/ and \wcd\/.  In Figure~\ref{seq_parallel} we plot
+  the runtime for \peace\/ and \wcd\/, again observing almost identical
+  results when run sequentially -- but signifcantly faster runtime for
+  PEACE on multiple processer when holding the EST/processor ratio
+  constant (ranging from a 65\% improvement for two processors too a
+  17\% improvement for 12 processors).}
 
 \begin{figure}
-  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2.40.eps}}
-  \caption{Comparisons of Sensitivity and Type 1 error, based on the
+  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.pdf}}
+  \caption{Comparisons of Sensitivity, Jiccard Index, Type 1 error and
+    Type 1 error, based on the
     average over 30 simulated Sanger Sequence ESTs sets derived from 100 zebra
@@
     tics = 95\% confidence intervals on estimates.  Intervals are not
-    presented for Type 1 error due to the extremely small
+    presented for Type 1 error due to the effective lack of
     variance.}\label{SeJiT1T2}
@@
 
-In applying the tools to real data, we started with a set of
-approximately $190,000$ clean Sanger ESTs derived from the {\it
-  Chlamydomonas reinhardtii} genome \cite{Liang2008}.  To compute
-sensitivity we used the {\it gmap} tool \cite{Wu05} to map the
-fragment set to the genome, taking this as our reference clustering.
-We see slight improvements in \peace\/ over \wcd\/ in both sensitivity
-and Type 1 error (both significantly outperforming \capthree).  Using
-the \mc{Arabidopsis} data set used in Hazelhurst {\it et al.}
-\cite{Hazelhurst08a} \peace\/ again shows slight improvements in
-sensitivity and Type 1 error rates, with an 18\% speedup for \peace\/ 
-(see Supplementary Materials, Section~D, for more details).
+\begin{figure}
+   \centerline{\includegraphics[width=3.35in]{pics.d/seq_parallel.pdf}}
+   \caption{Comparisons of Runtime: On left, we compare the sequential
+     runtime of \peace\/ (blue) and \wcd\/ (green) on simulate sets
+     ranging in size from 10K sequences to 100K sequences.  (\capthree\/
+     runtime is not reported, as the time spent on clustering cannot
+     be differentiated from the time spent on assembly.)  On right,
+     a comparison when run in parallel, ranging the number of
+     processors from 2 to 12 while holding the EST/Processor ratio
+     steady at the constant 8888.  All values represent the average
+     of 30 runs; vertical tics = 95\% confidence intervals on
+     estimates.  All runs were done on a 3.0 GHz Intel Xeon EM64T CPU with a 2 MB
+     cache and 800 MHz front side bus, model number Xeon LV 3.0 (2005).}\label{seq_parallel}
+\end{figure}
+
+\mc{In applying the tools to real data Sanger data, we used the Human
+Benchmark Dataset used to test \easycluster\/ \cite{Picardi09}, and
+the A076941.fa {\it Arabidopsis thaliana} dataset used to test WCD
+\cite{Hazelhurst08,Hazelhurst08a}, with results in
+Table~\ref{SangerReal}.  We notice essentially identical
+results for PEACE and WCD in quality, both significantly better than
+Cap3 in Sensitivity and Type 1 error rate, while slightly worse in the
+Jaccard Index and Type 2 error.  In runtime we see some inconsistency,
+with PEACE showing a 60\% runtime improvement in the smaller dataset,
+requiring 20\% more time in the Arabidopsis dataset.}
+
+\begin{table*}[tbp]
+\begin{center}
+\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
+\hline
+ & 
+& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
+  error \end{tabular} & \begin{tabular}{c} Type 2 \\
+  error \end{tabular} &
+\begin{tabular}{c} Number \\ of  \\
+    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
+    Singletons \end{tabular}
+    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
+\hline EasyCluster Human & \peace & 0.998 & 0.672 & 0.153 & 0.042 & 118 & 21 & 293 \\
+\cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
+\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1 & 0.001 & 2269 & 1827 & \mbox{NA} \\
+\hline
+\hline WCD A076941 & \peace & 0.932 & 0.475 & 0.351 & 0.027 & 18825 & 8951 & 1166\\
+\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
+\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
+\mbox{NA} \\
+\hline
+\end{tabular}
+\label{SangerReal}
+\end{center}
+\caption{Comparisons of runs on the EasyCluster human Benchmark
+  Dataset and the WCD A076941 Arabidopsis thaliana dataset using the
+  standard quality measurments.}
+\end{table*}
+
+% we started with a set of
+% approximately $190,000$ clean Sanger ESTs derived from the {\it
+%   Chlamydomonas reinhardtii} genome \cite{Liang2008}.  To compute
+% sensitivity we used the {\it gmap} tool \cite{Wu05} to map the
+% fragment set to the genome, taking this as our reference clustering.
+% We see slight improvements in \peace\/ over \wcd\/ in both sensitivity
+% and Type 1 error (both significantly outperforming \capthree).  Using
+% the \mc{Arabidopsis} data set used in Hazelhurst {\it et al.}
+% \cite{Hazelhurst08a} \peace\/ again shows slight improvements in
+% sensitivity and Type 1 error rates, with an 18\% speedup for \peace\/ 
+% (see Supplementary Materials, Section~D, for more details).
 
@@
 approximately 17.6 minutes to process the 7.9 Mb of data when run in
-sequential mode on a \mc{3.2 GHz Intel Xeon EM64T CPU with a 2 MB
-  cache and 800 MHz font side bus}.  When applied to a 64 Mb set of
+sequential mode on a .  When applied to a 64 Mb set of
 165,490 {\it

### 2010-04-19 peace/gui/src/org/peace_tools/core/job/AnalyzerWizardPage.java
I noticed that the command line for non-adaptive mode was using twoPassD2 with --dontAdapt flag which is no longer the case. So I fixed it to use just twoPassD2.
Touched up the description for non-adaptive two pass d2 to say it is primarily
for sanger sequences.

@@ -309,7 +309,7 @@ public class AnalyzerWizardPage extends GenericWizardPage implements ActionListe
 	 */
 	private static final String ANALYZERS[] = {
 		"Adaptive Two Pass D2 (safe & best choice)", 
-		"Non-adaptive Two Pass D2 (Sanger sequences)",
+		"Non-adaptive Two Pass D2 (for Sanger sequences)",
 		"D2 Optimized (similar to wcd)", 
 		"D2 Standard",
 		"CLU (similarity metric)"

### 2010-04-19 peace/gui/src/org/peace_tools/workspace/FWAnalyzer.java

@@ -227,7 +227,7 @@ public class FWAnalyzer {
 	 * @return Return the information as a command line parameter.
 	 */
 	public String toCmdLine() {
-		final String[] AnalyzerName = {"twopassD2", "twopassD2 --dontAdapt", "d2", "d2zim", "clu"};
+		final String[] AnalyzerName = {"twopassD2adapt", "twopassD2", "d2", "d2zim", "clu"};
 		String cmdLine = "--analyzer " + AnalyzerName[type.ordinal()];
 		cmdLine += " --frame " + windowSize;
 		cmdLine += " --word "  + wordSize;

### 2010-04-19 src/ParameterSetManager.h
Adjusted adaptive parameter settings.

@@ -214,8 +214,8 @@ public:
         of data sets.
     */
     static void setupParameters(int t1 = 20, int u1 = 4, int ws1 = 4,
-                                int t2 = 30, int u2 = 6, int ws2 = 8,
-                                int t3 = 40, int u3 = 8, int ws3 = 8);
+                                int t2 = 35, int u2 = 6, int ws2 = 8,
+                                int t3 = 50, int u3 = 8, int ws3 = 8);
 
 	/** Method to setup adaptive parameters to \i fast(er) settings.
 

### 2010-04-19 peace/src/ParameterSetManager.cpp
Adjusted adaptive parameter settings.

@@ -54,11 +54,11 @@ ParameterSetManager::setupParameters(int t1, int u1, int ws1, int t2,
 
     // Add the default parameter sets to the list of sets.
     // Line 1 is Two-Pass D2, Line 2 is heuristic parameters
-    ptrInstance->addParameterSet(new ParameterSet(-1, 150, 50, 1, 45, 45,
+    ptrInstance->addParameterSet(new ParameterSet(-1, 150, 50, 1, 40, 40,
                                                   t1, u1, ws1));
-    ptrInstance->addParameterSet(new ParameterSet(150, 400, 75, 25, 75, 100,
+    ptrInstance->addParameterSet(new ParameterSet(150, 400, 75, 25, 60, 100,
                                                   t2, u2, ws2));
-    ptrInstance->addParameterSet(new ParameterSet(400, INT_MAX, 100, 50, 105,
+    ptrInstance->addParameterSet(new ParameterSet(400, INT_MAX, 100, 50, 80,
                                                   130, t3, u3, ws3));
 }
 

### 2010-04-19 papers/NAR_web_server_2009/paper.tex
Changes to paper, response, and supplementary materials

@@
 fragments of varying length under different models of error
-(Supplementary Materials, Section~C.1), generating the fragments from
+(Supplementary Materials, Section~D.1), generating the fragments from
 the list of 100 zebra fish genes used in the \wcd\/ testing
@@
     tics = 95\% confidence intervals on estimates.  Intervals are not
-    presented for Type 1 error due to the extremely small
+    presented for Type 1 error due to the effective lack of
     variance.}\label{SeJiT1T2}
+\end{figure}
+
+\begin{figure}
+   \centerline{\includegraphics[width=3.35in]{pics.d/seq_parallel.pdf}}
+   \caption{Comparisons of Runtime: On left, we compare the sequential
+     runtime of \peace\/ (blue) and \wcd\/ (green) on simulate sets
+     ranging in size from 10K sequences to 100K sequences.  (\capthree\/
+     runtime is not reported, as the time spent on clustering cannot
+     be differentiated from the time spent on assembly.)  On right,
+     a comparison when run in parallel, ranging the number of
+     processors from 2 to 12 while holding the EST/Processor ratio
+     steady at the constant 8888.  All values represent the average
+     of 30 runs; vertical tics = 95\% confidence intervals on
+     estimates.  All runs were done on a 3.0 GHz Intel Xeon EM64T CPU with a 2 MB
+     cache and 800 MHz front side bus, model number Xeon LV 3.0 (2005).}\label{seq_parallel}
 \end{figure}
@@
 the A076941.fa {\it Arabidopsis thaliana} dataset used to test WCD
-\cite{Hazelhurst08,Hazelhurst08a}.  For the first, we find an almost
-exact match in results in quality between PEACE and WCD (both
-out-preforming Cap3), but a significant gain in runtime for PEACE over
-WCD, as shown in Tabe\ref{XXX}}.  For the A076941.fa...}
+\cite{Hazelhurst08,Hazelhurst08a}, with results in
+Table~\ref{SangerReal}.  We notice essentially identical
+results for PEACE and WCD in quality, both significantly better than
+Cap3 in Sensitivity and Type 1 error rate, while slightly worse in the
+Jaccard Index and Type 2 error.  In runtime we see some inconsistency,
+with PEACE showing a 60\% runtime improvement in the smaller dataset,
+requiring 20\% more time in the Arabidopsis dataset.}
+
+\begin{table*}[tbp]
+\begin{center}
+\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
+\hline
+ & 
+& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
+  error \end{tabular} & \begin{tabular}{c} Type 2 \\
+  error \end{tabular} &
+\begin{tabular}{c} Number \\ of  \\
+    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
+    Singletons \end{tabular}
+    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
+\hline EasyCluster Human & \peace & 0.998 & 0.672 & 0.153 & 0.042 & 118 & 21 & 293 \\
+\cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
+\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1 & 0.001 & 2269 & 1827 & \mbox{NA} \\
+\hline
+\hline WCD A076941 & \peace & 0.932 & 0.475 & 0.351 & 0.027 & 18825 & 8951 & 1166\\
+\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
+\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
+\mbox{NA} \\
+\hline
+\end{tabular}
+\label{SangerReal}
+\end{center}
+\caption{Comparisons of runs on the EasyCluster human Benchmark
+  Dataset and the WCD A076941 Arabidopsis thaliana dataset using the
+  standard quality measurments.}
+\end{table*}
 

### 2010-04-19 papers/NAR_web_server_2009/paper.sup.tex
Adding tex version of sup. materials

--- 2010-01-03 peace/papers/bioinf_apps_2009/bio-app.sup.tex
@@
 \input{epsf}
+
+\newcommand{\peace} {{\small PEACE}}
+\newcommand{\wcd} {{\small WCD}}
+\newcommand{\capthree} {{\small Cap3}}
+\newcommand{\easycluster} {{\small EasyCluster}}
 
@@
 
-\newlabel{sensitivityFig}{1}
+\newlabel{sensitivityFig}{3}
+\newlabel{SangerTable}{1}
 
@@
 
-\title{PEACE: {\underline P}arallel {\underline E}ST {\underline A}nalysis
-  and {\underline C}lustering {\underline E}ngine \\Supplementary
-  Materials}
+\title{PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly
+  and {\underline C}lustering of Gene {\underline E}xpression}
 
@@
 challenge, sets the threshold at $T=40$, hypothesizing that EST pairs
-with a $d^2$ distance of greater than 40 are unlikely to overlap.  We
-have set out threshold to $T=130$.
+with a $d^2$ distance of greater than 40 are unlikely to overlap.  \mc{We
+have set our threshold to the same $T=40$ (with discussion below on
+the effects of changing this).}
+
+\subsection{Adaptive $d^2$}
+
+In the above discussion of $d^2$ we assume sequences of at least 100 bp, and
+in pratice use a window of size 100.  While the technique should be,
+in principle, adaptable to short-read sequences (decreasing window
+size as necessary), there are some problems in the details.
+Diminishing window sizes implies both diminishing accuracy (requiring
+the thrshold be adjusted) and increasing runtime (as there are
+potenitally more window pairs).  More importantly, there is the
+problem of non-homogonous sequence lengths: short-read technologies
+tend to produce short reads, but can still include quite long reads in
+their output.  Hence it is not immediately clear how to compare a very
+short read against a very long one.
+
+We address these issues with our {\it adaptive $d^2$} strategy that
+works as follows: we put sequences, based on length, into three {\it
+  overlapping} groups (i.e. sequences less than length XXX, sequences
+between length YYY and ZZZ, and seqeunces greater than AAA).  For each
+group we pick a constant windowsize and threshold tailored to the
+length range of the group (as well as filtering heurstic constants),
+and use those values when computing the $d^2$ distance for two
+sequences in the same group.  We do not compute distances between
+sequences in different groups. Once computed, we assign to the MST
+edge the ratio of the $d^2$ score to the group threshold, thus
+normalizing the weight for comparisons ``between groups'' and
+effectivally giving us an edge-removal threhold of 1.
+
+By restricting $d^2$ comparisons to within a group, we minimize the
+effets of the hetrogenous sequences lengths by restricting our
+comparisons to only sequences of similar lengths.  By not restrcicting
+ourselves to a single window size, we allow ourselves to, when possible, avoid the
+preformance hit of using a small window.  As we are not comparing
+sequences from different groups we bypass the question of how to
+compare a largelong sequence against a short one.  And because the
+groups overlap in content, we still get implicit comparisons between
+long and short sequences in terms of the resulting path of the minimum
+spanning tree that makes use of these ``linking'' nodes.
+
+In the \textsc{peace} implementation, we subdevide into three groups:
+NEED INFO.
 
@@
 In {\it Hazelhurst et al.} [\cite{Hazelhurst08a}] the authors conduct
-an investigation of \textsc{wcd} against a number of clustering
+an investigation of \textsc{wcd} \mc{(version 0.5.1)} against a number of clustering
 tools.  As they make a convincing argument that \textsc{wcd} is returning
@@
 
-While our primary measurement of tool result is sensitivity
-(Figure~\ref{sensitivityFig}), a number of alternative measurements
-were investigated as well. The {\it Jaccard} index measures both
-sensitivity and specificity, calculated as the ratio of true positives
-to the sum of true positives, true negatives, and false positives ($tp
-/ (tp+fn+fp)$) [\cite{Hazelhurst08a}].  Notably, when we removed duplicated genes before
-generating the simulated EST sets, we found all three tools to have
-100\% specificity: no false positives.  In such a case the Jaccard
-Index is identical to the Sensitivity Index, hence \textsc{peace} exhibits the
-same improvement on both indexes relative to \textsc{wcd} and \textsc{cap3}.  Applying
-the tools to the original genes (in which 21\% of the genes were
-highly similar to genes in the remaining 79\%), we find only a slight
-difference in the Jaccard Index, and the same relative performance of
-the three tools (Figure~\ref{JiSe}).  The sensitivity of each tool to
-duplications is discussed below.
+\mc{ Our primary measurement tools if mesurement are sensitivity, the
+Jaccard Index, Type 1 error and Type 2 error
+(Figure~\ref{sensitivityFig}).  Sensitivity is the ratio for true
+positives to actual positives ($tp / (tp + fn)$) -- the fraction of
+sequences from the same transcript that are identified as such.  The
+{\it Jaccard} index combines both sensitivity and specificity,
+calculated as the ratio of true positives to the sum of true
+positives, true negatives, and false positives ($tp / (tp+fn+fp)$)
+[\cite{Hazelhurst08a}].  Notably, when we removed duplicated genes
+before generating the simulated EST sets, we found all three tools to
+have 100\% specificity: no false positives. }
 
-%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[tbp]
-\centerline{
-\includegraphics[scale=0.5]{4JiSe.eps}
-\label{JiSe}
-}
-\caption{Comparisons of Sensitivity and the Jaccard Index as a
-  function of base read error rate, based on 30 simulated EST sets
-  derived from a set of 79 genes with no significant similarity (left
-  column), and an additional 21 genes showing significant similarity
-  to the base set (right column).  Blue/Solid = \textsc{peace}, Green/Dash =
-  \textsc{wcd}, Black/Dot-Dash = \textsc{cap3}; vertical tics = 95\% confidence
-  intervals on estimates.}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%
-
-Type 1 and Type 2 errors, as defined in {\it Wang et al.}
+\mc{Type 1 and Type 2 errors, as defined in {\it Wang et al.}
 [\cite{Wang04}], measure quality at the level of the genes from which
@@
 Type 2 error we look at the fraction of clusters that contain two or
-more genes.   In Figure~\ref{t1t2} we compare the Type 1 and Type 2 error rates of
-the three tools for varying levels of simulated base read error.  For
-Type 1 error we find that while all tools do equally poorly when the
-EST is subjected to more than a $5\%$ error rate, \textsc{peace} does
-significantly better than \textsc{wcd} in the lower ranges (while
-\textsc{wcd} does correspondingly better than \textsc{cap3}).  When
-looking at Type 2 error rate, we find that \textsc{peace} does a
-generally better job than \textsc{cap3} in separating genes into
-different clusters.  We note, however, that Type 1 error is the more
-important measurement: genes incorrectly split between clusters cannot
-be later recovered (without re-examining the cluster as a whole), but
-there is the potential for an assembly program to later identify
-incorrectly merged clusters based only on local cluster information.
+more genes.   }
 
-%%%%%%%%%%%%%%
+\mc{A fifth metric is to look at the number of singletons produce by a
+  tool:  sequences that  could not  be  assigned to  any cluster.   In
+  practice such a sequence could represent compelte coverage of a very
+  short transcript (one that may be otherwise had to identify), but is
+  more likely to be an error in the clustering algorithm.  In order to
+  identify  the first  it is  important  to minimize  the second.   In
+  Figure~\ref{singletons} we look at the number of singletons produced
+  by each tool when run on  simulated data in which there should be no
+  singletons produced,  seeing a very slight improvment  of PEACE over
+  WCD in certain error ranges.}
+
 \begin{figure}[tbp]
-\centerline{\includegraphics[scale=0.5]{t1t2.eps}
-\label{t1t2}
+\centerline{
+\label{dups}
+\includegraphics[scale=0.35]{pics.d/singletons_40.pdf}
 }
-\caption{Type 1 error rate (fraction of genes partitioned) and Type 2
-  error rate (fraction of clusters containing multiple genes) as a
-  function of base read error rate.  Estimates averaged over
-  application to 30 simulated ESTs sets derived from the full 100
-  zebra fish gene set. Blue/Solid = \textsc{peace}, Green/Dash =
-  \textsc{wcd}, Black/Dot-Dash = \textsc{cap3}; vertical tics = 95\%
-  confidence intervals on estimates (not shown for Type 1 error due to
-  the extremely small variance in estimates).}
+\caption{Number of singletons produced by each tool in a simulated run
+  where a fully correct clustering would have no singletons.  Values
+  average over 30 triles.  Blue/Solid = \textsc{peace}, Green/Dashed =
+  \textsc{wcd}, Black/Dot-Dashed = \textsc{cap3}.}\label{singletons}
 \end{figure}
-%%%%%%%%%%%%%%
 
@@
 \label{dups}
-\includegraphics[scale=0.35]{duplicates.eps}
+\includegraphics[scale=0.35]{pics.d/duplicates_40.pdf}
+}
+\caption{Ability to distinguish duplicates as a function of
+  divergence.  Estimates averaged over 30 trials; one trial consists
+  of taking a random gene, copying it and stochastically changing bases
+  at the specified rate, then using the two genes as the bases for
+  generating a simulated set. Blue/Solid = \textsc{peace}, Green/Dashed = \textsc{wcd}, Black/Dot-Dashed = \textsc{cap3};
+  variance was too small for visible plotting of
+  confidence intervals.}
+\end{figure}
+
+\subsection{Raising the Threshold}
+\mc{ 
+  The \wcd\/ tool uses a $d^2$ threshold of $40$ to determine whether
+  two sequences are in the same cluster.  As this is equivilent to our
+  $T$ threshold for dismissing MST edges, we have set it to the same
+  default value when used for Sanger sequences.  However, we did
+  experiment with higher values, to interesting results.  A basic
+  analysis indicated that a $T$ value as high as 130 should not have a
+  significant impact on its effectivness as a descriminator.  And in
+  fact this was born out in simulation, as shown in
+  Figure~\label{SiJiT1T2130}.  In contrast with the equivilent results
+  for a threshold of $T=40$, as shown in the manuscript, we see
+  considerable improvement in \peace\/ over \wcd\/ in all metrics
+  except Type 2 error, in which \wcd\/ slightly
+  better. Unsurprisingly, this also makes a significant difference in
+  the tool's ability to distinguish duplications: we see fairly poor
+  performance in Figure~\ref{dups130}.  
+}
+
+\mc{
+  Most importantly, while the higher threshold appeared to be better
+  in theory and in analysis, it did not hold in practice.  In
+  Table~\ref{SangerTable130} we show the match to the manuscript
+  Table~\ref{SangerTable}, and observe the poor performance of
+  \peace\/ when looking at the Jaccard index.  It is clear that in the
+  real data, the level of similarity is consierably higher then in the
+  simulations, thus making the reduced specificity of the higher
+  threshold considerably more of a problem.
+XXX}
+
+\begin{figure}
+  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.pdf}}
+  \caption{Comparisons of Sensitivity, Jiccard Index, Type 1 error and
+    Type 1 error, based on the
+    average over 30 simulated Sanger Sequence ESTs sets derived from 100 zebra
+    fish genes  (see
+    Supplementary Materials, Section C, for more details).  Blue/Solid
+    = \peace, Green/Dash = \wcd, Black/Dot-Dash = \capthree; vertical
+    tics = 95\% confidence intervals on estimates.  Intervals are not
+    presented for Type 1 error due to the effective lack of
+    variance.}\label{SeJiT1T2130}
+\end{figure}
+
+\begin{figure}[tbp]
+\centerline{
+\label{dups130}
+\includegraphics[scale=0.35]{pics.d/duplicates_130.pdf}
 }
@@
 
+\mc{All tests were run on the University of Miami Redhawk cluster,
+  made up of 3.2 GHz Intel Xeon EM64T CPUs, each with a 2MB cache and
+  800 MHz side bus, model number Xeon LV 3.0 (2005).}
+
 %%%%%%%%%%%%%
@@
 \centerline{
-\includegraphics[scale=0.35]{seq_time.eps}
+\includegraphics[scale=0.35]{pics.d/seq_time.pdf}
 \label{seq_runtime}
@@
 \centerline{
-\includegraphics[scale=0.5]{par_time.eps}
+\includegraphics[scale=0.5]{pics.d/par_time.pdf}
 \label{par_runtime}
@@
 
-\section{Applications to Real Data Sets}
-
-In Table~\ref{chlamy} we lay out more comprehensive results for the
-application of our tool to an EST set collected from the {\it
-  Chlamydomonas reinhartdii} genome, while in Table~\ref{mouse} 
-we see the results for the mouse set used for the \textsc{wcd} analysis
-[\cite{Hazelhurst08a}].  We note that for the Chlamydomonas genome the
-correct clustering is not known; our estimates of sensitivity are
-based on as assignment of ESTs to the sequenced genomes by the Gmap tool
-[\cite{Wu05}], leaving some potential for error in the  reference.
-Further, we found a number of ligations in the data (ESTs derived by
-incorrectly combining multiple transcripts),  which were removed before testing.
-
-Interestingly, where in the simulated results we saw significant
-improvement in \textsc{peace} over \textsc{wcd} in terms of result quality and only a
-light improvement in runtime, here we see the reverse. The two tools are
-roughly comparable  in terms of sensitivity, Type 1 error and Type 2
-error, while \textsc{peace} shows a 10\% and 20\% improvement in runtime (both
-showing significant improvements over \textsc{cap3}).  Predictably , the full
-assembly tool does somewhat better in terms of specificity (the Jaccard
-Index and Type 2 error), though scores here are worse for all three
-tools due to the increased number of homologous gene pairs as compared
-to our simulated sets.
+\section{Raising the Threshold Real Data Sets}
 
@@
 \begin{center}
-\begin{tabular} {| c || c | c | c | c | c | c | c | c |}
+\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
 \hline
+ & 
 & Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
   error \end{tabular} & \begin{tabular}{c} Type 2 \\
-  error \end{tabular} & \begin{tabular}{c} Number \\ of  \\
-    Clusters \end{tabular} & Singletons & \begin{tabular}{c} Single
-      \\ processor \\ runtime (s) \end{tabular}\\
-\hline \hline \textsc{peace} & 0.958 & 0.386 & 0.623 & 0.026 & 19649 & 10493 & 8799 \\
-\hline \textsc{wcd} & 0.94 & 0.500 & 0.654 & 0.021 & 22433 & 13000 & 9913 \\
-\hline \textsc{cap3} & 0.77 & 0.740 & 0.737 & 0.015 & 33771 & 22169 &  \\
+  error \end{tabular} &
+\begin{tabular}{c} Number \\ of  \\
+    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
+    Singletons \end{tabular}
+    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular} \\
+\hline EasyCluster Human & \peace & 0.999 & 0.106 & 0.027 & 0.030 & 67 & 2 & 295 \\
+\cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
+\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1 & 0.001 & 2269 & 1827 & \mbox{NA} \\
+\hline
+\hline WCD A076941 & \peace & 0.937 & 0.345 & 0.337 & 0.0342 & 18256 & 8135 & 1293 \\
+\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
+\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
+\mbox{NA} \\
 \hline
 \end{tabular}
-\label{chlamy}
+\label{SangerReal130}
 \end{center}
-\caption{{\bf Chlamydomonas reinhartdii}: 189975 ESTs, average length
-  553 bp, estimated 9886 actual genes represented.}
-\end{table}
-
-\begin{table}[tbp]
-\begin{center}
-\begin{tabular} {| c || c | c | c | c | c | c | c | c |}
-\hline
-& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
-  error \end{tabular} & \begin{tabular}{c} Type 2 \\
-  error \end{tabular} & \begin{tabular}{c} Number \\ of  \\
-    Clusters \end{tabular} & Singletons & \begin{tabular}{c} Single
-      \\ processor \\ runtime (s) \end{tabular}\\
-\hline \hline \textsc{peace} & 0.936 & 0.348 & 0.340 & 0.034 & 18338 & 8206 & 820\\
-\hline \textsc{wcd} & 0.932 & 0.476 & 0.350 & 0.0275 & 18787 & 8554 & 1000\\
-\hline \textsc{cap3} & 0.830 & 0.805 & 0.486 & 0.0135 & 25040 & 14925 & \\
-\hline
-\end{tabular}
-\caption{{\bf A076951 (Mouse EST set)}: 76914 ESTs, average length 427,
-  13240 actual genes represented.}
-\label{mouse}
-\end{center}
+\caption{Comparisons of runs on the EasyCluster human Benchmark
+  Dataset and the WCD A076941 Arabidopsis thaliana dataset using the
+  standard quality measurments with $T=130$.}
 \end{table}

### 2010-04-19 papers/NAR_web_server_2009/paper.tex
Modifications to paper; addition of natbibOrig.sty file

@@
 with PEACE showing a 60\% runtime improvement in the smaller dataset,
-requiring 20\% more time in the Arabidopsis dataset.}
+requiring 20\% more time in the Arabidopsis dataset.  In Section~D.1
+of the Supplementary Materials we present runtimes several more
+sets, observing that while \peace\/ appears to be significantly more
+faster on the smaller sets, \wcd\/ does overtake it for larger sets.
 
@@
     & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
+\hline
 \hline EasyCluster Human & \peace & 0.998 & 0.672 & 0.153 & 0.042 & 118 & 21 & 293 \\
 \cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
-\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1 & 0.001 & 2269 & 1827 & \mbox{NA} \\
+\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1.000 & 0.001 & 2269 & 1827 & \mbox{NA} \\
 \hline
@@
 % (see Supplementary Materials, Section~D, for more details).
+
+\mc{We tested the three tools on short-read data using the MetaSim tool of
+{\it Richter et al.} [\cite{Richter2008}].  Encoded into MetaSim are
+modules configured to emulate the outpspecific tecnology,
+including the 454 and Illumina sequences.  For 454, the simulator  }
 

### 2010-04-19 peace/papers/NAR_web_server_2009/paper.sup.tex
Modifications to paper; addition of natbibOrig.sty file

--- 2010-04-19 papers/NAR_web_server_2009/paper.sup.tex
@@
 
+%\usepackage[pdftex,
+%colorlinks=true, %makes the links show up in color, rather than in a box
+%citecolor=black, %color of the in-text citation numbers
+%urlcolor=blue %color of the url links
+%]{hyperref}
+
 \sloppy 
@@
 \usepackage{color}
-\definecolor{blue}{rgb}{0,0,1}
-\newcommand{\mc}[1]{\textcolor{blue}{#1}}
+\definecolor{red}{rgb}{0,0,1}
+\newcommand{\mc}[1]{\textcolor{red}{#1}}
 %\newcommand{\mc}[1]{#1}   % Replace above line with this to eliminate all coloring
@@
 
-In the above discussion of $d^2$ we assume sequences of at least 100 bp, and
-in pratice use a window of size 100.  While the technique should be,
-in principle, adaptable to short-read sequences (decreasing window
-size as necessary), there are some problems in the details.
+In the above discussion of $d^2$ we assume sequences of at least 100
+bp, and in pratice use a window of size 100.  While the technique
+should be, in principle, adaptable to short-read sequences (decreasing
+window size as necessary), there are some problems in the details.
 Diminishing window sizes implies both diminishing accuracy (requiring
@@
 problem of non-homogonous sequence lengths: short-read technologies
-tend to produce short reads, but can still include quite long reads in
-their output.  Hence it is not immediately clear how to compare a very
-short read against a very long one.
+tend to produce short reads, but some technologiges can still include
+long reads in their output.  Hence it is not immediately clear how to
+compare a very short read against a very long one.
 
 We address these issues with our {\it adaptive $d^2$} strategy that
-works as follows: we put sequences, based on length, into three {\it
-  overlapping} groups (i.e. sequences less than length XXX, sequences
-between length YYY and ZZZ, and seqeunces greater than AAA).  For each
+works as follows: we partition sequences by length into disjoint groups.  For each
 group we pick a constant windowsize and threshold tailored to the
@@
 sequences in the same group.  We do not compute distances between
-sequences in different groups. Once computed, we assign to the MST
+sequences in non-adjaecent groups (i.e. a member of the ``shortest
+groups'' against a member of the ``longest groups'' if there is an
+interviening sized group).  Once computed, we assign to the MST
 edge the ratio of the $d^2$ score to the group threshold, thus
@@
 In the \textsc{peace} implementation, we subdevide into three groups:
-NEED INFO.
+\begin{itemize}
+\item Less than 150 ($w=50$, $T=40$, $t=20$, $u=4$).
+\item Between 150 and 400 ($w=75$, $T=60$, $t=35$, $u=6$)
+\item Greater than 400 ($w=100$, $T=80$, $t=50$, $t=50$).
+\end{itemize}
 
@@
 
-\mc{ Our primary measurement tools if mesurement are sensitivity, the
+\mc{Our primary measurement tools if mesurement are sensitivity, the
 Jaccard Index, Type 1 error and Type 2 error
@@
 before generating the simulated EST sets, we found all three tools to
-have 100\% specificity: no false positives. }
+have 100\% specificity: no false positives.}
 
@@
 Type 2 error we look at the fraction of clusters that contain two or
-more genes.   }
+more genes.}
 
@@
 \subsection{Raising the Threshold}
+\label{sim_thresh}
 \mc{ 
@@
   threshold considerably more of a problem.
-XXX}
+}
 
@@
 
-\section{Raising the Threshold Real Data Sets}
+\section{Real Data}
+
+\subsection{Runtimes}
+
+In Table~\ref{RunTime} we show the runtime of \peace\/ and \wcd\/ on
+five different benckmark sets obtained from the WCD and EasyCluster
+studies [\cite{Hazelhurst08a,Picardi09}].
+
+\begin{table}[tbp]
+\begin{center}
+\begin{tabular} {|c|| c | c || c | c |}
+\hline
+ Bencmark & \begin{tabular}{c} Number of \\ Sequences \end{tabular}
+ & \begin{tabular}{c} Number of \\ Bases \end{tabular} & Peace Runtme (s)&
+ WCD Runtime (s) \\
+\hline
+\hline
+ Human Cluster & 17733 & 11.7 Mb & 293 & 804 \\
+\hline
+ Cotton & 29,995 & 17.1 Mb & 222 & 339 \\
+\hline
+ Ricinus & 57,690 & 40.9 Mb & 1344 & 2356 \\
+\hline
+ A076941 & 76,941 & 32.8 Mb & 1166 & 966 \\
+\hline
+\hline
+ A686903 & 686,904 & 294.7 Mb & 9449 & 2481 \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Runtime for \peace\/ and \wcd\/ on benchmarks from the
+  WCD and EasyCluster studies [\cite{Hazelhurst08a, Picardi09}].
+  Runtimes for the first four sets are sequential, while that last was
+  run over thirty processors.}\label{RunTime}
+\end{table}
+
+\subsection{Higer Threshold}
+
+In Table~\ref{SangerReal130} we see the effect on the application of
+\peace\/ to the data sets from Table~\ref{SangerReal} when raising the
+threshold $T$ to 130 (see pervious discussion in
+Section~\ref{sim_thresh}.  While doing this appeared to be fairly
+effective in simulation, it did not hold up in pracice.  We see
+essentially no change in the sensitivity, but significant changes in
+the Jaccard Index (and a corresponding decrease in the total number of
+clusters).  We suspect that this is due to a higher number of more
+closely related genes than are present in the simulated data.  We can
+conclude that, without considerably more investigation, we should be
+settig $T$ to 40.
 
@@
     Singletons \end{tabular}
-    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular} \\
+    & \begin{tabular}{c} Single \\ processor \\ runtime
+      (s) \end{tabular} \\
+\hline
 \hline EasyCluster Human & \peace & 0.999 & 0.106 & 0.027 & 0.030 & 67 & 2 & 295 \\
 \cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
-\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1 & 0.001 & 2269 & 1827 & \mbox{NA} \\
+\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1.000 & 0.001 & 2269 & 1827 & \mbox{NA} \\
 \hline

### 2010-04-19 papers/NAR_web_server_2009/paper.tex
Modifications to resubmission

--- 2010-04-19 papers/NAR_web_server_2009/paper.tex
@@
 
-\history{Received Feb. 13, 2010}
+\history{Received Feb. 13, 2010 \\ Resubmitted April 20, 2010}
 
@@
   false positives by looking at the ratio of true positives to the sum
-  of true positives, false positives, and true negitives), {\it Type 1
+  of true positives, false positives, and true negatives), {\it Type 1
     error} (the fraction of genes that were divided between clusters),
@@
   materials.  In Figure~\ref{SeJiT1T2} we plot these four tests as a
-  function of erro rate, and observe the almost identical results
+  function of error rate, and observe the almost identical results
   between \peace\/ and \wcd\/.  In Figure~\ref{seq_parallel} we plot
   the runtime for \peace\/ and \wcd\/, again observing almost identical
-  results when run sequentially -- but signifcantly faster runtime for
-  PEACE on multiple processer when holding the EST/processor ratio
+  results when run sequentially -- but significantly faster runtime for
+  PEACE on multiple processor when holding the EST/processor ratio
   constant (ranging from a 65\% improvement for two processors too a
@@
   \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.pdf}}
-  \caption{Comparisons of Sensitivity, Jiccard Index, Type 1 error and
+  \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
     Type 1 error, based on the
@@
 \mc{In applying the tools to real data Sanger data, we used the Human
-Benchmark Dataset used to test \easycluster\/ \cite{Picardi09}, and
-the A076941.fa {\it Arabidopsis thaliana} dataset used to test WCD
-\cite{Hazelhurst08,Hazelhurst08a}, with results in
+Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
+the A076941.fa {\it Arabidopsis thaliana} data set used to test WCD
+\cite{Hazelhurst08a,Hazelhurst08b}, with results in
 Table~\ref{SangerReal}.  We notice essentially identical
@@
 Jaccard Index and Type 2 error.  In runtime we see some inconsistency,
-with PEACE showing a 60\% runtime improvement in the smaller dataset,
-requiring 20\% more time in the Arabidopsis dataset.}
+with PEACE showing a 60\% runtime improvement in the smaller data set,
+requiring 20\% more time in the Arabidopsis data set.  In Section~D.1
+of the Supplementary Materials we present runtimes several more
+sets, observing that while \peace\/ appears to be significantly more
+faster on the smaller sets, \wcd\/ does overtake it for larger sets.}
 
@@
     & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
-\hline EasyCluster Human & \peace & 0.998 & 0.672 & 0.153 & 0.042 & 118 & 21 & 293 \\
+\hline
+\hline Bicyclist Human & \peace & 0.998 & 0.672 & 0.153 & 0.042 & 118 & 21 & 293 \\
 \cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
-\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1 & 0.001 & 2269 & 1827 & \mbox{NA} \\
+\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1.000 & 0.001 & 2269 & 1827 & \mbox{NA} \\
 \hline
@@
 \caption{Comparisons of runs on the EasyCluster human Benchmark
-  Dataset and the WCD A076941 Arabidopsis thaliana dataset using the
-  standard quality measurments.}
+  Data Set and the WCD A076941 Arabidopsis thaliana data set using the
+  standard quality measurements.}
 \end{table*}
@@
 
-We were unable to run \wcd\/ on short-read sequences, as the
-distributed implementation is unable to handle such data without an
-adjustment to the code.  Hence, while \peace\/ was run on both
-simulated and real NGS fragment sets, there was no cluster tool to
-which the results could be compared.  \peace\/ did well, achieving a
-sensitivity of 0.75 when applied to a simulated short-read set
-subjected to a sequencing error rate of 2\% (with comparable {\it
-  Jaccard Index} values reflecting both sensitivity {\it and}
-specificity \cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was
-approximately 17.6 minutes to process the 7.9 Mb of data when run in
-sequential mode on a .  When applied to a 64 Mb set of
-165,490 {\it
- Chlamydomonas reinhardtii} transcript fragments
-sequenced by 454 pyrosequencing technology (ranging in size from 50
-bases to several hundred), \peace\/ completed the job in 49 minutes
-using 30 of the Xeon processors.
+\mc{We tested the three tools on short-read data using the MetaSim
+  tool of {\it Richter et al.} \cite{Richter2008}.  Encoded into
+  MetaSim are sequence generation and error models corresponding to
+  several technologies, including the 454 short-read technology.
+  Leaving \capthree\/ and \metasim\/ at their default values, but
+  employing the {\it adaptive $d^2$} strategy for short read sequences
+  (see Section~C in the supplementary materials), we find a that
+  \peace\/ performs significantly better than the other two tools on
+  454 data using the default MetaSim model (with sequence sizes
+  ranging in size from 200bp to 370bp, at an average size of 250bp and
+  a standard deviation of 17bp).  Over thirty runs, we see an average
+  sensitivity 0.884 for \peace, 0.331\/ for \wcd, and 0.004 for
+  \capthree.  For the Jaccard Index we get essentially the same
+  numbers.  For 100 transcript \peace creates 128 clusters (as opposed
+  to 6,370 and 19245 clusters for \wcd and \capthree), and creates
+  only 3 singletons (as opposed to 5,850 and 15432).  We find it
+  likely that the \wcd results could be considerably improved by the
+  simple adjustment of a few parameters, but were unable to determine
+  how to do that.  (Runtime of \peace\/ was somewhat slower than that
+  of \wcd, but given the significant improvement in quality results we
+  do not find this to be very troubling.)}
 
@@
 % LocalWords:  ssembly xpression GenBank dbEST pyrosequencing ESTsim sim mRNAs
-% LocalWords:  NGS polyA set's GHz Xeon Bikun
+% LocalWords:  NGS polyA set's GHz Xeon Bikun Arabidopsis JVM Jaccard LV
+% LocalWords:  Data Set thaliana data set EasyCluster MetaSim

### 2010-04-20 src/EST.cpp
Changes for ticket:60 :
I am done with incorporating the changes. For some reason I also had to 
move the earlier version of `MSTClusterMaker` class as well into the new 
code-path to achieve the earlier performance. I took a look at the differences 
but they seem minor; yet there was a notable performance difference that could
be bridged only with the earlier version of `MSTClusterMaker` that is now in 
the main branch under `OldMSTClusterMaker`. The command line to run all the 
earlier version of the code path is:

{{{

./peace --clusterMaker na-mst --analyzer oldTwopassD2 --heuristics oldTV

@@ -41,6 +41,7 @@
 #include <cctype>
 #include <cstring>
 #include <cstdio>
+#include <cstdlib>
 
 // The shared static list of ESTs currently available.
 std::vector<EST*> EST::estList;
@@ -60,7 +61,7 @@ EST::EST(const int idValue, const char *information, const char* seq,
 }
 
 EST::EST(const int idValue, const char *information, const bool maskBases,
-         const char* seq, const int fileOffset) :
+         const bool randomizeNbases, const char* seq, const int fileOffset) :
     id(idValue), sequenceLen(strlen(seq)), offset(fileOffset),
     customData(NULL) {
     // Duplicate the information and sequnece data as needed.
@@ -69,7 +70,7 @@ EST::EST(const int idValue, const char *information, const bool maskBases,
     similarity = 0;
     processed  = false;
     // Now normalize the base
-    normalizeBases(sequence, maskBases);
+    normalizeBases(sequence, maskBases, randomizeNbases);
 }
 
 EST::~EST() {
@@ -112,7 +113,8 @@ EST::getLine(FILE *fastaFile) {
 
 EST*
 EST::create(const int id, const char *info, const char* sequence,
-            const long offset, const bool maskBases) {
+            const long offset, const bool maskBases,
+            const bool randomizeNbase) {
     if (id != (int) estList.size()) {
         // This id is not acceptable. Sorry.
         return NULL;
@@ -120,7 +122,8 @@ EST::create(const int id, const char *info, const char* sequence,
     // Compute new max EST length
     maxESTlen = std::max(maxESTlen, strlen(sequence));
     // Instantiate new EST
-    EST *newEST = new EST(id, info, maskBases, sequence, offset);
+    EST *newEST = new EST(id, info, maskBases, randomizeNbase,
+                          sequence, offset);
     // Add est to end of the est list.
     estList.push_back(newEST);
     // return the newly created EST back tot he caller
@@ -128,7 +131,8 @@ EST::create(const int id, const char *info, const char* sequence,
 }
 
 EST*
-EST::create(FILE* fastaFile, int& lineNum, const bool maskBases) {
+EST::create(FILE* fastaFile, int& lineNum, const bool maskBases,
+            const bool randomizeNbase) {
     // Some basic validation on the fastFile first.
     if (feof(fastaFile) || ferror(fastaFile)) {
         // Can't read information from the fasta file.
@@ -173,7 +177,7 @@ EST::create(FILE* fastaFile, int& lineNum, const bool maskBases) {
[…]

### 2010-04-20 src/ESTAnalyzer.cpp

@@ -43,11 +43,12 @@
 #define NO_ERROR 0
 
 // The static instance variables for command line arguments.
-bool  ESTAnalyzer::readAhead      = false;
-char* ESTAnalyzer::estFileName    = NULL;
-char* ESTAnalyzer::sffFileName    = NULL;
-bool  ESTAnalyzer::htmlLog        = false;
-bool  ESTAnalyzer::noMaskBases    = false;
+bool  ESTAnalyzer::readAhead       = false;
+char* ESTAnalyzer::estFileName     = NULL;
+char* ESTAnalyzer::sffFileName     = NULL;
+bool  ESTAnalyzer::htmlLog         = false;
+bool  ESTAnalyzer::noMaskBases     = false;
+bool  ESTAnalyzer::randomizeNbases = false;
 
 // The common set of arguments for all EST analyzers
 arg_parser::arg_record ESTAnalyzer::commonArgsList[] = {
@@ -150,7 +151,8 @@ ESTAnalyzer::loadFASTAFile(const char *fileName, const bool unpopulate) {
     int lineNum = 1;
     // Repeatedly read EST's from the file.
     while (!feof(fastaFile)) {
-        EST *est = EST::create(fastaFile, lineNum, !noMaskBases);
+        EST *est = EST::create(fastaFile, lineNum, !noMaskBases,
+                               randomizeNbases);
         if ((est == NULL) && (!feof(fastaFile) || ferror(fastaFile))) {
             // An error occured when reading EST.
             fclose(fastaFile);
@@ -189,7 +191,7 @@ ESTAnalyzer::loadSFFFile(const char *fileName, const bool unpopulate) {
     }
     // Repeatedly read EST's from the file.
     while (sff.getPendingReads() > 0) {
-        EST *est = sff.getNextRead(!noMaskBases);
+        EST *est = sff.getNextRead(!noMaskBases, randomizeNbases);
         if ((est == NULL) || (!sff.isValid())) {
             // An error occured when reading EST.
             std::cerr << analyzerName << ": Error loading reads from "

### 2010-04-20 src/ESTAnalyzer.h

@@ -406,6 +406,27 @@ public:
         be written.
     */
     virtual void displayStats(std::ostream& UNREFERENCED_PARAMETER(os)) {}
+
+    /** Indicate how 'N' base characters in ESTs must be handled.
+
+        <p>Typically 'N' characters in reads are retained and are
+        masked over by the analyzers. However, in certain analyzers,
+        the assumption is that these 'N' characters are randomly
+        converted to a suitable base character.  This flag controls
+        the behavior whether 'N' characters should be randomly
+        converted.</p>
+
+        \param[in] flag If this value is \c false the \c 'N'
+        characters are \b not converted.  However, if the flag is set
+        to \c true then the 'N' characters are randomly converted to
+        regular bases.
+
+        \note This method must be called before the \c initialize
+        method in this class is invoked.
+    */
+    static void setRandomizeNbases(const bool flag) {
+        randomizeNbases = flag;
+    }
     
     /** The destructor.
 
@@ -486,7 +507,22 @@ protected:
         does the conversions.
     */
     static bool noMaskBases;
+
+	/** Flag to indicate if all 'N's must be randomly converted to
+        a valid base.
 	
+        Typically 'N' characters in reads are retained and are masked
+        over by the analyzers. However, in certain analyzers, the
+        assumption is that these 'N' characters are randomly converted
+        to a suitable base character.  This flag controls the behavior
+        whether 'N' characters should be randomly converted.  If this
+        value is \c false (the default) the characters are \b not
+        converted.  However, if the flag is set to \c true (either via
+        a command line parameter or through the static method), then
+        the 'N' characters are randomly converted to regular bases.
+    */
+    static bool randomizeNbases;
+
     /** The index of the reference EST in a given file. 
 
         This member object is used to hold the index of a reference

### 2010-04-20 src/ESTCodec.h

@@ -151,8 +151,8 @@ public:
     /** A functor to generate a encoded word (serves as a hash entry).
         
         This functor must be used to generate an encoded word from a
-	"normal" (rather than reverse complement) fragment. This
-	me thod handles 'n' entries in the EST in the following manner:
+        "normal" (rather than reverse complement) fragment. This
+        me thod handles 'n' entries in the EST in the following manner:
 
         <ul>
 
@@ -203,11 +203,25 @@ public:
     */
     template <const int& Shift, const int& Mask>
     struct NormalEncoder : public std::binary_function<int, char, int> {
+        /**
+           The newer-version of this method that explicitly tracks and
+           sets igoreMask to enable ignoring 'N' bases.
+        */
         inline int operator()(const int w, const char bp, int& ignoreMask) const {
             // Setup the ignore mask if base pair is 'n' otherwise
             // drop off the lowest 2 bits.
             ignoreMask = (bp != 'N') ? (ignoreMask >> 2) : Mask;
             // Compute the hash for the word. Encoding for 'n' is 0.
+            return ((w >> 2) | (ESTCodec::encode(bp) << Shift)) & Mask;
+        }
+
+        /**
+           The earlier-version of this method does not track ignoring
+           bases.  This method assumes that all 'N' characters have
+           been suitably replaced with a random base letter.
+        */
+        inline int operator()(const int w, const char bp) const {
+            // Compute the hash for the word.
             return ((w >> 2) | (ESTCodec::encode(bp) << Shift)) & Mask;
         }
     };
@@ -267,11 +281,24 @@ public:
     */
     template <const int& Shift, const int& Mask>
     struct RevCompEncoder : public std::binary_function<int, char, int> {
+        /**
+           The newer-version of this method that explicitly tracks and
+           sets igoreMask to enable ignoring 'N' bases.
+        */        
         inline int operator()(const int w, const char bp, int& ignoreMask) const {
             // Setup the ignore mask if base pair is 'n' otherwise
             // drop off the lowest 2 bits.
             ignoreMask = (bp != 'N') ? (ignoreMask >> 2) : Mask;
             // Compute the hash for the word. 'n' becomes 0.
+            return ((w << 2) | ESTCodec::encode2rc(bp)) & Mask;
+        }
+        /**
+           The earlier-version of this method does not track ignoring
+           bases.  This method assumes that all 'N' characters have
+           been suitably replaced with a random base letter.
+        */
[…]

### 2010-04-20 src/HeuristicFactory.cpp

@@ -37,6 +37,7 @@
 #include "HeuristicFactory.h"
 #include "UVSampleHeuristic.h"
 #include "TVHeuristic.h"
+#include "OldTVHeuristic.h"
 #include "arg_parser.h"
 
 void
@@ -70,6 +71,8 @@ HeuristicFactory::create(const char* name, const int refESTidx,
         return new UVSampleHeuristic("uv", outputFileName);
     } else if (!strcmp("tv", name)) {
         return new TVHeuristic(outputFileName);
+    } else if (!strcmp("oldTV", name)) {
+        return new OldTVHeuristic(outputFileName);
     }
         
     // invalid heuristic name!

### 2010-04-20 src/MSTCache.h

@@ -206,13 +206,41 @@ public:
         \param[out] alignmentData The alignment data associated with
         the srcESTidx and the destESTidx.
 
-	\param[out] directionData The direction data associated with
-	the srcESTidx and the destESTidx.
+		\param[out] directionData The direction data associated with
+		the srcESTidx and the destESTidx.
     */
     virtual void getBestEntry(int& srcESTidx, int& destESTidx,
                               float& metric, int& alignmentData,
-			      int& directionData) const = 0;
+							  int& directionData) const = 0;
 
+    /** Obtains the top-most similar entry from the MSTCache.
+
+        This method searches the similarity metrics stored in this
+        cache for various ESTs to locate the best entry with the
+        highest similarity.  It populates the parameters with the
+        appropriate value.  Note that the parameters are initialized
+        to -1, -1, -1.0f, and -1 respectively.
+
+        \param[out] srcESTidx The source EST index from where the
+        similarity metric is being measured.  The srcESTidx is already
+        present in the MST.
+
+        \param[out] destESTidx The destination EST index that is the
+        best choice to be added to the MST (based on the local
+        information).
+
+        \param[out] metric The similarity/distance metric between the
+        srcESTidx and the destESTidx.
+
+        \param[out] alignmentData The alignment data associated with
+        the srcESTidx and the destESTidx.
+    */
+    inline void getBestEntry(int& srcESTidx, int& destESTidx,
+                             float& metric, int& alignmentData) const {
+        int direction;
+        getBestEntry(srcESTidx, destESTidx, metric, alignmentData, direction);
+    }
+    
     /** Display cache usage statistics.
 
         This method can be used to print the current cache usage

### 2010-04-20 src/OldTVHeuristic.h

@@ -1,0 +1,300 @@
+#ifndef OLD_TV_HEURISTIC_H
+#define OLD_TV_HEURISTIC_H
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include "OldUVHeuristic.h"
+#include "EST.h"
+#include <utility>
+
+/** Heuristic based upon the T/V heuristic used in WCD, a type of
+    common word heuristic.
+
+    The idea of the <i>t/v</i>-heuristic is to require the common
+    words in a pair of ESTs to be found reasonably close to each other
+    but not too close. The rule of this heuristic is as follows:
+
+    <ol>
+
+    <li>Assume we are given two ESTs to analyze, say e<sub>i</sub> and
+    e<sub>j</sub> and a threshold \em t.</li>
+
+    <li>Consider all \em v words appearing in e<sub>i</sub>.</li>
+
+    <li>At least \em t of these \em v words must appear in \em j so
+    that they do not overlap (their starting positions must be at
+    least \em v base pairs different) and are at least 100 base pairs
+    of each other.</li>
+
+    <li>If there are at least \em t \em v words the huristic
[…]

### 2010-04-20 src/OldTwoPassD2.h

@@ -1,0 +1,554 @@
+#ifndef OLD_TWOPASS_D2_H
+#define OLD_TWOPASS_D2_H
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include "FWAnalyzer.h"
+#include <string>
+#include <vector>
+
+// Forward declaration to keep compiler happy
+class EST;
+class ResultLog;
+
+/** \brief EST Analyzer that uses the D2 algorithm in both its
+	symmetric and asymmetric variations to compute distance between
+	two ESTs.
+
+    <p>This analyzer provides the mechanism to use D2 algorithm to
+    compute the distance values between a pair of ESTs. For improved
+	performance, the analyzer uses D2 asymmetric and symmetric together.
+	It runs a fast "first pass" using D2 asymmetric and finds the minimum
+	D2 score, as well as the two windows between which the score was
+	computed.  The analyzer then computes bounds and runs a bounded
+	version of D2 symmetric.  Ideally this analyzer will provide
+	faster runtime because of the asymmetric pass, and not sacrifice any
+	precision because of the bounded symmetric pass.<p>
+
+    \note This D2 analyzer uses a word size of 8 base pairs.  This is
[…]

### 2010-04-20 src/OldUVHeuristic.cpp

@@ -1,0 +1,280 @@
+#ifndef OLD_UV_HEURISTIC_CPP
+#define OLD_UV_HEURISTIC_CPP
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include "OldUVHeuristic.h"
+#include "HeuristicChain.h"
+#include "ESTCodec.h"
+#include "EST.h"
+
+#include <cstdio>
+
+// default params
+int OldUVHeuristic::u           = 6;
+int OldUVHeuristic::v           = 8;
+int OldUVHeuristic::wordShift   = 8;
+int OldUVHeuristic::BitMask     = 0;
+int OldUVHeuristic::passes      = 3;
+
+// Instance variable to track number of bits to shift for building
+// hash values.  This is set to 2*(v-1) in initialize method.
+int OldUVHeuristic::bitsToShift = 0;
+
+// The set of arguments for this class.
+arg_parser::arg_record OldUVHeuristic::argsList[] = {
+    {"--uv_u", "u (number of v-word matches) (default=4)",
[…]

### 2010-04-20 src/OldUVHeuristic.h

@@ -1,0 +1,311 @@
+#ifndef OLD_UV_HEURISTIC_H
+#define OLD_UV_HEURISTIC_H
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include "arg_parser.h"
+#include "Heuristic.h"
+#include "HeuristicFactory.h"
+#include "Utilities.h"
+#include "HashMap.h"
+
+#include <string>
+#include <vector>
+
+/** \typedef HashMap<int, std::vector<short> > UVHashTable;
+
+    \brief A hash map to provide rapid access to the set of
+    pre-computed hash values for a given EST.
+
+    This typedef provides a convenient short cut to refer to a hash
+    map that is used to cache pre-computed hash values for a given
+    EST.  The zero-based index of the EST is used as the key into this
+    hash map.  Each entry in the hash map contains a std::vector that
+    essentially contains the array of hash values.
+
+    \note Refer to the additional documentation with the instance
+    variable uvCache for motivation for this hash map.
+*/
+typedef HashMap<int, std::vector<unsigned short> > UVHashTable;
+
+/** Heuristic based upon the "u/v sample heuristic" used in WCD,
[…]

### 2010-04-20 src/SFFReader.cpp

@@ -143,7 +143,7 @@ SFFReader::loadSFFHeader() {
     }
 
 EST*
-SFFReader::getNextRead(const bool maskBases) {
+SFFReader::getNextRead(const bool maskBases, const bool randomizeNbases) {
     if (!isValid() || (pendingReads <= 0)) {
         // Can't read more data from the SFF file.
         return NULL;
@@ -217,7 +217,7 @@ SFFReader::getNextRead(const bool maskBases) {
     sequence[rightIndex] = '\0';     // Logically terminate C-string
     // Convert the information to a valid EST entry.
     return EST::create(EST::getESTCount(), &name[0], &sequence[leftIndex],
-                       offset, maskBases);
+                       offset, maskBases, randomizeNbases);
 }
 
 #endif

### 2010-04-20 src/SFFReader.h

@@ -142,7 +142,12 @@ public:
 		\param[in] maskBases If this flag is true, then all lowercase
 		bases are converted to 'N' rather than uppercase characters,
 		causing them to be ignored by downstream processing.
-		
+
+		\param[in] randomizeNbase If this value is \c false the \c 'N'
+        characters are \b preserved.  However, if the flag is set to
+        \c true then the 'N' characters are randomly converted to a
+        regular base character from the list \c ATCG.
+
         \return A newly created (on the heap) EST object containing
         the next read in the SFF file.  The number of pending reads is
         decremented after this method call.  On errors this method
@@ -150,7 +155,8 @@ public:
         invalidating the file (causing isValid() method to return \c
         false).
     */
-    EST* getNextRead(const bool maskBases = true);
+    EST* getNextRead(const bool maskBases = true,
+					 const bool randomizeNbases = false);
     
 protected:
     /** Helper method to read the SFF common header.

### 2010-04-20 peace/src/ClusterMakerFactory.cpp
Changes for ticket:60 :
I am done with incorporating the changes. For some reason I also had to 
move the earlier version of `MSTClusterMaker` class as well into the new 
code-path to achieve the earlier performance. I took a look at the differences 
but they seem minor; yet there was a notable performance difference that could
be bridged only with the earlier version of `MSTClusterMaker` that is now in 
the main branch under `OldMSTClusterMaker`. The command line to run all the 
earlier version of the code path is:

{{{

./peace --clusterMaker na-mst --analyzer oldTwopassD2 --heuristics oldTV

@@ -39,6 +39,7 @@
 #include "arg_parser.h"
 
 #include "MSTClusterMaker.h"
+#include "OldMSTClusterMaker.h"
 #include "TransMSTClusterMaker.h"
 
 void
@@ -47,6 +48,8 @@ ClusterMakerFactory::displayList(std::ostream &os) {
     // easier.
     arg_parser::arg_record dummy_args[] = {
         {"mst", "MST-based Cluster Maker",
+         NULL, arg_parser::STRING},
+        {"na-mst", "Non-adaptive MST-based Cluster Maker for Sanger Sequences",
          NULL, arg_parser::STRING},
         {"tmst", "MST-based Cluster Maker with Transitivity",
          NULL, arg_parser::STRING},
@@ -78,6 +81,8 @@ ClusterMakerFactory::create(const char* name, ESTAnalyzer *analyzer,
         return new MSTClusterMaker(analyzer, refESTidx, outputFileName);
     } else if (!strcmp("tmst", name)) {
         return new TransMSTClusterMaker(analyzer, refESTidx, outputFileName);
+    } else if (!strcmp("na-mst", name)) {
+        return new OldMSTClusterMaker(analyzer, refESTidx, outputFileName);
     }
     
     // invalid analyzer name!

### 2010-04-20 peace/src/EST.h

@@ -118,7 +118,12 @@ public:
 		\param[in] maskBases If this flag is true, then all lowercase
 		bases are converted to 'N' rather than uppercase characters,
 		causing them to be ignored by downstream processing.
-		
+
+        \param[in] randomizeNbase If this value is \c false the \c 'N'
+        characters are \b preserved.  However, if the flag is set to
+        \c true then the 'N' characters are randomly converted to
+        a regular base character from the list \c ATCG.
+        
         \return If the id is valid and a duplicate EST with the same
         ID is not present, then this method creates a new EST and
         returns a pointer to that EST back to the caller.
@@ -126,7 +131,8 @@ public:
     static EST* create(const int id, const char *info,
                        const char* sequence = NULL,
                        const long offset    = -1,
-					   const bool maskBases = true);
+					   const bool maskBases = true,
+                       const bool randomizeNbases = false);
 
     /** Loads data from a FASTA file to create an EST.
 
@@ -147,13 +153,19 @@ public:
 		\param[in] maskBases If this flag is true, then all lowercase
 		bases are converted to 'N' rather than uppercase characters,
 		causing them to be ignored by downstream processing.
-		
+
+        \param[in] randomizeNbase If this value is \c false the \c 'N'
+        characters are \b preserved.  However, if the flag is set to
+        \c true then the 'N' characters are randomly converted to a
+        regular base character from the list \c ATCG.
+        
         \note At the end of this method the fastaFile's file pointer
         will point at the beginning of the next EST (if any) in the
         file.
     */
     static EST* create(FILE* fastaFile, int& lineNum,
-                       const bool maskBases = true);
+                       const bool maskBases = true,
+                       const bool randomizeNbases = false);
 	
     /** Obtain the list of ESTs.
 
@@ -166,17 +178,17 @@ public:
 
     /** Obtain count of ESTs that have been flagged as being processed.
 
-	This method can be used to determine the number of ESTs that
-	have been flagged as being processed. Subtracting this number
-	from the total number of ESTs indicates the number of ESTs to
-	be processed.
-	
-	\note This method iterates over the list of ESTs to determine
-	the current number of processed ESTs. So use this method
-	sparingly.
+        This method can be used to determine the number of ESTs that
+        have been flagged as being processed. Subtracting this number
+        from the total number of ESTs indicates the number of ESTs to
+        be processed.
+        
+        \note This method iterates over the list of ESTs to determine
[…]

### 2010-04-20 peace/src/ESTAnalyzerFactory.cpp

@@ -43,6 +43,7 @@
 #include "D2Zim.h"
 #include "AdaptiveTwoPassD2.h"
 #include "TwoPassD2.h"
+#include "OldTwoPassD2.h"
 #include "MatrixFileAnalyzer.h"
 
 void
@@ -60,9 +61,9 @@ ESTAnalyzerFactory::displayList(std::ostream &os) {
          NULL, arg_parser::STRING},
         {"d2zim", "Use D2 (Zimmerman) distance metric generation algorithm",
          NULL, arg_parser::STRING},
-	{"twopassD2", "Use two-pass asymmetric/bounded symmetric D2",
-	 NULL, arg_parser::STRING},
-	{"twopassD2adapt", "Use two-pass asymmetric/bounded symmetric D2, adaptive mode",
+        {"twopassD2", "Use two-pass asymmetric/bounded symmetric D2",
+         NULL, arg_parser::STRING},
+        {"twopassD2adapt", "Use two-pass asymmetric/bounded symmetric D2, adaptive mode",
          NULL, arg_parser::STRING},
         //        {"d2sim", "Use special version of D2 for simulation project",
         //         NULL, arg_parser::STRING},
@@ -96,8 +97,10 @@ ESTAnalyzerFactory::create(const char* name, const int refESTidx,
         return new D2Zim(refESTidx, outputFileName);        
     } else if (!strcmp("twopassD2", name)) {
         return new TwoPassD2(refESTidx, outputFileName);
+    } else if (!strcmp("oldTwopassD2", name)) {
+        return new OldTwoPassD2(refESTidx, outputFileName);
     } else if (!strcmp("twopassD2adapt", name)) {
-	return new AdaptiveTwoPassD2(refESTidx, outputFileName);
+        return new AdaptiveTwoPassD2(refESTidx, outputFileName);
         //    } else if (!strcmp("d2sim", name)) {
         //        return new SimulationD2(refESTidx, outputFileName);
     }

### 2010-04-20 peace/src/OldMSTClusterMaker.cpp

@@ -1,0 +1,769 @@
+#ifndef OLD_MST_CLUSTER_MAKER_CPP
+#define OLD_MST_CLUSTER_MAKER_CPP
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include "OldMSTClusterMaker.h"
+#include "ESTAnalyzer.h"
+#include "MSTMultiListCache.h"
+#include "MSTHeapCache.h"
+#include "EST.h"
+#include "MPIStats.h"
+#include "HeuristicChain.h"
+#include "Heuristic.h"
+#include <fstream>
+#include <sstream>
+
+// A define to remove magic 0 (zero) in code
+#define MANAGER_RANK 0
+
+// Another define to remove magic 0 (zero) success checks.
+#define NO_ERROR 0
+
+extern char DefCacheType[10];
+
+// The static instance variables for command line arguments.
+int    OldMSTClusterMaker::cacheSize     = 128;
+bool   OldMSTClusterMaker::noCacheRepop  = true;
+bool   OldMSTClusterMaker::strictOrder   = false;
[…]

### 2010-04-20 peace/src/OldMSTClusterMaker.h

@@ -1,5 +1,5 @@
-#ifndef MST_CLUSTER_MAKER_H
-#define MST_CLUSTER_MAKER_H
+#ifndef OLD_MST_CLUSTER_MAKER_H
+#define OLD_MST_CLUSTER_MAKER_H
 
 //--------------------------------------------------------------------
 //
@@ -36,9 +36,8 @@
 
 #include "ClusterMaker.h"
 #include "MSTCache.h"
-#include "MSTCluster.h"
 #include "MST.h"
-
+#include "MSTCluster.h"
 #include <fstream>
 
 /** A Minimum Spanning Tree (MST) based parallel cluster maker.
@@ -54,7 +53,7 @@
     documentation on the various method for detailed description on
     their functionality and usage.
 */
-class MSTClusterMaker : public ClusterMaker {
+class OldMSTClusterMaker : public ClusterMaker {
     friend class ClusterMakerFactory;
 public:
     /** The set of tags exchanged between various processes.
@@ -65,16 +64,16 @@ public:
         parallel/distributed manner.
     */
     enum MessageTags{REPOPULATE_REQUEST, COMPUTE_SIMILARITY_REQUEST,
-                     SIMILARITY_LIST, SIMILARITY_COMPUTATION_DONE,
-                     COMPUTE_MAX_SIMILARITY_REQUEST, MAX_SIMILARITY_RESPONSE,
-                     ADD_EST, TRANSITIVITY_LIST, COMPUTE_TOTAL_ANALYSIS_COUNT};
+		     SIMILARITY_LIST, SIMILARITY_COMPUTATION_DONE,
+		     COMPUTE_MAX_SIMILARITY_REQUEST, MAX_SIMILARITY_RESPONSE,
+		     ADD_EST, TRANSITIVITY_LIST, COMPUTE_TOTAL_ANALYSIS_COUNT};
     
     /** The destructor.
         
         The destructor frees up all any dynamic memory allocated by
         this object for its operations.
     */
-    virtual ~MSTClusterMaker();
+    virtual ~OldMSTClusterMaker();
 
     /** Display valid command line arguments for this cluster maker.
 
@@ -133,7 +132,7 @@ public:
     */
     virtual void displayStats(std::ostream& os);
 
-    /** Add a dummy cluster to the cluster maker.
+	    /** Add a dummy cluster to the cluster maker.
         
         This method can be used to add a dummy cluster to the cluster
         maker. The dummy clusters are added as direct descendants of
[…]

### 2010-04-20 peace/src/OldTVHeuristic.cpp

@@ -1,0 +1,158 @@
+#ifndef OLD_TV_HEURISTIC_CPP
+#define OLD_TV_HEURISTIC_CPP
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include "OldTVHeuristic.h"
+#include "ESTCodec.h"
+#include "EST.h"
+
+// Define the static parameters
+int OldTVHeuristic::t = 65;
+
+int OldTVHeuristic::windowLen = 100;
+
+// The set of arguments for this class.  Note that some of the base
+// class static instance variables are reused here so that the values
+// are consistently set.
+arg_parser::arg_record OldTVHeuristic::argsList[] = {
+    {"--tv_t", "t (number of v-word matches) (default=65)",
+     &OldTVHeuristic::t, arg_parser::INTEGER},
+    {"--tv_win", "Window size for t/v heuristics (default=100)",
+     &OldTVHeuristic::windowLen, arg_parser::INTEGER},    
[…]

### 2010-04-20 peace/src/OldTwoPassD2.cpp

@@ -1,0 +1,422 @@
+#ifndef OLD_TWOPASS_D2_CPP
+#define OLD_TWOPASS_D2_CPP
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include "OldTwoPassD2.h"
+#include "EST.h"
+#include "ESTCodec.h"
+#include "ParameterSetManager.h"
+#include <algorithm>
+
+// Skip parameter for d2 asymmetric.
+int OldTwoPassD2::frameShift = 50;
+
+// The bitmak to be used when build hash values.
+int OldTwoPassD2::BitMask   = 0;
+
+// Instance variable to store the number of bits to be shifted to
+// create hash values. This value is initialized to 2*(wordSize-1)
+int OldTwoPassD2::bitShift  = 0;
+
+// The threshold score below which two ESTs are considered
+// sufficiently similar to be clustered.
+int OldTwoPassD2::threshold = 0;
+
+// The threshold score above which two ESTs are considered
[…]

### 2010-04-20 peace/papers/NAR_web_server_2009/paper.sup.tex
Changed to manuscript

@@
 \end{tabular}
-\label{SangerReal130}
 \end{center}
@@
   Dataset and the WCD A076941 Arabidopsis thaliana dataset using the
-  standard quality measurments with $T=130$.}
+  standard quality measurments with $T=130$.}\label{SangerReal130}
+
 \end{table}

### 2010-04-20 papers/NAR_web_server_2009/paper.tex
Changed to manuscript

@@
 \newcommand{\easycluster} {{\small EasyCluster}}
+\newcommand{\metasim} {{\small MetaSim}}
 
@@
 \begin{abstract}
-  We present \peace, a stand alone tool for the high-throughput {\it
-    ab initio} clustering of transcript fragment sequences produced by
-  Next Generation or Sanger Sequencing technologies, freely available
-  from \href{http://www.peace-tools.org}{www.peace-tools.org}.
-  Installed and run through an easily downloaded GUI, \peace\/ can
+  \mc{We present \peace, a stand alone tool for the high-throughput
+    {\it ab initio} clustering of transcript fragment sequences
+    produced by Next Generation or Sanger Sequencing technologies,
+    freely available from
+    \href{http://www.peace-tools.org}{www.peace-tools.org}.  Installed
+    and managed through a download-able user-friendly GUI, \peace\/ can
     process large data sets of transcript fragments of length 50 bases
-  or greater, grouping the fragments by gene association and achieving
-  a greater sensitivity (without sacrificing runtime) than the
-  competing clustering tools.  Once clustered, the user can employ the
-  GUI to collect statistics and single out specific clusters for more
-  comprehensive study.  \peace\/, when used as a ``pre-assembly''
-  tool, enables any assembly tool to work with individual small
-  clusters of closely related sequences.  For a 5\% sequencing error
-  rate, we see a 52\% improvement in sensitivity over the leading
-  cluster tool, \wcd\/, and 184\% improvement in sensitivity over the
-  assembly tool \capthree.  When applied to a 361 Mb set of
-  Arabidopsis transcriptome fragments, we see a 40\% speedup over
-  \wcd.
+    or greater, grouping the fragments by gene associations with a
+    sensitivity compatible to leading clustering tools.  Once
+    clustered, the user can employ the GUI's analysis functions,
+    allowing the easy collection of statistics and single out specific
+    clusters for more comprehensive study or assembly.  Using a novel
+    minimum spanning tree based clustering method, \peace is the equal of
+    other leading tools in the literature.  It produces results of
+    quality virtually identical to those of the \wcd\/ tool when
+    applied to Sanger sequences, significantly improved results over
+    \wcd when applied the products Next Generate Sequencing
+    Technology, and significantly improved results over \capthree in
+    both cases.  In short, \peace provides an easily accessible user interface
+    for interface with a clustering engine that is the match of the
+    leading clustering tools in the literature.}
+
 \end{abstract}
@@
 information, it also provides immensely larger data sets -- and thus a
-need for faster, more efficient analysis algorithms.
+need for fast, efficient analysis algorithms.
 
@@
 runtime.  Nor are any of these tools designed for the ease of installation
-and use that \peace provides.
+and use that \peace\/ provides.
 
@@
 \begin{figure}
-  \centerline{\includegraphics[width=3in]{pics.d/workflow.eps}}
+  \centerline{\includegraphics[width=3in]{pics.d/workflow.png}}
   \caption{Overview of the procedure for clustering and analysis using
@@
   \begin{minipage}{3in}
-    \includegraphics[width=3in]{install_page_big.eps}
+    \includegraphics[width=3in]{screen.d/install_page_big.png}
     \centerline{\small{(a)}}
@@
     \vspace{0.47in}
-    \includegraphics[width=2in]{add_server_big.eps}
+    \includegraphics[width=2in]{screen.d/add_server_big.png}
     \centerline{\small{(b)}}
@@
     \vspace{0.47in}
-    \includegraphics[width=2in]{processor_assignment_big.eps}
+    \includegraphics[width=2in]{screen.d/processor_assignment_big.png}
     \centerline{\small{(c)}}
@@
   \begin{minipage}{3in}
-    \includegraphics[width=3in]{cluster_list_big.eps}
+    \includegraphics[width=3in]{screen.d/cluster_list_big.png}
     \centerline{\small{(d)}}
@@
   \begin{minipage}{3in}
-    \includegraphics[width=3in]{classifier_big.eps}
+    \includegraphics[width=3in]{screen.d/classifier_big.png}
     \centerline{\small{(e)}}
@@
 Sanger Sequencing technologies, comparing results against those
-produced by the \wcd\/ clustering tool \mc{(version
-  0.5.1)}\cite{Hazelhurst08a} and the \capthree\/ assembly tool
-\cite{Huang99} (the latter of which implicitly calculates a clustering
-in the process of assembly).  For our simulation tests we used the
-{\bf ESTSim} tool \cite{Hazelhurst03} to generate simulated transcript
-fragments of varying length under different models of error
-(Supplementary Materials, Section~D.1), generating the fragments from
-the list of 100 zebra fish genes used in the \wcd\/ testing
-\cite{Hazelhurst08a}.  Tool parameters were taken to match, as closely
-as possible, those used in the \wcd\/ study (see Supplementary
-Materials).  \mc{The most important method of quality assessment is
-  {\it sensitivity} (the fraction of fragment pairs from the same gene
-  that were correctly clustered together).  But we also look at the
-  {\it Jaccard Index} (which balances sensitivity with the number of
-  false positives by looking at the ratio of true positives to the sum
-  of true positives, false positives, and true negatives), {\it Type 1
-    error} (the fraction of genes that were divided between clusters),
-  and {\it Type 2 error} (the fraction of clusters containing two or
-  more genes).  Specificity was not an issue in simulated tests, as
-  both \peace\/ and \wcd\/ were completely successful in separating
-  unrelated ESTs (save when dealing with recently duplicated genes).
-  Other measurements are also discussed in the supplementary
-  materials.  In Figure~\ref{SeJiT1T2} we plot these four tests as a
-  function of error rate, and observe the almost identical results
-  between \peace\/ and \wcd\/.  In Figure~\ref{seq_parallel} we plot
-  the runtime for \peace\/ and \wcd\/, again observing almost identical
-  results when run sequentially -- but significantly faster runtime for
-  PEACE on multiple processor when holding the EST/processor ratio
-  constant (ranging from a 65\% improvement for two processors too a
-  17\% improvement for 12 processors).}
+produced by the \wcd\/ clustering tool \cite{Hazelhurst08a} and the
+\capthree\/ assembly tool \cite{Huang99} (the latter of which
+implicitly calculates a clustering in the process of assembly).  For
+our simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03}
+to generate simulated transcript fragments of varying length under
+different models of error (Supplementary Materials, Section~D.1),
+generating the fragments from the list of 100 zebra fish genes used in
+the \wcd\/ testing \cite{Hazelhurst08a}.  Tool parameters were taken
+to match, as closely as possible, those used in the \wcd\/ study (see
+Supplementary Materials).  \mc{The most important method of quality
+  assessment is {\it sensitivity} (the fraction of fragment pairs from
+  the same gene that were} \mc{correctly clustered together).  But we also
+  look at the {\it Jaccard Index} (which balances sensitivity with the
+  number of false positives by looking at the ratio of true positives
+  to the sum of true positives, false positives, and true negatives),
+  {\it Type 1 error} (the fraction of genes that were divided between
+  clusters), and {\it Type 2 error} (the fraction of clusters
+  containing two or more genes).  Specificity was not an issue in
+  simulated tests, as both \peace\/ and \wcd\/ were completely
+  successful in separating unrelated ESTs (save when dealing with
+  recently duplicated genes).  In Figure~\ref{SeJiT1T2} we plot
+  these four tests as a function of error rate, and observe the almost
+  identical results between \peace\/ and \wcd\/.  In
+  Figure~\ref{seq_parallel} we plot the runtime for \peace\/ and
+  \wcd\/, again observing almost identical results when run
+  sequentially -- but significantly faster runtime for PEACE on
+  multiple processor when holding the EST/processor ratio constant
+  (ranging from a 65\% improvement for two processors too a 17\%
+  improvement for 12 processors).}
+
+\mc{We further note, as a point of interest that, we can significantly
+  improve the quality of \peace simulation results through the
+  increase of the threshold value (see Supplementary Materials,
+  Section C) -- achieving a significant improved in PEACE sensitivity
+  without an adverse effect on the Jaccard index (i.e. a significant
+  increase in the incorrect merging of clusters).  However, the
+  improvement does not carry over to the application of real data,
+  where we observe a significant drop in the Jaccard Index.  Due to
+  this drop we have opted to leave the lower threshold value as the
+  default, but users might consider the higher value, thus decreasing
+  the risk of incorrectly splitting clusters, if planning on
+  using on the subsequent use of a assembly tool that capable of
+  breaking up the incorrectly merged clusters during transcript reconstruction.}
 
@@
   \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
-    Type 1 error, based on the
-    average over 30 simulated Sanger Sequence ESTs sets derived from 100 zebra
-    fish genes  (see
+    Type 1 error, based on the average over 30 simulated Sanger
+    Sequence ESTs sets derived from 100 zebra fish genes (see
     Supplementary Materials, Section C, for more details).  Blue/Solid
-    = \peace, Green/Dash = \wcd, Black/Dot-Dash = \capthree; vertical
-    tics = 95\% confidence intervals on estimates.  Intervals are not
-    presented for Type 1 error due to the effective lack of
-    variance.}\label{SeJiT1T2}
+    = \peace, Green/Dash = \wcd \mc{(version 0.5.1)}, Black/Dot-Dash =
+    \capthree; vertical tics = 95\% confidence intervals on estimates.
+    Intervals are not presented for Type 1 error due to the effective
+    lack of variance.}\label{SeJiT1T2}
 \end{figure}
@@
 the A076941.fa {\it Arabidopsis thaliana} data set used to test WCD
-\cite{Hazelhurst08a,Hazelhurst08b}, with results in
-Table~\ref{SangerReal}.  We notice essentially identical
+\cite{Hazelhurst08a,Hazelhurst08b}, (see Table~\ref{SangerReal}).  We notice essentially identical
 results for PEACE and WCD in quality, both significantly better than
@@
 Jaccard Index and Type 2 error.  In runtime we see some inconsistency,
-with PEACE showing a 60\% runtime improvement in the smaller data set,
-requiring 20\% more time in the Arabidopsis data set.  In Section~D.1
+with PEACE showing a 60\% runtime improvement in the first data set,
+but requiring 20\% more time in the Arabidopsis data set.  In Section~D.1
 of the Supplementary Materials we present runtimes several more
@@
 \end{tabular}
-\label{SangerReal}
 \end{center}
@@
   Data Set and the WCD A076941 Arabidopsis thaliana data set using the
-  standard quality measurements.}
+  standard quality measurements.}\label{SangerReal}
 \end{table*}
@@
 
-\mc{We tested the three tools on short-read data using the MetaSim
+\mc{We tested the three tools on short-read data using the \metasim\/
   tool of {\it Richter et al.} \cite{Richter2008}.  Encoded into
-  MetaSim are sequence generation and error models corresponding to
+  \metasim\/ are sequence generation and error models corresponding to
   several technologies, including the 454 short-read technology.
-  Leaving \capthree\/ and \metasim\/ at their default values, but
+  Leaving \capthree\/ and \wcd\/ at their default values, but
   employing the {\it adaptive $d^2$} strategy for short read sequences
@@
   \peace\/ performs significantly better than the other two tools on
-  454 data using the default MetaSim model (with sequence sizes
+  454 data using the default \metasim\/ model (with sequence sizes
   ranging in size from 200bp to 370bp, at an average size of 250bp and
@@
   \capthree.  For the Jaccard Index we get essentially the same
-  numbers.  For 100 transcript \peace creates 128 clusters (as opposed
-  to 6,370 and 19245 clusters for \wcd and \capthree), and creates
-  only 3 singletons (as opposed to 5,850 and 15432).  We find it
-  likely that the \wcd results could be considerably improved by the
-  simple adjustment of a few parameters, but were unable to determine
-  how to do that.  (Runtime of \peace\/ was somewhat slower than that
-  of \wcd, but given the significant improvement in quality results we
-  do not find this to be very troubling.)}
+  numbers.  For 100 transcript \peace\/ creates 128 clusters (as
+  opposed to 6,370 and 19245 clusters for \wcd\/ and \capthree), and
+  creates only 3 singletons (as opposed to 5,850 and 15432).  We do
+  note that it is likely that the \wcd results could be considerably
+  improved by the adjustment of a few parameters, but were unable to
+  determine how to do that.  (Runtime of \peace\/ was somewhat slower
+  than that of \wcd, but given the significant improvement in quality
+  of results this is unimportant.)}
 
@@
 
-Here we have presented \peace, a stand alone tool for the
+\mc{Here we have presented \peace, a stand alone tool for the
 high-throughput clustering of transcript fragments capable of dealing
@@
 association, \peace\/ shows significant improvement in sensitivity,
-without sacrificing runtime, over the competing \wcd\/ tool
-\cite{Hazelhurst08a}, and shows an order of magnitude in improvement
-over the clustering performed in the course of assembly by the
-\capthree\/ tool \cite{Huang99}.
+over the competing \wcd\/ tool \cite{Hazelhurst08a} when applied to
+NGS reads, matches \wcd\/ when applied to Traditional Sanger
+Sequencing output, and shows an order of magnitude in improvement over
+the clustering performed in the course of assembly by the \capthree\/
+tool \cite{Huang99}.}
 
@@
 % LocalWords:  ssembly xpression GenBank dbEST pyrosequencing ESTsim sim mRNAs
-% LocalWords:  NGS polyA set's GHz Xeon Bikun Arabidopsis JVM Jaccard LV
+% LocalWords:  NGS polyA set's GHz Xeon Bikun Arabidopsis JVM Jaccard LV GUI's
 % LocalWords:  Data Set thaliana data set EasyCluster MetaSim

### 2010-04-20 papers/NAR_web_server_2009/paper.tex
Updated the abstract with a few edits to fix spacing and a couple of spots of changes. I was unable to regenerate the PDF as some of the images/graphs were reported to be missing.

@@
 \begin{abstract}
-  \mc{We present \peace, a stand alone tool for the high-throughput
-    {\it ab initio} clustering of transcript fragment sequences
-    produced by Next Generation or Sanger Sequencing technologies,
+  \mc{We present \peace, a stand alone tool for high-throughput {\it
+      ab initio} clustering of transcript fragment sequences produced
+    by Next Generation or Sanger Sequencing technologies.  It is
     freely available from
     \href{http://www.peace-tools.org}{www.peace-tools.org}.  Installed
-    and managed through a download-able user-friendly GUI, \peace\/ can
-    process large data sets of transcript fragments of length 50 bases
-    or greater, grouping the fragments by gene associations with a
-    sensitivity compatible to leading clustering tools.  Once
+    and managed through a download-able user-friendly GUI, \peace\/
+    can process large data sets of transcript fragments of length 50
+    bases or greater, grouping the fragments by gene associations with
+    a sensitivity compatible to leading clustering tools.  Once
     clustered, the user can employ the GUI's analysis functions,
-    allowing the easy collection of statistics and single out specific
+    facilitating easy collection of statistics and single out specific
     clusters for more comprehensive study or assembly.  Using a novel
-    minimum spanning tree based clustering method, \peace is the equal of
-    other leading tools in the literature.  It produces results of
+    minimum spanning tree based clustering method, \peace is the equal
+    of other leading tools in the literature.  It produces results of
     quality virtually identical to those of the \wcd\/ tool when
     applied to Sanger sequences, significantly improved results over
-    \wcd when applied the products Next Generate Sequencing
-    Technology, and significantly improved results over \capthree in
-    both cases.  In short, \peace provides an easily accessible user interface
-    for interface with a clustering engine that is the match of the
-    leading clustering tools in the literature.}
+    \wcd\/ when applied the products of Next Generation Sequencing
+    Technology, and significantly improved results over \capthree\/ in
+    both cases.  In short, \peace\/ provides an intuitive graphical
+    user interface and a feature-rich, parallel clustering engine that
+    collectively are as good, if not better, than the leading
+    clustering tools in the literature.}
 

### 2010-04-20 peace/papers/NAR_web_server_2009/paper.sup.tex
Modifications the the supp. materials; addition of needed figures

@@
 \usepackage{natbibOrig}
+
+\usepackage[pdftex,
+colorlinks=true, %makes the links show up in color, rather than in a box
+citecolor=black, %color of the in-text citation numbers
+urlcolor=blue %color of the url links
+]{hyperref}
+\usepackage{wrapfig}
+\bibliographystyle{unsrt}
 
@@
 \usepackage{graphicx}
-\bibliographystyle{natbib}
+%\bibliographystyle{natbib}
 
@@
 \usepackage{color}
-\definecolor{red}{rgb}{0,0,1}
+\definecolor{red}{rgb}{1,0,0}
 \newcommand{\mc}[1]{\textcolor{red}{#1}}
@@
 \newlabel{sensitivityFig}{3}
-\newlabel{SangerTable}{1}
+\newlabel{seq_runtime}{4}
+\newlabel{SangerReal}{1}
 
@@
 
-\subsection{The $d^2$ distance metric}
-$d^2$, as outlined by {\it Hide et al.} [\cite{Hide94}], is an {\it
+\subsection{The distance metric}
+$d^2$, as outlined by {\it Hide et al.} \cite{Hide94}, is an {\it
   alignment free} distance pseudo-metric which quantifies local
@@
 The \textsc{peace} implementation of $d^2$ was initially based on the description from
-{\it Hazelhurst} [\cite{Hazelhurst04}].  For parameters
-we adapted those used by the \textsc{wcd} clustering tool [\cite{Hazelhurst08a}]:
+{\it Hazelhurst} \cite{Hazelhurst04}.  For parameters
+we adapted those used by the \textsc{wcd} clustering tool \cite{Hazelhurst08a}:
 a word size $k=6$ and a window size $r=100$.
 
-\subsection{Two-pass $d^2$}
+\subsection{Improved Distance Matrix}
 Our ``two-pass $d^2$'' algorithm works by sampling a subset of window
@@
 \textsc{peace} uses the $u/v$ and $t/v$ heuristics, roughly as described by {\it
-  Hazelhurst et al.} [\cite{Hazelhurst08a}] as filters that allows us to avoid
+  Hazelhurst et al.} \cite{Hazelhurst08a} as filters that allows us to avoid
 over 99\% of the potential $d^2$ calculations.  Each heuristic takes two
@@
 knowledge this is the first such use for the  EST
-clustering problem [\cite{Jain99,Wan08}].
+clustering problem \cite{Jain99,Wan08}.
 
@@
 
-To calculate this MST we use Prim's algorithm [\cite{Prim57}].  For a
+To calculate this MST we use Prim's algorithm \cite{Prim57}.  For a
 graph of $n$ nodes and $e$ edges, Prim's can be implemented such that
@@
 practice that $e$ is a very small fraction of $n^2$.  In the
-runtime results (see Section~\ref{rt_section} and
-Figure~\ref{seq_runtime}), we find that when holding the EST size
+runtime results (Figure~4), we find that when holding the EST size
 distribution constant, the tool as a whole has a runtime of $O(n^2)$
@@
 
-\subsection{Adaptive $d^2$}
+\section{Adaptive $d^2$}
 
-In the above discussion of $d^2$ we assume sequences of at least 100
-bp, and in pratice use a window of size 100.  While the technique
+\mc{In the above discussion of $d^2$ we assume sequences of at least 100
+bp, and in practice use a window of size 100.  While the technique
 should be, in principle, adaptable to short-read sequences (decreasing
@@
 Diminishing window sizes implies both diminishing accuracy (requiring
-the thrshold be adjusted) and increasing runtime (as there are
-potenitally more window pairs).  More importantly, there is the
-problem of non-homogonous sequence lengths: short-read technologies
-tend to produce short reads, but some technologiges can still include
+the threshold be adjusted) and increasing runtime (as there are
+potentially more window pairs).  More importantly, there is the
+problem of non-homogeneous sequence lengths: short-read technologies
+tend to produce short reads, but some technologies can still include
 long reads in their output.  Hence it is not immediately clear how to
-compare a very short read against a very long one.
+compare a very short read against a very long one.}
 
-We address these issues with our {\it adaptive $d^2$} strategy that
+\mc{We address these issues with our {\it adaptive $d^2$} strategy that
 works as follows: we partition sequences by length into disjoint groups.  For each
-group we pick a constant windowsize and threshold tailored to the
-length range of the group (as well as filtering heurstic constants),
+group we pick a constant window size and threshold tailored to the
+length range of the group (as well as filtering heuristic constants),
 and use those values when computing the $d^2$ distance for two
 sequences in the same group.  We do not compute distances between
-sequences in non-adjaecent groups (i.e. a member of the ``shortest
+sequences in non-adjacent groups (i.e. a member of the ``shortest
 groups'' against a member of the ``longest groups'' if there is an
-interviening sized group).  Once computed, we assign to the MST
+intervening sized group).  Once computed, we assign to the MST
 edge the ratio of the $d^2$ score to the group threshold, thus
 normalizing the weight for comparisons ``between groups'' and
-effectivally giving us an edge-removal threhold of 1.
+effectively giving us an edge-removal threshold of 1.}
 
-By restricting $d^2$ comparisons to within a group, we minimize the
-effets of the hetrogenous sequences lengths by restricting our
-comparisons to only sequences of similar lengths.  By not restrcicting
+\mc{By restricting $d^2$ comparisons to within a group, we minimize the
+effects of the heterogeneous sequences lengths by restricting our
+comparisons to only sequences of similar lengths.  By not restricting
 ourselves to a single window size, we allow ourselves to, when possible, avoid the
-preformance hit of using a small window.  As we are not comparing
+performance hit of using a small window.  As we are not comparing
 sequences from different groups we bypass the question of how to
-compare a largelong sequence against a short one.  And because the
+compare a long sequence against a short one.  And because the
 groups overlap in content, we still get implicit comparisons between
 long and short sequences in terms of the resulting path of the minimum
-spanning tree that makes use of these ``linking'' nodes.
+spanning tree that makes use of these ``linking'' nodes.}
 
-In the \textsc{peace} implementation, we subdevide into three groups:
+\mc{In the \textsc{peace} implementation, we subdivide into three groups:
 \begin{itemize}
@@
 \item Greater than 400 ($w=100$, $T=80$, $t=50$, $t=50$).
-\end{itemize}
+\end{itemize}}
 
@@
 
-In {\it Hazelhurst et al.} [\cite{Hazelhurst08a}] the authors conduct
+In {\it Hazelhurst et al.} \cite{Hazelhurst08a} the authors conduct
 an investigation of \textsc{wcd} \mc{(version 0.5.1)} against a number of clustering
@@
 analysis to a comparison of \textsc{peace} against \textsc{wcd} and the \textsc{cap3} assembly
-tool [\cite{Huang99}] -- a tool which implicitly clusters while
+tool \cite{Huang99} -- a tool which implicitly clusters while
 performing assembly.  We do so by applying all three tools to a number
@@
 For simulated data sets we use the {\bf ESTsim} tool to generate
-simulated EST data sets [\cite{Hazelhurst03}], using the collection of
+simulated EST data sets \cite{Hazelhurst03}, using the collection of
 zebra fish genes that served as a basis for the \textsc{wcd} simulations
-[\cite{Hazelhurst08a}].  In generating the ESTs, ESTsim models three
+\cite{Hazelhurst08a}.  In generating the ESTs, ESTsim models three
 types of error (general base read errors, errors due to
@@
 from the application of the ESTsim tool to a set of 100 zebra fish
-gene sequences [\cite{Hazelhurst03}].  All confidence intervals are
+gene sequences \cite{Hazelhurst03}.  All confidence intervals are
 calculated at a 95\% level of significance.
@@
 
-\mc{Our primary measurement tools if mesurement are sensitivity, the
+\mc{Our primary measurement tools if measurement are sensitivity, the
 Jaccard Index, Type 1 error and Type 2 error
-(Figure~\ref{sensitivityFig}).  Sensitivity is the ratio for true
+(Figure~3).  Sensitivity is the ratio for true
 positives to actual positives ($tp / (tp + fn)$) -- the fraction of
@@
 positives, true negatives, and false positives ($tp / (tp+fn+fp)$)
-[\cite{Hazelhurst08a}].  Notably, when we removed duplicated genes
+\cite{Hazelhurst08a}.  Notably, when we removed duplicated genes
 before generating the simulated EST sets, we found all three tools to
-have 100\% specificity: no false positives.}
+have perfect specificity: no false positives.}
 
 \mc{Type 1 and Type 2 errors, as defined in {\it Wang et al.}
-[\cite{Wang04}], measure quality at the level of the genes from which
+\cite{Wang04}, measure quality at the level of the genes from which
 the ESTs were derived.  For Type 1 error, we look at the fraction of genes
@@
 
-\mc{A fifth metric is to look at the number of singletons produce by a
+\mc{A fifth metric is the number of singletons produce by a
   tool:  sequences that  could not  be  assigned to  any cluster.   In
-  practice such a sequence could represent compelte coverage of a very
+  practice such a sequence could represent complete coverage of a very
   short transcript (one that may be otherwise had to identify), but is
@@
   by each tool when run on  simulated data in which there should be no
-  singletons produced,  seeing a very slight improvment  of PEACE over
+  singletons produced,  seeing a very slight improvement  of PEACE over
   WCD in certain error ranges.}
@@
 \centerline{
-\label{dups}
 \includegraphics[scale=0.35]{pics.d/singletons_40.pdf}
@@
   where a fully correct clustering would have no singletons.  Values
-  average over 30 triles.  Blue/Solid = \textsc{peace}, Green/Dashed =
+  average over 30 trials.  Blue/Solid = \textsc{peace}, Green/Dashed =
   \textsc{wcd}, Black/Dot-Dashed = \textsc{cap3}.}\label{singletons}
@@
 or less.  \textsc{peace} and \textsc{wcd} are roughly comparable, both clearly able to
-separate duplicates out at a similarity level of $83\%$ -- but completely 
+separate duplicates out at a similarity level of $78\%$ -- but completely 
 unable to distinguish sharing a similarity of $~88\%$ or more.
@@
 \centerline{
-\label{dups}
 \includegraphics[scale=0.35]{pics.d/duplicates_40.pdf}
@@
   divergence.  Estimates averaged over 30 trials; one trial consists
-  of taking a random gene, copying it and stochastically changing bases
-  at the specified rate, then using the two genes as the bases for
-  generating a simulated set. Blue/Solid = \textsc{peace}, Green/Dashed = \textsc{wcd}, Black/Dot-Dashed = \textsc{cap3};
-  variance was too small for visible plotting of
-  confidence intervals.}
+  of taking a random gene, copying it and stochastically changing
+  bases at the specified rate, then using the two genes as the bases
+  for generating a simulated set. Blue/Solid = \peace, Green/Dashed =
+  \wcd\/ (mostly obscured by the blue line), Black/Dot-Dashed =
+  \capthree; variance was too small for visible plotting of
+  confidence intervals.}\label{dups}
 \end{figure}
@@
   The \wcd\/ tool uses a $d^2$ threshold of $40$ to determine whether
-  two sequences are in the same cluster.  As this is equivilent to our
+  two sequences are in the same cluster.  As this is equivalent to our
   $T$ threshold for dismissing MST edges, we have set it to the same
@@
   analysis indicated that a $T$ value as high as 130 should not have a
-  significant impact on its effectivness as a descriminator.  And in
+  significant impact on its effectiveness as a discriminator.  And in
   fact this was born out in simulation, as shown in
-  Figure~\label{SiJiT1T2130}.  In contrast with the equivilent results
+  Figure~\label{SiJiT1T2130}.  In contrast with the equivalent results
   for a threshold of $T=40$, as shown in the manuscript, we see
@@
   in theory and in analysis, it did not hold in practice.  In
-  Table~\ref{SangerTable130} we show the match to the manuscript
-  Table~\ref{SangerTable}, and observe the poor performance of
+  Table~\ref{SangerReal130} we show the match to the manuscript
+  Table~1, and observe the poor performance of
   \peace\/ when looking at the Jaccard index.  It is clear that in the
-  real data, the level of similarity is consierably higher then in the
+  real data, the level of similarity is considerably higher then in the
   simulations, thus making the reduced specificity of the higher
@@
 \centerline{
-\label{dups130}
 \includegraphics[scale=0.35]{pics.d/duplicates_130.pdf}
@@
   variance was too small for visible plotting of
-  confidence intervals.}
+  confidence intervals.}\label{dups130}
 \end{figure}
-
-\subsection{Runtime}
-\label{rt_section}
-
-In Figure~\ref{seq_runtime} we see give a runtime comparison of \textsc{peace} and \textsc{wcd},
-seeing a slight improvement in \textsc{peace}.  (As \textsc{cap3} performs both
-clustering and assembly, runtime comparisons are not meaningful.)  As
-stated, we see an almost perfectly quadratic curve ($r > 0.99$),
-indicating that the runtime is dominated by the time required to look
-at every EST pair (either filtering or computing $d^2$).
-As both \textsc{peace} and \textsc{wcd} can be run in a parallel mode, we also
-examined runtime as a function of the ratio of EST set sizes to number
-of processors used.  In Figure~\ref{par_runtime} we show two such ratios,
-again seeing an improvement in \textsc{peace} over \textsc{wcd}.
-
-\mc{All tests were run on the University of Miami Redhawk cluster,
-  made up of 3.2 GHz Intel Xeon EM64T CPUs, each with a 2MB cache and
-  800 MHz side bus, model number Xeon LV 3.0 (2005).}
-
-%%%%%%%%%%%%%
-\begin{figure}[tbp]
-\centerline{
-\includegraphics[scale=0.35]{pics.d/seq_time.pdf}
-\label{seq_runtime}
-}
-\caption{Single Processor Runtime as a function of input size,
-  averaged over 50 simulated EST sets; Blue/Solid = \textsc{peace},
-  Green/Dashed = \textsc{wcd}; vertical tics represent= 95\% confidence
-  intervals on estimates.}
-\end{figure}
-\begin{figure}[tbp]
-\centerline{
-\includegraphics[scale=0.5]{pics.d/par_time.pdf}
-\label{par_runtime}
-}
-\caption{Runtime of a function of input size when holding constant the
-  ratio of set size to number of processors used; averaged over 30
-  simulated EST sets.  Blue/Solid = \textsc{peace}, Green/Dashed = \textsc{wcd};
-  vertical tics represent= 95\% confidence intervals on estimates.}
-\end{figure}
-%%%%%%%%%%%%%
 
@@
 In Table~\ref{RunTime} we show the runtime of \peace\/ and \wcd\/ on
-five different benckmark sets obtained from the WCD and EasyCluster
-studies [\cite{Hazelhurst08a,Picardi09}].
+five different benchmark sets obtained from the WCD and EasyCluster
+studies \cite{Hazelhurst08a,Picardi09}.
 
@@
 \hline
- Bencmark & \begin{tabular}{c} Number of \\ Sequences \end{tabular}
+ Benchmark & \begin{tabular}{c} Number of \\ Sequences \end{tabular}
  & \begin{tabular}{c} Number of \\ Bases \end{tabular} & Peace Runtme (s)&
@@
 \caption{Runtime for \peace\/ and \wcd\/ on benchmarks from the
-  WCD and EasyCluster studies [\cite{Hazelhurst08a, Picardi09}].
+  WCD and EasyCluster studies \cite{Hazelhurst08a, Picardi09}.
   Runtimes for the first four sets are sequential, while that last was
@@
 
-\subsection{Higer Threshold}
+\subsection{Higher Threshold}
 
-In Table~\ref{SangerReal130} we see the effect on the application of
-\peace\/ to the data sets from Table~\ref{SangerReal} when raising the
-threshold $T$ to 130 (see pervious discussion in
-Section~\ref{sim_thresh}.  While doing this appeared to be fairly
-effective in simulation, it did not hold up in pracice.  We see
+\mc{ In Table~\ref{SangerReal130} we see the effect on the application
+  of \peace\/ to the data sets from Table~1 when
+  raising the threshold $T$ to 130 (see previous discussion in
+  Section~\ref{sim_thresh}).  While doing this appeared to be fairly
+  effective in simulation, it did not hold up in practice.  We see
   essentially no change in the sensitivity, but significant changes in
-the Jaccard Index (and a corresponding decrease in the total number of
-clusters).  We suspect that this is due to a higher number of more
-closely related genes than are present in the simulated data.  We can
-conclude that, without considerably more investigation, we should be
-settig $T$ to 40.
+  the Jaccard Index (and a corresponding decrease in the total number
+  of clusters).  We suspect that this is due to a higher number of
+  more closely related genes than are present in the simulated
+  data. Regardless, without considerably more investigation, these
+  results indicated we should be leaving $T$ at a default rate of 40.
+}
 
@@
   Dataset and the WCD A076941 Arabidopsis thaliana dataset using the
-  standard quality measurments with $T=130$.}\label{SangerReal130}
+  standard quality measurements with $T=130$.}\label{SangerReal130}
 
@@
 % LocalWords:  nalysis lustering ngine al substring Hazelhurst \textsc{wcd}  ESTsim ESTs
-% LocalWords:  polymearse Methodoloy Jaccard tp fn fp  wcd th Prim's Ns
-% LocalWords:  Chlamydomonas reinhartdii
+% LocalWords:  polymearse Methodoloy Jaccard tp fn fp  wcd th Prim's Ns ssembly
+% LocalWords:  Chlamydomonas reinhartdii nvironment xpression runtime bp Runtme
+% LocalWords:  stochastically Runtimes EasyCluster Ricinus Dataset

### 2010-04-20 peace/papers/NAR_web_server_2009/paper.tex
Adding singletons_40.pdf figure

@@
 \begin{abstract}
-  We present \peace, a stand alone tool for the high-throughput {\it
-    ab initio} clustering of transcript fragment sequences produced by
-  Next Generation or Sanger Sequencing technologies, freely available
-  from \href{http://www.peace-tools.org}{www.peace-tools.org}.
-  Installed and run through an easily downloaded GUI, \peace\/ can
-  process large data sets of transcript fragments of length 50 bases
-  or greater, grouping the fragments by gene association and achieving
-  a greater sensitivity (without sacrificing runtime) than the
-  competing clustering tools.  Once clustered, the user can employ the
-  GUI to collect statistics and single out specific clusters for more
-  comprehensive study.  \peace\/, when used as a ``pre-assembly''
-  tool, enables any assembly tool to work with individual small
-  clusters of closely related sequences.  For a 5\% sequencing error
-  rate, we see a 52\% improvement in sensitivity over the leading
-  cluster tool, \wcd\/, and 184\% improvement in sensitivity over the
-  assembly tool \capthree.  When applied to a 361 Mb set of
-  Arabidopsis transcriptome fragments, we see a 40\% speedup over
-  \wcd.
+  \mc{We present \peace, a stand alone tool for high-throughput {\it
+      ab initio} clustering of transcript fragment sequences produced
+    by Next Generation or Sanger Sequencing technologies.  It is
+    freely available from
+    \href{http://www.peace-tools.org}{www.peace-tools.org}.  Installed
+    and managed through a download-able user-friendly GUI, \peace\/
+    can process large data sets of transcript fragments of length 50
+    bases or greater, grouping the fragments by gene associations with
+    a sensitivity compatible to leading clustering tools.  Once
+    clustered, the user can employ the GUI's analysis functions,
+    facilitating easy collection of statistics and single out specific
+    clusters for more comprehensive study or assembly.  Using a novel
+    minimum spanning tree based clustering method, \peace is the equal
+    of other leading tools in the literature.  It produces results of
+    quality virtually identical to those of the \wcd\/ tool when
+    applied to Sanger sequences, significantly improved results over
+    \wcd\/ when applied the products of Next Generation Sequencing
+    Technology, and significantly improved results over \capthree\/ in
+    both cases.  In short, \peace\/ provides an intuitive graphical
+    user interface and a feature-rich, parallel clustering engine that
+    collectively are as good, if not better, than the leading
+    clustering tools in the literature.}
+
 \end{abstract}
@@
 Sanger Sequencing technologies, comparing results against those
-produced by the \wcd\/ clustering tool \mc{(version
-  0.5.1)}\cite{Hazelhurst08a} and the \capthree\/ assembly tool
-\cite{Huang99} (the latter of which implicitly calculates a clustering
-in the process of assembly).  For our simulation tests we used the
-{\bf ESTSim} tool \cite{Hazelhurst03} to generate simulated transcript
-fragments of varying length under different models of error
-(Supplementary Materials, Section~C.1), generating the fragments from
-the list of 100 zebra fish genes used in the \wcd\/ testing
-\cite{Hazelhurst08a}.  Tool parameters were taken to match, as closely
-as possible, those used in the \wcd\/ study (see Supplementary
-Materials).  \mc{The most important method of quality assessment is
-  {\it sensitivity} (the fraction of fragment pairs from the same gene
-  that were correctly clustered together).  But we also look at the
-  {\it Jaccard Index} (which balances sensitivity with the number of
-  false positives by looking at the ratio of true positives to the sum
-  of true positives, false positives, and true negitives), {\it Type 1
-    error} (the fraction of genes that were divided between clusters),
-  and {\it Type 2 error} (the fraction of clusters containing two or
-  more genes).  Specificity was not an issue in simulated tests, as
-  both \peace\/ and \wcd\/ were completely successful in separating
-  unrelated ESTs (save when dealing with recently duplicated genes).
-  Other measurements are also discussed in the supplementary
-  materials.  In Figure~\ref{SeJiT1T2} we plot these four tests as a
-  function of erro rate, and observe the almost identical results
-  between \peace\/ and \wcd\/.  In Figure~\ref{seq_parallel} we plot
-  the runtime for \peace\/ and \wcd\/, again observing almost identical
-  results when run sequentially -- but signifcantly faster runtime for
-  PEACE on multiple processer when holding the EST/processor ratio
-  constant (ranging from a 65\% improvement for two processors too a
-  17\% improvement for 12 processors).}
+produced by the \wcd\/ clustering tool \cite{Hazelhurst08a} and the
+\capthree\/ assembly tool \cite{Huang99} (the latter of which
+implicitly calculates a clustering in the process of assembly).  For
+our simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03}
+to generate simulated transcript fragments of varying length under
+different models of error (Supplementary Materials, Section~D.1),
+generating the fragments from the list of 100 zebra fish genes used in
+the \wcd\/ testing \cite{Hazelhurst08a}.  Tool parameters were taken
+to match, as closely as possible, those used in the \wcd\/ study (see
+Supplementary Materials).  \mc{The most important method of quality
+  assessment is {\it sensitivity} (the fraction of fragment pairs from
+  the same gene that were} \mc{correctly clustered together).  But we also
+  look at the {\it Jaccard Index} (which balances sensitivity with the
+  number of false positives by looking at the ratio of true positives
+  to the sum of true positives, false positives, and true negatives),
+  {\it Type 1 error} (the fraction of genes that were divided between
+  clusters), and {\it Type 2 error} (the fraction of clusters
+  containing two or more genes).  Specificity was not an issue in
+  simulated tests, as both \peace\/ and \wcd\/ were completely
+  successful in separating unrelated ESTs (save when dealing with
+  recently duplicated genes).  In Figure~\ref{SeJiT1T2} we plot
+  these four tests as a function of error rate, and observe the almost
+  identical results between \peace\/ and \wcd\/.  In
+  Figure~\ref{seq_parallel} we plot the runtime for \peace\/ and
+  \wcd\/, again observing almost identical results when run
+  sequentially -- but significantly faster runtime for PEACE on
+  multiple processor when holding the EST/processor ratio constant
+  (ranging from a 65\% improvement for two processors too a 17\%
+  improvement for 12 processors).}
+
+\mc{We further note, as a point of interest that, we can significantly
+  improve the quality of \peace simulation results through the
+  increase of the threshold value (see Supplementary Materials,
+  Section C) -- achieving a significant improved in PEACE sensitivity
+  without an adverse effect on the Jaccard index (i.e. a significant
+  increase in the incorrect merging of clusters).  However, the
+  improvement does not carry over to the application of real data,
+  where we observe a significant drop in the Jaccard Index.  Due to
+  this drop we have opted to leave the lower threshold value as the
+  default, but users might consider the higher value, thus decreasing
+  the risk of incorrectly splitting clusters, if planning on
+  using on the subsequent use of a assembly tool that capable of
+  breaking up the incorrectly merged clusters during transcript reconstruction.}
 
@@
   \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.pdf}}
-  \caption{Comparisons of Sensitivity, Jiccard Index, Type 1 error and
-    Type 1 error, based on the
-    average over 30 simulated Sanger Sequence ESTs sets derived from 100 zebra
-    fish genes  (see
+  \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
+    Type 1 error, based on the average over 30 simulated Sanger
+    Sequence ESTs sets derived from 100 zebra fish genes (see
     Supplementary Materials, Section C, for more details).  Blue/Solid
-    = \peace, Green/Dash = \wcd, Black/Dot-Dash = \capthree; vertical
-    tics = 95\% confidence intervals on estimates.  Intervals are not
-    presented for Type 1 error due to the effective lack of
-    variance.}\label{SeJiT1T2}
+    = \peace, Green/Dash = \wcd \mc{(version 0.5.1)}, Black/Dot-Dash =
+    \capthree; vertical tics = 95\% confidence intervals on estimates.
+    Intervals are not presented for Type 1 error due to the effective
+    lack of variance.}\label{SeJiT1T2}
 \end{figure}
@@
 \mc{In applying the tools to real data Sanger data, we used the Human
-Benchmark Dataset used to test \easycluster\/ \cite{Picardi09}, and
-the A076941.fa {\it Arabidopsis thaliana} dataset used to test WCD
-\cite{Hazelhurst08,Hazelhurst08a}, with results in
-Table~\ref{SangerReal}.  We notice essentially identical
-results for PEACE and WCD in quality, both significantly better than
-Cap3 in Sensitivity and Type 1 error rate, while slightly worse in the
-Jaccard Index and Type 2 error.  In runtime we see some inconsistency,
-with PEACE showing a 60\% runtime improvement in the smaller dataset,
-requiring 20\% more time in the Arabidopsis dataset.}
+  Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
+  the A076941.fa {\it Arabidopsis thaliana} data set used to test WCD
+  \cite{Hazelhurst08a,Hazelhurst08b}, (see Table~\ref{SangerReal}).
+  We notice essentially identical results for PEACE and WCD in
+  quality, both significantly better than Cap3 in Sensitivity and Type
+  1 error rate, while slightly worse in the Jaccard Index and Type 2
+  error.  In runtime we see some inconsistency, with PEACE showing a
+  60\% runtime improvement in the first data set, but requiring 20\%
+  more time in the Arabidopsis data set.  In Section~D.1 of the
+  Supplementary Materials we present runtimes several more sets,
+  observing that while \peace\/ appears to be significantly more
+  faster on the smaller sets, \wcd\/ does overtake it for larger
+  sets.}
 
@@
 \end{tabular}
-\label{SangerReal}
 \end{center}
 \caption{Comparisons of runs on the EasyCluster human Benchmark
-  Dataset and the WCD A076941 Arabidopsis thaliana dataset using the
-  standard quality measurments.}
+  Data Set and the WCD A076941 Arabidopsis thaliana data set using the
+  standard quality measurements.}\label{SangerReal}
 \end{table*}
@@
 
-We were unable to run \wcd\/ on short-read sequences, as the
-distributed implementation is unable to handle such data without an
-adjustment to the code.  Hence, while \peace\/ was run on both
-simulated and real NGS fragment sets, there was no cluster tool to
-which the results could be compared.  \peace\/ did well, achieving a
-sensitivity of 0.75 when applied to a simulated short-read set
-subjected to a sequencing error rate of 2\% (with comparable {\it
-  Jaccard Index} values reflecting both sensitivity {\it and}
-specificity \cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was
-approximately 17.6 minutes to process the 7.9 Mb of data when run in
-sequential mode on a .  When applied to a 64 Mb set of
-165,490 {\it
- Chlamydomonas reinhardtii} transcript fragments
-sequenced by 454 pyrosequencing technology (ranging in size from 50
-bases to several hundred), \peace\/ completed the job in 49 minutes
-using 30 of the Xeon processors.
+\mc{We tested the three tools on short-read data using the \metasim\/
+  tool of {\it Richter et al.} \cite{Richter2008}.  Encoded into
+  \metasim\/ are sequence generation and error models corresponding to
+  several technologies, including the 454 short-read technology.
+  Leaving \capthree\/ and \wcd\/ at their default values, but
+  employing the {\it adaptive $d^2$} strategy for short read sequences
+  (see Section~C in the supplementary materials), we find a that
+  \peace\/ performs significantly better than the other two tools on
+  454 data using the default \metasim\/ model (with sequence sizes
+  ranging in size from 200bp to 370bp, at an average size of 250bp and
+  a standard deviation of 17bp).  Over thirty runs, we see an average
+  sensitivity 0.884 for \peace, 0.331\/ for \wcd, and 0.004 for
+  \capthree.  For the Jaccard Index we get essentially the same
+  numbers.  For 100 transcript \peace\/ creates 128 clusters (as
+  opposed to 6,370 and 19245 clusters for \wcd\/ and \capthree), and
+  creates only 3 singletons (as opposed to 5,850 and 15432).  We do
+  note that it is likely that the \wcd results could be considerably
+  improved by the adjustment of a few parameters, but were unable to
+  determine how to do that.  (Runtime of \peace\/ was somewhat slower
+  than that of \wcd, but given the significant improvement in quality
+  of results this is unimportant.)}
 
@@
 % LocalWords:  ssembly xpression GenBank dbEST pyrosequencing ESTsim sim mRNAs
-% LocalWords:  NGS polyA set's GHz Xeon Bikun
+% LocalWords:  NGS polyA set's GHz Xeon Bikun Arabidopsis JVM Jaccard LV GUI's
+% LocalWords:  Data Set thaliana data set EasyCluster MetaSim

### 2010-04-20 papers/NAR_web_server_2009/paper.sup.tex
Modifications to reviewer response

@@
 
+\usepackage[pdftex,
+colorlinks=true, %makes the links show up in color, rather than in a box
+citecolor=black, %color of the in-text citation numbers
+urlcolor=blue %color of the url links
+]{hyperref}
+\usepackage{wrapfig}
+\bibliographystyle{unsrt}
+
+%\usepackage[pdftex,
+%colorlinks=true, %makes the links show up in color, rather than in a box
+%citecolor=black, %color of the in-text citation numbers
+%urlcolor=blue %color of the url links
+%]{hyperref}
+
 \sloppy 
 \usepackage{graphicx}
-\bibliographystyle{natbib}
+%\bibliographystyle{natbib}
 
@@
 \usepackage{color}
-\definecolor{blue}{rgb}{0,0,1}
-\newcommand{\mc}[1]{\textcolor{blue}{#1}}
+\definecolor{red}{rgb}{1,0,0}
+\newcommand{\mc}[1]{\textcolor{red}{#1}}
 %\newcommand{\mc}[1]{#1}   % Replace above line with this to eliminate all coloring
@@
 
-\subsection{Adaptive $d^2$}
+\section{Adaptive $d^2$}
 
-In the above discussion of $d^2$ we assume sequences of at least 100 bp, and
-in pratice use a window of size 100.  While the technique should be,
-in principle, adaptable to short-read sequences (decreasing window
-size as necessary), there are some problems in the details.
+\mc{In the above discussion of $d^2$ we assume sequences of at least 100
+bp, and in practice use a window of size 100.  While the technique
+should be, in principle, adaptable to short-read sequences (decreasing
+window size as necessary), there are some problems in the details.
 Diminishing window sizes implies both diminishing accuracy (requiring
-the thrshold be adjusted) and increasing runtime (as there are
-potenitally more window pairs).  More importantly, there is the
-problem of non-homogonous sequence lengths: short-read technologies
-tend to produce short reads, but can still include quite long reads in
-their output.  Hence it is not immediately clear how to compare a very
-short read against a very long one.
+the threshold be adjusted) and increasing runtime (as there are
+potentially more window pairs).  More importantly, there is the
+problem of non-homogeneous sequence lengths: short-read technologies
+tend to produce short reads, but some technologies can still include
+long reads in their output.  Hence it is not immediately clear how to
+compare a very short read against a very long one.}
 
-We address these issues with our {\it adaptive $d^2$} strategy that
-works as follows: we put sequences, based on length, into three {\it
-  overlapping} groups (i.e. sequences less than length XXX, sequences
-between length YYY and ZZZ, and seqeunces greater than AAA).  For each
-group we pick a constant windowsize and threshold tailored to the
-length range of the group (as well as filtering heurstic constants),
+\mc{We address these issues with our {\it adaptive $d^2$} strategy that
+works as follows: we partition sequences by length into disjoint groups.  For each
+group we pick a constant window size and threshold tailored to the
+length range of the group (as well as filtering heuristic constants),
 and use those values when computing the $d^2$ distance for two
 sequences in the same group.  We do not compute distances between
-sequences in different groups. Once computed, we assign to the MST
+sequences in non-adjacent groups (i.e. a member of the ``shortest
+groups'' against a member of the ``longest groups'' if there is an
+intervening sized group).  Once computed, we assign to the MST
 edge the ratio of the $d^2$ score to the group threshold, thus
 normalizing the weight for comparisons ``between groups'' and
-effectivally giving us an edge-removal threhold of 1.
+effectively giving us an edge-removal threshold of 1.}
 
-By restricting $d^2$ comparisons to within a group, we minimize the
-effets of the hetrogenous sequences lengths by restricting our
-comparisons to only sequences of similar lengths.  By not restrcicting
+\mc{By restricting $d^2$ comparisons to within a group, we minimize the
+effects of the heterogeneous sequences lengths by restricting our
+comparisons to only sequences of similar lengths.  By not restricting
 ourselves to a single window size, we allow ourselves to, when possible, avoid the
-preformance hit of using a small window.  As we are not comparing
+performance hit of using a small window.  As we are not comparing
 sequences from different groups we bypass the question of how to
-compare a largelong sequence against a short one.  And because the
+compare a long sequence against a short one.  And because the
 groups overlap in content, we still get implicit comparisons between
 long and short sequences in terms of the resulting path of the minimum
-spanning tree that makes use of these ``linking'' nodes.
+spanning tree that makes use of these ``linking'' nodes.}
 
-In the \textsc{peace} implementation, we subdevide into three groups:
-NEED INFO.
+\mc{In the \textsc{peace} implementation, we subdivide into three groups:
+\begin{itemize}
+\item Less than 150 ($w=50$, $T=40$, $t=20$, $u=4$).
+\item Between 150 and 400 ($w=75$, $T=60$, $t=35$, $u=6$)
+\item Greater than 400 ($w=100$, $T=80$, $t=50$, $t=50$).
+\end{itemize}}
 
@@
 
-\mc{ Our primary measurement tools if mesurement are sensitivity, the
+\mc{Our primary measurement tools if measurement are sensitivity, the
 Jaccard Index, Type 1 error and Type 2 error
-(Figure~\ref{sensitivityFig}).  Sensitivity is the ratio for true
+(Figure~3).  Sensitivity is the ratio for true
 positives to actual positives ($tp / (tp + fn)$) -- the fraction of
@@
 positives, true negatives, and false positives ($tp / (tp+fn+fp)$)
-[\cite{Hazelhurst08a}].  Notably, when we removed duplicated genes
+\cite{Hazelhurst08a}.  Notably, when we removed duplicated genes
 before generating the simulated EST sets, we found all three tools to
-have 100\% specificity: no false positives. }
+have perfect specificity: no false positives.}
 
 \mc{Type 1 and Type 2 errors, as defined in {\it Wang et al.}
-[\cite{Wang04}], measure quality at the level of the genes from which
+\cite{Wang04}, measure quality at the level of the genes from which
 the ESTs were derived.  For Type 1 error, we look at the fraction of genes
@@
 Type 2 error we look at the fraction of clusters that contain two or
-more genes.   }
+more genes.}
 
-\mc{A fifth metric is to look at the number of singletons produce by a
+\mc{A fifth metric is the number of singletons produce by a
   tool:  sequences that  could not  be  assigned to  any cluster.   In
-  practice such a sequence could represent compelte coverage of a very
+  practice such a sequence could represent complete coverage of a very
   short transcript (one that may be otherwise had to identify), but is
@@
   divergence.  Estimates averaged over 30 trials; one trial consists
-  of taking a random gene, copying it and stochastically changing bases
-  at the specified rate, then using the two genes as the bases for
-  generating a simulated set. Blue/Solid = \textsc{peace}, Green/Dashed = \textsc{wcd}, Black/Dot-Dashed = \textsc{cap3};
-  variance was too small for visible plotting of
-  confidence intervals.}
+  of taking a random gene zebra-fish gene, copying it and stochastically changing
+  bases at the specified rate, then using the two genes as the bases
+  for generating a simulated set. Blue/Solid = \peace, Green/Dashed =
+  \wcd\/ (mostly obscured by the blue line), Black/Dot-Dashed =
+  \capthree; variance was too small for visible plotting of
+  confidence intervals.}\label{dups}
 \end{figure}
@@
 \subsection{Raising the Threshold}
+\label{sim_thresh}
 \mc{ 
   The \wcd\/ tool uses a $d^2$ threshold of $40$ to determine whether
-  two sequences are in the same cluster.  As this is equivilent to our
+  two sequences are in the same cluster.  As this is equivalent to our
   $T$ threshold for dismissing MST edges, we have set it to the same
@@
   in theory and in analysis, it did not hold in practice.  In
-  Table~\ref{SangerTable130} we show the match to the manuscript
-  Table~\ref{SangerTable}, and observe the poor performance of
+  Table~\ref{SangerReal130} we show the match to the manuscript
+  Table~1, and observe the poor performance of
   \peace\/ when looking at the Jaccard index.  It is clear that in the
-  real data, the level of similarity is consierably higher then in the
+  real data, the level of similarity is considerably higher then in the
   simulations, thus making the reduced specificity of the higher
   threshold considerably more of a problem.
-XXX}
+}
 
@@
   divergence.  Estimates averaged over 30 trials; one trial consists
-  of taking a random gene, copying it and stochastically changing bases
+  of taking a random zebra-fish gene, copying it and stochastically changing bases
   at the specified rate, then using the two genes as the bases for
@@
   variance was too small for visible plotting of
-  confidence intervals.}
+  confidence intervals.}\label{dups130}
 \end{figure}
 
-\subsection{Runtime}
-\label{rt_section}
+\section{Real Data}
 
-In Figure~\ref{seq_runtime} we see give a runtime comparison of \textsc{peace} and \textsc{wcd},
-seeing a slight improvement in \textsc{peace}.  (As \textsc{cap3} performs both
-clustering and assembly, runtime comparisons are not meaningful.)  As
-stated, we see an almost perfectly quadratic curve ($r > 0.99$),
-indicating that the runtime is dominated by the time required to look
-at every EST pair (either filtering or computing $d^2$).
-As both \textsc{peace} and \textsc{wcd} can be run in a parallel mode, we also
-examined runtime as a function of the ratio of EST set sizes to number
-of processors used.  In Figure~\ref{par_runtime} we show two such ratios,
-again seeing an improvement in \textsc{peace} over \textsc{wcd}.
+\subsection{Runtimes}
 
-\mc{All tests were run on the University of Miami Redhawk cluster,
-  made up of 3.2 GHz Intel Xeon EM64T CPUs, each with a 2MB cache and
-  800 MHz side bus, model number Xeon LV 3.0 (2005).}
-
-%%%%%%%%%%%%%
-\begin{figure}[tbp]
-\centerline{
-\includegraphics[scale=0.35]{pics.d/seq_time.pdf}
-\label{seq_runtime}
-}
-\caption{Single Processor Runtime as a function of input size,
-  averaged over 50 simulated EST sets; Blue/Solid = \textsc{peace},
-  Green/Dashed = \textsc{wcd}; vertical tics represent= 95\% confidence
-  intervals on estimates.}
-\end{figure}
-\begin{figure}[tbp]
-\centerline{
-\includegraphics[scale=0.5]{pics.d/par_time.pdf}
-\label{par_runtime}
-}
-\caption{Runtime of a function of input size when holding constant the
-  ratio of set size to number of processors used; averaged over 30
-  simulated EST sets.  Blue/Solid = \textsc{peace}, Green/Dashed = \textsc{wcd};
-  vertical tics represent= 95\% confidence intervals on estimates.}
-\end{figure}
-%%%%%%%%%%%%%
-
-\section{Raising the Threshold Real Data Sets}
+In Table~\ref{RunTime} we show the runtime of \peace\/ and \wcd\/ on
+five different benchmark sets obtained from the WCD and EasyCluster
+studies \cite{Hazelhurst08a,Picardi09}.
 
@@
 \begin{center}
+\begin{tabular} {|c|| c | c || c | c |}
+\hline
+ Benchmark & \begin{tabular}{c} Number of \\ Sequences \end{tabular}
+ & \begin{tabular}{c} Number of \\ Bases \end{tabular} & Peace Runtme (s)&
+ WCD Runtime (s) \\
+\hline
+\hline
+ Human Cluster & 17733 & 11.7 Mb & 293 & 804 \\
+\hline
+ Cotton & 29,995 & 17.1 Mb & 222 & 339 \\
+\hline
+ Ricinus & 57,690 & 40.9 Mb & 1344 & 2356 \\
+\hline
+ A076941 & 76,941 & 32.8 Mb & 1166 & 966 \\
+\hline
+\hline
+ A686903 & 686,904 & 294.7 Mb & 9449 & 2481 \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Runtime for \peace\/ and \wcd\/ on benchmarks from the
+  WCD and EasyCluster studies \cite{Hazelhurst08a, Picardi09}.
+  Runtimes for the first four sets are sequential, while that last was
+  run over thirty processors.}\label{RunTime}
+\end{table}
+
+\subsection{Higher Threshold}
+
+\mc{ In Table~\ref{SangerReal130} we see the effect on the application
+  of \peace\/ to the data sets from Table~1 when
+  raising the threshold $T$ to 130 (see previous discussion in
+  Section~\ref{sim_thresh}).  While doing this appeared to be fairly
+  effective in simulation, it did not hold up in practice.  We see
+  essentially no change in the sensitivity, but significant changes in
+  the Jaccard Index (and a corresponding decrease in the total number
+  of clusters).  We suspect that this is due to a higher number of
+  more closely related genes than are present in the simulated
+  data. Regardless, without considerably more investigation, these
+  results indicated we should be leaving $T$ at a default rate of 40.
+}
+
+\begin{table}[tbp]
+\begin{center}
+\begin{tiny}
 \begin{tabular} {| c || c || c | c | c | c | c | c | c |}
@@
     Singletons \end{tabular}
-    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular} \\
-\hline EasyCluster Human & \peace & 0.999 & 0.106 & 0.027 & 0.030 & 67 & 2 & 295 \\
-\cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
-\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1 & 0.001 & 2269 & 1827 & \mbox{NA} \\
+    & \begin{tabular}{c} Single \\ processor \\ runtime
+      (s) \end{tabular} \\
 \hline
-\hline WCD A076941 & \peace & 0.937 & 0.345 & 0.337 & 0.0342 & 18256 & 8135 & 1293 \\
-\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
-\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
+\hline EasyCluster Human & PEACE & 0.999 & 0.106 & 0.027 & 0.030 & 67 & 2 & 295 \\
+\cline{2-9} Benchmark & WCD & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
+\cline{2-9} (111 Genes) & Cap3 & 0.657 & 0.643 & 1.000 & 0.001 & 2269 & 1827 & \mbox{NA} \\
+\hline
+\hline WCD A076941 & PEACE & 0.937 & 0.345 & 0.337 & 0.0342 & 18256 & 8135 & 1293 \\
+\cline{2-9} Benchmark & WCD & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
+\cline{2-9} (13240 genes) & Cap3 & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
 \mbox{NA} \\
@@
 \end{tabular}
-\label{SangerReal130}
+\end{tiny}
 \end{center}
@@
   Dataset and the WCD A076941 Arabidopsis thaliana dataset using the
-  standard quality measurments with $T=130$.}
+  standard quality measurements with $T=130$.}\label{SangerReal130}
+
 \end{table}

### 2010-04-20 peace/papers/NAR_web_server_2009/paper.tex
Modifications to paper.

@@
     clustered, the user can employ the GUI's analysis functions,
-    facilitating easy collection of statistics and single out specific
-    clusters for more comprehensive study or assembly.  Using a novel
-    minimum spanning tree based clustering method, \peace is the equal
-    of other leading tools in the literature.  It produces results of
+    facilitating easy collection of statistics and allowing them to
+    single out specific clusters for more comprehensive study or
+    assembly.  Using a novel minimum spanning tree based clustering
+    method, \peace\/ is the equal of other leading tools in the
+    literature in quality and interface.  It produces results of
     quality virtually identical to those of the \wcd\/ tool when
@@
     user interface and a feature-rich, parallel clustering engine that
-    collectively are as good, if not better, than the leading
-    clustering tools in the literature.}
+    proves to be a valuable addition to the leading cDNA clustering
+    tools.}
 
@@
   \caption{Overview of the procedure for clustering and analysis using
-    PEACE.}\label{fig:workflow}
+    \peace.}\label{fig:workflow}
 \end{figure}
@@
 \begin{figure}
-  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.pdf}}
+  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.png}}
   \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
@@
     Supplementary Materials, Section C, for more details).  Blue/Solid
-    = \peace, Green/Dash = \wcd \mc{(version 0.5.1)}, Black/Dot-Dash =
+    = \peace, Green/Dash = \wcd\/ \mc{(version 0.5.1)}, Black/Dot-Dash =
     \capthree; vertical tics = 95\% confidence intervals on estimates.
@@
 \begin{figure}
-   \centerline{\includegraphics[width=3.35in]{pics.d/seq_parallel.pdf}}
+   \centerline{\includegraphics[width=3.35in]{pics.d/seq_parallel.png}}
    \caption{Comparisons of Runtime: On left, we compare the sequential
@@
   Supplementary Materials we present runtimes several more sets,
-  observing that while \peace\/ appears to be significantly more
+  observing that while \peace\/ appears to be significantly 
   faster on the smaller sets, \wcd\/ does overtake it for larger

### 2010-04-20 papers/NAR_web_server_2009/paper.sup.tex
Manuscript changes

@@
 distribution constant, the tool as a whole has a runtime of $O(n^2)$
--- appearing to dominated by the time required to apply
+-- appearing to be dominated by the time required to apply
 the filtering heuristics to every EST pair.

### 2010-04-20 papers/NAR_web_server_2009/paper.tex
Manuscript changes

@@
   assessment is {\it sensitivity} (the fraction of fragment pairs from
-  the same gene that were} \mc{correctly clustered together).  But we also
-  look at the {\it Jaccard Index} (which balances sensitivity with the
-  number of false positives by looking at the ratio of true positives
-  to the sum of true positives, false positives, and true negatives),
-  {\it Type 1 error} (the fraction of genes that were divided between
-  clusters), and {\it Type 2 error} (the fraction of clusters
-  containing two or more genes).  Specificity was not an issue in
-  simulated tests, as both \peace\/ and \wcd\/ were completely
-  successful in separating unrelated ESTs (save when dealing with
-  recently duplicated genes).  In Figure~\ref{SeJiT1T2} we plot
-  these four tests as a function of error rate, and observe the almost
+  the same gene that were} \mc{correctly clustered together).  But we
+  also look at the {\it Jaccard Index} (which balances sensitivity
+  with the number of false positives by looking at the ratio of true
+  positives to the sum of true positives, false positives, and true
+  negatives), {\it Type 1 error} (the fraction of genes that were
+  divided between clusters), and {\it Type 2 error} (the fraction of
+  clusters containing two or more genes).  Specificity was not an
+  issue in simulated tests, as both \peace\/ and \wcd\/ were
+  completely successful in separating unrelated ESTs (with the
+  exception of recently duplicated genes, as discussed in the
+  Supplementary Materials).  In Figure~\ref{SeJiT1T2} we plot these
+  four tests as a function of error rate, and observe the almost
   identical results between \peace\/ and \wcd\/.  In
@@
   the A076941.fa {\it Arabidopsis thaliana} data set used to test WCD
-\cite{Hazelhurst08a,Hazelhurst08b}, (see Table~\ref{SangerReal}).  We notice essentially identical
-results for PEACE and WCD in quality, both significantly better than
-Cap3 in Sensitivity and Type 1 error rate, while slightly worse in the
-Jaccard Index and Type 2 error.  In runtime we see some inconsistency,
-with PEACE showing a 60\% runtime improvement in the first data set,
-but requiring 20\% more time in the Arabidopsis data set.  In Section~D.1
-of the Supplementary Materials we present runtimes several more
-sets, observing that while \peace\/ appears to be significantly more
-faster on the smaller sets, \wcd\/ does overtake it for larger sets.}
+  \cite{Hazelhurst08a,Hazelhurst08b}, (see Table~\ref{SangerReal}).
+  We notice essentially identical results for PEACE and WCD in
+  quality, both significantly better than Cap3 in Sensitivity and Type
+  1 error rate, while slightly worse in the Jaccard Index and Type 2
+  error.  In runtime we see some inconsistency, with PEACE showing a
+  60\% runtime improvement in the first data set, but requiring 20\%
+  more time in the Arabidopsis data set.  In Section~D.1 of the
+  Supplementary Materials we present runtimes several more sets,
+  observing that while \peace\/ appears to be significantly 
+  faster on the smaller sets, \wcd\/ does overtake it for larger
+  sets.}
 
@@
 \hline
-\hline Bicyclist Human & \peace & 0.998 & 0.672 & 0.153 & 0.042 & 118 & 21 & 293 \\
+\hline EasyCluster Human & \peace & 0.998 & 0.672 & 0.153 & 0.042 & 118 & 21 & 293 \\
 \cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
@@
 
+\begin{table*}[tbp]
+\begin{center}
+\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
+\hline
+ & 
+& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
+  error \end{tabular} & \begin{tabular}{c} Type 2 \\
+  error \end{tabular} &
+\begin{tabular}{c} Number \\ of  \\
+    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
+    Singletons \end{tabular}
+    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
+\hline
+\hline EasyCluster Human & \peace & 0.871 & 0.809 &
+\cline{2-9} Benchmark & \wcd &  0.020 & 0.020 &
+\cline{2-9} (111 Genes) & \capthree & 0.005 & 0.004 &
+\hline
+\hline WCD A076941 & \peace & 0.932 & 0.475 & 0.351 & 0.027 & 18825 & 8951 & 1166\\
+\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
+\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
+\mbox{NA} \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Comparisons of runs on the EasyCluster human Benchmark
+  Data Set and the WCD A076941 Arabidopsis thaliana data set using the
+  standard quality measurements.}\label{SangerReal}
+\end{table*}
+
 \section{Conclusions}
@@
 \mc{Here we have presented \peace, a stand alone tool for the
-high-throughput clustering of transcript fragments capable of dealing
-with sequences as short as 50 bases.  \peace\/, available at
-\href{http://www.peace-tools.org}{www.peace-tools.org}, is open-source
-and managed through a user friendly GUI that enables both local and
-remote installation and execution in sequential or parallel mode.
-Based on a novel algorithm for the clustering of the fragments by gene
-association, \peace\/ shows significant improvement in sensitivity,
-over the competing \wcd\/ tool \cite{Hazelhurst08a} when applied to
-NGS reads, matches \wcd\/ when applied to Traditional Sanger
-Sequencing output, and shows an order of magnitude in improvement over
-the clustering performed in the course of assembly by the \capthree\/
+  high-throughput clustering of transcript fragments capable of
+  dealing with sequences as short as 50 bases.  \peace\/, available at
+  \href{http://www.peace-tools.org}{www.peace-tools.org}, is
+  open-source and managed through a user friendly GUI that enables
+  both local and remote installation and execution in sequential or
+  parallel mode.  Based on a novel minimum-spanning-tree based
+  algorithm for the clustering of the fragments by gene association,
+  \peace\/ shows significant improvement in sensitivity, over the
+  competing \wcd\/ tool \cite{Hazelhurst08a} when applied to NGS
+  reads, matches \wcd\/ when applied to Traditional Sanger Sequencing
+  output, and shows an order of magnitude in improvement over the
+  clustering performed in the course of assembly by the \capthree\/
   tool \cite{Huang99}.}
@@
 whole) results in a significant reduction in overall assembly time.
+
+Peace can be downloaded from
+\href{http://www.peace-tools.org}{www.peace-tools.org}, where we are
+committed to maintaining and improving the tool.  The underlying PEACE
+architecture is configured to support a variety of tasks, including the
+possibility of a minimum spanning tree based assembly tool we hope to
+add to the tool in a future study.
 

### 2010-04-21 papers/NAR_web_server_2009/paper.sup.tex
Semi-final changes to the resubmission

@@
 \begin{itemize}
-\item Less than 150 ($w=50$, $T=40$, $t=20$, $u=4$).
-\item Between 150 and 400 ($w=75$, $T=60$, $t=35$, $u=6$)
-\item Greater than 400 ($w=100$, $T=80$, $t=50$, $t=50$).
+\item Less than 150 ($w=50$, $T=50$, $t=20$, $u=4$).
+\item Between 150 and 400 ($w=75$, $T=75$, $t=35$, $u=6$)
+\item Greater than 400 ($w=100$, $T=100$, $t=50$, $t=50$).
 \end{itemize}}
@@
 \begin{figure}
-  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.pdf}}
+  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x130.pdf}}
   \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
-    Type 1 error, based on the
-    average over 30 simulated Sanger Sequence ESTs sets derived from 100 zebra
-    fish genes  (see
-    Supplementary Materials, Section C, for more details).  Blue/Solid
-    = \peace, Green/Dash = \wcd, Black/Dot-Dash = \capthree; vertical
-    tics = 95\% confidence intervals on estimates.  Intervals are not
-    presented for Type 1 error due to the effective lack of
-    variance.}\label{SeJiT1T2130}
+    Type 1 error using the elevated threhold ($T=130$).  Based on the average
+    over 30 simulated Sanger Sequence ESTs sets derived from 100 zebra
+    fish genes (see Supplementary Materials, Section C, for more
+    details).  Blue/Solid = \peace, Green/Dash = \wcd, Black/Dot-Dash
+    = \capthree; vertical tics = 95\% confidence intervals on
+    estimates.  Intervals are not presented for Type 1 error due to
+    the effective lack of variance.}\label{SeJiT1T2130}
 \end{figure}
@@
 
+\section{Short Read Analysis}
+
+\begin{table}[tbp]
+\begin{center}
+\begin{tiny}
+\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
+\hline
+ & 
+& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
+  error \end{tabular} & \begin{tabular}{c} Type 2 \\
+  error \end{tabular} &
+\begin{tabular}{c} Number \\ of  \\
+    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
+    Singletons \end{tabular}
+    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
+\hline
+\hline   & \peace & 0.871 & 0.809 & 0.186 & 0.049 & 129 & 4 & 129 \\
+\cline{2-9}  MetaSim 454 & \wcd(1) & 0.073 & 0.073 & 0.014 & 0.082 & 7028.500 & 5296.533 & 74 \\
+\cline{2-9}  Simulation & \wcd(2) &  0.020 & 0.020 & 0.008 & 0.198 & 13085 & 10610 & 64 \\
+\cline{2-9}  & \capthree & 0.005 & 0.005 & 0.005 & 0.565 & 19222 & 15425 & 132 \\
+\hline
+\hline Metasim  & \peace & 0.583 & 0.372 & 0.230 & 0.111 & 392.914 & 149.257 & 3454 \\
+\cline{2-9} Illumina  & \capthree & 0.006 & 0.006 & 0.002 & 1.766 & 51663.057 & 30107.429 & 291 \\
+\hline
+\end{tabular}
+\end{tiny}
+\end{center}
+\caption{Comparisons of runs on on short read data, as generated by
+  \metasim \cite{Richter2008}; each figure is an average over 30 runs.
+  For \wcd(1) we use wcd default parameters, while for \wcd(2) we set
+  it to match the \peace\/ parameters ($w=75$, $T=60$); \wcd\/ was
+  unable to return any results for Illumina data, hence has been
+  omitted.  Each 454 simulation produced 25,000 reads from the 100
+  zebra-fish transcripts, ranging  in size from 200bp to 270bp (aveage = 250bp,
+  standard deviation = 17bp).  Each Illumina simulation produce 150K
+  reads from the 100 zebra-fish transcript, each exactly 62bp.}\label{SRSim}
+\end{table}
+
 %Bib TeX

### 2010-04-21 peace/papers/NAR_web_server_2009/paper.tex
Semi-final changes to the resubmission

@@
 
-\section{\peace\/: Installation and Use}
+\section{\peace: Installation and Use}
 
@@
 
-\mc{We further note, as a point of interest that, we can significantly
-  improve the quality of \peace simulation results through the
+\mc{In the course of our simulations we also invetigated the memory
+  footprint -- of particular concern for large data sets.  We find
+  that, as with \wcd, the required memory for an execution is linear
+  in the size of the dataset.}
+
+  \mc{We note, as a point of interest, that we can significantly
+    improve the quality of \peace\/ simulation results through the
     increase of the threshold value (see Supplementary Materials,
-  Section C) -- achieving a significant improved in PEACE sensitivity
-  without an adverse effect on the Jaccard index (i.e. a significant
-  increase in the incorrect merging of clusters).  However, the
-  improvement does not carry over to the application of real data,
-  where we observe a significant drop in the Jaccard Index.  Due to
-  this drop we have opted to leave the lower threshold value as the
-  default, but users might consider the higher value, thus decreasing
-  the risk of incorrectly splitting clusters, if planning on
-  using on the subsequent use of a assembly tool that capable of
-  breaking up the incorrectly merged clusters during transcript reconstruction.}
+    Section C) -- achieving a significant improved in PEACE
+    sensitivity without an adverse effect on the Jaccard index (i.e. a
+    significant increase in the incorrect merging of clusters).
+    However, the improvement does not carry over to the application of
+    real data (Supplementary Materials, Section E.2), where we observe
+    a significant increase in the incorrect merging of clusters.
+    While this might be acceptable to a user planning to employ an
+    assembl capbable of breaking up the clusters, we have choosed to
+    base our analysis (and set out default values) at the lower threshold.}
 
@@
   \metasim\/ are sequence generation and error models corresponding to
-  several technologies, including the 454 short-read technology.
-  Leaving \capthree\/ and \wcd\/ at their default values, but
-  employing the {\it adaptive $d^2$} strategy for short read sequences
-  (see Section~C in the supplementary materials), we find a that
-  \peace\/ performs significantly better than the other two tools on
-  454 data using the default \metasim\/ model (with sequence sizes
-  ranging in size from 200bp to 370bp, at an average size of 250bp and
-  a standard deviation of 17bp).  Over thirty runs, we see an average
-  sensitivity 0.884 for \peace, 0.331\/ for \wcd, and 0.004 for
-  \capthree.  For the Jaccard Index we get essentially the same
-  numbers.  For 100 transcript \peace\/ creates 128 clusters (as
-  opposed to 6,370 and 19245 clusters for \wcd\/ and \capthree), and
-  creates only 3 singletons (as opposed to 5,850 and 15432).  We do
-  note that it is likely that the \wcd results could be considerably
-  improved by the adjustment of a few parameters, but were unable to
-  determine how to do that.  (Runtime of \peace\/ was somewhat slower
-  than that of \wcd, but given the significant improvement in quality
-  of results this is unimportant.)}
+  several technologies, including the 454 short-read technology
+  (producing reads ranging in size from 200bp to 370bp with a mean of
+  250bp and a standard deviation of 17bp), and the Illumina short-read
+  technology (producing reads of exactly 62bp).  Experiements were run
+  using two sets of \wcd\/ parameters (the default paramters, and
+  setting window size to 75 and threshold to 75 -- thus matching the
+  \peace\/ parameters for this size squence), and using default values
+  of \capthree.  Only with \peace were we able to get reasonable
+  results, seeing a sensitivity of 0.871 for the 454 reads and
+  0.583 or the shorter Illumina reads, a Jaccard Index of 0.809 and
+  0.372 respectivally, and other comperable figures.  (See Table~S3 in
+  the Supplementary Materials.)  In short, \peace performs quite well
+  for 454 reads and provides some useful information for Illumina
+  reads, while \wcd\. and \capthree are unable to produce working
+  results.}
 
@@
 whole) results in a significant reduction in overall assembly time.
+
+Peace can be downloaded from
+\href{http://www.peace-tools.org}{www.peace-tools.org}, where we are
+committed to keep maintaining and improving the tool in the future.
+Meanwhile, we are developing our own assembly tool that can seamlessly
+integrate with PEACE.  The underlying modular design of PEACE offers
+users many possibilities to expand and incorporate the MST algorithm
+for other bioinformatics application.
 

### 2010-04-21 peace/papers/NAR_web_server_2009/paper.tex
Modifications to paper abstract

@@
     clustered, the user can employ the GUI's analysis functions,
-    facilitating easy collection of statistics and allowing them to
-    single out specific clusters for more comprehensive study or
-    assembly.  Using a novel minimum spanning tree based clustering
-    method, \peace\/ is the equal of other leading tools in the
-    literature in quality and interface.  It produces results of
-    quality virtually identical to those of the \wcd\/ tool when
-    applied to Sanger sequences, significantly improved results over
-    \wcd\/ when applied the products of Next Generation Sequencing
-    Technology, and significantly improved results over \capthree\/ in
-    both cases.  In short, \peace\/ provides an intuitive graphical
-    user interface and a feature-rich, parallel clustering engine that
-    proves to be a valuable addition to the leading cDNA clustering
-    tools.}
+    facilitating the easy collection of statistics and allowing them
+    to single out specific clusters for more comprehensive study or
+    assembly.  Using a novel {\it minimum spanning tree} based
+    clustering method, \peace\/ is the equal of other leading tools in
+    the literature in quality with an interface making it
+    accessible to any user.  It produces results of quality virtually
+    identical to those of the \wcd\/ tool when applied to Sanger
+    sequences, significantly improved results over \wcd\/ when applied
+    the products of Next Generation Sequencing Technology, and
+    significantly improved results over \capthree\/ in both cases.  In
+    short, \peace\/ provides an intuitive graphical user interface and
+    a feature-rich, parallel clustering engine that proves to be a
+    valuable addition to the leading cDNA clustering tools.}
 

### 2010-04-21 peace/papers/NAR_web_server_2009/paper.tex
Final version of manuscript

@@
 and accurately find clusters of ESTs expressed from the same gene
-without reference to a sequenced genome.  Compared against \wcd, the
-leading clustering tool in the literature \cite{Hazelhurst08a}, as well as
-other tools designed for the same purpose 
+without reference to a sequenced genome.  \mc{Compared against
+  clustering tool in the literature
   \cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
-  Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be both more
-sensitive and more robust to sequencing error without sacrificing
-runtime.  Nor are any of these tools designed for the ease of installation
-and use that \peace\/ provides.
+    Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be of
+  equivalent quality to the \wcd\/ tool \cite{Hazelhurst08a}, more
+  sensitive and robust than the other tools, and more user friendly
+  than any published clustering tool.  Nor are any of} \mc{these tools
+  designed for the ease of installation and use that \peace\/
+  provides.}
 
@@
 large without the need to apply the full $d^2$ algorithm (see
-Sections~A and B of the Supplementary Materials for more details).
+Sections~A.3 of the Supplementary Materials for more details).
 
@@
 \capthree\/ assembly tool \cite{Huang99} (the latter of which
-implicitly calculates a clustering in the process of assembly).  For
-our simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03}
-to generate simulated transcript fragments of varying length under
-different models of error (Supplementary Materials, Section~D.1),
-generating the fragments from the list of 100 zebra fish genes used in
-the \wcd\/ testing \cite{Hazelhurst08a}.  Tool parameters were taken
-to match, as closely as possible, those used in the \wcd\/ study (see
-Supplementary Materials).  \mc{The most important method of quality
-  assessment is {\it sensitivity} (the fraction of fragment pairs from
-  the same gene that were} \mc{correctly clustered together).  But we
-  also look at the {\it Jaccard Index} (which balances sensitivity
-  with the number of false positives by looking at the ratio of true
-  positives to the sum of true positives, false positives, and true
-  negatives), {\it Type 1 error} (the fraction of genes that were
-  divided between clusters), and {\it Type 2 error} (the fraction of
-  clusters containing two or more genes).  Specificity was not an
-  issue in simulated tests, as both \peace\/ and \wcd\/ were
-  completely successful in separating unrelated ESTs (with the
-  exception of recently duplicated genes, as discussed in the
-  Supplementary Materials).  In Figure~\ref{SeJiT1T2} we plot these
+implicitly calculates a clustering in the process of assembly).
+\mc{For our simulation tests we used the {\bf ESTSim} tool to generate
+  simulated Sanger sequenced transcript fragments, and the \metasim\/
+  tool to generate simulated short read sequences from 454 and Illumina
+  technologies \cite{Hazelhurst03,Richter2008}.  (See Supplementary
+  Materials, Section~D.1, for details on the sequence size and error
+  models.)  Fragments were generated from the list of 100 zebra fish
+  genes used in the \wcd\/ testing \cite{Hazelhurst08a}.  Tool
+  parameters were taken to match, as closely as possible, those used
+  in the \wcd\/ study (see Supplementary Materials).  }
+
+\mc{The most important method of quality assessment is {\it
+    sensitivity} (the fraction of fragment pairs from the same gene
+  that were} \mc{correctly clustered together).  We also look at the
+  {\it Jaccard Index} (which balances sensitivity with the number of
+  false positives), {\it Type 1 error} (the fraction of genes that
+  were divided between clusters), and {\it Type 2 error} (the fraction
+  of clusters containing two or more genes)
+  \cite{Hazelhurst08a,Wang04}.  In Figure~\ref{SeJiT1T2} we plot these
   four tests as a function of error rate, and observe the almost
@@
   multiple processor when holding the EST/processor ratio constant
-  (ranging from a 65\% improvement for two processors too a 17\%
+  (ranging from a 65\% improvement for two processors to a 17\%
   improvement for 12 processors).}
 
-\mc{In the course of our simulations we also invetigated the memory
-  footprint -- of particular concern for large data sets.  We find
-  that, as with \wcd, the required memory for an execution is linear
-  in the size of the dataset.}
+\mc{In the course of our simulations we also investigated the memory
+  footprint -- of particular concern in the large data sets commonly
+  encountered.  In short, we find that, as with \wcd, the required
+  memory for an execution is linear in the size of the data set --
+  requiring slightly more memory than \wcd\/ and considerably less
+  than \capthree.  We present relevant results in Section D.4 and
+  Figure~S3 of the Supplementary Materials.}
 
-  \mc{We note, as a point of interest, that we can significantly
-    improve the quality of \peace\/ simulation results through the
-    increase of the threshold value (see Supplementary Materials,
-    Section C) -- achieving a significant improved in PEACE
-    sensitivity without an adverse effect on the Jaccard index (i.e. a
-    significant increase in the incorrect merging of clusters).
-    However, the improvement does not carry over to the application of
-    real data (Supplementary Materials, Section E.2), where we observe
-    a significant increase in the incorrect merging of clusters.
-    While this might be acceptable to a user planning to employ an
-    assembl capbable of breaking up the clusters, we have choosed to
-    base our analysis (and set out default values) at the lower threshold.}
+% Appologizes to anyone working with this: the coloring would not
+% presist between columns, and the break point kept changing on me.
+% Modifying the \mc macro will allow for color removal without having
+% to take all of this out.
+\mc{We note, as a point of interest, that we can significantly improve
+  the quality of \peace\/} \mc{simulation} \mc{results} \mc{through} \mc{the} \mc{increase} \mc{of}
+  \mc{the} \mc{threshold} \mc{value} \mc{(see} \mc{Supplementary} \mc{Materials,} \mc{Section~D.5) --}
+  \mc{achieving} \mc{a} \mc{significant} \mc{improvement} \mc{in PEACE sensitivity without an
+  adverse effect on the Jaccard Index.  However, the improvement does
+  not carry over to the application of real data (Supplementary
+  Materials, Section E.2), where we observe a significant increase in
+  the incorrect merging of clusters.  While this might be acceptable
+  to a user planning to employ an assembly tool capable of breaking
+  up the clusters (but unable to re-merge a split cluster), for the
+  purposes of clustering assessment it makes sense base our analysis (and
+  set out default values) at the lower threshold.}
 
@@
     Sequence ESTs sets derived from 100 zebra fish genes (see
-    Supplementary Materials, Section C, for more details).  Blue/Solid
+    Supplementary Materials, Section D, for more details).  Blue/Solid
     = \peace, Green/Dash = \wcd\/ \mc{(version 0.5.1)}, Black/Dot-Dash =
@@
   Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
-  the A076941.fa {\it Arabidopsis thaliana} data set used to test WCD
-  \cite{Hazelhurst08a,Hazelhurst08b}, (see Table~\ref{SangerReal}).
-  We notice essentially identical results for PEACE and WCD in
-  quality, both significantly better than Cap3 in Sensitivity and Type
+  the A076941 {\it Arabidopsis thaliana} data set used to test wcd\/
+  \cite{Hazelhurst08a,Hazelhurst08b} (see Table~\ref{SangerReal}).
+  We notice essentially identical results for \peace\/ and wcd\/ in
+  quality, both significantly better than \capthree\/ in Sensitivity and Type
   1 error rate, while slightly worse in the Jaccard Index and Type 2
-  error.  In runtime we see some inconsistency, with PEACE showing a
-  60\% runtime improvement in the first data set, but requiring 20\%
-  more time in the Arabidopsis data set.  In Section~D.1 of the
-  Supplementary Materials we present runtimes several more sets,
+  error.  In runtime we see some inconsistency, with \peace\/ showing a
+  60\% runtime improvement over \wcd\/ in the first data set, but requiring 20\%
+  more time than \wcd\/ in the Arabidopsis data set.  (We cannot make
+  a reasonable runtime comparison to \capthree, as the runtime of that
+  tool reflects both clustering and assembly.)  In Table~S1 of the
+  Supplementary Materials we present runtimes of several more sets,
   observing that while \peace\/ appears to be significantly 
@@
   250bp and a standard deviation of 17bp), and the Illumina short-read
-  technology (producing reads of exactly 62bp).  Experiements were run
-  using two sets of \wcd\/ parameters (the default paramters, and
+  technology (producing reads of exactly 62bp).  Experiments were run
+  using two sets of \wcd\/ parameters (the default parameters, and
   setting window size to 75 and threshold to 75 -- thus matching the
-  \peace\/ parameters for this size squence), and using default values
-  of \capthree.  Only with \peace were we able to get reasonable
+  equivalent \peace\/ parameters), and using default values
+  of \capthree.  Only with \peace\/ were we able to get reasonable
   results, seeing a sensitivity of 0.871 for the 454 reads and
   0.583 or the shorter Illumina reads, a Jaccard Index of 0.809 and
-  0.372 respectivally, and other comperable figures.  (See Table~S3 in
+  0.372 respectively, and other comparably figures.  (See Table~S3 in
   the Supplementary Materials.)  In short, \peace performs quite well
   for 454 reads and provides some useful information for Illumina
-  reads, while \wcd\. and \capthree are unable to produce working
+  reads, while \wcd\/ and \capthree\/ are unable to produce working
   results.}
@@
   high-throughput clustering of transcript fragments capable of
-  dealing with sequences as short as 50 bases.  \peace\/, available at
-  \href{http://www.peace-tools.org}{www.peace-tools.org}, is
+  dealing with sequences as short as 50 bases.  \peace\/ is
   open-source and managed through a user friendly GUI that enables
   both local and remote installation and execution in sequential or
-  parallel mode.  Based on a novel minimum-spanning-tree based
+  parallel mode.  Based on a novel minimum spanning tree based
   algorithm for the clustering of the fragments by gene association,
-  \peace\/ shows significant improvement in sensitivity, over the
-  competing \wcd\/ tool \cite{Hazelhurst08a} when applied to NGS
-  reads, matches \wcd\/ when applied to Traditional Sanger Sequencing
-  output, and shows an order of magnitude in improvement over the
+  \peace\/ shows significant improvement in sensitivity over the
+  competing \wcd\/} \mc{tool} \cite{Hazelhurst08a} \mc{when} \mc{applied} \mc{to} \mc{NGS}
+  \mc{reads,} \mc{matches} \mc{\wcd\/} \mc{when} \mc{applied} \mc{to} \mc{Sanger} \mc{sequencing}
+  \mc{output,} \mc{and} \mc{shows} \mc{an} \mc{order} \mc{of} \mc{magnitude} \mc{in improvement over the
   clustering performed in the course of assembly by the \capthree\/
@@
 poly(A)/(T) tails), and the presence of low-complexity repeats can
-cause similar effects in these clustering tools.  The problems can be
+cause similar effects in these clustering tools.  These problems can be
 handled through the application of the assembler, and the ability to
@@
 
-Peace can be downloaded from
+\mc{Peace can be downloaded from
 \href{http://www.peace-tools.org}{www.peace-tools.org}, where we are
 committed to keep maintaining and improving the tool in the future.
-Meanwhile, we are developing our own assembly tool that can seamlessly
-integrate with PEACE.  The underlying modular design of PEACE offers
-users many possibilities to expand and incorporate the MST algorithm
-for other bioinformatics application.
+Meanwhile, we are developing our own MST-based assembly tool that can
+seamlessly integrate with PEACE.  The underlying modular design of
+PEACE offers users many possibilities to expand and incorporate the
+MST algorithm for other bioinformatics application.}
 
@@
 % LocalWords:  NGS polyA set's GHz Xeon Bikun Arabidopsis JVM Jaccard LV GUI's
-% LocalWords:  Data Set thaliana data set EasyCluster MetaSim
+% LocalWords:  Data Set thaliana data set EasyCluster MetaSim Illumina

### 2010-04-21 papers/NAR_web_server_2009/paper.sup.tex
Final version of supplementary materials

@@
 
-Once the MST is has been calculated, our last step is to remove all
+Once the MST has been calculated, our last step is to remove all
 edges exceeding a threshold weight $T$, taking the resulting
-components as our clusters.  \textsc{wcd}, when faced with a similar
+components as our clusters.  \wcd, when faced with a similar
 challenge, sets the threshold at $T=40$, hypothesizing that EST pairs
@@
 
-\mc{By restricting $d^2$ comparisons to within a group, we minimize the
-effects of the heterogeneous sequences lengths by restricting our
-comparisons to only sequences of similar lengths.  By not restricting
-ourselves to a single window size, we allow ourselves to, when possible, avoid the
-performance hit of using a small window.  As we are not comparing
-sequences from different groups we bypass the question of how to
-compare a long sequence against a short one.  And because the
-groups overlap in content, we still get implicit comparisons between
-long and short sequences in terms of the resulting path of the minimum
-spanning tree that makes use of these ``linking'' nodes.}
+\mc{By restricting $d^2$ comparisons to within a group, we minimize
+  the effects of the heterogeneous sequences lengths by restricting
+  our comparisons to only sequences of similar lengths.  By not
+  restricting ourselves to a single window size, we allow ourselves
+  to, when possible, avoid the performance hit of using a small
+  window.  As we are not comparing sequences from groups corresponding
+  to radically different sizes we bypass the question of how to
+  compare a long sequence against a short one.  And because we allow
+  for the comparison of sequences in neighboring groups, we still get
+  implicit comparisons between long and short sequences in terms of
+  the resulting path of the minimum spanning tree that makes use of
+  these ``linking'' nodes.}
 
@@
 \begin{itemize}
-\item Less than 150 ($w=50$, $T=50$, $t=20$, $u=4$).
-\item Between 150 and 400 ($w=75$, $T=75$, $t=35$, $u=6$)
-\item Greater than 400 ($w=100$, $T=100$, $t=50$, $t=50$).
+\item Less than 150bp ($w=50$, $T=50$, $t=20$, $u=4$).
+\item Between 150bp and 400bp ($w=75$, $T=75$, $t=35$, $u=6$)
+\item Greater than 400bp ($w=100$, $T=100$, $t=50$, $t=50$).
 \end{itemize}}
@@
 (Figure~3).  Sensitivity is the ratio for true
-positives to actual positives ($tp / (tp + fn)$) -- the fraction of
-sequences from the same transcript that are identified as such.  The
-{\it Jaccard} index combines both sensitivity and specificity,
-calculated as the ratio of true positives to the sum of true
+positives to actual positives ($tp / (tp + fn)$) -- the} \mc{fraction} \mc{of}
+\mc{sequences} \mc{from} \mc{the} \mc{same} \mc{transcript} \mc{that} \mc{are} \mc{identified} \mc{as} \mc{such.}  \mc{The}
+\mc{{\it Jaccard}} \mc{index} \mc{combines} \mc{both} \mc{sensitivity} \mc{and} \mc{specificity,}
+\mc{calculated} \mc{as} \mc{the} \mc{ratio} \mc{of} \mc{true} \mc{positives} \mc{to} \mc{the sum of true
 positives, true negatives, and false positives ($tp / (tp+fn+fp)$)
-\cite{Hazelhurst08a}.  Notably, when we removed duplicated genes
-before generating the simulated EST sets, we found all three tools to
-have perfect specificity: no false positives.}
+\cite{Hazelhurst08a}.}
 
@@
   divergence.  Estimates averaged over 30 trials; one trial consists
-  of taking a random gene zebra-fish gene, copying it and stochastically changing
+  of taking a random gene zebra fish gene, copying it and stochastically changing
   bases at the specified rate, then using the two genes as the bases
@@
   confidence intervals.}\label{dups}
+\end{figure}
+
+\subsection{Runtime and Memory Footprint}
+
+\mc{Runtime of the tool is analyzed in Figure~4 of the main paper.  In
+Figure~\ref{memory} we present a plot of the memory footprint of each tool
+as a function of input size.  Data points were generated by the application of
+each tool to simulated sets of a specified size (generated by ESTSim),
+measuring the maximum amount of memory required for the run.  We see that while
+all tools require memory linear in the input size, \wcd\/ requires
+slightly less than does \peace, while both require significantly less
+memory than \capthree.  In general we find the footprint of \peace\/
+to be relativily small, as compared to other clustering tools
+(discussed in {\it Hazelhurst et al.} \cite{Hazelhurst08a}).}
+
+\begin{figure}[tbp]
+\centerline{
+\includegraphics[scale=0.35]{pics.d/memory.pdf}
+}
+\caption{Plots of memory footprint against sequence set size for a
+  single processor run on simulated data for each tool: \peace\/
+  (blue), \wcd\/ (green) and \capthree\/ (balck), along with the
+  lesat-squares fit for each.  For each fit we have an associated
+  regression coefficient of $r > 0.99$.}\label{memory}
 \end{figure}
@@
   $T$ threshold for dismissing MST edges, we have set it to the same
-  default value when used for Sanger sequences.  However, we did
-  experiment with higher values, to interesting results.  A basic
-  analysis indicated that a $T$ value as high as 130 should not have a
+  default value when used for Sanger} \mc{sequences.}  \mc{However,} \mc{we} \mc{did}
+  \mc{experiment} \mc{with} \mc{higher} \mc{values,} \mc{with} \mc{interesting} \mc{results.  An
+  analysis of the preformance of $d^2$ indicated that a $T$ value as high as 130 should not have a
   significant impact on its effectiveness as a discriminator.  And in
   fact this was born out in simulation, as shown in
-  Figure~\label{SiJiT1T2130}.  In contrast with the equivalent results
+  Figure~\ref{SeJiT1T2130}.  In contrast with the equivalent results
   for a threshold of $T=40$, as shown in the manuscript, we see
@@
 
-\mc{
-  Most importantly, while the higher threshold appeared to be better
-  in theory and in analysis, it did not hold in practice.  In
-  Table~\ref{SangerReal130} we show the match to the manuscript
-  Table~1, and observe the poor performance of
-  \peace\/ when looking at the Jaccard index.  It is clear that in the
-  real data, the level of similarity is considerably higher then in the
-  simulations, thus making the reduced specificity of the higher
-  threshold considerably more of a problem.
-}
+\mc{ Most importantly, while the higher threshold appeared to be
+  better in theory and in analysis, the result did not hold in
+  practice.  In Table~\ref{SangerReal130} we show the figures
+  corresponding to the manuscript Table~1 when raising the threshold
+  to $130$, and observe the poor performance of \peace\/ when looking
+  at the Jaccard index.  It is clear that in the real data, the level
+  of transcript similarity is considerably higher then in the simulations, thus
+  making the reduced specificity of the higher threshold considerably
+  more of a problem.  }
 
@@
   \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
-    Type 1 error using the elevated threhold ($T=130$).  Based on the average
-    over 30 simulated Sanger Sequence ESTs sets derived from 100 zebra
-    fish genes (see Supplementary Materials, Section C, for more
-    details).  Blue/Solid = \peace, Green/Dash = \wcd, Black/Dot-Dash
-    = \capthree; vertical tics = 95\% confidence intervals on
-    estimates.  Intervals are not presented for Type 1 error due to
-    the effective lack of variance.}\label{SeJiT1T2130}
+    Type 1 error using the elevated threhold ($T=130$), providing a
+    contrast against Figure 3 ($T=40$).  Based on the average over 30
+    simulated Sanger Sequence ESTs sets derived from 100 zebra fish
+    genes.  Blue/Solid = \peace, Green/Dash = \wcd, Black/Dot-Dash =
+    \capthree; vertical tics = 95\% confidence intervals on estimates.
+    Intervals are not presented for Type 1 error due to the effective
+    lack of variance.}\label{SeJiT1T2130}
 \end{figure}
@@
   divergence.  Estimates averaged over 30 trials; one trial consists
-  of taking a random zebra-fish gene, copying it and stochastically changing bases
+  of taking a random zebra fish gene, copying it and stochastically changing bases
   at the specified rate, then using the two genes as the bases for
@@
 \hline
- Human Cluster & 17733 & 11.7 Mb & 293 & 804 \\
+ Human Cluster & 17,733 & 11.7 Mb & 293 & 804 \\
 \hline
@@
   raising the threshold $T$ to 130 (see previous discussion in
-  Section~\ref{sim_thresh}).  While doing this appeared to be fairly
-  effective in simulation, it did not hold up in practice.  We see
+  Section~\ref{sim_thresh}).  While raising $T$ is  
+  effective in simulations, it did not hold up in practice.  We see
   essentially no change in the sensitivity, but significant changes in
@@
 
+\mc{In Table~\ref{SRSim} we present the results of application of the
+tools to \metasim\/ generated simulated short read data, as discussed in
+the results section of the paper.}
+
 \begin{table}[tbp]
@@
 \hline
-\hline   & \peace & 0.871 & 0.809 & 0.186 & 0.049 & 129 & 4 & 129 \\
-\cline{2-9}  MetaSim 454 & \wcd(1) & 0.073 & 0.073 & 0.014 & 0.082 & 7028.500 & 5296.533 & 74 \\
-\cline{2-9}  Simulation & \wcd(2) &  0.020 & 0.020 & 0.008 & 0.198 & 13085 & 10610 & 64 \\
-\cline{2-9}  & \capthree & 0.005 & 0.005 & 0.005 & 0.565 & 19222 & 15425 & 132 \\
+\hline   & PEACE & 0.871 & 0.809 & 0.186 & 0.049 & 129 & 4 & 129 \\
+\cline{2-9}  MetaSim 454 & WCD(1) & 0.073 & 0.073 & 0.014 & 0.082 & 7028.500 & 5296.533 & 74 \\
+\cline{2-9}  Simulation & WCD(2) &  0.020 & 0.020 & 0.008 & 0.198 & 13085 & 10610 & 64 \\
+\cline{2-9}  & Cap3 & 0.005 & 0.005 & 0.005 & 0.565 & 19222 & 15425 & 132 \\
 \hline
-\hline Metasim  & \peace & 0.583 & 0.372 & 0.230 & 0.111 & 392.914 & 149.257 & 3454 \\
-\cline{2-9} Illumina  & \capthree & 0.006 & 0.006 & 0.002 & 1.766 & 51663.057 & 30107.429 & 291 \\
+\hline Metasim  & PEACE & 0.583 & 0.372 & 0.230 & 0.111 & 392.914 & 149.257 & 3454 \\
+\cline{2-9} Illumina  & Cap3 & 0.006 & 0.006 & 0.002 & 1.766 & 51663.057 & 30107.429 & 291 \\
 \hline

### 2010-04-21 papers/NAR_web_server_2009/paper.sup.tex
Should be the final version of all components of the resubmission.

@@
 to be relativily small, as compared to other clustering tools
-(discussed in {\it Hazelhurst et al.} \cite{Hazelhurst08a}).}
+(discussed in {\it Hazelhurst et al.} \cite{Hazelhurst08a}).
+
+All runs were conducted on 3.0 GHz Intel Xeon EM64T CPUs with 2 MB
+caches and a 800 MHz front side bus, model number Xeon LV 3.0 (2005)}
 

### 2010-04-21 papers/NAR_web_server_2009/paper.tex
Should be the final version of all components of the resubmission.

@@
 
-\history{Received Feb. 13, 2010 \\ Resubmitted April 20, 2010}
+\history{Received Feb. 13, 2010 \\ Resubmitted April 21, 2010}
 
@@
 and accurately find clusters of ESTs expressed from the same gene
-without reference to a sequenced genome.  Compared against \wcd, the
-leading clustering tool in the literature \cite{Hazelhurst08a}, as well as
-other tools designed for the same purpose 
+without reference to a sequenced genome.  \mc{Compared against
+  clustering tool in the literature
   \cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
-  Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be both more
-sensitive and more robust to sequencing error without sacrificing
-runtime.  Nor are any of these tools designed for the ease of installation
-and use that \peace\/ provides.
+    Pertea2003,Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be of
+  equivalent quality to the \wcd\/ tool \cite{Hazelhurst08a}, more
+  sensitive and robust than the other tools, and more user friendly
+  than any published clustering tool.  Nor are any of} \mc{these tools
+  designed for the ease of installation and use that \peace\/
+  provides.}
 
@@
   multiple processor when holding the EST/processor ratio constant
-  (ranging from a 65\% improvement for two processors too a 17\%
+  (ranging from a 65\% improvement for two processors to a 17\%
   improvement for 12 processors).}
 
-\mc{We further note, as a point of interest that, we can significantly
-  improve the quality of \peace simulation results through the
-  increase of the threshold value (see Supplementary Materials,
-  Section C) -- achieving a significant improved in PEACE sensitivity
-  without an adverse effect on the Jaccard index (i.e. a significant
-  increase in the incorrect merging of clusters).  However, the
-  improvement does not carry over to the application of real data,
-  where we observe a significant drop in the Jaccard Index.  Due to
-  this drop we have opted to leave the lower threshold value as the
-  default, but users might consider the higher value, thus decreasing
-  the risk of incorrectly splitting clusters, if planning on
-  using on the subsequent use of a assembly tool that capable of
-  breaking up the incorrectly merged clusters during transcript reconstruction.}
+\mc{In the course of our simulations we also investigated the memory
+  footprint -- of particular concern in the large data sets commonly
+  encountered.  In short, we find that, as with \wcd, the required
+  memory for an execution is linear in the size of the data set --
+  requiring slightly more memory than \wcd\/ and considerably less
+  than \capthree.  We present relevant results in Section D.4 and
+  Figure~S3 of the Supplementary Materials.}
+
+% Appologizes to anyone working with this: the coloring would not
+% presist between columns, and the break point kept changing on me.
+% Modifying the \mc macro will allow for color removal without having
+% to take all of this out.
+\mc{We note, as a point of interest, that we can significantly improve
+  the quality of \peace\/} \mc{simulation} \mc{results} \mc{through} \mc{the} \mc{increase} \mc{of}
+  \mc{the} \mc{threshold} \mc{value} \mc{(see} \mc{Supplementary} \mc{Materials,} \mc{Section~D.5) --}
+  \mc{achieving} \mc{a} \mc{significant} \mc{improvement} \mc{in PEACE sensitivity without an
+  adverse effect on the Jaccard Index.  However, the improvement does
+  not carry over to the application of real data (Supplementary
+  Materials, Section E.2), where we observe a significant increase in
+  the incorrect merging of clusters.  While this might be acceptable
+  to a user planning to employ an assembly tool capable of breaking
+  up the clusters (but unable to re-merge a split cluster), for the
+  purposes of clustering assessment it makes sense base our analysis (and
+  set out default values) at the lower threshold.}
 
@@
      estimates.  All runs were done on a 3.0 GHz Intel Xeon EM64T CPU with a 2 MB
-     cache and 800 MHz front side bus, model number Xeon LV 3.0 (2005).}\label{seq_parallel}
+     cache and a 800 MHz front side bus, model number Xeon LV 3.0 (2005).}\label{seq_parallel}
 \end{figure}
@@
   \metasim\/ are sequence generation and error models corresponding to
-  several technologies, including the 454 short-read technology.
-  Leaving \capthree\/ and \wcd\/ at their default values, but
-  employing the {\it adaptive $d^2$} strategy for short read sequences
-  (see Section~C in the supplementary materials), we find a that
-  \peace\/ performs significantly better than the other two tools on
-  454 data using the default \metasim\/ model (with sequence sizes
-  ranging in size from 200bp to 370bp, at an average size of 250bp and
-  a standard deviation of 17bp).  Over thirty runs, we see an average
-  sensitivity 0.884 for \peace, 0.331\/ for \wcd, and 0.004 for
-  \capthree.  For the Jaccard Index we get essentially the same
-  numbers.  For 100 transcript \peace\/ creates 128 clusters (as
-  opposed to 6,370 and 19245 clusters for \wcd\/ and \capthree), and
-  creates only 3 singletons (as opposed to 5,850 and 15432).  We do
-  note that it is likely that the \wcd results could be considerably
-  improved by the adjustment of a few parameters, but were unable to
-  determine how to do that.  (Runtime of \peace\/ was somewhat slower
-  than that of \wcd, but given the significant improvement in quality
-  of results this is unimportant.)}
-
-\begin{table*}[tbp]
-\begin{center}
-\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
-\hline
- & 
-& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
-  error \end{tabular} & \begin{tabular}{c} Type 2 \\
-  error \end{tabular} &
-\begin{tabular}{c} Number \\ of  \\
-    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
-    Singletons \end{tabular}
-    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
-\hline
-\hline EasyCluster Human & \peace & 0.871 & 0.809 &
-\cline{2-9} Benchmark & \wcd &  0.020 & 0.020 &
-\cline{2-9} (111 Genes) & \capthree & 0.005 & 0.004 &
-\hline
-\hline WCD A076941 & \peace & 0.932 & 0.475 & 0.351 & 0.027 & 18825 & 8951 & 1166\\
-\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
-\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
-\mbox{NA} \\
-\hline
-\end{tabular}
-\end{center}
-\caption{Comparisons of runs on the EasyCluster human Benchmark
-  Data Set and the WCD A076941 Arabidopsis thaliana data set using the
-  standard quality measurements.}\label{SangerReal}
-\end{table*}
+  several technologies, including the 454 short-read technology
+  (producing reads ranging in size from 200bp to 370bp with a mean of
+  250bp and a standard deviation of 17bp), and the Illumina short-read
+  technology (producing reads of exactly 62bp).  Experiments were run
+  using two sets of \wcd\/ parameters (the default parameters, and
+  setting window size to 75 and threshold to 75 -- thus matching the
+  equivalent \peace\/ parameters), and using default values
+  of \capthree.  Only with \peace\/ were we able to get reasonable
+  results, seeing a sensitivity of 0.871 for the 454 reads and
+  0.583 or the shorter Illumina reads, a Jaccard Index of 0.809 and
+  0.372 respectively, and other comparably figures.  (See Table~S3 in
+  the Supplementary Materials.)  In short, \peace performs quite well
+  for 454 reads and provides some useful information for Illumina
+  reads, while \wcd\/ and \capthree\/ are unable to produce working
+  results.}
 
@@
 
-Peace can be downloaded from
+\mc{Peace can be downloaded from
 \href{http://www.peace-tools.org}{www.peace-tools.org}, where we are
-committed to maintaining and improving the tool.  The underlying PEACE
-architecture is configured to support a variety of tasks, including the
-possibility of a minimum spanning tree based assembly tool we hope to
-add to the tool in a future study.
+committed to keep maintaining and improving the tool in the future.
+Meanwhile, we are developing our own MST-based assembly tool that can
+seamlessly integrate with PEACE.  The underlying modular design of
+PEACE offers users many possibilities to expand and incorporate the
+MST algorithm for other bioinformatics application.}
 

### 2010-04-21 EAST/C++/Param.h
Add an option which can specify whether or not generating output files for SNP analysis. If the option is set to be 1, a series of output files with each corresponding to one consensus sequence will be produced for SNP analysis.

@@ -69,4 +69,9 @@ const int BAND_WIDTH_SW=100;
 //1-use, 0-not use. Use bounded version may reduce the precision.
 const int USE_BOUNDED_SW=0;
 
+//If generate output files for SNP analysis. 0: not, 1: generate.
+const int OUTPUT_SNP=0;
+//The prefix of the output files for SNP analysis. The number of SNP output files is equal to the number of consensus sequences.
+//They are named as snaAnalysis.0, snaAnalys.1, ...
+//std::string SNP_FILE="snpAnalysis";
 #endif

### 2010-04-21 EAST/C++/Reconstruction.cpp

@@ -40,6 +40,7 @@ bool operator< (const LeftEnd& k1, const LeftEnd& k2) {
 Reconstruction::Reconstruction() {
 	g = NULL;
 	incNodes = NULL;
+	idxOfSNPFile = 0;
 }
 
 Reconstruction::Reconstruction(Graph* graph, vector<SixTuple*> align, vector<SixTuple*> leftEnds, InclusionNodes* inc, const std::string& con, const std::string& sing, const std::string& numF, int longestEstLen) {
@@ -51,6 +52,7 @@ Reconstruction::Reconstruction(Graph* graph, vector<SixTuple*> align, vector<Six
 	singletonFileName = sing;
 	numOfUsedESTsFileName = numF;
 	COMPARISON_LENGTH = longestEstLen;
+	idxOfSNPFile = 0;
 
 	usedNodes = vector<int> (g->graphNodes.size(), 0);
 
@@ -562,6 +564,11 @@ vector<string> Reconstruction::reconstructSeq(vector<StartPos>& a) {
 	tConsensus = tConsensus.substr(0, totalLen);
 
 	ret[0] = replace(tConsensus, "P", "");
+
+	if (OUTPUT_SNP == 1) { //write output files for SNP analysis
+		writeToSNPFile(bases, totalLen);
+	}
+
 	//release bases
 	for (int i=0; i<bases.size(); i++) {
 		delete bases[i];
@@ -584,6 +591,20 @@ int Reconstruction::getNumUsedNodes(vector<StartPos>& a) {
 	return addedNodes.size();
 }
 
+void Reconstruction::writeToSNPFile(vector<SingleBase*> bases, int totalLen) {
+	if (totalLen <= 0) return;
+	stringstream ss;
+	ss << idxOfSNPFile++;
+	string outFileName = "snpAnalysis." + ss.str();
+	ofstream outFile1;
+	outFile1.open(outFileName.c_str(),ios::trunc);
+
+	outFile1 << "A\tG\tC\tT" << endl;
+	for (int i=0; i<totalLen; i++) {
+		outFile1 << bases[i]->numA << "\t" << bases[i]->numG << "\t" << bases[i]->numC << "\t" << bases[i]->numT << endl;
+	}
+	outFile1.close();
+}
 
 string Reconstruction::getCurConsensus(vector<SingleBase*> bases) {
 	int len = bases.size();

### 2010-04-21 peace/EAST/C++/Reconstruction.h
Add an option which can specify whether or not generating output files for SNP analysis. If the option is set to be 1, a series of output files with each corresponding to one consensus sequence will be produced for SNP analysis.

@@ -70,6 +70,7 @@ public:
 
 private:
 	int COMPARISON_LENGTH;
+	int idxOfSNPFile; //used to record the index of the current output file for SNP analysis
 	void printConsensus();
 	std::vector<std::string> reconstruct();
 	std::vector<std::string> reconstuctForEachCluster();
@@ -85,6 +86,7 @@ private:
 	std::string replace(std::string str, const std::string& old, const std::string& newstr);
 	std::string printLeftEndInfo(int leftEnd);
 	std::vector<std::string> reconstructSeq(std::vector<StartPos>& a);
+	void writeToSNPFile(std::vector<SingleBase*> bases, int totalLen);
 
 	DefGraph graphForPrim; //this is a graph which is used for generating MST. It is generated once in genDGraph(), then used multiple times.
 };

### 2010-05-10 peace/papers/NAR_web_server_2009/paper.tex
Final version of supplementary materials
Modification of manuscript for final submission

@@
 \definecolor{red}{rgb}{1,0,0}
-\newcommand{\mc}[1]{\textcolor{red}{#1}}
+\newcommand{\mc}[1]{#1}
 
@@
 \newcommand{\metasim} {{\small MetaSim}}
+\newcommand{\tgicl} {{\small TGICL}}
 
@@
     identical to those of the \wcd\/ tool when applied to Sanger
-    sequences, significantly improved results over \wcd\/ when applied
+    sequences, significantly improved results over \wcd\/ and \tgicl\/ when applied
     the products of Next Generation Sequencing Technology, and
@@
 
-In the past, access to a transcriptome sequence was primarily
-through the use of Expressed Sequence Tags (ESTs), single-pass cDNA
-sequences derived from transcribed mRNAs and sequenced by Sanger
-Sequencing technology.  More recently, Next Generation Sequencing
-(NGS) technology has begun to rapidly replace Sanger Sequencing,
-allowing for more comprehensive coverage of the transcriptome.  For
+In the past, access to a transcriptome sequence was primarily through
+the use of Expressed Sequence Tags (ESTs), single-pass cDNA sequences
+derived from transcribed mRNAs and sequenced by Sanger Sequencing
+technology.  More recently, Next Generation Sequencing (NGS)
+technology has begun to rapidly replace Sanger Sequencing.  For
 example, ESTs now being added to the GenBank dbEST are increasingly
@@
 enables the sequencing of novel and rare transcripts at a considerably
-higher rate of coverage than Sanger Sequencing
-\cite{Cheung2006,Emrich2007}.  From a computational perspective, this
+higher \cite{Cheung2006,Emrich2007}.  From a computational perspective, this
 is a mixed blessing: while NGS provides immense quantities of new
@@
 Given a set of transcript fragments sampled from across the genome, a
-necessary first step of the set's analysis is that of clustering: separating
+necessary first step of the set's analysis is clustering: separating
 the fragments according to the transcript from which they were
@@
 
-However, clustering is a computationally challenging problem.  Even
-with the smaller number of ESTs produced using Sanger Sequencing, the
-runtime and memory requirements to cluster on the basis of pair-wise
-sequence alignments make such an approach infeasible in practice.  The
-much larger data-set size produced by NGS technologies exacerbates
-this problem.  To deal with this, \peace\/ combines our own version of
-the $d^2$ alignment-free sequence distance function \cite{Hide94} and
-the concept of a {\it minimum spanning tree} \cite{Prim57} to quickly
-and accurately find clusters of ESTs expressed from the same gene
-without reference to a sequenced genome.  \mc{Compared against
-  clustering tool in the literature
+Clustering is a computationally challenging problem; the runtime and
+memory requirements to cluster on the basis of pair-wise sequence
+alignments make such an approach infeasible in practice.  To deal with
+this, \peace\/ combines our own version of the $d^2$ alignment-free
+sequence distance function \cite{Hide94} and the concept of a {\it
+  minimum spanning tree} \cite{Prim57} to quickly and accurately find
+clusters of ESTs expressed from the same gene without reference to a
+sequenced genome.  Compared against clustering tool in the literature
 \cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
-    Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be of
-  equivalent quality to the \wcd\/ tool \cite{Hazelhurst08a}, more
-  sensitive and robust than the other tools, and more user friendly
-  than any published clustering tool.  Nor are any of} \mc{these tools
-  designed for the ease of installation and use that \peace\/
-  provides.}
+  Pertea2003,Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be
+of equivalent quality to the \wcd\/ and \tgicl tools
+\cite{Hazelhurst08a,Pertea2003}, and more sensitive and robust than
+other tools.  From the user perspective, o tool in the literature can
+match \peace for ease of instillation, use, or the post-clustering
+analysis tools \peace provides.
 
@@
 
-The \peace\/ GUI can be launched by downloading and executing the
+The \peace\/ GUI can be launched by through the
 Jar file available on the \peace\/ website
-(\href{http://www.peace-tools.org}{www.peace-tools.org}) to any
+(\href{http://www.peace-tools.org}{www.peace-tools.org}) on any
 machine running the standard Java Virtual Machine (JVM).  Once
@@
 \end{itemize}
-Extensive documentation for the tool has been posted on the \peace\/ website, as well as links to several tutorial videos
-demonstrating \peace\/ use and capabilities.
+Extensive documentation for the tool has been posted on the \peace\/
+website, as well as links to several tutorial videos demonstrating
+\peace\/ use and capabilities.
 
@@
      estimates.  All runs were done on a 3.0 GHz Intel Xeon EM64T CPU with a 2 MB
-     cache and 800 MHz front side bus, model number Xeon LV 3.0 (2005).}\label{seq_parallel}
+     cache and a 800 MHz front side bus, model number Xeon LV 3.0 (2005).}\label{seq_parallel}
 \end{figure}
 
-\mc{In applying the tools to real data Sanger data, we used the Human
+While \peace\/ clearly matches \wcd\/ on long-read sequence, comparisons
+against \tgicl\/ \cite{Pertea2003}  are more complicated.  In short, we find that when
+parameters are picked appropriately, \tgicl\/ is more sensitive at lower
+error rate -- but \peace\/ is more robust to error, in all cases
+has an improved Type 2 error rates and is better able to
+distinguish duplicated genes.  (See the Supplementary Materials for a
+more detailed analysis.)
+
+In applying the tools to real data Sanger data, we used the Human
 Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
 the A076941 {\it Arabidopsis thaliana} data set used to test wcd\/
-  \cite{Hazelhurst08a,Hazelhurst08b} (see Table~\ref{SangerReal}).
-  We notice essentially identical results for \peace\/ and wcd\/ in
-  quality, both significantly better than \capthree\/ in Sensitivity and Type
-  1 error rate, while slightly worse in the Jaccard Index and Type 2
-  error.  In runtime we see some inconsistency, with \peace\/ showing a
-  60\% runtime improvement over \wcd\/ in the first data set, but requiring 20\%
-  more time than \wcd\/ in the Arabidopsis data set.  (We cannot make
-  a reasonable runtime comparison to \capthree, as the runtime of that
-  tool reflects both clustering and assembly.)  In Table~S1 of the
-  Supplementary Materials we present runtimes of several more sets,
-  observing that while \peace\/ appears to be significantly 
-  faster on the smaller sets, \wcd\/ does overtake it for larger
-  sets.}
+\cite{Hazelhurst08a,Hazelhurst08b} (see Table~\ref{SangerReal}).  We
+notice essentially identical results for \peace\/ and wcd\/ in
+quality, both significantly better than \capthree\/ in Sensitivity and
+Type 1 error rate, while slightly worse in the Jaccard Index and Type
+2 error.  In runtime we see some inconsistency, with \peace\/ showing
+a 60\% runtime improvement over \wcd\/ in the first data set, but
+requiring 20\% more time than \wcd\/ in the Arabidopsis data set.  (We
+cannot make a reasonable runtime comparison to \capthree, as the
+runtime of that tool reflects both clustering and assembly.)  In
+Table~S1 of the Supplementary Materials we present runtimes of several
+more sets, observing that while \peace\/ appears to be significantly
+faster on the smaller sets, \wcd\/ does overtake it for larger sets.
 
@@
 handled through the application of the assembler, and the ability to
-apply any assembler to small cluster (as opposed to the data set as a
+apply any assembler to small clusters (as opposed to the data set as a
 whole) results in a significant reduction in overall assembly time.
 
-\mc{Peace can be downloaded from
+Peace can be downloaded from
 \href{http://www.peace-tools.org}{www.peace-tools.org}, where we are
@@
 PEACE offers users many possibilities to expand and incorporate the
-MST algorithm for other bioinformatics application.}
+MST algorithm for other bioinformatics application.
 

### 2010-05-10 papers/NAR_web_server_2009/paper.sup.tex
Update of manuscript for final submission

--- 2010-04-21 papers/NAR_web_server_2009/paper.sup.tex
@@
 \definecolor{red}{rgb}{1,0,0}
-\newcommand{\mc}[1]{\textcolor{red}{#1}}
-%\newcommand{\mc}[1]{#1}   % Replace above line with this to eliminate all coloring
+%\newcommand{\mc}[1]{\textcolor{red}{#1}}
+\newcommand{\mc}[1]{#1}   % Replace above line with this to eliminate all coloring
 
@@
 
+\mc{All runs were conducted on 3.0 GHz Intel Xeon EM64T CPUs with 2 MB
+caches and a 800 MHz front side bus, model number Xeon LV 3.0 (2005)}
+
 \begin{figure}[tbp]
@@
 
-\begin{table}[tbp]
+\begin{table}[htbp]
 \begin{center}
@@
 \hline
-\hline   & PEACE & 0.871 & 0.809 & 0.186 & 0.049 & 129 & 4 & 129 \\
-\cline{2-9}  MetaSim 454 & WCD(1) & 0.073 & 0.073 & 0.014 & 0.082 & 7028.500 & 5296.533 & 74 \\
-\cline{2-9}  Simulation & WCD(2) &  0.020 & 0.020 & 0.008 & 0.198 & 13085 & 10610 & 64 \\
-\cline{2-9}  & Cap3 & 0.005 & 0.005 & 0.005 & 0.565 & 19222 & 15425 & 132 \\
+\hline   454 & PEACE & 0.871 & 0.809 & 0.186 & 0.049 & 129 & 4 & 129 \\
+\cline{2-9}    Simulation & WCD(1) & 0.459 & 0.448 & 0.946 & 0.003 & 599 & 235 & 76 \\
+\cline{2-9}  & TGICL & 0.297 & 0.288 & 0.984 & 0.005 & 886 & 0 & 68 \\
 \hline
-\hline Metasim  & PEACE & 0.583 & 0.372 & 0.230 & 0.111 & 392.914 & 149.257 & 3454 \\
-\cline{2-9} Illumina  & Cap3 & 0.006 & 0.006 & 0.002 & 1.766 & 51663.057 & 30107.429 & 291 \\
+\hline Illumina         & PEACE & 0.384 & 0.343 & 0.230 & 0.018 & 673  & 299 & 3463 \\
+\cline{2-9} Simulation  & TGICL & 0.131 & 0.127 & 0.977 & 0.008 & 1522 & 0   & 373 \\
 \hline
@@
   \metasim \cite{Richter2008}; each figure is an average over 30 runs.
-  For \wcd(1) we use wcd default parameters, while for \wcd(2) we set
-  it to match the \peace\/ parameters ($w=75$, $T=60$); \wcd\/ was
-  unable to return any results for Illumina data, hence has been
-  omitted.  Each 454 simulation produced 25,000 reads from the 100
-  zebra-fish transcripts, ranging  in size from 200bp to 270bp (aveage = 250bp,
+  For \wcd we use parameters suggested by the authors  (``-H 1 -T 160''). \wcd\/ was
+  unable to return any reasonable results for Illumina data, while
+  \capthree\/ was unable to return results for either sumulation.  Each
+  454 simulation produced 25,000 reads from the 100 zebra-fish
+  transcripts, ranging  in size from 200bp to 270bp (aveage = 250bp, 
   standard deviation = 17bp).  Each Illumina simulation produce 150K

### 2010-05-10 papers/NAR_web_server_2009/paper.tex
Update of manuscript for final submission

@@
 
-However, clustering is a computationally challenging problem.  Even
-with the smaller number of ESTs produced using Sanger Sequencing, the
-runtime and memory requirements to cluster on the basis of pair-wise
-sequence alignments make such an approach infeasible in practice.  The
-much larger data-set size produced by NGS technologies exacerbates
-this problem.  To deal with this, \peace\/ combines our own version of
-the $d^2$ alignment-free sequence distance function \cite{Hide94} and
-the concept of a {\it minimum spanning tree} \cite{Prim57} to quickly
-and accurately find clusters of ESTs expressed from the same gene
-without reference to a sequenced genome.  \mc{Compared against
-  clustering tool in the literature
+Clustering is a computationally challenging problem; the runtime and
+memory requirements to cluster on the basis of pair-wise sequence
+alignments make such an approach infeasible in practice.  To deal with
+this, \peace\/ combines our own version of the $d^2$ alignment-free
+sequence distance function \cite{Hide94} and the concept of a {\it
+  minimum spanning tree} \cite{Prim57} to quickly and accurately find
+clusters of ESTs expressed from the same gene without reference to a
+sequenced genome.  Compared against clustering tool in the literature
 \cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
-    Pertea2003,Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be of
-  equivalent quality to the \wcd\/ tool \cite{Hazelhurst08a}, more
-  sensitive and robust than the other tools, and more user friendly
-  than any published clustering tool.  Nor are any of} \mc{these tools
-  designed for the ease of installation and use that \peace\/
-  provides.}
+  Pertea2003,Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be
+of equivalent quality to the \wcd\/ and \tgicl tools
+\cite{Hazelhurst08a,Pertea2003}, and more sensitive and robust than
+other tools.  From the user perspective, o tool in the literature can
+match \peace for ease of instillation, use, or the post-clustering
+analysis tools \peace provides.
 
@@
 
-\mc{In applying the tools to real data Sanger data, we used the Human
+While \peace\/ clearly matches \wcd\/ on long-read sequence, comparisons
+against \tgicl\/ \cite{Pertea2003}  are more complicated.  In short, we find that when
+parameters are picked appropriately, \tgicl\/ is more sensitive at lower
+error rate -- but \peace\/ is more robust to error, in all cases
+has an improved Type 2 error rates and is better able to
+distinguish duplicated genes.  (See the Supplementary Materials for a
+more detailed analysis.)
+
+In applying the tools to real data Sanger data, we used the Human
 Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
 the A076941 {\it Arabidopsis thaliana} data set used to test wcd\/
-  \cite{Hazelhurst08a,Hazelhurst08b} (see Table~\ref{SangerReal}).
-  We notice essentially identical results for \peace\/ and wcd\/ in
-  quality, both significantly better than \capthree\/ in Sensitivity and Type
-  1 error rate, while slightly worse in the Jaccard Index and Type 2
-  error.  In runtime we see some inconsistency, with \peace\/ showing a
-  60\% runtime improvement over \wcd\/ in the first data set, but requiring 20\%
-  more time than \wcd\/ in the Arabidopsis data set.  (We cannot make
-  a reasonable runtime comparison to \capthree, as the runtime of that
-  tool reflects both clustering and assembly.)  In Table~S1 of the
-  Supplementary Materials we present runtimes of several more sets,
-  observing that while \peace\/ appears to be significantly 
-  faster on the smaller sets, \wcd\/ does overtake it for larger
-  sets.}
+\cite{Hazelhurst08a,Hazelhurst08b} (see Table~\ref{SangerReal}).  We
+notice essentially identical results for \peace\/ and wcd\/ in
+quality, both significantly better than \capthree\/ in Sensitivity and
+Type 1 error rate, while slightly worse in the Jaccard Index and Type
+2 error.  In runtime we see some inconsistency, with \peace\/ showing
+a 60\% runtime improvement over \wcd\/ in the first data set, but
+requiring 20\% more time than \wcd\/ in the Arabidopsis data set.  (We
+cannot make a reasonable runtime comparison to \capthree, as the
+runtime of that tool reflects both clustering and assembly.)  In
+Table~S1 of the Supplementary Materials we present runtimes of several
+more sets, observing that while \peace\/ appears to be significantly
+faster on the smaller sets, \wcd\/ does overtake it for larger sets.
 
@@
 
-\mc{We tested the three tools on short-read data using the \metasim\/
-  tool of {\it Richter et al.} \cite{Richter2008}.  Encoded into
-  \metasim\/ are sequence generation and error models corresponding to
-  several technologies, including the 454 short-read technology
-  (producing reads ranging in size from 200bp to 370bp with a mean of
-  250bp and a standard deviation of 17bp), and the Illumina short-read
-  technology (producing reads of exactly 62bp).  Experiments were run
-  using two sets of \wcd\/ parameters (the default parameters, and
-  setting window size to 75 and threshold to 75 -- thus matching the
-  equivalent \peace\/ parameters), and using default values
-  of \capthree.  Only with \peace\/ were we able to get reasonable
-  results, seeing a sensitivity of 0.871 for the 454 reads and
-  0.583 or the shorter Illumina reads, a Jaccard Index of 0.809 and
-  0.372 respectively, and other comparably figures.  (See Table~S3 in
-  the Supplementary Materials.)  In short, \peace performs quite well
-  for 454 reads and provides some useful information for Illumina
-  reads, while \wcd\/ and \capthree\/ are unable to produce working
-  results.}
+We tested the three tools on short-read data using the \metasim\/ tool
+of {\it Richter et al.} \cite{Richter2008}.  Encoded into \metasim\/
+are sequence generation and error models corresponding to several
+technologies, including the 454 short-read technology (producing reads
+ranging in size from 200bp to 370bp with a mean of 250bp and a
+standard deviation of 17bp), and the Illumina short-read technology
+(producing reads of exactly 62bp).  In the 454 experiments \peace\/
+achieved a sensitivity of 0.871, an 89.7\% improvement in sensitivity
+over \wcd\/ and a 193\% improvement in sensitivity over \tgicl, with
+comparable values for the Jaccard Index.  \wcd\/ was unable to handle
+the Illumina data, while \peace\/ sensitivity, at a value of 0.384,
+was nearly twice as good as Illumina.  (For more details, see Table~S3
+in the Supplementary Materials.)  In short, \peace\/ performs quite
+well for 454 reads and provides some useful information for Illumina
+reads, with considerably better results than \wcd\/ or \tgicl\/.
 

### 2010-05-11 papers/NAR_web_server_2009/paper.sup.tex
Manuscript update: final(?) version

@@
 better results then the tools against which they compare, we limit our
-analysis to a comparison of \textsc{peace} against \textsc{wcd} and the \textsc{cap3} assembly
-tool \cite{Huang99} -- a tool which implicitly clusters while
-performing assembly.  We do so by applying all three tools to a number
-of data sets, both simulated and real.
+analysis to a comparison of \textsc{peace} against \textsc{wcd},
+\tgicl \cite{Pertea2003}, and the \textsc{cap3} assembly
+tool \cite{Huang99} -- the last of which implicitly clusters while
+performing assembly.
 
@@
 take the form of substitutions, deletions, and insertions (of bases
-and Ns).  See the paper for a discussion of the probability
+and Ns).  See the ESTSim paper for a discussion of the probability
 distributions and default parameters.  For generating our simulated
@@
 explicitly being subjected to variation in our experiments --
-paralleling the testing of the \textsc{wcd} tool.
+parallelising the testing of the \textsc{wcd} tool.
 
@@
 
-Each estimate given in the main paper, or in the following, is
+Each estimate given in the main paper, as well as the following, is
 averaged over 30 trials.  Each trial consists of the application of
-all three tools to a simulated data set, the set having been derived
+all tools to a simulated data set, the set having been derived
 from the application of the ESTsim tool to a set of 100 zebra fish
@@
 
-\mc{Our primary measurement tools if measurement are sensitivity, the
-Jaccard Index, Type 1 error and Type 2 error
-(Figure~3).  Sensitivity is the ratio for true
-positives to actual positives ($tp / (tp + fn)$) -- the} \mc{fraction} \mc{of}
-\mc{sequences} \mc{from} \mc{the} \mc{same} \mc{transcript} \mc{that} \mc{are} \mc{identified} \mc{as} \mc{such.}  \mc{The}
-\mc{{\it Jaccard}} \mc{index} \mc{combines} \mc{both} \mc{sensitivity} \mc{and} \mc{specificity,}
-\mc{calculated} \mc{as} \mc{the} \mc{ratio} \mc{of} \mc{true} \mc{positives} \mc{to} \mc{the sum of true
-positives, true negatives, and false positives ($tp / (tp+fn+fp)$)
-\cite{Hazelhurst08a}.}
+\mc{Our primary measurement tools are sensitivity, the Jaccard Index,
+  Type 1 error and Type 2 error (see Figure~3 from the paper).
+  Sensitivity is the ratio for true positives to actual positives ($tp
+  / (tp + fn)$) -- the} fraction of sequences from the same transcript
+that are identified as such.  The {\it Jaccard} index combines both
+sensitivity and specificity, calculated as the ratio of true positives
+to the sum of true positives, true negatives, and false positives ($tp
+/ (tp+fn+fp)$) \cite{Hazelhurst08a}.
 
@@
 memory than \capthree.  In general we find the footprint of \peace\/
-to be relativily small, as compared to other clustering tools
+to be relatively small, as compared to other clustering tools
 (discussed in {\it Hazelhurst et al.} \cite{Hazelhurst08a}).}
@@
 \mc{All runs were conducted on 3.0 GHz Intel Xeon EM64T CPUs with 2 MB
-caches and a 800 MHz front side bus, model number Xeon LV 3.0 (2005)}
+caches and a 800 MHz front side bus, model number E5520 (2005)}
 
@@
   single processor run on simulated data for each tool: \peace\/
-  (blue), \wcd\/ (green) and \capthree\/ (balck), along with the
-  lesat-squares fit for each.  For each fit we have an associated
+  (blue), \wcd\/ (green) and \capthree\/ (black), along with the
+  least-squares fit for each.  For each fit we have an associated
   regression coefficient of $r > 0.99$.}\label{memory}
@@
   \mc{experiment} \mc{with} \mc{higher} \mc{values,} \mc{with} \mc{interesting} \mc{results.  An
-  analysis of the preformance of $d^2$ indicated that a $T$ value as high as 130 should not have a
+  analysis of the performance of $d^2$ indicated that a $T$ value as high as 130 should not have a
   significant impact on its effectiveness as a discriminator.  And in
@@
 
-\begin{figure}
-  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x130.pdf}}
+\begin{figure}[tbh]
+  \centerline{\includegraphics[width=3.25in]{pics.d/SeJiT1T2x130.pdf}}
   \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
-    Type 1 error using the elevated threhold ($T=130$), providing a
-    contrast against Figure 3 ($T=40$).  Based on the average over 30
+    Type 1 error using the elevated threshold ($T=130$), providing a
+    contrast against Figure~3 ($T=40$).  Based on the average over 30
     simulated Sanger Sequence ESTs sets derived from 100 zebra fish
@@
 
-\begin{figure}[tbp]
+\begin{figure}[tbh]
 \centerline{
@@
 
-\section{Real Data}
+\subsection{Comparison against \tgicl}
+
+In comparing against \wcd\/ we see equivalent results,
+while comparisons against \capthree\/ show improvement in most of the
+quality measures (at the cost of an ability to distinguish
+duplicates).  Comparisons against \tgicl\/ \cite{Pertea2003} are less
+straight forward.  In Figure~\ref{TGICL1} and Figure~\ref{TGICL2} we
+repeat out plots from Figure 3 of the manuscript at each threshold
+value.  We see that \peace\/ is more competitive with \tgicl\/ at higher
+error rates in terms of sensitivity.  More generally, we see \peace\/
+does a better job at specificity, as reflected both in the Jaccard index
+(where the shift in the \tgicl\/ curve indicates a problem with type 2 errors) and
+the higher Type 2 error rate.  We also see this reflected in the
+ability of \tgicl\/ to distinguish duplicated genes, appearing to be unable
+to consistently do so at a similarity of more than 72\% (as opposed to
+about 80\% for \peace), shown in Figure~\ref{tgicl_dups}.
+
+In a sequential execution, \tgicl\/ is clearly the faster tool --
+though this does not necessarily not compensate for the lower
+specificity and improved results of \peace\/ on short-read data.  While
+both tools enable parallal computations, they are implemented on
+different libraries: \peace\/ uses the standard OpenMPI library, while
+\tgicl\/ is built on the PVM.  While MPI has been adapted as the
+standard by IEEE and is generally availabel by default, PVM is not so
+wide spread -- making \tgicl\/ less portable to new
+systems in its parallel form.
+
+\begin{figure}[tbh]
+\centerline{
+  \centerline{\includegraphics[width=3.35in]{pics.d/TGICL_SeJiT1T2x40.png}}
+}
+\caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
+  Type 1 error, based on the average over 30 simulated Sanger Sequence
+  ESTs sets derived from 100 zebra fish genes, using a \peace\/ threshold
+  of $T=40$.  Blue/Solid = \peace, Green/Dash = \wcd\/ \mc{(version
+    0.5.1)}, Black/Dot-Dash = \capthree; Purple/Long-Dash = TGICL;
+  vertical tics = 95\% confidence intervals on estimates.  Intervals
+  are not presented for Type 1 error due to the effective lack of
+  variance.}
+\label{TGICL1}
+\end{figure}
+
+\begin{figure}[tbh]
+\centerline{
+  \centerline{\includegraphics[width=3.35in]{pics.d/TGICL_SeJiT1T2x130.png}}
+}
+\caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
+  Type 1 error, based on the average over 30 simulated Sanger Sequence
+  ESTs sets derived from 100 zebra fish genes, using a \peace\/ threshold
+  of $T=130$.  Blue/Solid = \peace, Green/Dash = \wcd\/ \mc{(version
+    0.5.1)}, Black/Dot-Dash = \capthree; Purple/Long-Dash = TGICL;
+  vertical tics = 95\% confidence intervals on estimates.  Intervals
+  are not presented for Type 1 error due to the effective lack of
+  variance.}
+\label{TGICL2}
+\end{figure}
+
+\begin{figure}[tbh]
+\centerline{
+  \centerline{\includegraphics[width=3.35in]{pics.d/tgicl_duplicates_40.png}}
+}
+\caption{Ability to distinguish duplicates as a function of divergence
+  (\peace\/ v. \mbox{\tgicl}).  Estimates averaged over 30 trials; one trial
+  consists of taking a random zebra fish gene, copying it and
+  stochastically changing bases at the specified rate, then using the
+  two genes as the bases for generating a simulated set. Blue/Solid =
+  \peace, Purple/LongDash = \tgicl; variance was too small for visible
+  plotting of confidence intervals.}\label{tgicl_dups}
+\end{figure}
+
+\section{Benchmark Data}
 
@@
 
-\begin{table}[tbp]
+\begin{table}[tbh]
 \begin{center}
@@
   unable to return any reasonable results for Illumina data, while
-  \capthree\/ was unable to return results for either sumulation.  Each
+  \capthree\/ was unable to return results for either simulation.  Each
   454 simulation produced 25,000 reads from the 100 zebra-fish
-  transcripts, ranging  in size from 200bp to 270bp (aveage = 250bp, 
+  transcripts, ranging  in size from 200bp to 270bp (average = 250bp, 
   standard deviation = 17bp).  Each Illumina simulation produce 150K
@@
 % LocalWords:  Chlamydomonas reinhartdii nvironment xpression runtime bp Runtme
-% LocalWords:  stochastically Runtimes EasyCluster Ricinus Dataset
+% LocalWords:  stochastically Runtimes EasyCluster Ricinus Dataset  GHz Xeon LV
+% LocalWords:  CPUs TGICL Arabidopsis thaliana dataset Illumina ESTSim LongDash
+% LocalWords:  parallelising

### 2010-05-11 papers/NAR_web_server_2009/paper.tex
Manuscript update: final(?) version

@@
   Pertea2003,Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be
-of equivalent quality to the \wcd\/ and \tgicl tools
+produce results of quality competitive with the \wcd\/ and \tgicl\/ tools
 \cite{Hazelhurst08a,Pertea2003}, and more sensitive and robust than
-other tools.  From the user perspective, o tool in the literature can
-match \peace for ease of instillation, use, or the post-clustering
+other tools.  From the user perspective, mo tool in the literature can
+match \peace\/ for ease of instillation or use, or the post-clustering
 analysis tools \peace provides.
@@
 
-The \peace\/ GUI can be launched by through the
-Jar file available on the \peace\/ website
-(\href{http://www.peace-tools.org}{www.peace-tools.org}) on any
-machine running the standard Java Virtual Machine (JVM).  Once
-running, the user can employ the GUI to install the clustering engine
-and perform a clustering of a data file in {\sc fasta} format, view an
-initial analysis of the clusters, and produce files containing subsets
-of the clusters as input to assembly tools such as \capthree\/ 
-\cite{Huang99}.  A typical (first) use of \peace\/ must be
-performed in the following manner (see Figure~\ref{fig:workflow}):
+The \peace\/ GUI, available as a {\small JAR} file from the \peace\/
+website (\href{http://www.peace-tools.org}{www.peace-tools.org}),
+can be run on any machine supporting the standard Java Virtual
+Machine (JVM).  The user can employ the GUI to install
+the clustering engine and perform a clustering of a data file in {\sc
+  fasta} format, view an analysis of the clusters, and produce
+files containing subsets of the clusters as input to assembly tools
+such as \capthree\/ \cite{Huang99}.  A typical (first) use of \peace\/
+must be performed in the following manner (see
+Figure~\ref{fig:workflow}):
 
@@
   footprint -- of particular concern in the large data sets commonly
-  encountered.  In short, we find that, as with \wcd, the required
-  memory for an execution is linear in the size of the data set --
-  requiring slightly more memory than \wcd\/ and considerably less
-  than \capthree.  We present relevant results in Section D.4 and
-  Figure~S3 of the Supplementary Materials.}
+  encountered.  We find that, as with \wcd, the memory used for
+  execution is linear in the size of the data set, requiring
+  slightly more memory than \wcd\/ and considerably less than
+  \capthree.  (See Section D.4 and Figure~S3 of the Supplementary
+  Materials.}
 
@@
   up the clusters (but unable to re-merge a split cluster), for the
-  purposes of clustering assessment it makes sense base our analysis (and
+  purposes of clustering assessment it makes sense to base our analysis (and
   set out default values) at the lower threshold.}
@@
     = \peace, Green/Dash = \wcd\/ \mc{(version 0.5.1)}, Black/Dot-Dash =
-    \capthree; vertical tics = 95\% confidence intervals on estimates.
-    Intervals are not presented for Type 1 error due to the effective
-    lack of variance.}\label{SeJiT1T2}
+    \capthree; vertical tics = 95\% confidence intervals on estimates.}\label{SeJiT1T2}
 \end{figure}
@@
      estimates.  All runs were done on a 3.0 GHz Intel Xeon EM64T CPU with a 2 MB
-     cache and a 800 MHz front side bus, model number Xeon LV 3.0 (2005).}\label{seq_parallel}
+     cache and a 800 MHz front side bus, model number E5520 (2005).}\label{seq_parallel}
 \end{figure}
 
-While \peace\/ clearly matches \wcd\/ on long-read sequence, comparisons
-against \tgicl\/ \cite{Pertea2003}  are more complicated.  In short, we find that when
-parameters are picked appropriately, \tgicl\/ is more sensitive at lower
-error rate -- but \peace\/ is more robust to error, in all cases
-has an improved Type 2 error rates and is better able to
-distinguish duplicated genes.  (See the Supplementary Materials for a
-more detailed analysis.)
+Comparisons against \tgicl\/ \cite{Pertea2003} are more complicated.
+We find that when \tgicl\/ is more sensitive at lower error rates,
+\peace\/ is more robust to error, and in all cases \peace\/ has an
+improved Type 2 error rates and is better able to distinguish
+duplicated genes.  (See Supplementary Materials Section D.6.)
 
@@
 the A076941 {\it Arabidopsis thaliana} data set used to test wcd\/
-\cite{Hazelhurst08a,Hazelhurst08b} (see Table~\ref{SangerReal}).  We
-notice essentially identical results for \peace\/ and wcd\/ in
-quality, both significantly better than \capthree\/ in Sensitivity and
-Type 1 error rate, while slightly worse in the Jaccard Index and Type
-2 error.  In runtime we see some inconsistency, with \peace\/ showing
-a 60\% runtime improvement over \wcd\/ in the first data set, but
-requiring 20\% more time than \wcd\/ in the Arabidopsis data set.  (We
-cannot make a reasonable runtime comparison to \capthree, as the
-runtime of that tool reflects both clustering and assembly.)  In
-Table~S1 of the Supplementary Materials we present runtimes of several
-more sets, observing that while \peace\/ appears to be significantly
-faster on the smaller sets, \wcd\/ does overtake it for larger sets.
+\cite{Hazelhurst08a} (see Table~\ref{SangerReal}).  We notice mixed
+results, with \peace, \wcd, and \tgicl\/ showing comparable similarity,
+\peace\/ and \wcd\/ showing some superiority to \tgicl\/ in Type 1 error
+rate but with a poorer showing in Type 2 error, and with all three
+tools coming considerably closer to the correct number of clusters
+than does \capthree.  In runtime we see some inconsistency, with
+\peace\/ showing a 60\% runtime improvement over \wcd\/ in the first
+data set and a time comparable to \tgicl, but requiring 20\% more time
+than \wcd\/ in the Arabidopsis data set and losing to \tgicl\/ by a
+large margin.  In Table~S1 of the Supplementary Materials we present
+runtimes of several more sets for the distance-based tool, observing
+that while \peace\/ appears to be significantly faster on the smaller
+sets, \wcd\/ does overtake it for larger sets.
 
@@
 \cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1.000 & 0.001 & 2269 & 1827 & \mbox{NA} \\
+\cline{2-9}             & \tgicl    & 0.998 & 0.949 & 0.568 & 0.018 & 221 & 86 & 278 \\
 \hline
-\hline WCD A076941 & \peace & 0.932 & 0.475 & 0.351 & 0.027 & 18825 & 8951 & 1166\\
+\hline A076941 & \peace           & 0.932 & 0.475 & 0.351 & 0.027 & 18825 & 8951 & 1166\\
 \cline{2-9} Benchmark & \wcd          & 0.933 & 0.476 & 0.350 & 0.027 & 18787 & 8553 & 966 \\
-\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
-\mbox{NA} \\
+\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 & \mbox{NA} \\
+\cline{2-9} & \tgicl                  & 0.939 & 0.209 & 0.401 & 0.020 & 20248 & 1065 & 425 \\
 \hline
@@
 \end{center}
-\caption{Comparisons of runs on the EasyCluster human Benchmark
+\caption{Comparisons of runs on the EasyCluster Human Benchmark
   Data Set and the WCD A076941 Arabidopsis thaliana data set using the
@@
 
-We tested the three tools on short-read data using the \metasim\/ tool
+We tested the tools on short-read data using the \metasim\/ tool
 of {\it Richter et al.} \cite{Richter2008}.  Encoded into \metasim\/
@@
 the Illumina data, while \peace\/ sensitivity, at a value of 0.384,
-was nearly twice as good as Illumina.  (For more details, see Table~S3
+was nearly twice as good as TGICL.  (For more details, see Table~S3
 in the Supplementary Materials.)  In short, \peace\/ performs quite
@@
 
-\mc{Here we have presented \peace, a stand alone tool for the
-  high-throughput clustering of transcript fragments capable of
-  dealing with sequences as short as 50 bases.  \peace\/ is
-  open-source and managed through a user friendly GUI that enables
-  both local and remote installation and execution in sequential or
-  parallel mode.  Based on a novel minimum spanning tree based
-  algorithm for the clustering of the fragments by gene association,
-  \peace\/ shows significant improvement in sensitivity over the
-  competing \wcd\/} \mc{tool} \cite{Hazelhurst08a} \mc{when} \mc{applied} \mc{to} \mc{NGS}
-  \mc{reads,} \mc{matches} \mc{\wcd\/} \mc{when} \mc{applied} \mc{to} \mc{Sanger} \mc{sequencing}
-  \mc{output,} \mc{and} \mc{shows} \mc{an} \mc{order} \mc{of} \mc{magnitude} \mc{in improvement over the
-  clustering performed in the course of assembly by the \capthree\/
-  tool \cite{Huang99}.}
+Here we have presented \peace, a stand alone tool for the
+high-throughput clustering of transcript fragments capable of dealing
+with sequences as short as 50 bases.  \peace\/ is open-source and
+managed through a user friendly GUI that enables both local and remote
+installation and execution in parallel mode.  Using a novel minimum
+spanning tree based algorithm for the clustering of fragments by gene
+association, \peace\/ shows significant improvement in sensitivity
+over the competing \wcd\/ tool and \tgicl tool when applied to NGS
+reads \cite{Hazelhurst08a,Pertea2003}, matches \wcd\/ when applied to
+Sanger sequencing output, and shows an order of magnitude in
+improvement over the clustering performed in the course of assembly by
+the \capthree\/ tool \cite{Huang99}.
 
 As a clustering tool based on sequence distance, \peace\/ faces
-certain inherent limitations. For example, \peace\/ cannot handle
-duplicate genes; like \wcd\/, it is unable to separate clusters
-corresponding to genes with a greater than 88\% similarity.
-Similarly, other natural biological effects (e.g. the trans-splicing of
-transcripts), effects from poorly cleaned transcript data (e.g. the
-failure to remove sequencing adapters or post-transcriptional
-poly(A)/(T) tails), and the presence of low-complexity repeats can
-cause similar effects in these clustering tools.  These problems can be
-handled through the application of the assembler, and the ability to
-apply any assembler to small clusters (as opposed to the data set as a
-whole) results in a significant reduction in overall assembly time.
+certain inherent limitations. \peace\/ cannot handle duplicate genes;
+like \wcd\/, it is unable to separate clusters corresponding to genes
+with a greater than 88\% similarity.  Similarly, other natural
+biological effects (e.g. the trans-splicing of transcripts), effects
+from poorly cleaned transcript data (e.g. the failure to remove
+sequencing adapters or post-transcriptional poly(A)/(T) tails), and
+the presence of low-complexity repeats can cause similar effects in
+these clustering tools.  These problems can be handled through the
+application of the assembler, and the ability to apply any assembler
+to small clusters results in a significant reduction in overall
+assembly time.
 
@@
 % LocalWords:  NGS polyA set's GHz Xeon Bikun Arabidopsis JVM Jaccard LV GUI's
-% LocalWords:  Data Set thaliana data set EasyCluster MetaSim Illumina
+% LocalWords:  Data Set thaliana data set EasyCluster MetaSim Illumina tgicl

### 2010-11-22 peace/papers/EAST/east.tex
Adding Trey's thesis

\NeedsTeXFormat{LaTeX2e}[1995/12/01]
\documentclass[10pt]{bmc_article}

% Load packages
\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\urlstyle{rm}

%%%%%%%%%%%%%%%%%%%%
%% Karro macros   %%
%%%%%%%%%%%%%%%%%%%%
\newcommand{\peace} {{\small PEACE}}
\newcommand{\wcd} {{\small WCD}}
\newcommand{\capthree} {{\small Cap3}}
\newcommand{\easycluster} {{\small EasyCluster}}
\newcommand{\estsim}{{\small ESTSim}}
\newcommand{\metasim} {{\small MetaSim}}
\newcommand{\tgicl} {{\small TGICL}}
\newcommand{\east} {{\small EAST}}
\newcommand{\velvet}{{\small Velvet}}
\newcommand{\mira}{{\small Mira3}}
\newcommand{\pave} {{\small PAVE}}
\newcommand{\peast}{{\small PEACE+EAST}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%	
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%   
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %% 
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %% 
%%  submitted article.                         %% 
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{graphicx}
%\def\includegraphic{}
%\def\includegraphics{}

\setlength{\topmargin}{0.0cm}
\setlength{\textheight}{21.5cm}
\setlength{\oddsidemargin}{0cm} 
\setlength{\textwidth}{16.5cm}
\setlength{\columnsep}{0.6cm}

\newboolean{publ}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                              %%
%% You may change the following style settings  %%
%% Should you wish to format your article       %%
%% in a publication style for printing out and  %%
%% sharing with colleagues, but ensure that     %%
%% before submitting to BMC that the style is   %%
%% returned to the Review style setting.        %%
%%                                              %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Review style settings
\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}

%Publication style settings
%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}

% Begin ...
\begin{document}
\begin{bmcformat}

\title{EAST: {\underline E}xpression Fragment {\underline A}ssembly from {\underline S}panning {\underline T}rees}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Ensure \and is entered between all but   %%
%% the last two authors. This will be       %%
%% replaced by a comma in the final article %%
%%                                          %%
%% Ensure there are no trailing spaces at   %% 
%% the ends of the lines                    %%     	
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author{Yuan Zhang$^1$
  \email{Yuan Zhang - zhangy9@muohio.edu}
  \and
  James C Moler$^1$
  \email{James Moler - molerjc@muohio.edu}
  \and
  Dhananjai M Rao$^1$
  \email{Dhananjai Rao - raodm@muohio.edu}
  \and
  Mufit Ozden$^1$
  \email{Mufit Ozden - ozdenm@muohio.edu}
  \and
  Chun Liang\correspondingauthor$^{1,2}$
  \email{Chun Liang\correspondingauthor - liangc@muohio.edu}
  and
  John E. Karro\correspondingauthor$^{1,3}$
  \email{John E. Karro\correspondingauthor - karroje@muohio.edu}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address{
    \iid(1)Department of Computer Science and Software Engineering, Miami University, Oxford OHIO, USA \\
    \iid(2)Department of Botanty, Miami University, Oxford OHIO, USA \\
    \iid(3)Department of Microbiology, Miami University, Oxford OHIO, USA
}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% The Section headings here are those for  %%
%% a Research article submitted to a        %%
%% BMC-Series journal.                      %%  
%%                                          %%
%% If your article is not of this type,     %%
%% then refer to the Instructions for       %%
%% authors on http://www.biomedcentral.com  %%
%% and change the section headings          %%
%% accordingly.                             %%   
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
\end{abstract}

\ifthenelse{\boolean{publ}}{\begin{multicols}{2}}{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% The Section headings here are those for  %%
%% a Research article submitted to a        %%
%% BMC-Series journal.                      %%  
%%                                          %%
%% If your article is not of this type,     %%
%% then refer to the instructions for       %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and change the section headings          %%
%% accordingly.                             %% 
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Introduction}
The {\it de novo} assembly of RNA transcripts is an important and
computationally challanging problem, made more difficult by the advent
of Next Generation Sequencing technologies such as 454 or Illumina
[CITE].  Given the completion of large-scale seqencing run of multiple
mature transcripts of a given organism, the invetigator is left with a
large set of sequence fragments, but no ordering or even transcript
association information; these fragments must be clustered by
transcript then assembled before the sequencing output can be truely
useful.  If the genome has not been sequenced, this must be done {\it
  de novo} -- using only the sequence information extrapolated from
the sequences themselves.  Even when the genomic sequence is known,
there are a number of advanatages to ignoring the known sequence
information and starting from scratch.  Using an existing squence as a
template for sequencing in population studies has the potential to
introduce bias.  And when looking specifically for genes, a known
sequence may be of limited use: the target transcripts may be
undiscovered, and the splicing of introns before sequencing can complicate
the approach of mapping fragemtns to the genomic sequence.  In short,
there is significant value in tools that can reassemble fragement data
into the orginial assembly without requiring access to genomic
information.

The problem of {\it de novo} transcript assembly has been addressed a number
of times in the literature, with initial approaches consentrating on
the longer reads produced by traditional Sanger Sequencing techniques,
and newer approaches looking at produces of Next Generation Sequencing
(NGS) technologies.  A list of the tools most successfully addressing
these problems include \capthree, \tgicl, \velvet, and \mira [CITE],
each excelling within their own context.  \capthree\/ and \tgicl\/ use
alignment-based strategies that produce high quality results when
applied to the output of Sanger sequencing technologies.  \velvet,
based on the modeling of the problem with a {\it de bruijn} graphs,
was designed to handle short read sequences.  [SOMETHING ABOUT MIRA --
INCLUDING ABIITY TO HANDLE HYBRID DATA.]

In this paper we present \east, a tool for the assembly of both
Sanger, NGS and hybrid data with the capacity to incorperate quality
scores when available.  \east\/ is based on a novel minimum spanning
tree (MST) apraoch to compute assembly.  The tool works in tandem with
the \peace\/ clustering tool [CITE] -- allowing \peace\/ to cluster
fragements by transcript and produce the initial MSTs that will serve
as the basis for the \east\/ algorithm.  In the following we present a
comparison of \east\/ against those of \capthree, \tgicl, \velvet, and
\mira on both Sanger and NGS sequences, showing that [???]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Results and Discussion %%
%%
\section*{Results and Discussion}

In assessing result quality, we must view the \peace\/ clustering tool
\cite{Rao10} and \east\/ as a combined process.  While most assembly
tools integrate these two jobs, \east\/ assumes clustering has been
preformed and requires the \peace\/ MST output.  Hence for purposes of
the following section, we consider a single tool \peast, and
assess the quality of these results.

In order to assess the quality of the \east\/ (\peast\/) results, we compared
against four prominent assembly tools: \capthree, \tgicl,
\mira, and \velvet [CITE].  Tools were tested on data from three
sequencing technologies (Sanger, 454 and Illumina), using both
simulated output obtained from the \estsim\/ and \metasim\/ tools
[CITE] and benchmark sets obtained from GenBank [CITE].

\subsection*{Measuring Quality}
Our primary measure of quality was the {\bf normalized A-score}, based
on the {\it assembly score} (A-score) used in {\it Liang et al.}
\cite{Liang00}.  For an assembly of some originial sequence, the
A-score is defined as:
$$\mbox{($2\times$sequence length) - ($5\times$ number of
mismatches) - ($15 \times$ number of indels)}$$
with a ``perfect'' score equal to twice the sequence length.  Our
normalized A-score modifies this definition as:
$$\frac{\mbox{($2\times$assembly length) - ($5\times$ number of mismatches)
  - ($15\times$ number of indels)}}{\mbox{($2\times$ sequence length)}}$$
with a perfect score equal to 1.  This normalization allows for the
comparison of assembly quality between sequences of differenct
lengths. Further, using this score we can more accurately quantify the
quality of a partial reconstrction: if we are able to reconstruct only
a portion of a fragment, the normalized A-score will take full
consideration of the quality of the reconstruction, but then bound the
score by the relative size of the reconstruction.  (e.g. Perfectly
reconstructiong one fourth of the original sequence will lead to an a
score of $0.25$.)

In addition to A-score...

\subsection*{Simulation Results}
Using simulated fragement sets allows us to both compare assemblies to
a known correct solution and control parameters in their model
(e.g. base-call error rate) to test the effect of variation.  All sets
were generated from a set of 100 zebra fish transcripts obtained from
{\it Hazelhurst et al.}  \cite{Hazelhurst08}.  For simulated Sanger
sequences we used the \estsim\/ tool, using its default model of EST
generation and error rate when not otherwise specified.  For simulated
454 and Illumina sets we used the \metasim\/ tool, again using their
default model to simulate the errors associated with those
technologies.

\noindent {\bf Sanger Sequencing:} We start by looking at the assembly
quality of each tool, when applied to simulated Sanger Sequences, as a
function of error rate for a range of coverage depths.  For each
parameter combination we applied each tool to 30 independently
randomly generated simulated sets, derived from 15 zebra fish genes,
reporting the average normalized A-score.  In
Figure~\ref{sangerAscore} we look at normaized A-score as a function
of error rate, fixing coverage at 50, 30 and 10 and varying base-read
error rate from 0\% to 6\% [WHAT IS INDEL RATE?] for each of the five
tools being tested.  While the three illustrated tools preform well
when applied error free (i.e. impossible) sequencing technoloy, we
find \east\/ to be considerably more robust to base-call errors than the
other tools.  Taking the mid-range coverage value of 30 as an example,
at a 1\% error rate we see \east, \capthree, and \tgicl\/ all achieving an
essentially perfect normalize A-score, while \mira\/ ranges around a 0.5
and \velvet\/ does quite poorly.  At a 3\% error rate \east\/ maintains its
almost perfect normalized A-score, showing a 2\% improvement of \tgicl,
a 21\% improvlement over \capthree, and a 133\% improvement over \velvet.
At a 4\% error rate \east\/ still maintains its near-perfect score
(showing a 54\% and 85\% improvement over \capthree\/ and \tgicl\/
respecitvally), and even at a 6\% error rate is still achieving an
normalized A-score of 0.95 (a 110\% improvement over the next closest
tool, \tgicl\/).

\begin{figure}[htb]
\centerline{\includegraphics[width=6in]{pics.d/ascore_sanger.png}}
\label{sangerAscore}
\caption{Normalized a-score as a function of base error rate for
  simmulated sanger sequencing.  \east=blue, \capthree=Black,
  \tgicl=green, \velvet=red, \mira=purple.}
\end{figure}

Another useful metric is the number of contigs and number of singleton
fragments in the final result.  In Figure~\ref{contigs.sanger} is a
plot of the average number of contigs per gene seqence (left column), ad
the number of singletons (right column).  For the first we observe that all tools have
comperable results for lower error rates, but only \east\/ is able to
maintain the low score with increasing error rates.  In minimizing the
number of singletons, we find that \east\/ does a better job than
\capthree\/ for error rates larger than $2\%$, while TGICL does a
better job that \east\/ for error rates larger than $4\%$ to $5\%$.

\begin{figure}[htb]
\begin{minipage}{3in}
\includegraphics[width=3in]{pics.d/numContigs_sanger.png}
\end{minipage}
\begin{minipage}{3in}
\includegraphics[width=3in]{pics.d/numSingle_sanger.png}
\end{minipage}
\label{contigs.sanger}
\caption{Number of contigs and singles as a function of base error
  rate for simmulated sanger sequencing.  \east=blue, \capthree=Black,
  \tgicl=green, \velvet=red, \mira=purple.  (Number of singles for
  \velvet\/ and \mira\/ not shown.)}
\end{figure}

[ADD SOMETHING ABOUT COVERAGE?  The situation is strange.]

\noindent {\bf Next Generation Sequences:}

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions}

%%%%%%%%%%%%%%%%%%
\section*{Methods}

The algorithm underlying \east\/ is based on three components: minimum
spanning trees (MSTs)\cite{Prim57}, the $d^2$ sequence distance
measure \cite{Hide94}, and standard sequence alignment
\cite{Needleman70,Smith81}.  While we would ideally be using sequence
alignment to identify overlapping ends and reassembling on that basis,
the quadratic run time of alignment algorithms coupled with the large
number of sequences makes such an approach infeasible.  To avoid excessive
applications of such algorithms, we make use MSTs to provide a {\it
  guide} through the EST set, dictating which pairs we aligned.  A
modified version of the $d^2$ measure provides the edge weights for
the underlying graph, allowing us to define the MST and also providing
some speedup to the alignment algorithm.

\east\/ is technically an assembly tool, and require clustering be
done as a pre-processing step, as well as requiring that $d^2$-based
MSTs be provided for each cluster.  Conveniently, the \peace\/
clustering tool \cite{Rao10} provides exactly this information.  While
\peace\/ an \east\/ are technically two tools, \east\/ was designed
specifically for \peace\/ output.  For purposes of this paper, it
makes sense to consider them as one.  In the following we briefly
outline the \peace\/ algorithm for clustering (as is relevant to
\east\/), and then discuss the steps taken by \east\/ for assembly of
the \peace\/ clusters.

\subsection*{Clustering}

Given a set of fragments from two or more transcripts, we must first
cluster the fragments by transcript association -- leaving us with a
single cluster of fragments for each transcript.  Details of the
\peace\/ algorithm can be found in {\it Rao et al.} \cite{Rao10}, but
the general idea is as follows: we model the problem as a weighted,
undirected graph, representing each fragment with a node and assigning
edge weights based on a $d^2$ comparison of the incident fragments
(explained in more detail below).  By employing the $u/v$ and $t/v$
filtering heuristics of {\it Hazelhurst et al.} \cite{Hazelhurst08}, we
can quickly dismiss most of the edges as irrelevant (i.e. connecting
nodes that belong in different clusters), then use Prim's algorithm to
compute an MST \cite{Prim57}.  We then remove all
edges exceeding a specified threshold value, and take each component
of the resulting forest as a cluster.  As a result \peace\/ can pass
to \east\/ a set of clusters, an MST for each cluster, and the
orientation of each fragment relative to the cluster.  Hence \east\/
has not only the clusters, but the \peace\/ MST and may quickly adjust
the sequences to ensure all are on the same strand.

\subsection*{Overlap Detection}

Given the \peace\/ generated MST for a cluster, the challenge is to
identify and align the ends of overlapping fragments without taking
the significant run time hit needed to align every pair.  We
accomplish this using two methods: we reduce the number of times an
alignment is calculated aligning only those sequences adjacent in the
MST, and we reduce the amount of  time spent per alignment by aligning
only those portions indicated by an application of a modified form of
the $d^2$ sequence distance measure.

The presence of an edge in the tree is a heuristic indication that the
incident sequences overlap; were there not sufficient sequence
similarity to indicate an overlap, the edge would have been assigned a
high $d^2$ value and would have been removed by \peace\/.  To
determine how the nodes overlap, we calculate a modified $d^2$ score
as follows.  The standard calculation of $d^2$, as define by {\it Hide
  et al.} \cite{Hide94}, requires looking at every pair of length $w$
sub-strings (windows) from the two sequences, calculating a distance
between these two windows based on the number of shared words
(6-mers), and taking the score of the lowest-scoring window pairs as
the $d^2$ distance.  However, when looking for overlap of fragments
$s_1$ and $s_2$, we are looking only for end-overlap.  That is, we
wish to determine of the left end of $s_1$ overlaps the right end of
$s_2$, or the reverse (recalling that we know them to be on the same
strand).  Hence we need only consider the left-most and right-most
windows of $s_1$ against each window of $s_2$, reducing the number of
window comparisons from quadratic to linear in the size of $s_2$.

In Figure~\ref{fig:overlap} we see illustrate the use of modified $d^2$,
using it to take two sequences (a), find the window on $s_2$ most
resembling the right-most window of $s1$ (red section of (b)),
extending to the end of $s_2$ (green section of (c)) and preforming a
{\it global} alignment on the overlapping portion.  We then compute the
{\it overlap distance} ($1 - s/l$, where $s$ is the alignment score and
$l$ is the length of the overlap segment), a value that is inversely
proportional to the probability that this is a legitimate overlap.

\begin{figure}
\centerline{\includegraphics[width=3in]{overlap_alignment.pdf}}
\caption{(a) Two sequences potentially overlapping at the ends.  (b)
  The window of $s_2$ pairing with the right-most window of $s_1$ to
  minimize the $d^2$ score.  (c) An extension of the $s_2$ window to
  the end of the sequence, and the corresponding extension on $s_1$.
  (d) A global alignment of the overlapping areas determined from
  (c).}\label{fig:overlap}
\end{figure}

We note that both \peace\/ and \east\/ make use of an {\it adaptive
  $d^2$} strategy to handle hybrid data sets.  The $d^2$ measure (and
our modification) is flexible enough to handle a range of sequence
sizes, requiring only a modification of the parameters (i.e. window
size and threshold values).  However, it runs into problems when
comparing sequences of significantly different sizes.  Comparing a 62
base Illumina read against a 1000 base Sanger read requires the use of
parameters appropriate to the shorter fragment -- which are
considerably less accurate when applied to the longer read.  Our
solution is to partition fragments by sizes (e.g. into groups of
small, medium and large fragments), and compute distances only
between segments in the name or neighboring groups.  By skipping the
comparisons between sequences of significantly different sizes we
avoid the pitfalls of such comparisons.

\subsection*{Ordering and Reconstruction}

Having defined overlap distance, ordering the fragments is
straight-forward.  We pick an arbitrary node on the cluster's MST and
transverse the tree, using the overlap distance measure to check each
node against its two neighbors.  Those nodes with no identified
left-overlap are checked against nodes further out in the tree, and
any remaining with no identified left-overlap are designated as a contig
left end.

Having identified the distance of each node from its two closest
overlapping sequences, we now construct a {\it directed} graph, with
each edge pointing from a sequence to its right neighbor and weighted
with the overlap distance.  Using this graph we calculate a new
minimum spanning tree, the starting from the root of the tree
(necessarily a left-end of the contig), we transverse the tree and
assign the first base of each sequence a position relative to its
parent.  Such an assignment gives us an implicit ordering of the
fragments, by scanning through them in order and aligning each to
the next with the Smith-Waterman alignment we end up with a multiple
alignment of the fragments that allows us to derive a consensus
sequence.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Authors contributions}
    Text for this section \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
  \ifthenelse{\boolean{publ}}{\small}{}
  Text for this section \ldots

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%              
%%  Bmc_article.bst  will be used to                       %%
%%  create a .BBL file for submission, which includes      %%
%%  XML structured for BMC.                                %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %% 
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %% 
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
 \bibliographystyle{bmc_article}  % Style BST file
  \bibliography{bmc_article} }     % Bibliography file (usually '*.bib' )

%%%%%%%%%%%

\ifthenelse{\boolean{publ}}{\end{multicols}}{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Figures                       %%
%%                               %%
%% NB: this is for captions and  %%
%% Titles. All graphics must be  %%
%% submitted separately and NOT  %%
%% included in the Tex document  %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%
%% Do not use \listoffigures as most will included as separate files

\section*{Figures}
  \subsection*{Figure 1 - Sample figure title}
      A short description of the figure content
      should go here.

  \subsection*{Figure 2 - Sample figure title}
      Figure legend text.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Tables                        %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - Sample table title}
    Here is an example of a \emph{small} table in \LaTeX\ using  
    \verb|\tabular{...}|. This is where the description of the table 
    should go. \par \mbox{}
    \par
    \mbox{
      \begin{tabular}{|c|c|c|}
        \hline \multicolumn{3}{|c|}{My Table}\\ \hline
        A1 & B2  & C3 \\ \hline
        A2 & ... & .. \\ \hline
        A3 & ..  & .  \\ \hline
      \end{tabular}
      }
  \subsection*{Table 2 - Sample table title}
    Large tables are attached as separate files but should
    still be described here.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                               %%
%% Additional Files              %%
%%                               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Additional Files}
  \subsection*{Additional file 1 --- Sample additional file title}
    Additional file descriptions text (including details of how to
    view the file, if it is in a non-standard format or the file extension).  This might
    refer to a multi-page table or a figure.

  \subsection*{Additional file 2 --- Sample additional file title}
    Additional file descriptions text.

\end{bmcformat}
\end{document}

% LocalWords:  Ozden Chun Liang Karro Botanty publ MSTs et al Hazelhurst Prim's
% LocalWords:  mers Illumina contig bmc multi

### 2010-11-22 peace/papers/NAR_web_server_2009/paper.tex
Adding Trey's thesis

@@
 
-\mc{We tested the three tools on short-read data using the \metasim\/
-  tool of {\it Richter et al.} \cite{Richter2008}.  Encoded into
-  \metasim\/ are sequence generation and error models corresponding to
-  several technologies, including the 454 short-read technology
-  (producing reads ranging in size from 200bp to 370bp with a mean of
-  250bp and a standard deviation of 17bp), and the Illumina short-read
-  technology (producing reads of exactly 62bp).  Experiments were run
-  using two sets of \wcd\/ parameters (the default parameters, and
-  setting window size to 75 and threshold to 75 -- thus matching the
-  equivalent \peace\/ parameters), and using default values
-  of \capthree.  Only with \peace\/ were we able to get reasonable
-  results, seeing a sensitivity of 0.871 for the 454 reads and
-  0.583 or the shorter Illumina reads, a Jaccard Index of 0.809 and
-  0.372 respectively, and other comparably figures.  (See Table~S3 in
-  the Supplementary Materials.)  In short, \peace performs quite well
-  for 454 reads and provides some useful information for Illumina
-  reads, while \wcd\/ and \capthree\/ are unable to produce working
-  results.}
+We tested the tools on short-read data using the \metasim\/ tool
+of {\it Richter et al.} \cite{Richter2008}.  Encoded into \metasim\/
+are sequence generation and error models corresponding to several
+technologies, including the 454 short-read technology (producing reads
+ranging in size from 200bp to 370bp with a mean of 250bp and a
+standard deviation of 17bp), and the Illumina short-read technology
+(producing reads of exactly 62bp).  In the 454 experiments \peace\/
+achieved a sensitivity of 0.871, an 89.7\% improvement in sensitivity
+over \wcd\/ and a 193\% improvement in sensitivity over \tgicl, with
+comparable values for the Jaccard Index.  \wcd\/ was unable to handle
+the Illumina data, while \peace\/ sensitivity, at a value of 0.384,
+was nearly twice as good as TGICL.  (For more details, see Table~S3
+in the Supplementary Materials.)  In short, \peace\/ performs quite
+well for 454 reads and provides some useful information for Illumina
+reads, with considerably better results than \wcd\/ or \tgicl\/.
 

### 2011-01-10 peace/papers/EAST/east.tex
First version of the EAST manuscript (FINALLY)

@@
   \and
-  James C Moler$^1$
-  \email{James Moler - molerjc@muohio.edu}
-  \and
   Dhananjai M Rao$^1$
@@
   \email{Mufit Ozden - ozdenm@muohio.edu}
+  \and
+  James C Moler$^1$
+  \email{James Moler - molerjc@muohio.edu}
   \and
@@
 The {\it de novo} assembly of RNA transcripts is an important and
-computationally challanging problem, made more difficult by the advent
+computationally challenging problem, made more difficult by the advent
 of Next Generation Sequencing technologies such as 454 or Illumina
-[CITE].  Given the completion of large-scale seqencing run of multiple
-mature transcripts of a given organism, the invetigator is left with a
-large set of sequence fragments, but no ordering or even transcript
-association information; these fragments must be clustered by
-transcript then assembled before the sequencing output can be truely
+\cite{Nagaraj07,Rao10}.  Given the completion of a large-scale sequencing run of
+multiple mature transcripts of a given organism, the investigator is
+left with a large set of sequence fragments, but no ordering or even
+transcript association information; these fragments must be clustered
+by transcript then assembled before the sequencing output can be truly
 useful.  If the genome has not been sequenced, this must be done {\it
@@
 the sequences themselves.  Even when the genomic sequence is known,
-there are a number of advanatages to ignoring the known sequence
-information and starting from scratch.  Using an existing squence as a
-template for sequencing in population studies has the potential to
+there are a number of advantages to ignoring the known sequence
+information and starting from scratch.  Using an existing sequence as
+a template for sequencing in population studies has the potential to
 introduce bias.  And when looking specifically for genes, a known
 sequence may be of limited use: the target transcripts may be
-undiscovered, and the splicing of introns before sequencing can complicate
-the approach of mapping fragemtns to the genomic sequence.  In short,
-there is significant value in tools that can reassemble fragement data
-into the orginial assembly without requiring access to genomic
-information.
+undiscovered, and the splicing of introns before sequencing can
+complicate the approach of mapping fragments to the genomic sequence.
+In short, there is significant value in tools that can reassemble
+fragment data into the original assembly without requiring access to
+genomic information.
 
-The problem of {\it de novo} transcript assembly has been addressed a number
-of times in the literature, with initial approaches consentrating on
-the longer reads produced by traditional Sanger Sequencing techniques,
-and newer approaches looking at produces of Next Generation Sequencing
-(NGS) technologies.  A list of the tools most successfully addressing
-these problems include \capthree, \tgicl, \velvet, and \mira [CITE],
-each excelling within their own context.  \capthree\/ and \tgicl\/ use
-alignment-based strategies that produce high quality results when
-applied to the output of Sanger sequencing technologies.  \velvet,
-based on the modeling of the problem with a {\it de bruijn} graphs,
-was designed to handle short read sequences.  [SOMETHING ABOUT MIRA --
-INCLUDING ABIITY TO HANDLE HYBRID DATA.]
+The problem of {\it de novo} transcript assembly has been addressed a
+number of times in the literature, with initial approaches
+concentrating on the longer reads produced by traditional Sanger
+Sequencing techniques, and newer approaches looking at produces of
+Next Generation Sequencing (NGS) technologies.  A list of the tools
+most successfully addressing these problems include \capthree, \tgicl,
+\velvet, and \mira, each excelling within their own context
+\cite{Huang99,Pertea03,Chevreux04,Zerbino08}.  \capthree\/ and
+\tgicl\/ use alignment-based strategies that produce high quality
+results when applied to the output of Sanger sequencing technologies.
+\velvet, based on the modeling of the problem with a {\it de bruijn}
+graphs, was designed to handle short read sequences.  \mira\/ is a
+multi-pass DNA assembler based on an overlap-layout-consensus
+strategy.  [STILL NEED TO SAY SOMETHING ABOUT MIRA AND HYBRID SETS.]
 
 In this paper we present \east, a tool for the assembly of both
-Sanger, NGS and hybrid data with the capacity to incorperate quality
+Sanger, NGS and hybrid data with the capacity to incorporate quality
 scores when available.  \east\/ is based on a novel minimum spanning
-tree (MST) apraoch to compute assembly.  The tool works in tandem with
-the \peace\/ clustering tool [CITE] -- allowing \peace\/ to cluster
-fragements by transcript and produce the initial MSTs that will serve
+tree (MST) approach to compute assembly.  The tool works in tandem with
+the \peace\/ clustering tool \cite{Rao10}: \peace\/ clusters
+fragments by transcript and produces the initial MSTs that will serve
 as the basis for the \east\/ algorithm.  In the following we present a
 comparison of \east\/ against those of \capthree, \tgicl, \velvet, and
-\mira on both Sanger and NGS sequences, showing that [???]
+\mira on both Sanger and NGS sequences, showing that \peast\/
+generates the best quality results in the fastest runtime for Sanger
+sequences, and that while certain tools out-preform \peast\/ under specific
+conditions, it is the only tool able to produce constant quality results
+for Sanger, 454, Illumina, {\it and} hybrid data sets.
 
@@
 tools integrate these two jobs, \east\/ assumes clustering has been
-preformed and requires the \peace\/ MST output.  Hence for purposes of
-the following section, we consider a single tool \peast, and
-assess the quality of these results.
+preformed and requires the \peace\/ MST output.  While other cluster
+tools could be substituted in to preform the actual clustering
+(e.g. \wcd or \easycluster \cite{Hazelhurst08,Picardi09}), only
+\peace\/ provides the requires MST output.  Hence for purposes of the
+following section, we consider a single tool \peast, and assess the
+quality of these results.
 
-In order to assess the quality of the \east\/ (\peast\/) results, we compared
+In order to assess the quality of the (\peace\/ +) \east\/ results, we compared
 against four prominent assembly tools: \capthree, \tgicl,
-\mira, and \velvet [CITE].  Tools were tested on data from three
+\mira, and \velvet\/ \cite{Huang99,Pertea03,Chevreux04,Zerbino08}.  Tools were tested on data from three
 sequencing technologies (Sanger, 454 and Illumina), using both
 simulated output obtained from the \estsim\/ and \metasim\/ tools
-[CITE] and benchmark sets obtained from GenBank [CITE].
+\cite{Hazelhurst03,Richter08} and benchmark sets obtained from GenBank \cite{Benson10}.
 
 \subsection*{Measuring Quality}
-Our primary measure of quality was the {\bf normalized A-score}, based
-on the {\it assembly score} (A-score) used in {\it Liang et al.}
-\cite{Liang00}.  For an assembly of some originial sequence, the
-A-score is defined as:
-$$\mbox{($2\times$sequence length) - ($5\times$ number of
-mismatches) - ($15 \times$ number of indels)}$$
+Our primary measure of quality was the {\bf normalized a-score}, based
+on the {\it assembly score} (a-score) used in {\it Liang et al.}
+\cite{Liang00}.  For an assembly of some original sequence, the
+a-score is defined as:
+$$\mbox{($2\times$sequence length) - ($5\times$number of
+mismatches) - ($15 \times$number of indels)}$$
 with a ``perfect'' score equal to twice the sequence length.  Our
-normalized A-score modifies this definition as:
-$$\frac{\mbox{($2\times$assembly length) - ($5\times$ number of mismatches)
-  - ($15\times$ number of indels)}}{\mbox{($2\times$ sequence length)}}$$
-with a perfect score equal to 1.  This normalization allows for the
-comparison of assembly quality between sequences of differenct
-lengths. Further, using this score we can more accurately quantify the
-quality of a partial reconstrction: if we are able to reconstruct only
-a portion of a fragment, the normalized A-score will take full
-consideration of the quality of the reconstruction, but then bound the
-score by the relative size of the reconstruction.  (e.g. Perfectly
-reconstructiong one fourth of the original sequence will lead to an a
-score of $0.25$.)
-
-In addition to A-score...
+normalized a-score modifies this definition as:
+$$\frac{\mbox{($2\times$assembly length) - ($5\times$number of mismatches)
+    - ($15\times$number of indels)}}{\mbox{($2\times$sequence
+    length)}}$$ with a perfect score equal to 1.  This normalization
+allows for the comparison of assembly quality between sequences of
+different lengths. Further, using this score we can more accurately
+quantify the quality of a partial reconstruction: if we are able to
+reconstruct only a portion of a fragment, the normalized a-score will
+take full consideration of the quality of the reconstruction, but then
+bound the score by the relative size of the reconstruction.
+(e.g. Perfectly reconstruction one fourth of the original sequence
+will lead to an a score of $0.25$.)  In addition to a-score we also
+look at: the number of contigs produced by a tool (which, in the
+perfect solution, will be exactly equal to the number of transcripts
+represented by the data set), normalized by the number of transcripts;
+the number of singleton ESTs in the solution (i.e. ESTs unassigned to
+any contig and thus, most cases, effectively thrown away), and the runtime of the tool.
 
 \subsection*{Simulation Results}
-Using simulated fragement sets allows us to both compare assemblies to
+Using simulated fragment sets allows us to both compare assemblies to
 a known correct solution and control parameters in their model
@@
 sequences we used the \estsim\/ tool, using its default model of EST
-generation and error rate when not otherwise specified.  For simulated
-454 and Illumina sets we used the \metasim\/ tool, again using their
+generation when not otherwise specified.  For simulated
+454 and Illumina sets we used the \metasim\/ tool \cite{Richter08}, again using their
 default model to simulate the errors associated with those
@@
 
-\noindent {\bf Sanger Sequencing:} We start by looking at the assembly
-quality of each tool, when applied to simulated Sanger Sequences, as a
-function of error rate for a range of coverage depths.  For each
-parameter combination we applied each tool to 30 independently
-randomly generated simulated sets, derived from 15 zebra fish genes,
-reporting the average normalized A-score.  In
-Figure~\ref{sangerAscore} we look at normaized A-score as a function
+\noindent {\bf Sanger Sequencing:} We first compare the assembly
+quality of the tools on Sanger Sequencing, using \east, \capthree, and
+\tgicl\/ (the tools designed to work with Sanger sequences), looking
+at the a-score as a function of error rate for a range of coverage
+depths.  For each parameter combination we applied each tool to 30
+independently randomly generated simulated sets, derived from 15 zebra
+fish genes, reporting the average normalized a-score.  In
+Figure~\ref{sangerAscore} we look at normalized a-score as a function
 of error rate, fixing coverage at 50, 30 and 10 and varying base-read
-error rate from 0\% to 6\% [WHAT IS INDEL RATE?] for each of the five
-tools being tested.  While the three illustrated tools preform well
-when applied error free (i.e. impossible) sequencing technoloy, we
-find \east\/ to be considerably more robust to base-call errors than the
-other tools.  Taking the mid-range coverage value of 30 as an example,
-at a 1\% error rate we see \east, \capthree, and \tgicl\/ all achieving an
-essentially perfect normalize A-score, while \mira\/ ranges around a 0.5
-and \velvet\/ does quite poorly.  At a 3\% error rate \east\/ maintains its
-almost perfect normalized A-score, showing a 2\% improvement of \tgicl,
-a 21\% improvlement over \capthree, and a 133\% improvement over \velvet.
-At a 4\% error rate \east\/ still maintains its near-perfect score
-(showing a 54\% and 85\% improvement over \capthree\/ and \tgicl\/
-respecitvally), and even at a 6\% error rate is still achieving an
-normalized A-score of 0.95 (a 110\% improvement over the next closest
-tool, \tgicl\/).
+error rate from 0\% to 6\% for each of the three tools designed for
+Sanger Sequencing.  While all three tools preform well when applied to
+error free (i.e. impossible) sequencing technology, we find \east\/ to
+be considerably more robust to base-call errors than the other tools.
+Taking the mid-range coverage value of 30 as an example, at a 1\%
+error rate we see \east, \capthree, and \tgicl\/ all achieving an
+essentially perfect normalize a-score .  At a 3\% error rate \east\/
+maintains its almost perfect normalized a-score, showing a 2\%
+improvement over \tgicl\/ and a 21\% improvement over \capthree.  At a
+4\% error rate \east\/ still maintains its near-perfect score (showing
+a 54\% and 85\% improvement over \capthree\/ and \tgicl\/
+respectively), and even at a 6\% error rate is still achieving an
+normalized a-score of 0.95 (a 110\% improvement over the next closest
+tool, \tgicl\/), meaning that at most 5\% of the bases were left out of
+their contigs or were incorrectly reconstructed.
 
 \begin{figure}[htb]
-\centerline{\includegraphics[width=6in]{pics.d/ascore_sanger.png}}
+\centerline{\includegraphics[width=6in]{pics.d/ascore_sanger.pdf}}
+\caption{Average normalized a-score as a function of base error rate for
+  simulated Sanger sequencing.  \east=blue, \capthree=Black,
+  \tgicl=green, \velvet=red, \mira=purple.}
 \label{sangerAscore}
-\caption{Normalized a-score as a function of base error rate for
-  simmulated sanger sequencing.  \east=blue, \capthree=Black,
-  \tgicl=green, \velvet=red, \mira=purple.}
 \end{figure}
@@
 Another useful metric is the number of contigs and number of singleton
-fragments in the final result.  In Figure~\ref{contigs.sanger} is a
-plot of the average number of contigs per gene seqence (left column), ad
+fragments in the final result.  In Figure~\ref{contigsSanger} is a
+plot of the average number of contigs per gene sequence (left column), and
 the number of singletons (right column).  For the first we observe that all tools have
-comperable results for lower error rates, but only \east\/ is able to
-maintain the low score with increasing error rates.  In minimizing the
+comparable results for lower error rates, but only \east\/ is able to
+maintain the low score with increasing error rates -- generating an
+perfect score in almost every case.  In minimizing the
 number of singletons, we find that \east\/ does a better job than
-\capthree\/ for error rates larger than $2\%$, while TGICL does a
+\capthree\/ for error rates larger than $2\%$, while \tgicl does a
 better job that \east\/ for error rates larger than $4\%$ to $5\%$.
@@
 \begin{minipage}{3in}
-\includegraphics[width=3in]{pics.d/numContigs_sanger.png}
+\includegraphics[width=3in]{pics.d/numContigs_sanger.pdf}
 \end{minipage}
 \begin{minipage}{3in}
-\includegraphics[width=3in]{pics.d/numSingle_sanger.png}
+\includegraphics[width=3in]{pics.d/numSingle_sanger.pdf}
 \end{minipage}
-\label{contigs.sanger}
-\caption{Number of contigs and singles as a function of base error
-  rate for simmulated sanger sequencing.  \east=blue, \capthree=Black,
+\caption{Number of contigs (normalized to the number of transcripts)
+  and number of singletons as a function of base call error
+  rate for simulated Sanger sequencing.  \east=blue, \capthree=Black,
   \tgicl=green, \velvet=red, \mira=purple.  (Number of singles for
   \velvet\/ and \mira\/ not shown.)}
+\label{contigsSanger}
 \end{figure}
 
-[ADD SOMETHING ABOUT COVERAGE?  The situation is strange.]
+\noindent {\bf Next Generation Sequences:} In
+Figure~\ref{nextgenAscore} we compare the normalized a-score for
+simulated 454 and Illumina outputs (with sequence lengths averaging
+250 and 62 bp respectively).  For these tests we varied coverage
+levels but otherwise took the default Metasim model.  On 454 data we
+see significant improvement in results quality in \east\/ for all
+levels of coverage.  On Illumina data we see that \east\/ is
+competitive with \velvet\/ at higher coverages, though is unable to
+match that tool when coverage is sparse.
 
-\noindent {\bf Next Generation Sequences:}
+\begin{figure}[htb]
+\centerline{\includegraphics[width=6in]{pics.d/ascore_nxtgen.pdf}}
+\caption{Comparative normalized a-scores for 454 and Illumina sequence
+  output, using the \metasim default model at different levels
+  of coverage.  \east=blue, \capthree=Black, \tgicl=green,
+  \velvet=red, \mira=purple.  }
+\label{nextgenAscore}
+\end{figure}
+
+\noindent {\bf Hybrid:} In Figure~\ref{hybridAscore} we look at the
+quality of results derived from applying each tool to a hybrid data
+set of Sanger (40\%), 454 (40\%) and Illumina (20\%) sequences, again
+showing normalized a-score as a function of base call error rate in
+the Sanger portion.  Here we observe \east matching or outperforming
+\capthree\/ and \mira\/ at moderate to high coverages.  At a coverage
+level of 50, \east\/ maintains a normalized a-score hovering around
+0.85 at all error levels (showing a 54\% improvement over \mira\/ at a
+3\% base-call error rate).  \east\/ actually improves its performance
+at a coverage level of 30 (hovering around a normalized a-score of
+0.95, showing a 142\% improvement of Mira at the 3\% base-call error
+rate), an oddity we attribute to the decreased absolute number of NGS
+sequences with the smaller coverage.  At a coverage level of 10
+\east's normalized a-score is around 0.37 -- doing considerably worse
+that \mira\/ at low error rates (e.g. 0.94 v. 0.40 for error-free
+bases), but showing a 52\% improvement over \mira\/ at the 3\% error
+rate mark and in general showing much more robustness to error.
+Remarkably, it is only in this situation that \capthree\/ is
+competitive, with that tool achieving a normalized a-score of 0.62
+(v. \east's 0.37) for the 3\% base call error rate, but loosing out to
+\east at greater than a 4.5\% base call error rate -- a result that we
+attribute to \east's problems with the 62 bp Illumina sequences (see below).
+
+\begin{figure}[htb]
+\centerline{\includegraphics[width=6in]{pics.d/ascore_hybrid.pdf}}
+\caption{Comparative normalized a-scores a hybrid set of Sanger, 454
+  and Illumina sequence output, varying the base call error rate for
+  Sanger sequences and using the MetaSim default model for the 454 and
+  Illumina sequences.  Each data set tested was composed of 40\%
+  Sanger Sequences, 40\% 454 Sequences, and 20\% Illumina sequences.
+  \east=blue, \capthree=Black, \tgicl=green,
+  \velvet=red,\mira=purple.}
+\label{hybridAscore}
+\end{figure}
+
+\noindent {\bf Tool Runtime:} As \east\/ assembly requires the
+\peace\/ cluster (as opposed to the competing tools that do both
+jobs), for \east\/ runtime we actually include the \peace\/
+runtime. In Figures~\ref{runtime.fixed} we see the average runtime for
+each of the experiments reflected in Figures
+Figures~\ref{sangerAscore}-\ref{hybridAscore} for a coverage of 30,
+while in Figure~\ref{runtime.coverage} we look at the average runtime
+for each tool when applied to simulated data sets generated from a
+single zebra fish gene at varying levels of coverage -- effectively
+plotting the runtime as a function of fragment set size.  For Sanger
+sequences, we see marked improvement of (\peace+)\east\/ over all
+other tools for almost all coverage levels: for the smallest set
+(coverage = 10) \east\/ has 1.9 speedup over \capthree\/ and a 3.9
+speedup over \tgicl, while at a more moderate size (coverage=50) we
+see speedups of 4.7 and 8.4.  This returns back down to 3.0 and 5.2 at the
+largest set tested (coverage = 100), as can be seen in the bump around
+a coverage of 85 in Figure~\ref{runtime.coverage}.  We are unable to
+superficially explain this bump, but have noticed the same effect in
+data sets derived from other genes as well.
+
+\east\/ is less competitive on shorter sequence (an effect of the
+$d^2$ bias of our algorithm, as discussed below), reducing its runtime
+improvement for the 250 bp 454 sequences, severely reducing it for the
+62 bp Illumina sequences, and also affecting the hybrid data sets (also
+due to the presence of the Illumina sequence).  We argue that the high
+quality of our results still makes \east\/ an appealing option for 454
+and hybrid data (while acknowledging that cannot complete with
+\velvet\/ on pure Illumina data), and also observe that while it has
+not been implemented, the basic \east\. algorithm should easily admit
+paralilization -- thus increasing the practical limit on data set size.
+
+\begin{figure}[htb]
+\centerline{\includegraphics[width=6in]{pics.d/runtime_fixedsize_sanger.pdf}}
+\caption{Average runtime per trial for when generating data for
+  Figures~\ref{sangerAscore}-\ref{hybridAscore}  when coverage = 30).
+  Note that the exceptionally high speed of \velvet\/ on Illumina sequences make the
+  red (center) bars in that plot essentially indistinguishable.
+  (\peace+)\east=blue, \capthree=Black, \tgicl=green,
+  \velvet=red,\mira=purple.}
+\label{runtime.fixed}
+\end{figure}
+
+\begin{figure}[htb]
+\centerline{\includegraphics[width=6in]{pics.d/runtime_bycoverage_all.pdf}}
+\caption{The average runtime when applying each tool to a simulated
+  data set derived from a specific zebrafish transcript of length
+  14,625 bp.  As we decrease coverage the number of fragments drops,
+  hence illustrating the relationship between tool runtime and
+  fragment set size.  Each data point represents the average runtime
+  from the application of the tool to at least 30 randomly generated
+  simulated sets coming from the transcript at the given level of coverage.
+  (\peace + )\east=blue, \capthree=Black, \tgicl=green,
+  \velvet=red,\mira=purple.}
+\label{runtime.coverage}
+\end{figure}
 
@@
 \section*{Conclusions}
+
+In this manuscript we have described the \east\/ algorithm and run a
+comparison of the \peast combination against the best preforming tools
+in the literature.  When compared against the fine benchmark tools on
+normalized a-score and runtime metrics, the \peast\/ combination of
+tools prove to be both the best tool for Sanger sequencing, the
+best tool overall, and by far the most robust against sequencing
+error.  While certain tools may do better than it in specific circumstances,
+there is no tool that does better than it on multiple sequencing
+technology platforms.  While its speed is slower when applied to short
+sequence fragments, the quality of its results more then compensate.
+
+As observed in the results section, when applying \east, \capthree,
+and \tgicl\/ to Sanger sequenced transcripts with normal to high error
+rates, \east\/ consistently out preforms all other tools in terms of
+the normalized a-score metric when applied to transcript fragments
+resulting from Sanger sequencing (Figures~\ref{sangerAscore},
+\ref{runtime.fixed}, and \ref{runtime.coverage}).  We also find that
+the number of contigs produced is almost exactly equal to the number
+of transcripts represented in the set, regardless of error rate, while
+the number of singleton ESTs remains low up to about a 5\% error rate
+(Figures~\ref{contigsSanger}).  The high normalized a-score indicates
+that \east\/ is both reconstructing larger portions of each transcript
+and doing so more faithfully, while the low number of contigs means
+that we are getting complete transcripts.  Finally, runtimes for
+\east\/ are considerably faster than other tools for larger sets
+(Figures~\ref{runtime.fixed} and \ref{runtime.coverage}, with
+(\peace+)\east\/ running a minimum of two to three times as fast as
+the other tools, with speedups over \capthree\/ (the faster tool)
+ranging from 1.9 to 6.0, depending on the coverage level.  In short,
+when used for Sanger Sequences \east\/ is a faster tool that produces
+the high quality reconstructions versus comparing tools and is
+significantly more robust to error than other tools.
+
+When involving shorter sequences, the runtime of \east\/ takes a
+significant hit.  As explained in our methods section, \east\/ is
+dependent on the $d^2$ sequence distance function \cite{Hide94}, whose
+parameters must be adjusted when applied to smaller sequences to
+preserve quality, at the cost of runtime.  We describe below our {\it
+  adoptive $d^2$} algorithm that allows us to target more specific
+parameters on a case-by-case basis, but large numbers of short
+sequences will still lead to slower runtimes.  In our tests our
+simulated 454 sequences averaged 250 bp in length, while our Illumina
+sequences were uniformly 62 bp, accounting for the slightly reduced
+improvement of \east, as compared to other tool runtimes, in the 454 tests and
+the significant reduction for the Illumina and Hybrid test sets
+containing the short Illumina sequences.
+
+Despite the slower runtime, it can still be argued that the \east\/ is
+competitive, given the quality of its output, excepting when compared
+to \velvet\/ on Illumina data.  For 454 sequences, no tool comes even
+close to matching the \east\/ results.  \east is able to match
+\velvet for high-coverage Illumina data, though loses out at lower
+coverages (and is not even remotely competitive in terms of runtime).
+Looking at hybrid sets, which the \mira\/ tool is specifically geared
+to address, we again find \east\/ to produce significantly superior
+results at moderate to high levels of coverage.  At low coverage
+levels \east\/ does suffer (another effect of the short
+Illumina sequences), but its robustness then leads to
+better results when the Sanger sequence portion of the set is subject
+to elevated error rates.
 
@@
 determine how the nodes overlap, we calculate a modified $d^2$ score
-as follows.  The standard calculation of $d^2$, as define by {\it Hide
+as follows.  The standard calculation of $d^2$, as defined by {\it Hide
   et al.} \cite{Hide94}, requires looking at every pair of length $w$
@@
 
-In Figure~\ref{fig:overlap} we see illustrate the use of modified $d^2$,
+In Figure~\ref{fig:overlap} we illustrate the use of modified $d^2$,
 using it to take two sequences (a), find the window on $s_2$ most
@@
 comparisons between sequences of significantly different sizes we
-avoid the pitfalls of such comparisons.
+avoid the pitfalls of such comparisons, while still picking up the
+connections through transitivity.
 
@@
  \bibliographystyle{bmc_article}  % Style BST file
-  \bibliography{bmc_article} }     % Bibliography file (usually '*.bib' )
+  \bibliography{all} }     % Bibliography file (usually '*.bib' )
 
@@
 
-\section*{Figures}
-  \subsection*{Figure 1 - Sample figure title}
-      A short description of the figure content
-      should go here.
+% \section*{Figures}
+%   \subsection*{Figure 1 - Sample figure title}
+%       A short description of the figure content
+%       should go here.
 
-  \subsection*{Figure 2 - Sample figure title}
-      Figure legend text.
+%   \subsection*{Figure 2 - Sample figure title}
+%       Figure legend text.
 
@@
 %%
-\section*{Tables}
-  \subsection*{Table 1 - Sample table title}
-    Here is an example of a \emph{small} table in \LaTeX\ using  
-    \verb|\tabular{...}|. This is where the description of the table 
-    should go. \par \mbox{}
-    \par
-    \mbox{
-      \begin{tabular}{|c|c|c|}
-        \hline \multicolumn{3}{|c|}{My Table}\\ \hline
-        A1 & B2  & C3 \\ \hline
-        A2 & ... & .. \\ \hline
-        A3 & ..  & .  \\ \hline
-      \end{tabular}
-      }
-  \subsection*{Table 2 - Sample table title}
-    Large tables are attached as separate files but should
-    still be described here.
+% \section*{Tables}
+%   \subsection*{Table 1 - Sample table title}
+%     Here is an example of a \emph{small} table in \LaTeX\ using  
+%     \verb|\tabular{...}|. This is where the description of the table 
+%     should go. \par \mbox{}
+%     \par
+%     \mbox{
+%       \begin{tabular}{|c|c|c|}
+%         \hline \multicolumn{3}{|c|}{My Table}\\ \hline
+%         A1 & B2  & C3 \\ \hline
+%         A2 & ... & .. \\ \hline
+%         A3 & ..  & .  \\ \hline
+%       \end{tabular}
+%       }
+%   \subsection*{Table 2 - Sample table title}
+%     Large tables are attached as separate files but should
+%     still be described here.
 
@@
 
-\section*{Additional Files}
-  \subsection*{Additional file 1 --- Sample additional file title}
-    Additional file descriptions text (including details of how to
-    view the file, if it is in a non-standard format or the file extension).  This might
-    refer to a multi-page table or a figure.
+% \section*{Additional Files}
+%   \subsection*{Additional file 1 --- Sample additional file title}
+%     Additional file descriptions text (including details of how to
+%     view the file, if it is in a non-standard format or the file extension).  This might
+%     refer to a multi-page table or a figure.
 
-  \subsection*{Additional file 2 --- Sample additional file title}
-    Additional file descriptions text.
+%   \subsection*{Additional file 2 --- Sample additional file title}
+%     Additional file descriptions text.
 
@@
 % LocalWords:  Ozden Chun Liang Karro Botanty publ MSTs et al Hazelhurst Prim's
-% LocalWords:  mers Illumina contig bmc multi
+% LocalWords:  mers Illumina contig bmc multi LaTeX xpression ssembly rees Rao
+% LocalWords:  Zhang Moler Dhananjai Mufit de novo introns NGS bruijn ABIITY bp
+% LocalWords:  GenBank indels contigs TGICL Metasim MetaSim Runtime runtime pre
+% LocalWords:  zebrafish runtimes ESTs paralilization

### 2011-06-08 peace/papers/EAST/east.tex
Major modifications to EAST manuscript.

@@
 \documentclass[10pt]{bmc_article}    
+\usepackage[usenames]{color}
 
@@
   \and
-  James C Moler$^1$
-  \email{James Moler - molerjc@muohio.edu}
+  Jens Mueller$^2$
+  \email{Jens Mueller - muellej@muohio.edu}
   \and
-  Chun Liang\correspondingauthor$^{1,2}$
+  Chun Liang\correspondingauthor$^{1,3}$
   \email{Chun Liang\correspondingauthor - liangc@muohio.edu}
   and
-  John E. Karro\correspondingauthor$^{1,3}$
+  John E. Karro\correspondingauthor$^{1,4.5}$
   \email{John E. Karro\correspondingauthor - karroje@muohio.edu}
@@
 \address{
-    \iid(1)Department of Computer Science and Software Engineering, Miami University, Oxford OHIO, USA \\
-    \iid(2)Department of Botanty, Miami University, Oxford OHIO, USA \\
-    \iid(3)Department of Microbiology, Miami University, Oxford OHIO, USA
+    \iid(1)Department of Computer Science and Software Engineering,  \\
+    \iid(2)Informational Technology Services
+    \iid(3)Department of Botanty, \\
+    \iid(4)Department of Microbiology, \\
+    \iid(5)Department of Statistics, Miami University, Oxford OHIO, USA
 }
@@
 \begin{abstract}
-        % Do not use inserted blank lines (ie \\) until main body of text.
 \end{abstract}
@@
 %%
-\section*{Introduction}
-The {\it de novo} assembly of RNA transcripts is an important and
-computationally challenging problem, made more difficult by the advent
-of Next Generation Sequencing technologies such as 454 or Illumina
-\cite{Nagaraj07,Rao10}.  Given the completion of a large-scale sequencing run of
-multiple mature transcripts of a given organism, the investigator is
-left with a large set of sequence fragments, but no ordering or even
-transcript association information; these fragments must be clustered
-by transcript then assembled before the sequencing output can be truly
-useful.  If the genome has not been sequenced, this must be done {\it
-  de novo} -- using only the sequence information extrapolated from
-the sequences themselves.  Even when the genomic sequence is known,
-there are a number of advantages to ignoring the known sequence
-information and starting from scratch.  Using an existing sequence as
-a template for sequencing in population studies has the potential to
-introduce bias.  And when looking specifically for genes, a known
-sequence may be of limited use: the target transcripts may be
-undiscovered, and the splicing of introns before sequencing can
-complicate the approach of mapping fragments to the genomic sequence.
-In short, there is significant value in tools that can reassemble
-fragment data into the original assembly without requiring access to
-genomic information.
+  \section*{Introduction} {\color{blue} The {\it de novo} assembly of
+    RNA transcripts is an important and computationally challenging
+    problem, made more difficult by the advent of Next Generation
+    Sequencing technologies such as 454 and Illumina
+    \cite{Nagaraj07,Rao10}.  Upon completion of
+    transcriptome-sequencing, the investigator is left with a large
+    set of sequence fragments but no ordering or transcript
+    association information.  These fragments must be clustered by
+    transcript and assembled before the sequencing output can be truly
+    useful.  {\it de novo} assembly tools assemble the
+    transcript out of the fragments without using genmic sequence
+    information.  {\it de novo} tools are obivously necessary when working with a genome that is
+    unsequenced or poorly annotated, but is also useful when there is
+    limited knowledge of gene splice sites, as well as for the discovery
+    of novel events \cite{Birol09,Robertson10}.}
+
+\vspace{3mm}
 
@@
 Sequencing techniques, and newer approaches looking at produces of
-Next Generation Sequencing (NGS) technologies.  A list of the tools
-most successfully addressing these problems include \capthree, \tgicl,
-\velvet, and \mira, each excelling within their own context
+Next Generation Sequencing (NGS) technologies.  A list of popular
+tools addressing these problems include \capthree, \tgicl, \velvet,
+and \mira, each excelling within their own context
 \cite{Huang99,Pertea03,Chevreux04,Zerbino08}.  \capthree\/ and
@@
 results when applied to the output of Sanger sequencing technologies.
-\velvet, based on the modeling of the problem with a {\it de bruijn}
+\velvet, based on the modeling of the problem with {\it de bruijn}
 graphs, was designed to handle short read sequences.  \mira\/ is a
-multi-pass DNA assembler based on an overlap-layout-consensus
-strategy.  [STILL NEED TO SAY SOMETHING ABOUT MIRA AND HYBRID SETS.]
+multi-pass DNA assembler based on an overlap-layout-consensus strategy
+that also has an increased ability to assemble {\it hybrid sets}: sets
+formed by the mixture of sequencing results from multiple technologies
+\cite{MiraWeb}.
 
-In this paper we present \east, a tool for the assembly of both
+\vspace{3mm}
+
+In this paper we present \east: a tool for the assembly of
 Sanger, NGS and hybrid data with the capacity to incorporate quality
-scores when available.  \east\/ is based on a novel minimum spanning
-tree (MST) approach to compute assembly.  The tool works in tandem with
-the \peace\/ clustering tool \cite{Rao10}: \peace\/ clusters
+scores when available.  \east\/ is based on a novel Minimum Spanning
+Tree (MST) approach to compute assembly.  The tool works in tandem
+with the \peace\/ clustering tool \cite{Rao10}: \peace\/ clusters
 fragments by transcript and produces the initial MSTs that will serve
-as the basis for the \east\/ algorithm.  In the following we present a
-comparison of \east\/ against those of \capthree, \tgicl, \velvet, and
-\mira on both Sanger and NGS sequences, showing that \peast\/
+as the basis for the \east\/ algorithm.  In this paper we present a
+comparison of \east\/ against \capthree, \tgicl, \velvet, and
+\mira\/ on both Sanger and NGS sequences, showing that \peast\/
 generates the best quality results in the fastest runtime for Sanger
-sequences, and that while certain tools out-preform \peast\/ under specific
-conditions, it is the only tool able to produce constant quality results
-for Sanger, 454, Illumina, {\it and} hybrid data sets.
+sequences.  Although certain tools out-perform \peace\/ under
+specific conditions, we show that it is the only tool able to produce consistently
+higher quality results for Sanger, 454, Illumina, {\it and} hybrid
+data sets.  In this paper we discuss the results and relevant
+computational methods.  \east\/ has been incorporated into the
+\peace\/ clustering tools \cite{Rao10}; it can be installed
+and run through a Java-based PEACE GUI available at
+www.peace-tools.org.
 
@@
 
-In assessing result quality, we must view the \peace\/ clustering tool
+\textcolor{blue}{In assessing result quality, we must view the \peace\/ clustering tool
 \cite{Rao10} and \east\/ as a combined process.  While most assembly
 tools integrate these two jobs, \east\/ assumes clustering has been
-preformed and requires the \peace\/ MST output.  While other cluster
-tools could be substituted in to preform the actual clustering
-(e.g. \wcd or \easycluster \cite{Hazelhurst08,Picardi09}), only
-\peace\/ provides the requires MST output.  Hence for purposes of the
-following section, we consider a single tool \peast, and assess the
-quality of these results.
+performed and requires the \peace\/ MST output.  Hence for purposes of
+the following section, we consider them as a single tool \peast, and
+assess the quality of these results.}
 
-In order to assess the quality of the (\peace\/ +) \east\/ results, we compared
+\vspace{3mm}
+
+In order to assess the quality of (\peace\/ +) \east\/ results, we compared
 against four prominent assembly tools: \capthree, \tgicl,
-\mira, and \velvet\/ \cite{Huang99,Pertea03,Chevreux04,Zerbino08}.  Tools were tested on data from three
+\mira, and \velvet\/ \cite{Huang99,Pertea03,Chevreux04,Zerbino08}.  The tools were tested on data from three
 sequencing technologies (Sanger, 454 and Illumina), using both
 simulated output obtained from the \estsim\/ and \metasim\/ tools
-\cite{Hazelhurst03,Richter08} and benchmark sets obtained from GenBank \cite{Benson10}.
+\cite{Hazelhurst03,Richter08}.
 
@@
 bound the score by the relative size of the reconstruction.
-(e.g. Perfectly reconstruction one fourth of the original sequence
-will lead to an a score of $0.25$.)  In addition to a-score we also
-look at: the number of contigs produced by a tool (which, in the
-perfect solution, will be exactly equal to the number of transcripts
-represented by the data set), normalized by the number of transcripts;
-the number of singleton ESTs in the solution (i.e. ESTs unassigned to
-any contig and thus, most cases, effectively thrown away), and the runtime of the tool.
+\textcolor{blue}{(For example, were a tool only able to reconstruct
+  one fourth of a transcript -- but able to reconstruct that fraction
+  perfectly -- the resulting contig's a-score would be $0.25$.)}  In
+addition we look at: the number of contigs produced by a tool
+normalized by the number of transcripts (a value equal to 1 in a
+perfect reconstruction); the number of singleton ESTs in the solution
+(i.e. ESTs unassigned to any contig and thus, most cases, effectively
+thrown away), and the runtime of the tool.
 
 \subsection*{Simulation Results}
-Using simulated fragment sets allows us to both compare assemblies to
+Using simulated fragment sets allows us to compare assemblies to
 a known correct solution and control parameters in their model
 (e.g. base-call error rate) to test the effect of variation.  All sets
-were generated from a set of 100 zebra fish transcripts obtained from
-{\it Hazelhurst et al.}  \cite{Hazelhurst08}.  For simulated Sanger
-sequences we used the \estsim\/ tool, using its default model of EST
-generation when not otherwise specified.  For simulated
-454 and Illumina sets we used the \metasim\/ tool \cite{Richter08}, again using their
-default model to simulate the errors associated with those
-technologies.
+were generated from a collection of 100 zebra fish transcripts
+obtained from {\it Hazelhurst et al.}  \cite{Hazelhurst08}.  For
+simulated Sanger sequences we used the \estsim\/ tool, using its
+default model of EST generation when not otherwise specified
+\cite{Hazelhurst03}.  For simulated 454 and Illumina sets we used the
+\metasim\/ tool, again using their default model to simulate the
+errors associated with those technologies \cite{Richter08}. {\bf
+  [COMMENT: Will try to add a discussion of, or reference to, the
+  default values here.]}
 
-\noindent {\bf Sanger Sequencing:} We first compare the assembly
-quality of the tools on Sanger Sequencing, using \east, \capthree, and
-\tgicl\/ (the tools designed to work with Sanger sequences), looking
-at the a-score as a function of error rate for a range of coverage
-depths.  For each parameter combination we applied each tool to 30
-independently randomly generated simulated sets, derived from 15 zebra
-fish genes, reporting the average normalized a-score.  In
+\vspace{3mm}
+
+\noindent {\bf Sanger Sequencing:} \textcolor{blue}{We first compare the
+  assembly quality of the tools on a Sanger Sequencing model, using
+  \east, \capthree, and \tgicl\/, the tools designed to work with
+  Sanger sequence.  The quality of any assembly will be directly
+  affected by the {\it covreage} of the input: the average number of
+  ESTs that cover a based.  Thus we investigate the a-score as a
+  function of error rate at different coverage levels.}  For each
+parameter combination we applied each tool to 30 independent, randomly
+generated simulated sets, derived from 15 randomly chosen members of
+our zebra fish gene set, reporting the average normalized a-score.  In
 Figure~\ref{sangerAscore} we look at normalized a-score as a function
-of error rate, fixing coverage at 50, 30 and 10 and varying base-read
+of error rate, fixing coverage at 50, 30 and 10 and varying base read
 error rate from 0\% to 6\% for each of the three tools designed for
-Sanger Sequencing.  While all three tools preform well when applied to
+Sanger Sequencing.  While all three tools perform well when applied to
 error free (i.e. impossible) sequencing technology, we find \east\/ to
@@
 error rate we see \east, \capthree, and \tgicl\/ all achieving an
-essentially perfect normalize a-score .  At a 3\% error rate \east\/
+essentially perfect normalized a-score.  At a 3\% error rate \east\/
 maintains its almost perfect normalized a-score, showing a 2\%
@@
 a 54\% and 85\% improvement over \capthree\/ and \tgicl\/
-respectively), and even at a 6\% error rate is still achieving an
+respectively).  Even at a 6\% error rate \east\/ is still achieving a
 normalized a-score of 0.95 (a 110\% improvement over the next closest
-tool, \tgicl\/), meaning that at most 5\% of the bases were left out of
-their contigs or were incorrectly reconstructed.
+tool, \tgicl\/), meaning that at most 5\% of the bases were left out
+of their contigs or were incorrectly reconstructed.
 
 \begin{figure}[htb]
-\centerline{\includegraphics[width=6in]{pics.d/ascore_sanger.pdf}}
+\centerline{\includegraphics[width=6in]{pics.d/ascore_sanger_norm.pdf}}
 \caption{Average normalized a-score as a function of base error rate for
   simulated Sanger sequencing.  \east=blue, \capthree=Black,
-  \tgicl=green, \velvet=red, \mira=purple.}
+  \tgicl=green. COMMENT: Need to change figure heading to Normalized a-score.}
 \label{sangerAscore}
@@
 
-Another useful metric is the number of contigs and number of singleton
-fragments in the final result.  In Figure~\ref{contigsSanger} is a
+In Figure~\ref{contigsSanger} is a
 plot of the average number of contigs per gene sequence (left column), and
@@
 comparable results for lower error rates, but only \east\/ is able to
-maintain the low score with increasing error rates -- generating an
+maintain the low score with increasing error rates -- generating a
 perfect score in almost every case.  In minimizing the
 number of singletons, we find that \east\/ does a better job than
-\capthree\/ for error rates larger than $2\%$, while \tgicl does a
+\capthree\/ for error rates larger than $2\%$, while \tgicl\/ does a
 better job that \east\/ for error rates larger than $4\%$ to $5\%$.
@@
 \caption{Number of contigs (normalized to the number of transcripts)
-  and number of singletons as a function of base call error
+  and number of singletons as a function of base-call error
   rate for simulated Sanger sequencing.  \east=blue, \capthree=Black,
-  \tgicl=green, \velvet=red, \mira=purple.  (Number of singles for
+  \tgicl=green.  (Number of singletons for
   \velvet\/ and \mira\/ not shown.)}
@@
 250 and 62 bp respectively).  For these tests we varied coverage
-levels but otherwise took the default Metasim model.  On 454 data we
-see significant improvement in results quality in \east\/ for all
-levels of coverage.  On Illumina data we see that \east\/ is
-competitive with \velvet\/ at higher coverages, though is unable to
-match that tool when coverage is sparse.
+levels but otherwise continued to use the default Metasim model
+\cite{Richter08}.  On 454 data we see significant improvement in
+results quality in \east\/ for all levels of coverage.  On Illumina
+data we see that \east\/ is competitive with \velvet\/ at higher
+coverages, though is unable to match that tool when coverage is
+sparse.
 
@@
   output, using the \metasim default model at different levels
-  of coverage.  \east=blue, \capthree=Black, \tgicl=green,
-  \velvet=red, \mira=purple.  }
+  of coverage.  \east=blue, \capthree=Black, \tgicl=green.  }
 \label{nextgenAscore}
@@
 set of Sanger (40\%), 454 (40\%) and Illumina (20\%) sequences, again
-showing normalized a-score as a function of base call error rate in
-the Sanger portion.  Here we observe \east matching or outperforming
+showing normalized a-score as a function of base-call error rate in
+the Sanger portion.  Here we observe \east\/ matching or outperforming
 \capthree\/ and \mira\/ at moderate to high coverages.  At a coverage
@@
 at a coverage level of 30 (hovering around a normalized a-score of
-0.95, showing a 142\% improvement of Mira at the 3\% base-call error
+0.95, showing a 142\% improvement over \mira\/ at the 3\% base-call error
 rate), an oddity we attribute to the decreased absolute number of NGS
 sequences with the smaller coverage.  At a coverage level of 10
-\east's normalized a-score is around 0.37 -- doing considerably worse
-that \mira\/ at low error rates (e.g. 0.94 v. 0.40 for error-free
-bases), but showing a 52\% improvement over \mira\/ at the 3\% error
-rate mark and in general showing much more robustness to error.
-Remarkably, it is only in this situation that \capthree\/ is
+\east's normalized a-score is around 0.37 -- performing considerably
+worse than \mira\/ at low error rates (e.g. 0.94 v. 0.40 for
+error-free bases), but showing a 52\% improvement over \mira\/ at the
+3\% error rate mark and in general showing increased robustness to
+error.  Remarkably, it is only in this situation that \capthree\/ is
 competitive, with that tool achieving a normalized a-score of 0.62
-(v. \east's 0.37) for the 3\% base call error rate, but loosing out to
-\east at greater than a 4.5\% base call error rate -- a result that we
-attribute to \east's problems with the 62 bp Illumina sequences (see below).
+(v. \east's 0.37) for the 3\% base-call error rate, but loosing out to
+\east\/ at greater than a 4.5\% base-call error rate -- a situation
+likely due to \east's problems with the 62 bp Illumina sequences (see
+below).
 
@@
 \centerline{\includegraphics[width=6in]{pics.d/ascore_hybrid.pdf}}
-\caption{Comparative normalized a-scores a hybrid set of Sanger, 454
-  and Illumina sequence output, varying the base call error rate for
-  Sanger sequences and using the MetaSim default model for the 454 and
-  Illumina sequences.  Each data set tested was composed of 40\%
-  Sanger Sequences, 40\% 454 Sequences, and 20\% Illumina sequences.
-  \east=blue, \capthree=Black, \tgicl=green,
-  \velvet=red,\mira=purple.}
+\caption{Comparative normalized a-scores for a hybrid set consisting
+  of Sanger, 454 and Illumina sequence output, varying the base-call
+  error rate for Sanger sequences and using the MetaSim default model
+  for the 454 and Illumina sequences.  Each data set tested was
+  composed of 40\% Sanger Sequences, 40\% 454 Sequences, and 20\%
+  Illumina sequences.  \east=blue, \capthree=Black, \tgicl=green,
+  \velvet=red, \mira=purple.}
 \label{hybridAscore}
@@
 
-\noindent {\bf Tool Runtime:} As \east\/ assembly requires the
-\peace\/ cluster (as opposed to the competing tools that do both
-jobs), for \east\/ runtime we actually include the \peace\/
-runtime. In Figures~\ref{runtime.fixed} we see the average runtime for
-each of the experiments reflected in Figures
-Figures~\ref{sangerAscore}-\ref{hybridAscore} for a coverage of 30,
-while in Figure~\ref{runtime.coverage} we look at the average runtime
-for each tool when applied to simulated data sets generated from a
-single zebra fish gene at varying levels of coverage -- effectively
-plotting the runtime as a function of fragment set size.  For Sanger
-sequences, we see marked improvement of (\peace+)\east\/ over all
-other tools for almost all coverage levels: for the smallest set
-(coverage = 10) \east\/ has 1.9 speedup over \capthree\/ and a 3.9
-speedup over \tgicl, while at a more moderate size (coverage=50) we
-see speedups of 4.7 and 8.4.  This returns back down to 3.0 and 5.2 at the
-largest set tested (coverage = 100), as can be seen in the bump around
-a coverage of 85 in Figure~\ref{runtime.coverage}.  We are unable to
+\noindent {\bf Tool Runtime:} In Figures~\ref{runtime.fixed} we see
+the average runtime for each of the experiments reflected in
+Figures~\ref{sangerAscore}-\ref{hybridAscore} for a coverage of 30
+(continuing to report the combined runtime of \peace+\east), while in
+Figure~\ref{runtime.coverage} we look at the average runtime for each
+tool when applied to simulated data sets generated from a single zebra
+fish gene at varying levels of coverage.  For Sanger sequences we
+see marked improvement of (\peace+)\east\/ over all other tools for
+almost all coverage levels: for the smallest set (coverage = 10)
+\east\/ has 1.9 speedup over \capthree\/ and a 3.9 speedup over
+\tgicl, while at a more moderate size (coverage=50) there are speedups of
+4.7 and 8.4.  This returns back down to 3.0 and 5.2 at the largest set
+tested (coverage = 100), as can be seen in the bump around a coverage
+of 85 in Figure~\ref{runtime.coverage}.  We are unable to
 superficially explain this bump, but have noticed the same effect in
@@
 
-\east\/ is less competitive on shorter sequence (an effect of the
-$d^2$ bias of our algorithm, as discussed below), reducing its runtime
-improvement for the 250 bp 454 sequences, severely reducing it for the
-62 bp Illumina sequences, and also affecting the hybrid data sets (also
-due to the presence of the Illumina sequence).  We argue that the high
-quality of our results still makes \east\/ an appealing option for 454
-and hybrid data (while acknowledging that cannot complete with
-\velvet\/ on pure Illumina data), and also observe that while it has
-not been implemented, the basic \east\. algorithm should easily admit
-paralilization -- thus increasing the practical limit on data set size.
+\vspace{3mm}
+
+\east\/ is less competitive on shorter sequence.  In our simulations,
+we find a reduced runtime improvement on the simulated 454 sequences
+(averaging 250 bp in length), and a significantly reduced runtime on
+the simulated Illumina sequences (exactly 62 bp in length).  However,
+as technology advances, the ``short-read'' sequencers are producing
+increasingly longer reads: 454 sequencing technology can produce
+reads exceeding 400 bp, while Illumina sequences can exceed 100 bp
+\cite{Eid09,Li10}.  The loss of runtime in our experiments on 454,
+Illumina and the hybrid sequences are due to the shorter sequence
+(specifically, the problems with applying the $d^2$ metric to shorter
+sequences -- as discussed below), hence are becoming less of an issue
+with this increase in short-read fragment length.
+
+\vspace{3mm}
+
+In short, we argue that the high quality of our results still makes
+\east\/ an appealing option for 454 and hybrid data (while
+acknowledging that \east\/ cannot currently compete with \velvet\/ on pure
+Illumina data), and also observe that while it has not been
+implemented, the basic \east\/ algorithm will easily admit
+parallelization -- thus increasing the practical limit on data set
+size.
 
@@
 \centerline{\includegraphics[width=6in]{pics.d/runtime_fixedsize_sanger.pdf}}
-\caption{Average runtime per trial for when generating data for
-  Figures~\ref{sangerAscore}-\ref{hybridAscore}  when coverage = 30).
+\caption{Average runtime per trial when when generating data for
+  Figures~\ref{sangerAscore}-\ref{hybridAscore}  when coverage = 30.
   Note that the exceptionally high speed of \velvet\/ on Illumina sequences make the
@@
   (\peace+)\east=blue, \capthree=Black, \tgicl=green,
-  \velvet=red,\mira=purple.}
+  \velvet=red, \mira=purple.}
 \label{runtime.fixed}
@@
   simulated sets coming from the transcript at the given level of coverage.
-  (\peace + )\east=blue, \capthree=Black, \tgicl=green,
-  \velvet=red,\mira=purple.}
+  \peace + \east=blue, \capthree=Black, \tgicl=green,
+  \velvet=red, \mira=purple.}
 \label{runtime.coverage}
+\end{figure}
+
+\noindent{\bf Implementation:} \east\/ has been integrated into the
+\peace\/ clustering package \cite{Rao10} and is available for
+download from the \peace\/ website ({\bf www.peace-tools.org}).
+Users can run the GUI provided by that site, and through it
+install both the \peace\/ cluster engine and \east\/ assembly engine
+on a local machine or remote server (see Figure~\ref{GUI-fig}).
+Initial EST data must be in fasta format, though the tool can print
+out the final assembly in fasta, ACE, SAM and BAM formats [NEED TO ADD
+CITATION TO THE CONVERSION TOOL].
+
+\begin{figure}[htb]
+\centerline{\includegraphics[width=6in]{pics.d/GUI1.pdf}}
+\caption{Screen shots of the PEACE GUI.  [COMMENT: Need to improve this.]}
+\label{GUI-fig}
 \end{figure}
@@
 
-In this manuscript we have described the \east\/ algorithm and run a
-comparison of the \peast combination against the best preforming tools
-in the literature.  When compared against the fine benchmark tools on
-normalized a-score and runtime metrics, the \peast\/ combination of
-tools prove to be both the best tool for Sanger sequencing, the
-best tool overall, and by far the most robust against sequencing
-error.  While certain tools may do better than it in specific circumstances,
-there is no tool that does better than it on multiple sequencing
+{\textcolor{blue}{In this manuscript we have described the \east\/ algorithm and we have run a
+comparison of the \peast\/ tool combination against popular tools
+from the literature.}  When compared against competing tools on simulated
+data using the normalized a-score and runtime metrics, the \peast\/ combination of
+tools produce the best sequence assemblies for Sanger sequence, the
+best tool overall [COMMENT: I need a better way to phrase this --
+something to indicate the do better as a whole, even if they don't do
+as well as some tools on some sets], and by far the most robust against sequencing
+error.  While certain tools may do better than \east\/ in specific circumstances,
+none of them do better than EAST on multiple sequencing
 technology platforms.  While its speed is slower when applied to short
-sequence fragments, the quality of its results more then compensate.
+sequence fragments, the quality of its results more then compensate.}
+\vspace{3mm}
 
@@
 and \tgicl\/ to Sanger sequenced transcripts with normal to high error
-rates, \east\/ consistently out preforms all other tools in terms of
-the normalized a-score metric when applied to transcript fragments
-resulting from Sanger sequencing (Figures~\ref{sangerAscore},
+rates, \east\/ consistently out performs all other tools in terms of
+the normalized a-score metric (Figures~\ref{sangerAscore},
 \ref{runtime.fixed}, and \ref{runtime.coverage}).  We also find that
@@
 and doing so more faithfully, while the low number of contigs means
-that we are getting complete transcripts.  Finally, runtimes for
+that we are assembling complete transcripts.  Finally, runtimes for
 \east\/ are considerably faster than other tools for larger sets
-(Figures~\ref{runtime.fixed} and \ref{runtime.coverage}, with
+(Figures~\ref{runtime.fixed} and \ref{runtime.coverage}), with
 (\peace+)\east\/ running a minimum of two to three times as fast as
-the other tools, with speedups over \capthree\/ (the faster tool)
-ranging from 1.9 to 6.0, depending on the coverage level.  In short,
+the other tools.  The speedup over \capthree\/ (the faster tool)
+ranged from 1.9 to 6.0, depending on the coverage level.  In short,
 when used for Sanger Sequences \east\/ is a faster tool that produces
@@
 significantly more robust to error than other tools.
+
+\vspace{3mm}
 
@@
 parameters must be adjusted when applied to smaller sequences to
-preserve quality, at the cost of runtime.  We describe below our {\it
-  adoptive $d^2$} algorithm that allows us to target more specific
-parameters on a case-by-case basis, but large numbers of short
-sequences will still lead to slower runtimes.  In our tests our
+preserve quality, at the cost of runtime.  We describe in the methods
+section our {\it adaptive $d^2$} algorithm that allows us to target
+more specific parameters on a case-by-case basis, but large numbers of
+short sequences will still lead to slower runtimes.  In our tests the
 simulated 454 sequences averaged 250 bp in length, while our Illumina
-sequences were uniformly 62 bp, accounting for the slightly reduced
-improvement of \east, as compared to other tool runtimes, in the 454 tests and
-the significant reduction for the Illumina and Hybrid test sets
-containing the short Illumina sequences.
+sequences were uniformly 62 bp.  These shorter lengths, and their
+interaction with the $d^2$ distance metric, account for the slightly
+reduced improvement of \east, as compared to other tool runtimes, in
+the 454 tests and the significant reduction for the Illumina and
+Hybrid test sets containing the short Illumina sequences, but are
+becoming less of an issue as short-read sequencing technologies are
+improved to produce longer fragments.  Looking at hybrid sets, which
+the \mira\/ tool is specifically geared to address, we again find
+\east\/ to produce significantly superior results at moderate to high
+levels of coverage.  At low coverage levels \east\/ does suffer (again
+an effect of the presence of the very short sequences), but its
+robustness to error leads to better results when the Sanger sequence
+portion of the set is subject to elevated error rates.
 
-Despite the slower runtime, it can still be argued that the \east\/ is
-competitive, given the quality of its output, excepting when compared
-to \velvet\/ on Illumina data.  For 454 sequences, no tool comes even
-close to matching the \east\/ results.  \east is able to match
-\velvet for high-coverage Illumina data, though loses out at lower
-coverages (and is not even remotely competitive in terms of runtime).
-Looking at hybrid sets, which the \mira\/ tool is specifically geared
-to address, we again find \east\/ to produce significantly superior
-results at moderate to high levels of coverage.  At low coverage
-levels \east\/ does suffer (another effect of the short
-Illumina sequences), but its robustness then leads to
-better results when the Sanger sequence portion of the set is subject
-to elevated error rates.
+\vspace{3mm}
+
+Finally, we note, but have not yet explored, the fact that \east\/ is
+amenable to to paralilization and the possibility of identifying
+isophormes resulting from alternative splicing.  For the first: the
+\east\/ tool is trivially parallelizable in that once the \peace\/
+portion has completed clustering, \east\/ can address the clusters in
+parallel.  More sophisticated paralilization techniques are applicable
+at different phased of the east assembly (e.g. the depth-first search
+of the MST, or the multiple independent applications of the $d^2$ metric).
 
@@
 applications of such algorithms, we make use MSTs to provide a {\it
-  guide} through the EST set, dictating which pairs we aligned.  A
+  guide} through the EST set, dictating which pairs are to be aligned.  A
 modified version of the $d^2$ measure provides the edge weights for
@@
 
-\east\/ is technically an assembly tool, and require clustering be
+\vspace{3mm}
+
+\east\/ is an assembly tool, and requires clustering be
 done as a pre-processing step, as well as requiring that $d^2$-based
@@
 (explained in more detail below).  By employing the $u/v$ and $t/v$
-filtering heuristics of {\it Hazelhurst et al.} \cite{Hazelhurst08}, we
-can quickly dismiss most of the edges as irrelevant (i.e. connecting
-nodes that belong in different clusters), then use Prim's algorithm to
-compute an MST \cite{Prim57}.  We then remove all
+filtering heuristics of {\it Hazelhurst et al.} \cite{Hazelhurst08},
+we can quickly dismiss most of the edges as irrelevant
+(i.e. connecting nodes that belong in different clusters), then use
+Prim's algorithm to compute an MST \cite{Prim57}.  We then remove all
 edges exceeding a specified threshold value, and take each component
-of the resulting forest as a cluster.  As a result \peace\/ can pass
-to \east\/ a set of clusters, an MST for each cluster, and the
-orientation of each fragment relative to the cluster.  Hence \east\/
-has not only the clusters, but the \peace\/ MST and may quickly adjust
-the sequences to ensure all are on the same strand.
+of the resulting forest as a cluster.  As a result, \peace\/ passes to
+\east\/ a set of clusters, an MST for each cluster, and the
+orientation of each fragment relative to the cluster.
 
@@
 the $d^2$ sequence distance measure.
+
+\vspace{3mm}
 
@@
 wish to determine of the left end of $s_1$ overlaps the right end of
-$s_2$, or the reverse (recalling that we know them to be on the same
+$s_2$, or the reverse (recalling that \peace\/ has ensured they are on the same
 strand).  Hence we need only consider the left-most and right-most
@@
 
+\vspace{3mm}
+
 In Figure~\ref{fig:overlap} we illustrate the use of modified $d^2$,
@@
 resembling the right-most window of $s1$ (red section of (b)),
-extending to the end of $s_2$ (green section of (c)) and preforming a
+extending to the end of $s_2$ (green section of (c)) and performing a
 {\it global} alignment on the overlapping portion.  We then compute the
@@
 small, medium and large fragments), and compute distances only
-between segments in the name or neighboring groups.  By skipping the
+between segments in the same or neighboring groups.  By skipping the
 comparisons between sequences of significantly different sizes we
@@
 
+\vspace{3mm}
+
 Having identified the distance of each node from its two closest
@@
 with the overlap distance.  Using this graph we calculate a new
-minimum spanning tree, the starting from the root of the tree
-(necessarily a left-end of the contig), we transverse the tree and
+minimum spanning tree, then starting from the root of the tree
+(necessarily a left-end of the contig) we transverse the tree and
 assign the first base of each sequence a position relative to its
@@
 alignment of the fragments that allows us to derive a consensus
-sequence.
+sequence.  During this proces quality score information can also be
+incorperated, reducing the scor contribution of a specific base-pair
+if the involved bases are of low qulaity.
+
+% Comments from Jenna:
+% The following paragraph is from a paper
+% (http://bioinformatics.oxfordjournals.org/content/26/5/589.abstract):
+% roche/454 sequencing technology has already
+% produced reads >400 bp in production, Illumina gradually increases
+% read length >100 bp, and Pacific Bioscience generates 1000 bp reads
+% in early testing (Eid et al., 2009). Reads coming from the new
+% sequencing technologies are not short any more, which effectively
+% rules out many of the new aligners exclusively designed for reads no
+% longer than 100 bp.
+
+% Do we need further argue that the performance in very short reads
+% (64bp) is not important due to the improvement in sequencing
+% technology?
+
+% I read through the manuscript and have some thoughts:
+
+% 1) We'd better mention that all the tests are made on the default
+% settings of these tools.
+% 2) You will add that PEACD+EAST supports different output format (ace,
+% sam) and have a nice user interface, right?
 
@@
 % LocalWords:  GenBank indels contigs TGICL Metasim MetaSim Runtime runtime pre
-% LocalWords:  zebrafish runtimes ESTs paralilization
+% LocalWords:  zebrafish runtimes ESTs paralilization mufit fasta BAM
+% LocalWords:  isophormes

### 2011-06-09 papers/EAST.bioinf/east-gb.tex
papers/EAST.bioinf: New version of the EAST paper, re-structured for the journal Bioinformatics.  Semi-final draft for submission.

--- 2011-06-08 peace/papers/EAST/east.tex
@@
   \and
+  Praveen Kumar Raj Kumar$^4$
+  \email{Praveen Kumar - rajkump@muohio.edu}
+  \and
   Chun Liang\correspondingauthor$^{1,3}$
@@
   and
-  John E. Karro\correspondingauthor$^{1,4.5}$
+  John E. Karro\correspondingauthor$^{1,4,5}$
   \email{John E. Karro\correspondingauthor - karroje@muohio.edu}
@@
   \iid(1)Department of Computer Science and Software Engineering,  \\
-    \iid(2)Informational Technology Services
+  \iid(2)Informational Technology Services\\
   \iid(3)Department of Botanty, \\
@@
 %%
-  \section*{Introduction} {\color{blue} The {\it de novo} assembly of
-    RNA transcripts is an important and computationally challenging
-    problem, made more difficult by the advent of Next Generation
-    Sequencing technologies such as 454 and Illumina
-    \cite{Nagaraj07,Rao10}.  Upon completion of
-    transcriptome-sequencing, the investigator is left with a large
-    set of sequence fragments but no ordering or transcript
-    association information.  These fragments must be clustered by
-    transcript and assembled before the sequencing output can be truly
-    useful.  {\it de novo} assembly tools assemble the
-    transcript out of the fragments without using genmic sequence
-    information.  {\it de novo} tools are obivously necessary when working with a genome that is
-    unsequenced or poorly annotated, but is also useful when there is
-    limited knowledge of gene splice sites, as well as for the discovery
-    of novel events \cite{Birol09,Robertson10}.}
+  \section*{Introduction}
+
+  The {\it de novo} assembly of RNA transcripts is an important and
+  computationally challenging problem, made more difficult by the
+  advent of Next Generation Sequencing technologies such as 454 and
+  Illumina \cite{Nagaraj07,Rao10}.  Upon completion of
+  transcriptome-sequencing, the investigator is left with a large set
+  of sequence fragments but no ordering or transcript association
+  information.  These fragments must be clustered by transcript and
+  assembled before the sequencing output can be truly useful.  {\it de
+    novo} assembly tools assemble the transcript out of the fragments
+  without using genomic sequence information.  {\it de novo} tools are
+  obivously necessary when working with a genome that is unsequenced
+  or poorly annotated, but are also useful when there is limited
+  knowledge of gene splice sites \cite{Birol09,Robertson10}, they can
+  aid in the discovery of novel biological processes
+  (e.g. transplicing and fusion transcript) \cite{Mitelman07,Li10b}, and
+    they are invaluable for annotating structural variations not
+    contained within the genome \cite{Li08}.
 
@@
 
-The problem of {\it de novo} transcript assembly has been addressed a
-number of times in the literature, with initial approaches
+  The problem of {\it de novo} transcript assembly has been addressed
+  a number of times in the literature, with initial approaches
   concentrating on the longer reads produced by traditional Sanger
@@
   \tgicl\/ use alignment-based strategies that produce high quality
-results when applied to the output of Sanger sequencing technologies.
-\velvet, based on the modeling of the problem with {\it de bruijn}
-graphs, was designed to handle short read sequences.  \mira\/ is a
-multi-pass DNA assembler based on an overlap-layout-consensus strategy
-that also has an increased ability to assemble {\it hybrid sets}: sets
-formed by the mixture of sequencing results from multiple technologies
-\cite{MiraWeb}.
+  results when applied to the output of Sanger sequencing
+  technologies.  \velvet, based on the modeling of the problem with
+  {\it de bruijn} graphs, was designed to handle short read sequences.
+  \mira\/ is a multi-pass DNA assembler based on an
+  overlap-layout-consensus strategy that also has an increased ability
+  to assemble {\it hybrid sets}: sets formed by the mixture of
+  sequencing results from multiple technologies \cite{MiraWeb}.
 
@@
 
-\textcolor{blue}{In assessing result quality, we must view the \peace\/ clustering tool
+In assessing result quality, we must view the \peace\/ clustering tool
 \cite{Rao10} and \east\/ as a combined process.  While most assembly
@@
 the following section, we consider them as a single tool \peast, and
-assess the quality of these results.}
+assess the quality of these results.
 
@@
 bound the score by the relative size of the reconstruction.
-\textcolor{blue}{(For example, were a tool only able to reconstruct
-  one fourth of a transcript -- but able to reconstruct that fraction
-  perfectly -- the resulting contig's a-score would be $0.25$.)}  In
-addition we look at: the number of contigs produced by a tool
-normalized by the number of transcripts (a value equal to 1 in a
-perfect reconstruction); the number of singleton ESTs in the solution
-(i.e. ESTs unassigned to any contig and thus, most cases, effectively
-thrown away), and the runtime of the tool.
+(For example, were a tool only able to reconstruct a
+  contig covering one fourth of a transcript -- but able to
+  reconstruct that fraction perfectly -- the resulting contig's
+  normalized a-score would be $0.25$.)  In addition we look at: the
+number of contigs produced by a tool normalized by the number of
+transcripts (a value equal to 1 in a perfect reconstruction); the
+number of singleton ESTs in the solution (i.e. ESTs unassigned to any
+contig and thus, most cases, effectively thrown away), and the runtime
+of the tool.
 
@@
 
-\noindent {\bf Sanger Sequencing:} \textcolor{blue}{We first compare the
+\noindent {\bf Sanger Sequencing:} We first compare the
   assembly quality of the tools on a Sanger Sequencing model, using
@@
   ESTs that cover a based.  Thus we investigate the a-score as a
-  function of error rate at different coverage levels.}  For each
+  function of error rate at different coverage levels.  For each
 parameter combination we applied each tool to 30 independent, randomly

### 2011-06-09 papers/EAST.bioinf/east.tex
papers/EAST.bioinf: New version of the EAST paper, re-structured for the journal Bioinformatics.  Semi-final draft for submission.

\documentclass{bioinfo}
\copyrightyear{2011}
\pubyear{2011}

\bibliographystyle{natbib}
\usepackage{url}  % Formatting web addresses

%%%%%%%%%%%%%%%%%%%%
%% Karro macros   %%
%%%%%%%%%%%%%%%%%%%%
\newcommand{\peace} {{\small PEACE}}
\newcommand{\wcd} {{\small WCD}}
\newcommand{\capthree} {{\small Cap3}}
\newcommand{\easycluster} {{\small EasyCluster}}
\newcommand{\estsim}{{\small ESTSim}}
\newcommand{\metasim} {{\small MetaSim}}
\newcommand{\tgicl} {{\small TGICL}}
\newcommand{\east} {{\small EAST}}
\newcommand{\velvet}{{\small Velvet}}
\newcommand{\mira}{{\small Mira3}}
\newcommand{\pave} {{\small PAVE}}
\newcommand{\peast}{{\small PEACE+EAST}}

\usepackage{graphicx}
%\def\includegraphic{}
%\def\includegraphics{}

\begin{document}

\firstpage{1}

\title[EAST transcript assembler]{EAST: \textit{de novo} {\underline E}xpression Fragment {\underline A}ssembly from {\underline S}panning {\underline T}rees}
\author[Zhang \textit{et~al.}]{Yuan Zhang$^1$, Dhananjai M Rao$^1$,
  Mufit Ozden$^1$, Jens Mueller$^2$, Praveen K. Raj Kumar$^4$, Chun
  Liang$^{*1,3}$, John E. Karro$^{*1,4,5}$}

\address{$^1$ Department of Computer Science and Software Engineering, \\
  $^2$ Information Technology Services,
  $^3$ Department of Botany, \\
  $^4$ Department of Microbiology, and
  $^5$ Department of Statistics \\ 
at Miami University, Oxford OHIO, USA
}
\history{Received on XXXXX}

\editor{Associate Editor: XXXXX}

%\title[EAST transcript assembler]

  %\email{Yuan Zhang - zhangy9@muohio.edu}
  %\and

  %\email{Dhananjai Rao - raodm@muohio.edu}
  %\and

  %\email{Mufit Ozden - ozdenm@muohio.edu}
  %\and

  %\email{Jens Mueller - muellej@muohio.edu}
  %\and

  %\email{Praveen Kumar - rajkump@muohio.edu}
  %\and

  %\email{Chun Liang\correspondingauthor - liangc@muohio.edu}
  %and

  %\email{John E. Karro\correspondingauthor - karroje@muohio.edu}

\maketitle

\textcolor{blue}{
\begin{abstract}
\section{Motivation:} The {\it de novo} assembly of RNA transcript is
computationally challenging problem who importance and difficulty have
both increased with the advent of next generation sequencing
technologies.  The ability to reassemble RNA sequences from a
extremely large dataset of sequence fragment allows not only access to
the transcriptome of unsequenced genomes, but also the exploration and
discovery of novel biological processes.  Hence software tools that
can produce reliable assemblies from large sets of short fragments
from error-prone sequencing technology are in high demand.
\section{Results:} We describe the \east\/ algorithm for the {\it de
  novo} assembly of traditional, short-read, and hybrid
transcript-derived fragments sets, and find that its implementation
results in a tool that appears to be of the most general use (with
respect to multiple sequencing technology platforms) of the tools
examined.  \east\/ consistently outperforms several popular tools on
Sanger sequencing data, out performs or matches competing tools on 454
and hybrid data, and is functional over Illumina data.  In addition,
\east\/ proves to be the most robust tool with respect to sequencing
error.
\section{Availability:} \east\/ can be installed and run through the
\peace\/ GUI, with the implementation and code freely available at
\url{http://www.peace-tools.org}.
\section{Contact:} karrroje@muohio.edu, liangc@muohio.edu
\end{abstract}
}
\section*{Introduction}

\textcolor{blue}{The {\it de novo} assembly of RNA transcripts is an important and
computationally challenging problem, made more difficult by the advent
of Next Generation Sequencing technologies such as 454 and Illumina
[\cite{Nagaraj07}].  Upon completion of transcriptome-sequencing, the
investigator is left with a large set of sequence fragments but no
ordering or transcript association information.  These fragments must
be clustered by transcript and assembled before the sequencing output
can be truly useful.  Access to the relevant organism's genomic
sequence can considerably reduce the complexity of this task, but
using this information is not always an option, or the best option.
{\it de novo} assembly tools assemble the transcript out of the
fragments without using genomic sequence information.  Such tools are
are necessary when working with a genome that is unsequenced
or poorly annotated, but are also useful when in a number of contexts
(for example: when there is limited knowledge of gene splice sites
[\cite{Birol09,Robertson10}]; as an aid in the discovery of novel
biological processes such as transplicing and transcript fusion
[\cite{Mitelman07,Li10b}]; in the annotation of structural variations
not contained within the genome [\cite{Li08}]).}

%\vspace{3mm}

The problem of {\it de novo} transcript assembly has been addressed a
number of times in the literature, with initial approaches
concentrating on the longer reads produced by traditional Sanger
Sequencing techniques, and newer approaches looking at produces of
Next Generation Sequencing (NGS) technologies.  A list of popular
tools addressing these problems include \capthree, \tgicl, \velvet,
and \mira, each excelling within their own context
[\cite{Huang99,Pertea03,Chevreux04,Zerbino08}].  \capthree\/ and
\tgicl\/ use alignment-based strategies that produce high quality
results when applied to the output of Sanger sequencing technologies.
\velvet, based on the modeling of the problem with {\it de bruijn}
graphs, was designed to handle short read sequences.  \mira\/ is a
multi-pass DNA assembler based on an overlap-layout-consensus strategy
that also has an increased ability to assemble {\it hybrid sets}: sets
formed by the mixture of sequencing results from multiple technologies
[\cite{MiraWeb}].

%\vspace{3mm}

\textcolor{blue}{Here we present a novel Minimum Spanning Tree (MST) approach to the
problem of assembling short read, traditional, and hybrid transcript
sequence sets, and show it to be a viable approach through its
implementation in our \east\/ assembly tool.  The tool has been
integrated into the \peace\/ clustering tool set [\cite{Rao10}],
freely available from \url{www.peace-tools.org}.  After a description
of the underlying assembly algorithm, we present a comparison of
\east\/ against \capthree, \tgicl, \velvet, and \mira\/ on multiple
sequencing technology platforms.}

%%%%%%%%%%%%%%%%%%
\section*{Methods}

The algorithm underlying \east\/ is based on three components: minimum
spanning trees (MSTs) [\cite{Prim57}], the $d^2$ sequence distance
measure [\cite{Hide94}], and standard sequence alignment
[\cite{Needleman70,Smith81}].  The objective is to find overlapping
fragments input set and assemble them into contiguous segments
(contigs).  While we would ideally be using sequence alignment to
identify these overlapping ends, the quadratic runtime of alignment
algorithms coupled with the large number of sequences in the input
makes such an approach infeasible.  To avoid excessive alignment
computation, we make use of MSTs to provide a {\it guide} through the
EST set, dictating which pairs are to be aligned.  A modified version
of $d^2$, an alignment-free method for estimating sequence distance,
provides the edge weights for the graph model, allowing us to
construct the MST.

%\vspace{3mm}

\east\/ is an assembly tool, and requires clustering be
done as a pre-processing step, as well as requiring that $d^2$-based
MSTs be provided for each cluster.  Conveniently, the \peace\/
clustering tool [\cite{Rao10}] provides exactly this information.  While
\peace\/ and \east\/ are technically two tools, \east\/ was designed
specifically for \peace\/ output.  For purposes of this paper, it
makes sense to consider them as one.  In the following we briefly
outline the \peace\/ algorithm for clustering (as is relevant to
\east\/), and then discuss the steps taken by \east\/ for assembly of
the \peace\/ clusters.

\subsection*{\peace Clustering}

Given a set of fragments from two or more transcripts, we must first
cluster the fragments by transcript association -- leaving us with a
single cluster of fragments for each transcript.  Details of the
\peace\/ algorithm can be found in {\it Rao et al.} [\cite{Rao10}],
with the general idea as follows: we model the problem as a weighted,
undirected graph, representing each fragment with a node and assigning
edge weights based on a $d^2$ comparison of the incident fragments
(explained in more detail below).  By employing the $u/v$ and $t/v$
filtering heuristics of {\it Hazelhurst et al.} [\cite{Hazelhurst08}],
we can quickly dismiss most of the edges as irrelevant
(i.e. connecting fragments with no overlap), then use Prim's algorithm
to compute an MST [\cite{Prim57}].  Upon completion of the MST we
remove all edges exceeding a specified threshold value, taking each
component of the resulting forest as a cluster.  As a result, \peace\/
passes to \east\/ a set of clusters, an MST for each cluster, and the
orientation of each fragment relative to the cluster.

\subsection*{Overlap Detection}

Given the \peace\/ generated MST for a cluster, the challenge is to
identify and align the ends of overlapping fragments without taking
the significant runtime hit needed to align every pair.  This is
accomplished using two methods: aligning only those sequences adjacent
in the MST, and restricting such alignments to portions identified by
a modified form of the $d^2$ heuristic.

%\vspace{3mm}

A \peace-generated MST represents a single transcript, with each edge
serving as a heuristic indication that the incident sequences overlap;
were there not sufficient sequence similarity to indicate an overlap,
the edge would have initially been assigned a high $d^2$ value and
would have been removed by \peace\/.  To determine exactly how the
nodes overlap, we calculate a modified $d^2$ score as follows.  The
standard calculation of $d^2$, requires looking at every pair of
length $w$ sub-strings (windows) from the two sequences, calculating a
distance between these two windows based on the number of shared words
(6-mers), and taking the score of the lowest-scoring window pairs as
the $d^2$ distance [\cite{Hide94}].  However, when looking for overlap
of fragments $s_1$ and $s_2$, we are looking only for end-overlap.
That is, we wish to determine if the left end of $s_1$ overlaps the
right end of $s_2$, or the reverse (recalling that \peace\/ has
ensured they are on the same strand).  Hence we need only consider the
left-most and right-most windows of $s_1$ against each window of
$s_2$, reducing the number of window comparisons from quadratic to
linear in the size of $s_2$.

%\vspace{3mm}

In Figure~\ref{fig:overlap} we illustrate the use of modified $d^2$,
using it to take two sequences (a), find the window on $s_2$ most
resembling the right-most window of $s1$ (red section of (b)),
extending to the end of $s_2$ (green section of (c)) and performing a
{\it global} alignment on the overlapping portion.  We then compute the
{\it overlap distance} ($1 - s/l$, where $s$ is the alignment score and
$l$ is the length of the overlap segment), a value that is inversely
proportional to the probability that this is a legitimate overlap.

\begin{figure}
\centerline{\includegraphics[width=3in]{pics.d/overlap_alignment.pdf}}
\caption{(a) Two sequences potentially overlapping at the ends.  (b)
  The window of $s_2$ pairing with the right-most window of $s_1$ to
  minimize the $d^2$ score.  (c) An extension of the $s_2$ window to
  the end of the sequence, and the corresponding extension on $s_1$.
  (d) A global alignment of the overlapping areas determined from
  (c).}\label{fig:overlap}
\end{figure}

In order to accommodate a range of potential fragment lengths, \peace\/
and \east\/ make use of an {\it adaptive $d^2$} strategy to handle
hybrid data sets.  The $d^2$ measure (and our modification) is
flexible enough to handle a range of sequence sizes, requiring only a
modification of the parameters (i.e. window size and threshold
values).  However, it runs into problems when comparing sequences of
significantly different sizes.  Comparing a 62 base Illumina read
against a 1000 base Sanger read requires the use of parameters
appropriate to the shorter fragment -- which are considerably less
accurate when applied to the longer read.  Our solution is to
partition fragments by sizes (e.g. into groups of small, medium and
large fragments), and compute distances only between segments in the
same or neighboring groups.  By skipping the comparisons between
sequences of significantly different sizes we avoid the pitfalls of
such comparisons, while still implicitly discovering those connections
through transitivity.

\subsection*{Ordering and Reconstruction}

Having defined overlap distance, ordering the fragments is
straight-forward.  We pick an arbitrary node on the cluster's MST and
transverse the tree, using the overlap distance measure to check each
node against its two neighbors.  Those nodes with no identified
left-overlap are checked against nodes further out in the tree, and
any then remaining with no identified left neighbor as designated as a contig
left end.

%\vspace{3mm}

Having identified the distance of each node from its two closest
overlapping sequences, we now construct a {\it directed} graph, with
each edge pointing from a sequence to a right neighbor and weighted
with the overlap distance.  Using this graph we calculate a new
minimum spanning tree, then starting from the root of the tree
(necessarily a left-end of the contig) we transverse the tree and
assign the first base of each sequence a position relative to its
parent.  Such an assignment gives us an implicit ordering of the
fragments.  By scanning through them in order and aligning each to
the next with the Smith-Waterman alignment we end up with a multiple
alignment of the fragments that allows us to derive a consensus
sequence.  During this process quality score information can also be
incorporated, reducing the score contribution of a specific base-pair
if the involved bases are of low quality.  A straightforward extension
would also allow the predictions of single nucleotide polymorphism by
identifying those positions whose covering ESTs vary more in content
than could reasonably be accounted for by the base-call rate.

\noindent{\bf Implementation:} \east\/ has been integrated into the
\peace\/ clustering package [\cite{Rao10}] and is available for
download from the \peace\/ website (\url{www.peace-tools.org}).
Users can run the GUI provided by that site, and through it
install both the \peace\/ cluster engine and \east\/ assembly engine
on a local machine or remote server.
Initial EST data must be in fasta format for fastaq format, though the tool can print
out the final assembly in fasta, ACE, SAM and BAM formats [\cite{Li09}].

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Results and Discussion %%
%%
\section*{Results}

In assessing result quality, we must view the \peace\/ clustering tool
[\cite{Rao10}] and \east\/ as a combined process.  Given a unordered
set of fragments representing multiple transcripts, \peace\/ is given
the task of (ideally) producing a minimum spanning tree for each transcript,
with edges representing a subset of the overlap information.  \east\/
then assembles each cluster into one or more contigs.  As most tools
integrate the clustering and assembly process, for purposes of
comparison we must consider our tools as a single tool \peast, and
assess the quality of these results.

%\vspace{3mm}

In order to assess the quality of (\peace\/ +) \east\/ results, we
compared against four prominent assembly tools: \capthree, \tgicl,
\mira, and \velvet\/ [\cite{Huang99,Pertea03,Chevreux04,Zerbino08}].
The tools were tested on data from three sequencing technologies
(Sanger, 454 and Illumina).  Test data was produced using a set of
zebra fish transcripts \cite{Hazelhurst08} and deriving simulated
simulated fragment sets using bot the \estsim\/ and \metasim\/ tools
[\cite{Hazelhurst03,Richter08}].

\subsection*{Measuring Quality}
Our primary measure of quality was the {\bf normalized a-score}, based
on the {\it assembly score} (a-score) used in {\it Liang et al.}
[\cite{Liang00}].  After reconstruction a (portion) of a transcript,
the a-score of that contig is defined as $2c - 5m - 15i$ where $c$ is
the length of the reconstructed contig, $m$ is the number of
miss-called bases, and $i$ is the number of indels.  A ``perfect''
score would be equal to twice the sequence length.  Our normalized
a-score normalizes of the length $t$ of the transcript being
reconstructed, resulting in a formula of:
$$\frac{2s - 5m - 15i}{2t}.$$
Thus a perfect score would be equal to 1.  This normalization allows
for the comparison of assembly quality between sequences of different
lengths. Further, using this score we can more accurately quantify the
quality of a partial reconstruction: if we are able to reconstruct
only a portion of a fragment, the normalized a-score will take full
consideration of the quality of the reconstruction, but then bound the
score by the relative size of the reconstruction.  (For example, were
a tool only able to reconstruct a contig covering one fourth of a
transcript -- but able to reconstruct that portion perfectly -- the
resulting contig's normalized a-score would be $0.25$.)  In addition
we look at: the number of contigs produced by a tool normalized by the
number of transcripts (a value equal to 1 in a perfect
reconstruction); the number of singleton ESTs in the solution
(i.e. ESTs unassigned to any contig and thus, most cases, effectively
thrown away); and the runtime of the tool.

\subsection*{Tool Comparisons}
Using simulated fragment sets allows us to compare assemblies to a
known correct solution and control parameters in their model
(e.g. base-call error rate) to test the effect of variation.  All sets
were generated from a collection of 100 zebra fish transcripts
obtained from {\it Hazelhurst et al.}  [\cite{Hazelhurst08}].  For
Sanger sequences we used the \estsim\/ tool to generated simulated
fragment sets, taking its default model of EST generation when not
otherwise specified [\cite{Hazelhurst03}].  For 454 and Illumina sets
we used the \metasim\/ tool, again using taking their default model to
simulate the errors associated with those technologies
[\cite{Richter08}].  Each tool models potential sequencing mistakes
such as base-call errors, insertion and deletions, using parameters
tailored to the target the technologies; see those publications for
details.

%\vspace{3mm}

\noindent {\bf Sanger Sequencing:} We first compare the assembly
quality of the tools on a Sanger Sequencing model, using \east,
\capthree, and \tgicl\/ (tools designed to work with Sanger
sequences).  The quality of any assembly will be directly affected by
the {\it coverage} of the input: the average number of ESTs that cover
a based.  Thus we investigate the a-score as a function of error rate
at different coverage levels.  For each parameter combination we
applied each tool to 30 independent sets, derived from 15 randomly
chosen members of our zebra fish transcript set, reporting the average
normalized a-score.  In Figure~\ref{sangerAscore} we look at
normalized a-score as a function of error rate, fixing coverage at 50,
30 and 10 and varying base read error rate from 0\% to 6\% for each of
the three tools designed for Sanger Sequencing.  While all three tools
perform well when applied to error free (i.e. impossible) sequencing
technology, we find \east\/ to be considerably more robust to
base-call errors than the other tools.  Taking the mid-range coverage
value of 30 as an example, at a 1\% error rate we see \east,
\capthree, and \tgicl\/ all achieving an essentially perfect
normalized a-score.  At a 3\% error rate \east\/ maintains its almost
perfect normalized a-score, showing a 2\% improvement over \tgicl\/
and a 21\% improvement over \capthree.  At a 4\% error rate \east\/
still maintains its near-perfect score (showing a 54\% and 85\%
improvement over \capthree\/ and \tgicl\/ respectively).  Even at a
6\% error rate \east\/ is still achieving a normalized a-score of 0.95
(a 110\% improvement over the next closest tool, \tgicl\/), meaning
that at most 5\% of the bases were left out of their contigs or were
incorrectly reconstructed.

\begin{figure*}[htb]
\centerline{\includegraphics[width=6in]{pics.d/ascore_sanger_norm.pdf}}
\caption{Average normalized a-score as a function of base error rate for
  simulated Sanger sequencing.  \east=blue, \capthree=Black,
  \tgicl=green. COMMENT: Need to change figure heading to Normalized a-score.}
\label{sangerAscore}
\end{figure*}

In Figures~\ref{contigsSanger} is a plot of the average number of
contigs per gene sequence as a function of error rate (top row), the
number of singletons is shown on the bottom row.  For the first we
observe that all tools have comparable results for lower error rates,
but only \east\/ is able to maintain the low score with increasing
error rates -- generating a perfect score in almost every case.  In
minimizing the number of singletons, we find that \east\/ does a
better job than \capthree\/ for error rates larger than $2\%$, while
\tgicl\/ does a better job that \east\/ for error rates larger than
$4\%$ to $5\%$.

\begin{figure*}[htb]
\includegraphics[width=6in]{pics.d/numContigs_sanger.pdf}
\includegraphics[width=6in]{pics.d/numSingle_sanger.pdf}
\caption{Number of contigs (normalized to the number of transcripts)
  and number of singletons as a function of base-call error
  rate for simulated Sanger sequencing.}
\label{contigsSanger}
\end{figure*}

\noindent {\bf Next Generation Sequences:} In
Figure~\ref{nextgenAscore} we compare the normalized a-score for 454
and Illumina outputs derived from the same zebra fish transcripts
(with sequence lengths averaging 250 and 62 bp respectively).  For
these tests we varied coverage levels but otherwise continued to use
the default Metasim model [\cite{Richter08}].  On 454 data we see
significant improvement in results quality in \east\/ for all levels
of coverage.  On Illumina data we see that \east\/ is competitive with
\velvet\/ at higher coverages, though is unable to match that tool
when coverage is sparse.

\begin{figure*}[htb]
\centerline{\includegraphics[width=5in]{pics.d/ascore_nxtgen.pdf}}
\caption{Comparative normalized a-scores for 454 and Illumina sequence
  output, using the \metasim default model at different levels
  of coverage.}
\label{nextgenAscore}
\end{figure*}

\noindent {\bf Hybrid:} In Figure~\ref{hybridAscore} we look at the
quality of results derived from applying each tool to a hybrid data
set of Sanger (40\%), 454 (40\%) and Illumina (20\%) sequences, again
showing normalized a-score as a function of base-call error rate in
the Sanger portion.  Here we observe \east\/ matching or outperforming
\capthree\/ and \mira\/ at moderate to high coverages.  At a coverage
level of 50, \east\/ maintains a normalized a-score hovering around
0.85 at all error levels (showing a 54\% improvement over \mira\/ at a
3\% base-call error rate).  \east\/ actually improves its performance
at a coverage level of 30 (hovering around a normalized a-score of
0.95, showing a 142\% improvement over \mira\/ at the 3\% base-call
error rate), an oddity we attribute to the decreased absolute number
of 62bp Illumina sequence.  At a coverage level of 10 \east's
normalized a-score is around 0.37 -- performing considerably worse
than \mira\/ at low error rates (e.g. 0.94 v. 0.40 for error-free
bases), but showing a 52\% improvement over \mira\/ at the 3\% error
rate mark and in general showing increased robustness to error.
Remarkably, it is only in this situation that \capthree\/ is
competitive, with that tool achieving a normalized a-score of 0.62
(v. \east's 0.37) for the 3\% base-call error rate.  However, it does
lose out to \east\/ at greater than a 4.5\% base-call error rate -- a
situation likely due to \east's problems with the 62 bp Illumina
sequences (see below).

\begin{figure*}[htb]
\centerline{\includegraphics[width=5in]{pics.d/ascore_hybrid.pdf}}
\caption{Comparative normalized a-scores for a hybrid set consisting
  of Sanger, 454 and Illumina sequence output, varying the base-call
  error rate for Sanger sequences and using the MetaSim default model
  for the 454 and Illumina sequences.  Each data set tested was
  composed of 40\% Sanger Sequences, 40\% 454 Sequences, and 20\%
  Illumina sequences.}
\label{hybridAscore}
\end{figure*}

\noindent {\bf Tool Runtime:} In Figures~\ref{runtime.fixed} we see
the average runtime for each of the experiments reflected in
Figures~\ref{sangerAscore}-\ref{hybridAscore} for a coverage of 30
(continuing to report the combined runtime of \peace+\east), while in
Figure~\ref{runtime.coverage} we look at the average runtime for each
tool when applied to simulated data sets generated from a single zebra
fish gene at varying levels of coverage.  For Sanger sequences we
see marked improvement of (\peace+)\east\/ over all other tools for
almost all coverage levels: for the smallest set (coverage = 10)
\east\/ has 1.9 speedup over \capthree\/ and a 3.9 speedup over
\tgicl, while at a more moderate size (coverage=50) there are speedups of
4.7 and 8.4.  This returns back down to 3.0 and 5.2 at the largest set
tested (coverage = 100), as can be seen in the bump around a coverage
of 85 in Figure~\ref{runtime.coverage}.  We are unable to
explain this bump, but have noticed the same effect in
data sets derived from other genes as well.

%\vspace{3mm}

\east\/ is less competitive on shorter sequence.  In our simulations,
we find a reduced runtime improvement on the simulated 454 sequences
(averaging 250 bp in length), and a significantly reduced runtime on
the simulated Illumina sequences (exactly 62 bp in length).  The
increased runtime is due to the problems with applying the $d^2$
metric to short sequences, and will become less of a problem as NGS
technology continues to increase output sequence length
[\cite{Eid09,Li10}].

However,
as technology advances, the ``short-read'' sequencers are producing
increasingly longer reads: 454 sequencing technology can produce
reads exceeding 400 bp, while Illumina sequences can exceed 100 bp
[\cite{Eid09,Li10}].  The loss of runtime in our experiments on 454,
Illumina and the hybrid sequences are due to the reduced length
(specifically, the problems with applying the $d^2$ metric to shorter
sequences -- as discussed below), hence are becoming less of an issue
with this increase in short-read fragment length.

%\vspace{3mm}

\begin{figure*}[htb]
\centerline{\includegraphics[width=6in]{pics.d/runtime_fixedsize_sanger.pdf}}
\caption{Average runtime per trial when when generating data for
  Figures~\ref{sangerAscore}-\ref{hybridAscore}  when coverage = 30.
  Note that the exceptionally high speed of \velvet\/ on Illumina sequences make the
  red (center) bars in that plot essentially indistinguishable.}
\label{runtime.fixed}
\end{figure*}

\begin{figure*}[htb]
\centerline{\includegraphics[width=6in]{pics.d/runtime_bycoverage_all.pdf}}
\caption{The average runtime when applying each tool to a simulated
  data set derived from a specific zebrafish transcript of length
  14,625 bp.  As we decrease coverage the number of fragments drops,
  hence illustrating the relationship between tool runtime and
  fragment set size.  Each data point represents the average runtime
  from the application of the tool to at least 30 randomly generated
  simulated sets coming from the transcript at the given level of
  coverage.}
\label{runtime.coverage}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions and Discussion}

\textcolor{blue}{It is said that nothing is perfect, and this statement is certainly
true when discussing assembly tools.  No one assembler can
consistently out performs all other tools, and a user must choose the
the tool that works best for their data.  Our experiments have shown
that the \peast\/ assembly tool produces the best assemblies for Sanger
data with a shorter runtime then competing tools, is the most robust
to sequencing error, and preforms better then any competing tool when
applied to data produced multiple sequencing technology platforms.
\east\/ also demonstrates the viability of the MST-based
approached to sequencing -- an approach we believe has considerable
potential for improvement.}

As observed, when applying \east, \capthree,
and \tgicl\/ to Sanger sequenced transcripts with normal to high error
rates, \east\/ consistently out performs all other tools in terms of
the normalized a-score metric (Figures~\ref{sangerAscore},
\ref{runtime.fixed}, and \ref{runtime.coverage}).  We also find that
the number of contigs produced is almost exactly equal to the number
of transcripts represented in the set, regardless of error rate, while
the number of singleton ESTs remains low up to about a 5\% error rate
(Figures~\ref{contigsSanger}).  The high normalized a-score indicates
that \east\/ is both reconstructing larger portions of each transcript
and doing so more faithfully, while the low number of contigs suggests
that we are assembling complete transcripts.  Finally, runtimes for
\east\/ are considerably faster than other tools for larger sets
(Figures~\ref{runtime.fixed} and \ref{runtime.coverage}), with
(\peace+)\east\/ running a minimum of two to three times as fast as
the other tools.  The speedup over \capthree\/ (the faster tool)
ranged from 1.9 to 6.0, depending on the coverage level.  In short,
when used for Sanger Sequences \east\/ is a faster tool that produces
the high quality reconstructions versus comparing tools and is
significantly more robust to error than other tools.

%\vspace{3mm}

\textcolor{blue}{For Next Generation Sequencing output we see a significant result
quality increase over other tools in many (though not all) situations
(Figures~\ref{nextgenAscore} and \ref{hybridAscore}) -- with \east\/
out performing every tool on 454 data but uncompetitive with \velvet\/
on Illumina data.  This quality does come at a price: the runtime of
\east\/ takes a significant hit when dealing with shorter sequences.
This is due to the dependence of \east\/ on the $d^2$ algorithm, which
has poor performance on shorter sequences.  In practice this does not
pose a large problem: the output sequence length of NGS technology
is continually increasing as the technology advances, leaving
our 62 bp sequence already outdated [\cite{Eid09,Li10}].   These
continuing improvements in runtime, and the improvements in result
quality, justify the \east algorithm as a good choice for assembly.}

Finally, we note, but have not yet explored, the fact that \east\/ is
amenable to to paralilization and the possibility of identifying
isoformes resulting from alternative splicing.  For the first: the
\east\/ tool is trivially parallelizable in that once the \peace\/
portion has completed clustering, \east\/ can address the clusters in
parallel.  More sophisticated paralilization techniques are applicable
at different phased of the east assembly (e.g. the depth-first search
of the MST, or the multiple independent applications of the $d^2$ metric).

\section*{Acknowledgements}
  Text for this section \ldots

\paragraph{Funding\textcolon} Text Text Text Text Text Text  Text Text.

\bibliography{all}

\end{document}

% LocalWords:  Ozden Chun Liang Karro Botanty publ MSTs et al Hazelhurst Prim's
% LocalWords:  mers Illumina contig bmc multi LaTeX xpression ssembly rees Rao
% LocalWords:  Zhang Moler Dhananjai Mufit de novo introns NGS bruijn ABIITY bp
% LocalWords:  GenBank indels contigs TGICL Metasim MetaSim Runtime runtime pre
% LocalWords:  zebrafish runtimes ESTs paralilization mufit fasta BAM Praveen
% LocalWords:  isophormes XXXXX transcriptome unsequenced transplicing fastaq
% LocalWords:  Kumar

### 2011-06-17 papers/EAST.bioinf/east.tex
Update to semi-final draft of EAST paper.

@@
 
-\textcolor{blue}{
 \begin{abstract}
-\section{Motivation:} The {\it de novo} assembly of RNA transcript is
-computationally challenging problem who importance and difficulty have
-both increased with the advent of next generation sequencing
-technologies.  The ability to reassemble RNA sequences from a
-extremely large dataset of sequence fragment allows not only access to
-the transcriptome of unsequenced genomes, but also the exploration and
+\section{Motivation:} The {\it de novo} assembly of RNA transcripts is a
+computationally challenging problem whose importance and difficulty have
+increased with the advent of next generation sequencing
+technologies.  The ability to reassemble RNA sequences from an
+extremely large dataset of cDNA fragments allows both access to
+the transcriptome of unsequenced genomes and  the exploration and
 discovery of novel biological processes.  Hence software tools that
@@
 from error-prone sequencing technology are in high demand.
-\section{Results:} We describe the \east\/ algorithm for the {\it de
-  novo} assembly of traditional, short-read, and hybrid
-transcript-derived fragments sets, and find that its implementation
-results in a tool that appears to be of the most general use (with
-respect to multiple sequencing technology platforms) of the tools
-examined.  \east\/ consistently outperforms several popular tools on
-Sanger sequencing data, out performs or matches competing tools on 454
-and hybrid data, and is functional over Illumina data.  In addition,
-\east\/ proves to be the most robust tool with respect to sequencing
-error.
+\section{Results:} We describe \east, an novel algorithm making use of
+{\it Minimum Spanning Trees} for the {\it de novo} assembly of
+short-read, traditional, and hybrid transcript-derived fragments sets,
+and find that its implementation results in a tool that appears to be
+of the most general use (with respect to multiple sequencing
+technology platforms) of the tools examined.  \east\/ consistently
+outperforms several popular tools on longer sequences (e.g. Sanger
+sequencing data and longer 454 sequences), and competitive against
+most tools on very short (e.g. 62 bp Illumina) data.  In addition, \east\/
+proves to be the most robust tool with respect to sequencing error,
+and is capable of utilizing base quality value to further improve
+assembly.
 \section{Availability:} \east\/ can be installed and run through the
@@
 \end{abstract}
-}
+
 \section*{Introduction}
 
-\textcolor{blue}{The {\it de novo} assembly of RNA transcripts is an important and
+The {\it de novo} assembly of RNA transcripts is an important and
 computationally challenging problem, made more difficult by the advent
@@
 sequence can considerably reduce the complexity of this task, but
-using this information is not always an option, or the best option.
+using this information is not always the best option (or an option at all).
 {\it de novo} assembly tools assemble the transcript out of the
@@
 are necessary when working with a genome that is unsequenced
-or poorly annotated, but are also useful when in a number of contexts
-(for example: when there is limited knowledge of gene splice sites
+or poorly annotated, but are also useful in a number of contexts, such as:
+when there is limited knowledge of gene splice sites
 [\cite{Birol09,Robertson10}]; as an aid in the discovery of novel
 biological processes such as transplicing and transcript fusion
-[\cite{Mitelman07,Li10b}]; in the annotation of structural variations
-not contained within the genome [\cite{Li08}]).}
+[\cite{Mitelman07,Li08}]; and in the annotation of structural variations
+not contained within the genome [\cite{Li10b}].
 
@@
 \tgicl\/ use alignment-based strategies that produce high quality
-results when applied to the output of Sanger sequencing technologies.
-\velvet, based on the modeling of the problem with {\it de bruijn}
-graphs, was designed to handle short read sequences.  \mira\/ is a
-multi-pass DNA assembler based on an overlap-layout-consensus strategy
-that also has an increased ability to assemble {\it hybrid sets}: sets
-formed by the mixture of sequencing results from multiple technologies
-[\cite{MiraWeb}].
+results for Sanger sequencing results.  \velvet, based on {\it de
+  bruijn} graphs, was designed to handle short read sequences.
+\mira\/ is a multi-pass DNA assembler based on an
+overlap-layout-consensus strategy that also has an increased ability
+to assemble {\it hybrid sets}: sets formed by the mixture of
+sequencing results from multiple technologies [\cite{MiraWeb}].
 
@@
 
-\textcolor{blue}{Here we present a novel Minimum Spanning Tree (MST) approach to the
+Here we present a novel Minimum Spanning Tree (MST) approach to the
 problem of assembling short read, traditional, and hybrid transcript
@@
 \east\/ against \capthree, \tgicl, \velvet, and \mira\/ on multiple
-sequencing technology platforms.}
+sequencing technology platforms.
 
@@
 measure [\cite{Hide94}], and standard sequence alignment
-[\cite{Needleman70,Smith81}].  The objective is to find overlapping
+[\cite{Smith81}].  The objective is to find overlapping
 fragments input set and assemble them into contiguous segments
@@
 identify these overlapping ends, the quadratic runtime of alignment
-algorithms coupled with the large number of sequences in the input
+coupled with the large number of sequences in the input
 makes such an approach infeasible.  To avoid excessive alignment
@@
 
-\east\/ is an assembly tool, and requires clustering be
-done as a pre-processing step, as well as requiring that $d^2$-based
-MSTs be provided for each cluster.  Conveniently, the \peace\/
+\east\/ is an assembly tool, requiring that clustering be
+done as a pre-processing step and that $d^2$-based
+MSTs be provided for each cluster.  The \peace\/
 clustering tool [\cite{Rao10}] provides exactly this information.  While
@@
 
-\subsection*{\peace Clustering}
+\subsection*{\peace\/ Clustering}
 
@@
 cluster the fragments by transcript association -- leaving us with a
-single cluster of fragments for each transcript.  Details of the
-\peace\/ algorithm can be found in {\it Rao et al.} [\cite{Rao10}],
-with the general idea as follows: we model the problem as a weighted,
+single cluster of fragments for each transcript [\cite{Rao10}].  The
+general idea is as follows: we model the problem as a weighted,
 undirected graph, representing each fragment with a node and assigning
@@
 (explained in more detail below).  By employing the $u/v$ and $t/v$
-filtering heuristics of {\it Hazelhurst et al.} [\cite{Hazelhurst08}],
+filtering heuristics of {\it Hazlehurst et al.} [\cite{Hazelhurst08}],
 we can quickly dismiss most of the edges as irrelevant
@@
 the significant runtime hit needed to align every pair.  This is
-accomplished using two methods: aligning only those sequences adjacent
-in the MST, and restricting such alignments to portions identified by
-a modified form of the $d^2$ heuristic.
+accomplished by aligning only those sequences adjacent in the MST, and
+restricting such alignments to portions identified by a modified form
+of the $d^2$ heuristic.
 
@@
 
-A \peace-generated MST represents a single transcript, with each edge
-serving as a heuristic indication that the incident sequences overlap;
-were there not sufficient sequence similarity to indicate an overlap,
-the edge would have initially been assigned a high $d^2$ value and
-would have been removed by \peace\/.  To determine exactly how the
-nodes overlap, we calculate a modified $d^2$ score as follows.  The
-standard calculation of $d^2$, requires looking at every pair of
-length $w$ sub-strings (windows) from the two sequences, calculating a
-distance between these two windows based on the number of shared words
-(6-mers), and taking the score of the lowest-scoring window pairs as
-the $d^2$ distance [\cite{Hide94}].  However, when looking for overlap
-of fragments $s_1$ and $s_2$, we are looking only for end-overlap.
-That is, we wish to determine if the left end of $s_1$ overlaps the
-right end of $s_2$, or the reverse (recalling that \peace\/ has
-ensured they are on the same strand).  Hence we need only consider the
-left-most and right-most windows of $s_1$ against each window of
-$s_2$, reducing the number of window comparisons from quadratic to
-linear in the size of $s_2$.
+A \peace-generated MST (ideally) represents a single gene, with each
+edge serving as a heuristic indication that the incident sequences
+overlap.  Were there not sufficient sequence similarity to indicate an
+overlap, the edge would have been removed by \peace\/ due to a high
+assigned a high $d^2$ value.  To determine exactly how the nodes
+overlap, we calculate a modified $d^2$ score as follows.  The standard
+$d^2$ requires examining every pair of length $w$ sub-strings
+(windows) from the two sequences, calculating a distance between these
+two windows based on the number of shared 6-mers, and taking the score
+of the lowest-scoring window pairs as the $d^2$ distance
+[\cite{Hide94}].  However, when looking for overlap of fragments $s_1$
+and $s_2$, we are looking only for end-overlap: two sequences sharing
+a common region contained strictly within the interior of both
+sequences cannot be derived from the same transcript, as there can be
+neither a neighbor- or containment-relationship between them.  Thus we
+need only determine if one end of $s_1$ overlaps the opposite end of
+$s_2$ (recalling that \peace\/ has ensured they are on the same
+strand).  We need consider only the left-most and right-most windows
+of $s_1$ against each window of $s_2$, reducing the number of window
+comparisons from quadratic to linear in the size of $s_2$.
 
@@
 
-In Figure~\ref{fig:overlap} we illustrate the use of modified $d^2$,
-using it to take two sequences (a), find the window on $s_2$ most
-resembling the right-most window of $s1$ (red section of (b)),
-extending to the end of $s_2$ (green section of (c)) and performing a
-{\it global} alignment on the overlapping portion.  We then compute the
-{\it overlap distance} ($1 - s/l$, where $s$ is the alignment score and
-$l$ is the length of the overlap segment), a value that is inversely
-proportional to the probability that this is a legitimate overlap.
+In Figure~\ref{fig:overlap} we use the modified $d^2$ to on two
+sequences (a), finding the window on $s_2$ most resembling the right-most
+window of $s_1$ (red section of (b)), extending to the end of $s_2$
+(green section of (c)) and performing a {\it global} alignment on the
+overlapping portion.  We then compute the {\it overlap distance} ($1 -
+s/l$), where $s$ is the alignment score and $l$ is the length of the
+overlap segment).
 
@@
 
-In order to accommodate a range of potential fragment lengths, \peace\/
-and \east\/ make use of an {\it adaptive $d^2$} strategy to handle
-hybrid data sets.  The $d^2$ measure (and our modification) is
-flexible enough to handle a range of sequence sizes, requiring only a
-modification of the parameters (i.e. window size and threshold
-values).  However, it runs into problems when comparing sequences of
-significantly different sizes.  Comparing a 62 base Illumina read
-against a 1000 base Sanger read requires the use of parameters
-appropriate to the shorter fragment -- which are considerably less
-accurate when applied to the longer read.  Our solution is to
-partition fragments by sizes (e.g. into groups of small, medium and
-large fragments), and compute distances only between segments in the
-same or neighboring groups.  By skipping the comparisons between
-sequences of significantly different sizes we avoid the pitfalls of
-such comparisons, while still implicitly discovering those connections
+In order to accommodate a range of potential fragment lengths,
+\peace\/ and \east\/ make use of an {\it adaptive $d^2$} strategy to
+handle hybrid data sets.  The $d^2$ measure can handle a range of
+sequence sizes (given proper adjustment to the parameters), but runs
+into problems when comparing sequences of significantly different
+sizes (e.g. a 62 base Illumina read against a 1000 base Sanger).  Our
+solution is to partition fragments by sizes, and compute distances
+only between segments in the same or neighboring partitions.  For
+example, we will compare two smalls segments, or a small segment to a
+medium segment, but skip any comparisons between a small segment and a
+large segment.  By skipping the comparisons between sequences of
+significantly different sizes we avoid the pitfalls of such
+comparisons, while still implicitly discovering those connections
 through transitivity.
@@
 node against its two neighbors.  Those nodes with no identified
-left-overlap are checked against nodes further out in the tree, and
-any then remaining with no identified left neighbor as designated as a contig
-left end.
+left-overlap are checked against nodes further out in the tree; any
+remaining node with no identified left neighbor as designated as
+a contig left end.
 
@@
 with the overlap distance.  Using this graph we calculate a new
-minimum spanning tree, then starting from the root of the tree
-(necessarily a left-end of the contig) we transverse the tree and
-assign the first base of each sequence a position relative to its
-parent.  Such an assignment gives us an implicit ordering of the
-fragments.  By scanning through them in order and aligning each to
-the next with the Smith-Waterman alignment we end up with a multiple
-alignment of the fragments that allows us to derive a consensus
-sequence.  During this process quality score information can also be
-incorporated, reducing the score contribution of a specific base-pair
-if the involved bases are of low quality.  A straightforward extension
-would also allow the predictions of single nucleotide polymorphism by
-identifying those positions whose covering ESTs vary more in content
-than could reasonably be accounted for by the base-call rate.
+minimum spanning tree, then starting from the root (necessarily a
+left-end of the contig) we transverse the tree and assign the first
+base of each sequence a position relative to its parent.  This results
+in an implicit ordering of the fragments.  By scanning through them in
+order and aligning each to the next with the Smith-Waterman alignment
+we end up with a multiple alignment of the fragments that allows us to
+derive a consensus sequence.  During this process quality score
+information can also be incorporated, reducing the score contribution
+of a specific base-pair if the involved bases are of low quality.  A
+straightforward extension would also allow the predictions of single
+nucleotide polymorphism by identifying positions whose covering ESTs
+vary more in content than can be accounted for by the base-call error
+rate.
 
@@
 set of fragments representing multiple transcripts, \peace\/ is given
-the task of (ideally) producing a minimum spanning tree for each transcript,
+the task of (ideally) producing a minimum spanning tree for each gene,
 with edges representing a subset of the overlap information.  \east\/
-then assembles each cluster into one or more contigs.  As most tools
+then assembles each cluster into one or more contigs, with the goal of
+creating one contig for each transcript.  As most tools
 integrate the clustering and assembly process, for purposes of
@@
 (Sanger, 454 and Illumina).  Test data was produced using a set of
-zebra fish transcripts \cite{Hazelhurst08} and deriving simulated
+zebra fish transcripts [\cite{Hazelhurst08}] and deriving simulated
 simulated fragment sets using bot the \estsim\/ and \metasim\/ tools
@@
 Our primary measure of quality was the {\bf normalized a-score}, based
-on the {\it assembly score} (a-score) used in {\it Liang et al.}
-[\cite{Liang00}].  After reconstruction a (portion) of a transcript,
+on the {\it assembly score} used in {\it Liang et al.}
+[\cite{Liang00}].  After reconstruction of a transcript,
 the a-score of that contig is defined as $2c - 5m - 15i$ where $c$ is
@@
 were generated from a collection of 100 zebra fish transcripts
-obtained from {\it Hazelhurst et al.}  [\cite{Hazelhurst08}].  For
-Sanger sequences we used the \estsim\/ tool to generated simulated
-fragment sets, taking its default model of EST generation when not
-otherwise specified [\cite{Hazelhurst03}].  For 454 and Illumina sets
-we used the \metasim\/ tool, again using taking their default model to
-simulate the errors associated with those technologies
-[\cite{Richter08}].  Each tool models potential sequencing mistakes
-such as base-call errors, insertion and deletions, using parameters
-tailored to the target the technologies; see those publications for
-details.
+[\cite{Hazelhurst08}].  For Sanger sequences we used the \estsim\/
+tool to generated simulated fragment sets, taking its default model of
+EST generation when not otherwise specified [\cite{Hazelhurst03}].
+For 454 and Illumina sets we used the \metasim\/ tool, again using
+taking their default model to simulate the errors associated with
+those technologies [\cite{Richter08}].  Each tool models potential
+sequencing mistakes such as base-call errors, insertion and deletions,
+using parameters tailored to the target the technologies; see those
+publications for details.
 
-%\vspace{3mm}
+\vspace{3mm}
 
@@
 the {\it coverage} of the input: the average number of ESTs that cover
-a based.  Thus we investigate the a-score as a function of error rate
+a base.  Thus we investigate the a-score as a function of error rate
 at different coverage levels.  For each parameter combination we
@@
 base-call errors than the other tools.  Taking the mid-range coverage
-value of 30 as an example, at a 1\% error rate we see \east,
+value of 30 as an example at a 1\% error rate we see \east,
 \capthree, and \tgicl\/ all achieving an essentially perfect
@@
 perfect normalized a-score, showing a 2\% improvement over \tgicl\/
-and a 21\% improvement over \capthree.  At a 4\% error rate \east\/
+and a 21\% improvement over \capthree, while at a 4\% error rate \east\/
 still maintains its near-perfect score (showing a 54\% and 85\%
@@
 better job than \capthree\/ for error rates larger than $2\%$, while
-\tgicl\/ does a better job that \east\/ for error rates larger than
+\tgicl\/ does a better job than \east\/ for error rates larger than
 $4\%$ to $5\%$.
@@
 
+\vspace{3mm}
+
 \noindent {\bf Next Generation Sequences:} In
@@
 and Illumina outputs derived from the same zebra fish transcripts
-(with sequence lengths averaging 250 and 62 bp respectively).  For
+(with sequence lengths averaging 250 bp and 62 bp respectively).  For
 these tests we varied coverage levels but otherwise continued to use
-the default Metasim model [\cite{Richter08}].  On 454 data we see
-significant improvement in results quality in \east\/ for all levels
-of coverage.  On Illumina data we see that \east\/ is competitive with
-\velvet\/ at higher coverages, though is unable to match that tool
-when coverage is sparse.
+the default Metasim model [\cite{Richter08}].  On 454 data (average
+250 bp lengths) we see significant improvement in results quality in
+\east\/ for all levels of coverage.  On Illumina data (average 62 bp
+length) we see that \east\/ is competitive with \velvet\/ at higher
+coverage levels, though is unable to match that tool when coverage is
+sparse.
 
@@
 
+\vspace{3mm}
+
 \noindent {\bf Hybrid:} In Figure~\ref{hybridAscore} we look at the
@@
 the Sanger portion.  Here we observe \east\/ matching or outperforming
-\capthree\/ and \mira\/ at moderate to high coverages.  At a coverage
+\capthree\/ and \mira\/ at moderate to high coverage levels.  At a coverage
 level of 50, \east\/ maintains a normalized a-score hovering around
@@
 
-\noindent {\bf Tool Runtime:} In Figures~\ref{runtime.fixed} we see
-the average runtime for each of the experiments reflected in
+\vspace{3mm}
+
+\noindent {\bf Tool Runtime:} In Figures~\ref{runtime.fixed} are
+the average runtime for the experiments reflected in
 Figures~\ref{sangerAscore}-\ref{hybridAscore} for a coverage of 30
 (continuing to report the combined runtime of \peace+\east), while in
-Figure~\ref{runtime.coverage} we look at the average runtime for each
+Figure~\ref{runtime.coverage} are at the average runtimes for each
 tool when applied to simulated data sets generated from a single zebra
@@
 
-\east\/ is less competitive on shorter sequence.  In our simulations,
+\east\/ is less competitive on shorter sequences.  In our simulations,
 we find a reduced runtime improvement on the simulated 454 sequences
-(averaging 250 bp in length), and a significantly reduced runtime on
+(averaging 250 bp in length), and a significantly increased runtime on
 the simulated Illumina sequences (exactly 62 bp in length).  The
@@
 
-However,
-as technology advances, the ``short-read'' sequencers are producing
-increasingly longer reads: 454 sequencing technology can produce
-reads exceeding 400 bp, while Illumina sequences can exceed 100 bp
-[\cite{Eid09,Li10}].  The loss of runtime in our experiments on 454,
-Illumina and the hybrid sequences are due to the reduced length
-(specifically, the problems with applying the $d^2$ metric to shorter
-sequences -- as discussed below), hence are becoming less of an issue
-with this increase in short-read fragment length.
-
 %\vspace{3mm}
@@
 \centerline{\includegraphics[width=6in]{pics.d/runtime_fixedsize_sanger.pdf}}
-\caption{Average runtime per trial when when generating data for
+\caption{Average runtime per trial when generating data for
   Figures~\ref{sangerAscore}-\ref{hybridAscore}  when coverage = 30.
@@
 
-\textcolor{blue}{It is said that nothing is perfect, and this statement is certainly
-true when discussing assembly tools.  No one assembler can
-consistently out performs all other tools, and a user must choose the
-the tool that works best for their data.  Our experiments have shown
-that the \peast\/ assembly tool produces the best assemblies for Sanger
-data with a shorter runtime then competing tools, is the most robust
-to sequencing error, and preforms better then any competing tool when
+No one assembler can consistently out performs all other tools, and a
+user must choose the tool that works best for their data.  Our
+experiments have shown that the \peast\/ assembly tool produces the
+best assemblies for data containing sufficiently long transcripts,
+with a shorter runtime then competing tools, is the most robust to
+sequencing error, and preforms better then any competing tool when
 applied to data produced multiple sequencing technology platforms.
-\east\/ also demonstrates the viability of the MST-based
-approached to sequencing -- an approach we believe has considerable
-potential for improvement.}
+\east\/ also demonstrates the viability of the MST-based approached to
+sequencing -- an approach we believe has considerable potential for
+improvement.
 
-As observed, when applying \east, \capthree,
-and \tgicl\/ to Sanger sequenced transcripts with normal to high error
-rates, \east\/ consistently out performs all other tools in terms of
-the normalized a-score metric (Figures~\ref{sangerAscore},
-\ref{runtime.fixed}, and \ref{runtime.coverage}).  We also find that
-the number of contigs produced is almost exactly equal to the number
-of transcripts represented in the set, regardless of error rate, while
-the number of singleton ESTs remains low up to about a 5\% error rate
+As observed, when applying our competing tools to transcripts of long
+to moderate length (e.g. Sanger sequencing and 454 output) subjected
+to normal to high error rates, \east\/ consistently out performs all
+other tools in terms of the normalized a-score metric
+(Figures~\ref{sangerAscore}, \ref{runtime.fixed}, and
+\ref{runtime.coverage}).  We also find that the number of contigs
+produced is almost exactly equal to the number of transcripts
+represented in the set, regardless of error rate, while the number of
+singleton ESTs remains low up to about a 5\% error rate
 (Figures~\ref{contigsSanger}).  The high normalized a-score indicates
@@
 (\peace+)\east\/ running a minimum of two to three times as fast as
-the other tools.  The speedup over \capthree\/ (the faster tool)
-ranged from 1.9 to 6.0, depending on the coverage level.  In short,
-when used for Sanger Sequences \east\/ is a faster tool that produces
-the high quality reconstructions versus comparing tools and is
-significantly more robust to error than other tools.
+the other tools on the longer sequences.  In short, for sufficiently
+long sequences \east\/ is a faster tool that produces the high quality
+reconstructions versus comparing tools, and is significantly more
+robust to error than other tools.
 
@@
 
-\textcolor{blue}{For Next Generation Sequencing output we see a significant result
-quality increase over other tools in many (though not all) situations
-(Figures~\ref{nextgenAscore} and \ref{hybridAscore}) -- with \east\/
-out performing every tool on 454 data but uncompetitive with \velvet\/
-on Illumina data.  This quality does come at a price: the runtime of
-\east\/ takes a significant hit when dealing with shorter sequences.
-This is due to the dependence of \east\/ on the $d^2$ algorithm, which
-has poor performance on shorter sequences.  In practice this does not
-pose a large problem: the output sequence length of NGS technology
-is continually increasing as the technology advances, leaving
-our 62 bp sequence already outdated [\cite{Eid09,Li10}].   These
-continuing improvements in runtime, and the improvements in result
-quality, justify the \east algorithm as a good choice for assembly.}
+For Next Generation Sequencing output we see a significant quality
+increase over other tools for those technologies producing
+significantly long sequences -- with \east\/ out performing every tool
+on 454 data but uncompetitive with \velvet\/ on Illumina data.  This
+quality does come at a price: the runtime of \east\/ takes a
+significant hit when dealing with shorter sequences.  This is due to
+the dependence of \east\/ on $d^2$, which has poor
+performance on shorter sequences.  In practice this does not pose a
+large problem: the output sequence length of NGS technology is
+continually increasing as the technology advances, leaving our 62 bp
+sequence already outdated [\cite{Eid09,Li10}].  These continuing
+improvements in runtime, and the improvements in result quality,
+justify the \east algorithm as a good choice for assembly.
 
 Finally, we note, but have not yet explored, the fact that \east\/ is
-amenable to to paralilization and the possibility of identifying
-isoformes resulting from alternative splicing.  For the first: the
+amenable to parallelization and the possibility of identifying
+isoforms resulting from alternative splicing.  For the first: the
 \east\/ tool is trivially parallelizable in that once the \peace\/
 portion has completed clustering, \east\/ can address the clusters in
-parallel.  More sophisticated paralilization techniques are applicable
-at different phased of the east assembly (e.g. the depth-first search
+parallel.  More sophisticated parallelization techniques are applicable
+at different phases of the east assembly (e.g. the depth-first search
 of the MST, or the multiple independent applications of the $d^2$ metric).
 
-\section*{Acknowledgements}
-  Text for this section \ldots
+\section*{Acknowledgements} We would like to thank Jyothilakshmi
+Somasundaram and James Moller for there help on this project.
 
-\paragraph{Funding\textcolon} Text Text Text Text Text Text  Text Text.
+\paragraph{Funding\textcolon} NSF Grant 0953215, PhARMA Research
+Starts Grant in Bioinformatics
 

### 2011-06-21 peace/papers/EAST.bioinf/east.tex
Updating paper submission.

--- 2011-06-17 papers/EAST.bioinf/east.tex
@@
 \begin{figure}
-\centerline{\includegraphics[width=3in]{pics.d/overlap_alignment.pdf}}
+\centerline{\includegraphics[width=3in]{overlap_alignment.pdf}}
 \caption{(a) Two sequences potentially overlapping at the ends.  (b)
@@
 \begin{figure*}[htb]
-\centerline{\includegraphics[width=6in]{pics.d/ascore_sanger_norm.pdf}}
+\centerline{\includegraphics[width=6in]{ascore_sanger_norm.pdf}}
 \caption{Average normalized a-score as a function of base error rate for
@@
 \begin{figure*}[htb]
-\includegraphics[width=6in]{pics.d/numContigs_sanger.pdf}
-\includegraphics[width=6in]{pics.d/numSingle_sanger.pdf}
+\includegraphics[width=6in]{numContigs_sanger.pdf}
+\includegraphics[width=6in]{numSingle_sanger.pdf}
 \caption{Number of contigs (normalized to the number of transcripts)
@@
 \begin{figure*}[htb]
-\centerline{\includegraphics[width=5in]{pics.d/ascore_nxtgen.pdf}}
+\centerline{\includegraphics[width=5in]{ascore_nxtgen.pdf}}
 \caption{Comparative normalized a-scores for 454 and Illumina sequence
@@
 \begin{figure*}[htb]
-\centerline{\includegraphics[width=5in]{pics.d/ascore_hybrid.pdf}}
+\centerline{\includegraphics[width=5in]{ascore_hybrid.pdf}}
 \caption{Comparative normalized a-scores for a hybrid set consisting
@@
 \begin{figure*}[htb]
-\centerline{\includegraphics[width=6in]{pics.d/runtime_fixedsize_sanger.pdf}}
+\centerline{\includegraphics[width=6in]{runtime_fixedsize_sanger.pdf}}
 \caption{Average runtime per trial when generating data for
@@
 \begin{figure*}[htb]
-\centerline{\includegraphics[width=6in]{pics.d/runtime_bycoverage_all.pdf}}
+\centerline{\includegraphics[width=6in]{runtime_bycoverage_all.pdf}}
 \caption{The average runtime when applying each tool to a simulated

### 2025-06-29 papers/hicomb19/initial/abstract.tex
This commit moves some of the changes from the internal SVN to github. The key changes in this commit are:
1. Checking-in the source for the hicomb-19 paper along with changes to PEACE.
2. Checking in changes for global alignment based clustering (using NW score)
   and a new structural similarity (score generated by TM-align) based
   clustering solution.

\begin{abstract}
  \underline{Motivation}: Clustering genomic data, including those
  generated via high-throughput sequencing, is an important
  preliminary step for assembly and analysis. However, clustering a
  large number of sequences is time-consuming. \underline{Methods}: In
  this paper, we discuss algorithmic performance improvements to our
  existing clustering system called \peace\/ via the following two new
  approaches: \ding{182} using Approximate Spanning Tree (AST) that is
  computed \emph{much faster} than the currently used Minimum Spanning
  Tree (MST) approach, and \ding{183} a novel Prime Numbers based
  Heuristic (PNH) for generating features and comparing them to
  further reduce comparison overheads. \underline{Results}:
  Experiments conducted using a variety of data sets show that the
  proposed method significantly improves performance for datasets with
  large clusters with only minimal degradation in clustering
  quality. We also compare our methods against \q{wcd-kaboom}, a
  state-of-the-art clustering software.  Our experiments show that
  with AST and PNH underperform \q{wcd-kaboom} for datasets that have
  many small clusters.  However, they significantly outperform
  \q{wcd-kaboom} for datasets with large clusters by a conspicuous
  \mytilde 550\texttimes with comparable clustering quality.  The
  results indicate that the proposed methods hold considerable promise
  for accelerating clustering of genomic data with large clusters.
\end{abstract}

\begin{IEEEkeywords}
  Clustering, Minimum Spanning Tree, d2
\end{IEEEkeywords}

### 2025-06-29 papers/hicomb19/initial/acknowledgment.tex
This commit moves some of the changes from the internal SVN to github. The key changes in this commit are:
1. Checking-in the source for the hicomb-19 paper along with changes to PEACE.
2. Checking in changes for global alignment based clustering (using NW score)
   and a new structural similarity (score generated by TM-align) based
   clustering solution.

\section*{ACKNOWLEDGMENT}

We thank Dr. Jens Muller for his help to run experiments on the Red
Hawk compute cluster.  We thank Dr. Scott Hazelhurst for his help with
using \q{wcd} for our experiments.

\section*{AUTHOR CONTRIBUTIONS}

D. M. Rao: A first coauthor, implemented AST and PNH in \peace,
involved in running the experiments, analyzing the data, and writing
the paper.

S. Sreeskandarajan: A first coauthor, designed PNH, involved in
running experiments, analyzing the data, and writing the paper.

C. Liang: Helped with implementation, testing, and analysis of MST
approach in \peace.

### 2025-06-29 papers/hicomb19/initial/ast.tex
This commit moves some of the changes from the internal SVN to github. The key changes in this commit are:
1. Checking-in the source for the hicomb-19 paper along with changes to PEACE.
2. Checking in changes for global alignment based clustering (using NW score)
   and a new structural similarity (score generated by TM-align) based
   clustering solution.

\section{Approximate Spanning Tree}\label{sec:ast}

\peace\/ uses a Minimum Spanning Tree (MST) based method for
clustering~\cite{rao-10}.  In this method, each read in a dataset is
deemed as a node in the MST.  The MST is constructed using Prims's
algorithm, with d2 scores serving as the edge weights.  Recollect
that, as discussed in Section~\ref{sec:background}, d2 scores tend to
zero for highly similar reads -- \ie\/ similar reads will be adjacent
or nearby branches in the MST.  Once an MST is constructed, clustering
is accomplished by cutting the MST at branches that are significantly
distant, based on a given clustering-threshold, currently set to 1.0
in \peace.  Note that this 1.0 is \emph{not} a normalized value but a
putative d2 score (with 10.0 corresponding to infinity) that indicates
sufficiently different reads.

The most time-consuming operation is the construction of MST.  The
most dominant issue is that MST construction is an $O(n^{2})$
algorithm (where $n$ is the number of reads), which is prohibitive for
large datasets~\cite{hazelhurst-11}.  MST construction results in
$n^{2}$ calls to d2-score computation, which in itself, is expensive
and requires few milliseconds.  Here, heuristics (than run in
microseconds) help by eliminating over 80\% of d2 comparisons.  The
proposed Prime Number based Heuristic (PNH) aims to help in this area
and is discussed in Section~\ref{sec:pnh}.  Nevertheless, the
remaining 20\% of d2 comparisons take substantial runtime.

The Approximate Spanning Tree (AST) aims to address the $O(n^{2})$
time complexity by rapidly adding nodes to the tree, thereby reducing
the time constants.  Note the AST does not change the asymptotic time
complexity of $O(n^{2})$ but rather aims to rapidly reduce $n$.  An
overview of the AST construction method is summarized in
Algorithm~\ref{alg:ast}.  The algorithm begins with a list of reads
(\q{estList}) and an arbitrarily chosen \q{root} read.  Next, given a
\q{root}, sufficiently similar reads are obtained via call to the
\q{getReads} method.  This method has an $O(n)$ time complexity and
uses heuristics to reduce the number of calls to the heavyweight
\q{d2} computation method.  Only reads that are below the specified
\q{ASTthresh} threshold are returned.

\input{ast_alg.tex}

Unlike MST construction that adds only the most similar read, the AST
adds all of the similar reads, thereby rapidly reducing the number of
pending reads (\ie\/ $n$).  Reads are considered similar if their d2
distance is below a given AST-threshold value.  Similar to d2 scores,
the AST-threshold value is not normalized and this value cannot exceed
the clustering-threshold, which is currently set to 1.0.  The
AST-threshold value is a critical factor that influences the overall
performance.  The chart in Figure~\ref{fig:ast-thresh} illustrates the
impact of changing the threshold on a Haemagglutinin (HA) dataset with
\mytilde 65K reads.  Larger AST-thresholds allow more reads to be
added to the AST, thereby rapidly reducing $n$.  This results in fast
runtime but the resulting spanning tree drifts further away from the
MST.  In contrast, smaller threshold values result in a tree closer to
an MST -- \ie\/ AST-threshold of 0.0 produces an MST.  However,
smaller thresholds increase runtime with an $O(n^{2})$ time complexity
as illustrated by the inset chart in Figure~\ref{fig:ast-thresh}.

\begin{figure}
  \centerline{\includegraphics[width=\linewidth]{../charts/ha_1500_maxUse_convergence/maxUse_convergence.pdf}}
  \caption{Impact of AST threshold}\label{fig:ast-thresh}
\end{figure}

A comparison of MST and its corresponding AST is shown in
Figure~\ref{fig:mst-ast}.  The MST has been generated using 500 (for
readability) randomly chosen reads.  The AST has been generated with
an AST-threshold of 1.0 (\ie\/ \q{AST=1.0} in
Figure~\ref{fig:ast-thresh}).  The nodes in the tree have been
color-coded based on the clusters they were assigned.  The key feature
of interest is the structure of the two trees.  The MST has many
deeper branches because nodes are added one at a time.  On the other
hand, the AST has a much shallower structure with some nodes having a
very high degree.  The high degree nodes are the key in improving
performance as they enable many reads to be added to the AST thereby
rapidly reducing $n$ as illustrated by the curve for \q{AST=1.0} in
Figure~\ref{fig:ast-thresh}.

\begin{figure}[h]
  \begin{minipage}{\linewidth}
    \centerline{\includegraphics[width=\linewidth]{../charts/mst_vs_ast/ha_500_digraph.pdf}}
    \centerline{\footnotesize{(a) MST}}
  \end{minipage}

  \begin{minipage}{\linewidth}
    \centerline{\includegraphics[width=\linewidth]{../charts/mst_vs_ast/ha_500_ast_digraph.pdf}}
    \centerline{\footnotesize{(b) AST}}
  \end{minipage}
  \caption{Comparison of MST and AST}\label{fig:mst-ast}
\end{figure}

The impact of varying the AST-threshold on the clustering quality is
illustrated by the charts in Figure~\ref{fig:ast-qual}.  As
illustrated by the NMI curve (see Section~\ref{sec:metrics} for
details on NMI), the quality of clustering degrades as the
AST-threshold is increased.  NMI degrades as the number of clusters
increases with increase in AST-threshold.  This is because in the AST
potentially shorter links are not captured.  This results in
additional edges to be cut causing more clusters to be generated.
However, as indicated by the chart, the purity of clustering does not
degrade -- \ie\/ the AST does not incorrectly cluster disparate reads
together.  The observation suggests that a cluster-merging operation
can optionally be used to reduce the number of clusters, if needed.

\begin{figure}[h]
  \centerline{\includegraphics[width=\linewidth]{../charts/param_analysis/maxuse_quality.pdf}}
  \caption{Impact of AST-threshold}\label{fig:ast-qual}
\end{figure}

### 2025-06-29 papers/hicomb19/initial/background.tex
This commit moves some of the changes from the internal SVN to github. The key changes in this commit are:
1. Checking-in the source for the hicomb-19 paper along with changes to PEACE.
2. Checking in changes for global alignment based clustering (using NW score)
   and a new structural similarity (score generated by TM-align) based
   clustering solution.

\section{BACKGROUND: \peace\/ and d2}\label{sec:background}

\peace\/ (Parallel Environment for Assembly and Clustering of Gene
Expression) is a user-friendly nucleotide sequence clustering tool
which is designed to cluster transcript reads obtained by Sanger and
NGS sequencing technologies~\cite{rao-10,rao-18}. \peace\/ clusters
nucleotide sequences based on a Minimum Spanning Tree (MST) method as
summarized in Figure~\ref{fig:peace}.  An MST is generated (via Prim's
algorithm) using pairwise sequence comparison between the reads in
alignment-free manner using the d2 algorithm -- \ie\/ the MST edges
are weighted using d2 pseudo-metric.

\begin{figure}[h]
  \includegraphics[width=\linewidth]{figures/peace_overview.pdf}
  \caption{Overivew of clustering in \peace}\label{fig:peace}
\end{figure}

The d2 algorithm works by comparing the frequency of words (strings of
a fixed length) appearing in a limited region of each read. \peace\/
uses a default word size of 6 base pairs. Fragments overlapping by a
sufficient length will share neighborhoods of enough similarity to
ensure a small distance (close to zero) even in the presence of a
moderate number of base errors.  Specifically, \peace\/ uses a sliding
window of fixed size $r$ (defaults to 100 in \peace) to generate d2
score for two sequences $x$ and $y$ ($|x| \ge r$, $|y| \ge r$) via:

$$
d2(x, y) = min\big\{d2(u, v) : u \sqsubseteq x, v \sqsubseteq y, |u|= |v| = r\big\}
$$

\noindent where $u \sqsubseteq x$ denotes that $u$ is a substring of
$x$. Defined as such, d2 is, in a mathematical sense, a pseudo-metric
-- \ie\/ $d2(x, y) = 0$ does not imply $x = y$.  However, as $d2(x, y)
\rightarrow 0$, $x$ is more similar to $y$ and vice versa.
Sufficiently similar reads, based on a user-defined threshold, are
clustered together.

Computing d2 distance is expensive requiring a few milliseconds for ~1
kilobase sequence.  Consequently, to minimize unsuccessful d2
comparisons, \peace\/ uses \emph{u/v} and \emph{t/v} filtering
heuristics that run in microseconds.  As indicated in
Figure~\ref{fig:peace}, the heuristics minimize the number of d2
comparisons by eliminating comparisons of sufficiently differing
sequences -- \ie\/ sequences that would result in a very high d2
distance.  A more detailed discussion on the heuristics is available
in the literature~\cite{rao-10}.

Using MST and d2, \peace\/ has shown to cluster transcript reads with
high accuracy and sensitivity. It can be run both sequentially and in
parallel, though in this study we are not exploring parallelization.
The ability to run PEACE using a Graphical User Interface and a robust
set of default/out-of-the-box parameters make it a very easy to use
tool for biologists.

\peace\/ is a highly modularized framework developed in C++ and
heavily relies on the language's object-oriented features.  The
framework consists of loosely coupled core subsystems. The subsystems
are further modularized into loosely coupled components.  Though its
modular design the framework provides the user with a wide choice of
filters, heuristics and clustering methods.  In this study, we have
used these features to implement our proposed Approximate Spanning
Tree (AST) and Prime Number-based Filter (PNF).

%% This research utilized the above mentioned PEACE framework’s
%% modularity and extensibility to develop proof of concepts for the
%% speed improvements to PEACE using Approximate Minimum Tree (AMST)
%% clustering approach and a novel Prime Number-based Filter (PNF) with a
%% few localized subsystem based additions and modifications to the PEACE
%% framework.  The clustering subsystem was modified to include the AMST
%% method (Section III.B.) and the PNF (Section III.C. ). PNF was added
%% as a component to heuristics interface and use before the use of TV
%% and UV heuristics in the heuristic chain. The improved were evaluated
%% based of runtime and clustering quality using publically available
%% nucleotide sequence data (Section III.D. ).

### 2025-06-29 papers/hicomb19/initial/conclusions.tex
This commit moves some of the changes from the internal SVN to github. The key changes in this commit are:
1. Checking-in the source for the hicomb-19 paper along with changes to PEACE.
2. Checking in changes for global alignment based clustering (using NW score)
   and a new structural similarity (score generated by TM-align) based
   clustering solution.

\section{CONCLUSIONS}

Clustering is a vital processing step for analysis of genomic data.
Several software tools have been proposed to enable fast clustering.
Nevertheless, continued advancement in clustering methods is necessary
to keep pace with the ongoing exponential growth of genomic data.
This study proposed two novel enhancements to an existing clustering
software system called \peace.  The first enhancement was the use of
Approximate Spanning Tree (AST) which enables \emph{much faster}
clustering than the current Minimum Spanning Tree (MST) approach.  In
addition, a novel Prime Numbers based Heuristic (PNH) is also
proposed.  The paper discussed these two enhancements in detail and
presented an empirical analysis of their effectiveness.  The empirical
data was collected from experiments conducted using two different
types of data sets.  Moreover, the paper also presented comparison
against \q{wcd-kaboom}, a \emph{fast}, state-of-the-art clustering
software.

The outcomes of this study show that the AST approach effectively
increase the performance of \peace\/ only for datasets with large
clusters (viral genomic sequences).  In the case of Influenza data
set, a dramatic 550\texttimes\/ performance improvement was observed.
In general, the AST approach did not significantly compromise the
quality of the clustering.  In most cases, the AST method generates
additional clusters, but did not impact purity.  These additional
clusters can be merged, if necessary.

The outcomes also indicate that the Prime Number based Heuristic (PNH)
is a promising approach for further increasing the speed, but only for
clustering long genomic sequences.  A detailed investigation of the
effects of hyperparameter choices for PNH is underway to increase its
broader applicability.  Furthermore, we are planning enhancements to
reduce memory footprint by using multi-tier heaps~\cite{higiro-17}.
Importantly, we plan to explore parallel clustering capabilities of
\peace\/ in conjunction with AST and the PNH.

Rapid clustering of viral and bacterial genomes forming large clusters
will provide insight into the genetic diversity. Viral genotypes such
as Influenza A subtypes can be identified by clustering unidentified
genomic sequences with known genotypes. This type of cluster-based
identification approaches has the potential to detect novel viral
strains by separating them into new clusters which does not consist
any known types.  The improved performance of \peace\/ with AST, with
its default parameter settings and without the need for additional
external tools (such as \q{mkesa} for \q{wcd}), makes it an ideal tool
for use by biologists for clustering large datasets.

\newpage

### 2025-06-29 papers/hicomb19/initial/introduction.tex
This commit moves some of the changes from the internal SVN to github. The key changes in this commit are:
1. Checking-in the source for the hicomb-19 paper along with changes to PEACE.
2. Checking in changes for global alignment based clustering (using NW score)
   and a new structural similarity (score generated by TM-align) based
   clustering solution.

\section{INTRODUCTION}

Clustering nucleotide sequence data has many applications ranging from
identification of gene expressions~\cite{hazelhurst-11}, reducing the
runtime of sequencing reads to the
genomes~\cite{rao-10,hazelhurst-08}, and inferring phylogenetic
relationships of organisms~\cite{cybis-18,solovyov-09,wei-12}.
Clustering approaches are being widely used to group or ``bin'' the
reads belonging to a taxonomic group together in metagenomics
analyses~\cite{sedlar-17}. Clustering approaches can be used to reduce
sequencing errors by pre-clustering reads in the initial stages of
data analysis~\cite{kozich-13}.  Clustering whole large sets of
chromosome, genomes and other larger nucleotide sequences using
alignment-free methods enable rapid identification of phylogenetic
relationships between organisms. An example of such an application
would be the clustering of large sets of viral and other microbial
genomes~\cite{delgado-15,solovyov-09,wei-12}. Clustering such
microbial data enables understanding the microbial phylogenetic
distribution in large datasets and potentially identifying novel
microbial taxa~\cite{wei-12}.

\subsection{Current state-of-the-art}

Classical sequence clustering approaches were based on sequence
alignments using dynamic
programming~\cite{morrison-15,sievers-14,thompson-03}. The major
drawback of alignment-based methods is runtime and memory consumption
-- \ie\/ they tend to be very slow and consume a lot of
memory~\cite{zielezinski-17}. Hence they are not preferred when
clustering large datasets. Moreover, even though alignment-based
approaches are considered to be highly accurate, the quality of the
results could be questionable due to several deficiencies as discussed
by Zielezinski \etal\/ (they specifically discuss five
cases)~\cite{zielezinski-17}.

%%They discusses five cases where alignment-based sequence analysis can
%%be deficient, including the inability to incorporate certain complex
%%sequence features such as duplication,
%%inversions~\cite{zielezinski-17} and secondary structural
%%features~\cite{morrison-15}.

Alignment-free approaches are more practical alternatives to dynamic
programming-based methods.  Instead of dynamic programming,
alignment-free approaches rely on partial comparisons and
pseudo-metrics for clustering.  For example, our preliminary
clustering software called \peace\/ used a well established
alignment-free, pseudo-metric called \q{d2} to estimate similarity or
``distance'' between pairs of reads.  \peace\/ uses the \q{d2} score
(more details in Section~\ref{sec:background}) to build a Minimum
Spanning Tree (MST) and then cuts the MST to form clusters.  \q{d2} is
also used by \q{wcd}~\cite{hazelhurst-08}.  Recently, Hazelhurst
\etal~\cite{hazelhurst-11} further enhanced \q{wcd} with a suffix
array based approach called \q{kaboom}, to significantly improve the
performance of \q{wcd}.  \q{wcd-kaboom} has shown to outperform (both
in runtime and cluster-quality) several mainstream clustering
software, including ESTate, xsact, PaCE, CAP3, and
TGICL~\cite{hazelhurst-11}.  Consequently, we use \q{wcd-kaboom} as
the reference for performance comparisons.

\subsection{Motivation for this research}

The primary advantage of alignment-free methods is that they run
\emph{much faster} and often have a small memory footprint but at the
cost of some degradation in clustering quality.  Consequently,
alignment-free methods are widely used for clustering large
datasets~\cite{zielezinski-17,vinga-14}.  However, the ongoing
exponential growth in data volumes, continue to pose challenges in
accomplishing fast and effective clustering.  Even with just ~1\%
pairwise comparisons~\cite{hazelhurst-11}, \q{wcd-kaboom} takes
\mytilde 149 minutes (on an Intel
Xeon\textsuperscript{\textregistered}\/ Gold 6126 CPU @ 2.6 GHz) to
cluster \mytilde65K Haemagglutinin (HA) reads.  Such a long runtime of
\mytilde 3.5 hours for a relatively small data set, with a
state-of-the-art tool, highlights the ongoing challenges, thereby
motivating the need for high-performance clustering solutions.

\subsection{Proposed solution: An overview}

This research proposes a high-performance clustering solution to
effectively cluster large datasets of both short and long genomic
data.  The objective is to reduce runtime without significantly
degrading clustering-quality.  We propose to improve performance using
two approaches: \ding{182} first we focus on improving performance by
changing the Minimum Spanning Tree (MST) clustering approach used in
\peace\/ with an Approximate Spanning Tree (AST), and \ding{183}
adding a Prime-Number based Heuristic (PNH) based on the previously
proposed prime number based scoring system for inverted repeats
detection~\cite{sreeskandarajan-14}.  We discuss the conceptual
underpinnings and algorithmic details of these two approaches in
Section~\ref{sec:ast} and Section~\ref{sec:pnh}.

The paper presents results from experiments conducted using a broad
spectrum of data sets in Section~\ref{sec:results}.  We also compare
performance of the proposed method with \q{wcd-kaboom} to highlight
the effectiveness of AST and PNH.  For example, with AST, the runtime
for clustering the HA dataset is just 16 seconds, instead of 8,940
seconds for {\ttfamily wcd-kaboom}, a 550\texttimes\/ performance
improvement.  This is a very substantial performance improvement with
clustering quality slightly better than {\ttfamily wcd-kaboom}.  This
study establishes the potential of the AST approach in providing a
high-performance clustering tool well suited for clustering datasets
with large clusters.  The PNH is a promising approach for further
increasing the speed of large sets of viral genomic data.

% It is hypothesized that the prime number-based filter will be very
% efficient in speeding up the clustering of viral genomic sequences in
% comparison to the Khaboom filter of WCD due to nature of viral genomic
% data: longer sequences, high similarity and few clusters.

%% For example, a popular Next Generation Sequencing technology Illumina
%% HiSeq2500 system is able to produce 100 nucleotides long reads, while
%% Helicos’s Heliscope is able to produce very short reads of average
%% length 30 nucleotides. The RSII platform of Pacific bioscience has the
%% ability to produce large reads of length more than 10,000
%% bases. Alignment-free clustering methods can aid the analysis of
%% sequencing reads data.

%% With the recent availability of large volumes of pathogenic viral
%% genomic data clustering tools based on alignment-free methods can aid
%% in the rapid analysis of pathogenic viral diversity and potentially
%% aid in the identification of newly emerging viral pathogens. Influenza
%% genome sequences are publically available via databases such as
%% Influenza Research Database~\cite{zhang-17}, GISAID~\cite{bogner-06}
%% and Influenza Virus Resource~\cite{bao-08}. Several other pathogenic
%% viral genomic datasets such as genomic sequences of Ebola, Dengue, and
%% Zika can be obtained from Virus Variation Resource~\cite{brister-14}
%% and Virus Pathogen Database~\cite{pickett-12}. Alignment-free
%% clustering approach enable effective clustering these large public
%% datasets aiding in the understand of the evolution of viral pathogens.

%% Based on testing performed of the modified version PEACE (PEACE 2) on
%% short nucleotide reads and large viral genomic the modified version
%% PEACE (PEACE 2) was found to be faster than WCDExpress and PEACE,
%% without much compromise in the accuracy of the clustering. Overall,
%% the AMAST modification was able to increase the speed of PEACE greatly
%% for sequencing reads clustering and viral genome clustering, while the
%% PNF was able effectively increase the speed effectively only for the
%% viral genomic data when combined with the AMAST.

### 2025-06-29 papers/hicomb19/initial/metrics.tex
This commit moves some of the changes from the internal SVN to github. The key changes in this commit are:
1. Checking-in the source for the hicomb-19 paper along with changes to PEACE.
2. Checking in changes for global alignment based clustering (using NW score)
   and a new structural similarity (score generated by TM-align) based
   clustering solution.

\section{ASSESSMENT METRICS}\label{sec:metrics}

This study primarily focuses on reducing runtime for clustering.
However, the quality of clustering is important~\cite{hazelhurst-11}.
In this study we have used two widely used metrics for assessing the
quality of clustering, namely Normalized Mutual Information (NMI) and
Purity.  NMI is a popular measure of clustering quality based on
Information Theory~\cite{strehl-03}. Values of NMI range from 0 to
1. High NMI values indicate good clustering.  In this work, NMI is the
primary measure of clustering quality.

In addition to NMI, purity has also been used as a measure to further
validate clustering quality. Purity values range from 0 to 1.0.
Higher purity values indicate that a cluster does not have reads from
other clusters mixed into it.  The R packages aricode
(\url{https://cran.r-project.org/web/packages/aricode/index.html}) and
IntNMF~\cite{chalise-17} have been used to calculate NMI and purity,
respectively.

The NMI and purity values have been calculated using the clustering
output from \peace\/ as the reference.  Using the output from \peace\/
is motivated by two reasons.  First, \peace\/ has been tested by
multiple authors and has shown to produce good quality
clustering~\cite{rao-10,hazelhurst-11}.  Second, our objective is to
improve the performance of \peace\/ without impacting clustering
quality.  Consequently, we have used the output of \peace\/ as the
reference to assess clustering quality.

### 2025-06-29 papers/hicomb19/initial/primes.tex
This commit moves some of the changes from the internal SVN to github. The key changes in this commit are:
1. Checking-in the source for the hicomb-19 paper along with changes to PEACE.
2. Checking in changes for global alignment based clustering (using NW score)
   and a new structural similarity (score generated by TM-align) based
   clustering solution.

\section{Prime Number-based Heuristic (PNH)}\label{sec:pnh}

A prime-numbers based heuristic is proposed to further reduce
filtering overheads.  The PNH is based on the previously proposed
prime-numbers based scoring system for perfect inverted repeats
detection~\cite{sreeskandarajan-14}. In PNH, each nucleotide is
represented by a prime number (or its negative value) and a nucleotide
sequence is represented as a vector of prime numbers (corresponding to
nucleotides). The prime number score or feature for a nucleotide is
calculated by summing the corresponding vector of prime numbers. A
given number of features, say $p$, is extracted from a given
nucleotide sequence by breaking the sequences into $p$ non-overlapping
fragments as shown in Figure~\ref{fig:pnh} (here, $p$ = 3).  For each
of the $p$ sub-sequences, a prime number score is computed via:

$$
S_{l} = \sum_{i=1}^{L} p_{i}, \textrm{where} \quad p_{i} = \left\{
\begin{array}{ll}
  P_{A}  & \textrm{if } i = \q{A}\\
  -P_{A} & \textrm{if } i = \q{T}\\
  P_{G}  & \textrm{if } i = \q{G}\\
  -P_{G} & \textrm{if } i = \q{C}\\
  0 & otherwise
\end{array}\right.
$$

In the example shown in Figure~\ref{fig:pnh}, $P_{A}$ = 71 and $P_{G}$
= 113.  The prime values have been empirically chosen to provide a
balance between sensitivity and specificity.  The above approach
results in a $p$ dimensional, primes-based feature vector to be
generated for a given nucleotide sequence.

Two nucleotide sequences are compared based on the Euclidean distance
between the $p$ dimensional primes features generated for each
sequence.  A given pair of sequences are candidates for pairwise
comparison via d2 only if the Euclidean distance is below a given
threshold.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/PrimeNumberScoring.pdf}
  \caption{Example of feature extraction in the PNH}\label{fig:pnh}
\end{figure}

\subsection{Choice of primes for PNH}

The PNH provides a computationally light approach to extract features
from a nucleotide sequence.  Each of the $p$ dimensional features
embody nucleotide frequencies which is sensitive to biologically
meaningful features such as: direct repeats, inverted repeats, and
skew patterns~\cite{hemert-18} but is insensitive to inversions.  This
is the intuition behind the usage of PNH.

The choice of prime numbers to encode nucleotide is a balance.  Small
prime numbers (such as 3, 17, 23, etc.) were avoided in an attempt to
increase the uniqueness of the feature values~\cite{ye-14}.
Conversely, very large values are not used to ensure that a few
differences do not skew the features.  The default values of $P_{A}$ =
113 and $P_{G}$ = 71 were identified via preliminary analysis on HIV
sequences~\cite{delgado-15} and were also used as guidance for the
prime number choices, but was not highly optimized for any particular
organisms' genome.  Moreover, smaller values permit the use of faster,
primitive datatypes without concerns about numeric overflows (within
reason) during summations (see Figure~\ref{fig:pnh}).

\subsection{Impact of hyperparameters}

The Prime Numbers-based Heuristic (PNH) involves two key
hyperparameters, namely -- \ding{182} the number of features $p$ used
for comparing two reads, and \ding{183} $\tau_{prime}$: the threshold
Euclidean distance used to decide if two reads are sufficiently
similar.  Currently, the value of $\tau_{prime}$ for a given reference
sequence is computed as the average Euclidean distance from a given
reference strain minus the standard deviation.

The number of $p$ features has been set to a default value of 4 based
on experimental results summarized in
Figure~\ref{fig:pri-qual}. Overall, the number of features $p$ does
not have a significant impact on the clustering quality -- \ie\/ NMI
and purity do not vary much.  However, as the number of features $p$
is increased the number of clusters increase due to too many
false-negatives as the reads are broken into too short fragments.
Moreover, the memory usage also increases due to the overhead of
storing the large features.  Consequently, in this study, we have set
the value of $p$ to be 4.

\begin{figure}[h]
  \begin{minipage}{\linewidth}
    \centerline{\includegraphics[width=\linewidth]{../charts/param_analysis/primes_features_quality.pdf}}
    \centerline{(a) Impact on quality}
  \end{minipage}
  \begin{minipage}{\linewidth}
    \centerline{\includegraphics[width=\linewidth]{../charts/param_analysis/primes_features_runtime.pdf}}    
    \centerline{(b) Impact on runtime}
  \end{minipage}
  \caption{Impact of varying number of PNH features $p$}\label{fig:pri-qual}
\end{figure}

### 2025-06-29 papers/hicomb19/initial/relatedwork.tex
This commit moves some of the changes from the internal SVN to github. The key changes in this commit are:
1. Checking-in the source for the hicomb-19 paper along with changes to PEACE.
2. Checking in changes for global alignment based clustering (using NW score)
   and a new structural similarity (score generated by TM-align) based
   clustering solution.

\section{RELATED WORKS}

Many alignment-free sequence clustering tool have been proposed to
cluster large sets biological sequences. CD-Hit~\cite{li-06},
WCD~\cite{hazelhurst-08}, \peace~\cite{rao-10}, and
MeShClust~\cite{james-18} are some good examples for alignment-free
clustering tools. The popular clustering tool CD-Hit has the ability
to cluster large sets of nucleotide sequences, while WCD and \peace\/
are less fast but more accurate sequence clustering
alternatives. CD-Hit clusters sequences using a distance measure based
on the minimum number of shared short strings (words) between two
sequences.  WCD and \peace\/ use a similar word-based sequence
comparison approach but differ from CD-Hit in the distance measure
used to compare two sequences. Both WCD and \peace\/ utilize the d2
distance measure. \peace\/ differs from WCD by using a novel Minimum
Spanning Tree (MST) based clustering approach where the tree branches
are weighted by d2 distances.  It has been shown that the speed of WCD
can be improved by the use of a filter-based preprocessing
approach. The proposed suffix array-based filter for WCD named Kaboom
was shown to make WCD much faster than \peace~\cite{hazelhurst-11}.
Essentially, the Kaboom filter significantly reduces the number of d2
comparisons to be performed, to \textless\/ 1\% in many cases.  The
filter processing runtime is amortized by reducing the comparatively
slower d2 comparison.  However, the Kaboom filter requires generation
of suffix trees using a separate utility called \q{mkesa}.  MeShClust
is a recently proposed tool which claims to cluster nucleotide
sequences with high accuracy and considerable speed~\cite{james-18}.

The popular tools CD-Hit and WCD are primarily made to cluster
expression data such generated by sequencing methods as expressed
sequence tags and DNA sequence reads. They are shown to be well suited
for clustering next-generation sequencing data and other similar data
consisting of short length sequences. The utility of CD-Hit and WCD in
clustering whole chromosomes and genomes (which are larger in length
than typical sequencing reads) is questionable and has not been
extensively tested. Even though MeShClust claims to be efficient in
clustering short sequencing reads and relatively larger whole genomes,
it was not tested extensively for its whole chromosome/genome and long
read clustering ability.

Minimum Spanning Tree (MST) is a widely used method in clustering
nucleotide sequences and inferring phylogenetic relationships between
the sequences. Examples of such uses of MST include clustering
nucleotide sequences to aid in sequence domain
searching~\cite{guan-98}, performing intraspecific phylogenetic
analysis~\cite{rohl-99} and inferring phylogeographic
relationships~\cite{baric-03}.  A major issue associated with the MST
is the time complexity of the tree generation. The MST generation has
the time complexity of $O(n^{2})$ (where n is the number of reads to
be clustered)~\cite{caiming-13}.  Several approximate MST models
approaches have been proposed to reduce the time complexity using
heuristics approaches.  An example of an approximate MST approach is
minimizing the comparison overhead by building MSTs on K-Means
partition of the considered dataset~\cite{caiming-13}.  Another
example is the use of a novel centroid-based nearest neighbor rule for
the fast approximate MST generation (with the time complexity of $O(n
^{\frac{3}{2}} log\,n)$~\cite{jothi-18}. Employing such heuristic
approaches can aid in the development of efficient MST-based
nucleotide clustering and phylogenetic tools~\cite{rohl-99}.

### 2025-06-29 papers/hicomb19/initial/results.tex
This commit moves some of the changes from the internal SVN to github. The key changes in this commit are:
1. Checking-in the source for the hicomb-19 paper along with changes to PEACE.
2. Checking in changes for global alignment based clustering (using NW score)
   and a new structural similarity (score generated by TM-align) based
   clustering solution.

\section{EXPERIMENTS \& DISCUSSIONS}\label{sec:results}

The effectiveness of the proposed Approximate Spanning Tree (AST) and
the Prime Number based Heuristic (PNH) has been empirically assessed
using two different types of datasets.  The first dataset was
Expressed Sequence Tags (ESTs) from different species as summarized in
Table~\ref{tab:est-ds}.  This dataset consists of characteristic
short reads (about 300 -- 500 nucleotides) that have been used by
several investigators in the past.

\input{est_dataset.tex}

\q{S8} is a synthetic data set from James \etal~\cite{james-18}.  The
\q{C08} is a synthetically generated test dataset generated using
ESTsim (see ~\cite{hazelhurst-11} for details).  The \q{A076941}
dataset contains a subset of \emph{Arabidopsis} \emph{Thaliana} ESTs
downloaded from Genbank.  The \q{C08} and \q{A076941} datasets were
primary used because they have been used by Hazelhurst
\etal~\cite{hazelhurst-11} for assessment of \q{wcd-kaboom}.

The second dataset was viral genome fragments summarized in
Table~\ref{tab:virus-ds}.  This datasets consists of \emph{much}
longer reads (1000s of nucleotides) when compared to the EST dataset.
The \q{HA} dataset consists of Haemagglutinin (HA) segments from
multiple Influenza serotypes (H1 -- H19) and should ideally result in
20 clusters.  This dataset has been downloaded from Influenza Research
Database~\cite{bogner-06}.  The \q{Dengue} dataset consists of the
full genome of all 4 serotypes of dengue.  These long genomes are
analogous to the long reads (\mytilde 10,000 bases) produced by the
RSII platform from Pacific bioscience.  It has been downloaded from
Virus Variation Resource~\cite{brister-14}, specifically from the
Virus Pathogen Database~\cite{pickett-12}.  Ideally, this dataset
should generate 4 clusters but due to high homology between the
serotypes, most clustering software yield just 1 cluster.

\input{virus_dataset.tex}

\subsection{Experimental Platform}

The experiments reported in this paper have been conducted on a
contemporary compute cluster running PBS.  Each compute node on the
cluster consists of two (dual socket) Intel
Xeon\textsuperscript{\textregistered}\/ CPUs (Gold 6126 @ 2.60 GHz)
with hyperthreading disabled.  Each CPU has 14 cores and 19 MiB of
shared L3 cache.  The cluster runs CentOS 7.5 (Linux kernel version
3.10.  All of the software compiled using GNU Compiler Collection
(GCC) version 6.3.0 at \q{-O2} optimization level.  The timings and
memory usage as been recorded using \q{/usr/bin/time} utility.

\subsection{Design of Experiments}

Prior to conducting experiments, the reads in each of the datasets
were randomly shuffled to eliminate any potential patterns that may be
inherently present when the datasets were downloaded.  Each tool
configuration was run as an independent PBS job with 1 core and 12 GB
of memory reserved for it.  The experimental observations for the EST
data and Virus data are shown in Table~\ref{tab:est} and
Table~\ref{tab:virus}.  In these two tables, the row \q{wcd-kbm}
corresponds to \q{wcd} with \q{kaboom} filter configuration.  The run
time data for \q{wcd-kaboom} includes the time taken to generate
suffix arrays as discussed by Hazelhurst \etal~\cite{hazelhurst-11}.
The row with name \q{MST} shows results for base case \peace\/ run
using its default MST.  The rows labeled \q{AST=1} and \q{AST=0.8}
show results for AST with AST-thresholds set to 1.0 and 0.8
respectively.  The rows with \q{PNH} prefix show results from using
the heuristic along with the default $u/v$ and $t/v$ heuristics in
\peace.

\subsection{Discussions}

The results from various experiments conducted using the EST datasets
from Table~\ref{tab:est-ds} are summarized in Table~\ref{tab:est}.
Similarly, the results for the virus datasets from
Table~\ref{tab:virus-ds} are summarized in Table~\ref{tab:virus}.  The
NMI and purity values have been computed using the clustering
generated by \peace-MST as the reference as discussed in
Section~\ref{sec:metrics}.

For the EST datasets, \q{wcd-kaboom} consistently outperformed
\peace-MST as reported by Hazelhurst \etal~\cite{hazelhurst-11}.
However, as shown by the results, the proposed AST approach was
considerably faster than \peace-MST without much compromise in the
quality in most of the cases.  However, \q{wcd-kaboom} outperformed
the AST approach for two out of the three datasets.  In these
datasets, the performance gains of the AST approach was muted due to
the large number of small clusters -- \ie\/ only a few reads were
similar and hence reads could not be rapidly added to the AST (see
discussion in Section~\ref{sec:ast}).  In other words, Kaboom filter
was more successful at eliminating redundant calls to heuristics and
the heavyweight d2-score generation when compared to AST.  In the case
of the \q{A076941} dataset, a dominant portion of the runtime was
consumed for running the $u/v$ and $t/v$ heuristics.  Analysis of the
runtime behaviors suggests that a threshold based on number of
successful reads would be beneficial here -- \ie\/ once some $s$ reads
below the AST-threshold have been found, stop checking further reads.
We plan to pursue this optimization in the near future to assess its
efficacy.  Some degradation was observed in the \q{C08} case, with the
AST approach generating more clusters.  This results in some
degradation of NMI with respect to \peace-MST, but the NMI of 0.837 is
comparable to \q{wcd}'s 0.847.

\input{results_table_est.tex}

The most conspicuous performance improvement was observed in the HA
dataset, where the AST method finished clustering in \mytilde 16
seconds, while \q{wcd-kaboom} and \peace-MST took 149 mins and 930
mins respectively.  This corresponds to a remarkable \textgreater\/
550\texttimes\/ and \textgreater\/ 3400\texttimes\/ performance
improvements!  In the case of the Dengue dataset, the performance
improvement of AST over \q{wcd-kaboom} was \mytilde 90\texttimes.  The
results suggest that the AST method will yield good performance
improvements for datasets with large clusters.

\textbf{An unforeseen benefit}: Our experiments revealed an
interesting unforeseen benefit of the AST method for the HA dataset.
The dataset has 19 different serotypes and a few unclassified
reads. However, the serotypes are highly homologous and hence, the
more accurate MST-based approach results in just 8 clusters.  However,
with the AST approach results in detection of these serotypes as shown
in Table~\ref{tab:virus}.  For example, with a threshold of 0.95, the
AST method generates 20 clusters separating out the serotypes which
could be a desirable result.  This behavior will require further
investigation.

\input{results_table_virus.tex}

The Prime Numbers based Heuristic (PNH) was able to further boost
runtime performance in the \q{HA} dataset, but at the cost of
reduction in NMI.  For this dataset prime heuristics was able to
provide another 18\% improvement (see Table~\ref{tab:virus}.  It also
increased performance in the \q{Dengue} dataset by another 50\% with
NMI comparable to AST.  For the datasets having relatively shorter
sequences and a large number of clusters (\ie\/ \q{S8}, \q{C08}, and
\q{A076941}) the PNH was slower.  This is attributed to the additional
computational overhead of the PNH and this overhead is not effectively
amortized. Overall, prime heuristics provided clustering with lower
NMIs, except in the case of \q{S8} dataset.  These results suggested
that the PNH is effective in clustering viral genomic data possessing
highly similar, large nucleotide sequences. This type of clustering
has a several biological applications including analysis of microbial
genome variations, Influenza A subtype identification, and identifying
novel viral strains.

\subsection{Memory consumption}

A drawback inherited from \peace\/ is the increased memory usage for
clustering.  For example, in the case of the \q{C08} dataset, the
default MST-based clustering in \peace\/ consumes over 12 GB of RAM
while the AST approach consumes about 5 GB.  In contrast,
\q{wcd-kaboom} consumes only 26 MB (even when the raw dataset size if
48 MB).  The low memory footprint of \q{wcd-kaboom} is attributed to
the fact that it only maintains the suffix array in memory while
processing 1 read at a time.

The increased memory footprint of \peace\/ arises from 2 factors.
First, \peace\/ experiments were conducted by holding all the reads in
memory.  It does have an option to load reads on-demand, but this
option has not been used to minimize runtime.  The largest fraction of
memory usage arises from the caches maintained by \peace.  The cache
has been implemented using a standard binary heap.  It serves as a
priority queue to identify the next read to be added to the MST/AST.
This cache is pruned, but not aggressively, to reduce runtime.
Performing a periodic, deep pruning of this cache will reduce memory
footprint, but at the cost of some increase in runtime.  In addition,
alternative data structures such as 3-tier heap~\cite{higiro-17} can
be utilized to enable efficient pruning.  We are planning to explore
such enhancements to \peace\/ in the near future.

### 2025-06-29 papers/hicomb19/initial/virus_dataset.tex
This commit moves some of the changes from the internal SVN to github. The key changes in this commit are:
1. Checking-in the source for the hicomb-19 paper along with changes to PEACE.
2. Checking in changes for global alignment based clustering (using NW score)
   and a new structural similarity (score generated by TM-align) based
   clustering solution.

\begin{table}
  \caption{Summary of Virus dataset used in experiments}
  \begin{center}
    \begin{tabular}{|l|r|c|r|p{1in}|}
      \hline

      {\bf Name} & {\bf \#Reads} & {\bf Avg Length} & {\bf Size}  & {\bf Description}  \\
           &         & {\bf (mean$\pm$ SD)} & {\bf (MB)}  &   \\ \hline

      HA   & 65,069 & 1698.8$\pm$7 & 117 & Influenza Haemmaglutinin  (HA) \\ \hline

      Dengue  & 5,286 & 10,588$\pm$181 & 55 & Degune strains (4 serotypes) \\

      \hline
    \end{tabular}\label{tab:virus-ds}
  \end{center}
\end{table}

### 2025-06-29 papers/hicomb19/charts/ha_1500_maxUse_convergence/maxUse_job.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+#PBS -j oe
+#PBS -l "nodes=1:ppn=1"
+#PBS -l "mem=4gb"
+#PBS -N "maxUse_0.10"
+
+maxUse=0.10
+
+cd $PBS_O_WORKDIR
+
+/usr/bin/time -v ~/research/peace/src/peace --print-progress --fastaFile /shared/raodm/sutharzan/datasets/HA_1500_rnd.fa --clusterMaker mst --maxUse ${maxUse}
+

### 2025-06-29 papers/hicomb19/charts/mst_vs_ast/comp_degree.sh

@@ -1,0 +1,30 @@
+#!/bin/bash
+
+# Convenience script to print the average node degree from an MST
+if [ $# -ne 1 ]; then
+    echo "Specify the MST/AST file to be used"
+    exit 2
+fi
+
+# Setup a variable to make the script readable
+mst="$1"
+
+# First get the nodes that have child-nodes associated with them
+parentNodes=( `grep "^[0-9]" "${mst}" | cut -d"," -f1 | sort -nu` )
+echo "Number of parent nodes: ${#parentNodes[@]}"
+
+# Count the degree of each of the parent nodes to determine average
+# degree.
+degreeSum=0
+for id in ${parentNodes[@]}
+do
+    nodeDegree=`grep -c "^${id}," "${mst}"`
+    # echo "${id}, ${nodeDegree}"
+    degreeSum=$(( degreeSum + nodeDegree ))
+done
+
+# Compute average degree
+avgDegree=`echo "${degreeSum} / ${#parentNodes[@]}" | bc -l`
+printf "Average node degree: %.2f\n" ${avgDegree}
+
+# End of script

### 2025-06-29 papers/hicomb19/charts/mst_vs_ast/make_trees.sh

@@ -1,0 +1,34 @@
+#!/bin/bash
+
+# Script to generate plots of MST
+~/miami/research/peace/src/peace --fastaFiles HA_500.fa --output-mst-file HA_500.mst --output-cls-file HA_500.cls --clusterMaker mst --flat-print 2> /dev/null
+
+# Create a DOT output using mst and cls files
+echo "digraph HA_500_mst {" > "ha_500_digraph.dot"
+echo 'node[label="",width=0.6,shape=circle];' >>  "ha_500_digraph.dot"
+grep "^[0-9]" HA_500.mst | cut -d"," -f1,2 | sed 's/,/ -> /g' | sed 's/$/ [dir=none];/g' >> "ha_500_digraph.dot"
+# Add coloring information to the dot file.
+cut -d"," -f1,3 HA_500.cls | tr "," "=" | cut -d"=" -f2,4 | tr "=" " " | awk '{ print $2 " [style=filled,colorscheme=", ($1 < 13) ? "set312,color=" $1 : "rdylbu11,color=" $1-12, "];"}' >> "ha_500_digraph.dot"
+echo "}" >> "ha_500_digraph.dot"
+
+# Generate PDF
+neato -Tpdf -Gsize=6,6\! ha_500_digraph.dot -o ha_500_digraph.pdf
+
+#-----------------------------------------------------------
+# Script to generate plots of AST
+~/miami/research/peace/src/peace --fastaFiles HA_500.fa --output-mst-file HA_500_ast.mst --output-cls-file HA_500_ast.cls --clusterMaker mst --flat-print --maxUse 1.0 2> /dev/null
+
+# Create a DOT output using mst and cls files
+echo "digraph HA_500_ast {" > "ha_500_ast_digraph.dot"
+echo 'node[label="",width=0.3,shape=circle];' >>  "ha_500_ast_digraph.dot"
+grep "^[0-9]" HA_500_ast.mst | cut -d"," -f1,2 | sed 's/,/ -> /g' | sed 's/$/ [dir=none];/g' >> "ha_500_ast_digraph.dot"
+# Add coloring information to the dot file.
[…]

### 2025-06-29 papers/hicomb19/initial/embed_fonts.sh

@@ -1,0 +1,1 @@
+gs -dCompatibilityLevel=1.4 -dPDFSETTINGS=/screen -dCompressFonts=true -dSubsetFonts=true -dNOPAUSE -dBATCH -sDEVICE=pdfwrite -sOutputFile=main_fonts_embedded.pdf -c ".setpdfwrite <</NeverEmbed [ ]>> setdistillerparams" -f main.pdf

### 2025-06-29 src/clustering/AlignmentAnalyzer.cpp

@@ -1,0 +1,418 @@
+#ifndef ALIGNMENT_ANALYZER_CPP
+#define ALIGNMENT_ANALYZER_CPP
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include <vector>
+#include <algorithm>
+#include "ArgParser.h"
+#include "AlignmentAnalyzer.h"
+#include "ESTList.h"
+
+AlignmentAnalyzer::AlignmentAnalyzer()
+    : ESTAnalyzer("align") {
+}
+
+AlignmentAnalyzer::~AlignmentAnalyzer() {
+    // Nothing else to be done for now.
+}
+
+void
+AlignmentAnalyzer::addCommandLineArguments(ArgParser& argParser) {
+    // Let base class add common parameters.
+    ESTAnalyzer::addCommandLineArguments(argParser);
+    // Now setup arguments for this class
+    const ArgParser::ArgRecord LocalArgsList[] = {
+        {"--matchScore", "Score for matching bases in 2 reads",
+         &matchScore, ArgParser::INTEGER},
[…]

### 2025-06-29 src/clustering/AlignmentAnalyzer.h

@@ -1,0 +1,256 @@
+#ifndef ALIGNMENT_ANALYZER_H
+#define ALIGNMENT_ANALYZER_H
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include "ESTAnalyzer.h"
+#include <string>
+#include <vector>
+
+// Forward declaration to keep compiler happy
+class EST;
+class ResultLog;
+class ESTList;
+
+/** An analyzer to report similarity between 2 strains based on global
+   alignment between a pair of strains.
+
+   This is a more heavy-weight (that is, uses more CPU-time) analyzer
+   that uses alignment information between a pair of reads to estimate
+   similarity between strains.  This analyzer uses the NW score and
+   normalizes the score from 0 (identical) to 1.0 (completely
+   different) based on the aligned length.
+*/
+class AlignmentAnalyzer : public ESTAnalyzer {
+    friend class ESTAnalyzerFactory;
+public:
+    /** The destructor.
+        
+        The destructor frees up all any dynamic memory allocated by
+        this object for its operations.
+    */
+    virtual ~AlignmentAnalyzer();
+
+    /** Display valid command line arguments for this analyzer.
+
[…]

### 2025-06-29 src/clustering/TMAlignAnalyzer.cpp

@@ -1,0 +1,73 @@
+#ifndef TMALIGN_ANALYZER_CPP
+#define TMALIGN_ANALYZER_CPP
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include <vector>
+#include <algorithm>
+#include "ArgParser.h"
+#include "TMAlignAnalyzer.h"
+#include "ESTList.h"
+
+// Forward declaration for method defined in TMalign.cpp
+double getTMalignScore(const std::string& xname, const std::string& yname);
+
+TMAlignAnalyzer::TMAlignAnalyzer()
+    : ESTAnalyzer("align") {
+}
+
+TMAlignAnalyzer::~TMAlignAnalyzer() {
+    // Nothing else to be done for now.
+}
+
+void
+TMAlignAnalyzer::addCommandLineArguments(ArgParser& argParser) {
+    // Let base class add common parameters.
+    ESTAnalyzer::addCommandLineArguments(argParser);
+    // No arguments to be added.
+}
+
+float
+TMAlignAnalyzer::getMetric(const PDB* otherPDB) {
[…]

### 2025-06-29 src/clustering/TMAlignAnalyzer.h

@@ -1,0 +1,199 @@
+#ifndef TMALIGN_ANALYZER_H
+#define TMALIGN_ANALYZER_H
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include "ESTAnalyzer.h"
+#include <string>
+#include <vector>
+
+// Forward declaration to keep compiler happy
+class EST;
+class ResultLog;
+class ESTList;
+
+/** An alias to make this source contextually meaningful */
+using PDB = EST;
+
+/** An analyzer to report similarity between 2 strains based on
+    structural similarity.
+
+   This is a more heavy-weight (that is, uses more CPU-time) analyzer
+   that uses structural similarity score reported by TM-Align as the
+   distance between a pair of reads.  TM-Align is a 3rd-party program
+   that uses structural information from CIF files to report
+   structural similarity. TM-Align scores are in the range 0.0
+   (dissimilar) to 1.0 (identical).
+
+   \see TMAlign.cpp
+*/
+class TMAlignAnalyzer : public ESTAnalyzer {
+    friend class ESTAnalyzerFactory;
+public:
+    /** The destructor.
+        
[…]

### 2025-06-29 src/clustering/TMalign.cpp

@@ -1,0 +1,5417 @@
+/* TM-align: sequence-independent structure alignment of monomer proteins by
+ * TM-score superposition. Please report issues to zhanglab@zhanggroup.org
+ * 
+ * References to cite:
+ * Y Zhang, J Skolnick. Nucl Acids Res 33, 2302-9 (2005)
+ *
+ * DISCLAIMER:
+ *  Permission to use, copy, modify, and distribute the Software for any
+ *  purpose, with or without fee, is hereby granted, provided that the
+ *  notices on the head, the reference information, and this copyright
+ *  notice appear in all copies or substantial portions of the Software.
+ *  It is provided "as is" without express or implied warranty.
+ *
+ * ==========================
+ * How to install the program
+ * ==========================
+ * The following command compiles the program in your Linux computer:
+ *
+ *     g++ -static -O3 -ffast-math -lm -o TMalign TMalign.cpp
+ *
+ * The '-static' flag should be removed on Mac OS, which does not support
+ * building static executables.
+ *
+ * ======================
+ * How to use the program
+ * ======================
+ * You can run the program without argument to obtain the document.
+ * Briefly, you can compare two structures by:
+ *
+ *     ./TMalign structure1.pdb structure2.pdb
+ *
+ * ==============
+ * Update history
+ * ==============
+ * 2012/01/24: A C/C++ code of TM-align was constructed by Jianyi Yang
+ * 2016/05/21: Several updates of this program were made by Jianji Wu:
+ *            (1) fixed several compiling bugs
+ *            (2) made I/O of C/C++ version consistent with the Fortran version
+ *            (3) added outputs including full-atom and ligand structures
+ *            (4) added options of '-i', '-I' and '-m'
+ * 2016/05/25: Fixed a bug on PDB file reading
+ * 2018/06/04: Several updates were made by Chengxin Zhang, including
+ *            (1) Fixed bug in reading PDB files with negative residue index,
+ *            (2) Implemented the fTM-align algorithm (by the '-fast' option)
+ *                as described in R Dong, S Pan, Z Peng, Y Zhang, J Yang
+ *                (2018) Nucleic acids research. gky430.
+ *            (3) Included option to perform TM-align against a whole 
[…]

### 2025-06-29 src/input/PDBListFile.cpp

@@ -1,0 +1,114 @@
+#ifndef PDB_LIST_FILE_CPP
+#define PDB_LIST_FILE_CPP
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include <sstream>
+#include "PDBListFile.h"
+
+PDBListFile::PDBListFile(const std::string& fileName) : InputFile(fileName) {
+    // Open the FASTA file to be read and processed
+    pdbListFile.open(fileName.c_str());
+}
+
+PDBListFile::~PDBListFile() {
+    pdbListFile.close();
+}
+
+bool
+PDBListFile::hasNextEntry() {
+    return !pdbListFile.eof() && (pdbListFile.peek() != EOF);
+}
+
+size_t
+PDBListFile::getCurrPos() {
+    return pdbListFile.tellg();
+}
+
+bool
+PDBListFile::setCurrPos(const size_t position) {
+    pdbListFile.seekg(position, std::ios::beg);
+    if (!pdbListFile.fail()) {
+        // The seek operation was successful. If the stream was not in
+        // a good state earlier (because we hit EOF), we reset the EOF
+        // flag here to permit smooth reading after this seek.
[…]

### 2025-06-29 src/input/PDBListFile.h

@@ -1,0 +1,178 @@
+#ifndef PDB_LIST_FILE_H
+#define PDB_LIST_FILE_H
+
+//--------------------------------------------------------------------
+//
+// This file is part of PEACE.
+// 
+// PEACE is free software: you can redistribute it and/or modify it
+// under the terms of the GNU General Public License as published by
+// the Free Software Foundation, either version 3 of the License, or
+// (at your option) any later version.
+// 
+// PEACE is distributed in the hope that it will be useful, but
+// WITHOUT ANY WARRANTY; without even the implied warranty of
+// MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the GNU
+// General Public License for more details.
+// 
+// You should have received a copy of the GNU General Public License
+// along with PEACE.  If not, see <http://www.gnu.org/licenses/>.
+// 
+// Miami University makes no representations or warranties about the
+// suitability of the software, either express or implied, including
+// but not limited to the implied warranties of merchantability,
+// fitness for a particular purpose, or non-infringement.  Miami
+// University shall not be liable for any damages suffered by licensee
+// as a result of using, result of using, modifying or distributing
+// this software or its derivatives.
+//
+// By using or copying this Software, Licensee agrees to abide by the
+// intellectual property laws, and all other applicable laws of the
+// U.S., and the terms of GNU General Public License (version 3).
+//
+// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+//
+//---------------------------------------------------------------------
+
+#include <fstream>
+#include "InputFile.h"
+
+/** The class for reading a list of PDB file paths from a given text file.
+
+    This class serves as a concrete implementation to read a list of
+    PDB file paths from a given input text file.  The text file format
+    is kept very simple as follows:
+
+    <ul>
+
+    <li>All lines that begin with a '#' sign are ignored. So are empty
+    lines. </ul>
+
+    <li>All other lines are assumed to have a valid path to a PDB file
+    that will be used for analysis.</li>
+
+    </ul>
+
+    This class implements the various API methods stipulated by the
+    InputFile base class.
+
+    \note This class is not really meant to be directly instantiated.
+    Instead, the InputFileFactory class is designed to serve as the
+    primary entry point for instantiating a suitable, derived
+    InputFile object.
+*/
+class PDBListFile : public InputFile {
+public:    
+    /** Constructor to create an InputFile for reading data from a
+        FASTA text file.
+
[…]

### 2025-06-29 src/clustering/D2.cpp
Checking in changes that got missed in the previous commit -- most of the changes related to adding the following 2 types of clustering approaches:
1. A global alignment (using NW score) based clustering

2. A structural similarity (using TM-align score) based clustering

@@ -158,7 +158,10 @@ D2::runD2(const EST* estS2) {
     // reduced based on hints from the <i>t/v</i> heuristic.
     int sq1Start = 0, sq1End = s1WordTable.size() + wordSize - 1;
     int sq2Start = 0, sq2End = s2WordTable.size() + wordSize - 1;
-    
+
+    // Temorarily added code to track the window with minimum score.
+    int sq1MinPos = 0, sq2MinPos = 0;
+
     // Initialize the delta tables.
     memset(delta, 0, sizeof(int) * (1 << (wordSize * 2)));
     int score = 0;
@@ -197,7 +200,8 @@ D2::runD2(const EST* estS2) {
             // the word at s2Win is moving out as we move window
             // associated with EST #2 from left-to-right.
             updateWindow(s2WordTable[s2Win + numWordsInWindow],
-                         s2WordTable[s2Win], score, minScore);
+                         s2WordTable[s2Win], score, minScore,
+                         s1Win, s2Win, sq1MinPos, sq2MinPos);
         }
         // Break if the window on s1 cannot be shifted any further right
         if (s1Win >= LastWordInSq1) {
@@ -221,7 +225,8 @@ D2::runD2(const EST* estS2) {
             // the word at s2Win is moving out as we move window
             // associated with EST #2 from right-to-left.
             updateWindow(s2WordTable[s2Win - numWordsInWindow],
-                         s2WordTable[s2Win], score, minScore);
+                         s2WordTable[s2Win], score, minScore,
+                         s1Win, s2Win, sq1MinPos, sq2MinPos);
         }
         // Break if the window on s1 cannot be shifted any further right
         if ((s1Win+1) >= LastWordInSq1) {
@@ -233,7 +238,8 @@ D2::runD2(const EST* estS2) {
         // EST #1.
         updateWindow(s1WordTable[s1Win + 1],
                      s1WordTable[s1Win + 1 + numWordsInWindow],
-                     score, minScore);
+                     score, minScore, s1Win, FirstWordInSq2,
+                     sq1MinPos, sq2MinPos);
         // Break out of this loop if we have found a a potential match
         if (minScore <= threshold) {
             break;

### 2025-06-29 src/clustering/D2.h

@@ -431,6 +431,60 @@ private:
         minScore = std::min(score, minScore);
     }
 
+    /** Helper method to update the scores based on a sliding window
+        and track the index positions of the minimum scores.
+
+        This method is invoked from several different spots from the
+        runD2() method to update the d2 scores as the window slides
+        across the two sequences being analyzed. The hash values of
+        the words moving into and out of the window are used to update
+        the scores.
+
+        \param[in] wordIn The hash value of the word moving into the
+        window.
+
+        \param[in] wordOut The hash value of the word moving out of
+        the window.
+
+        \param[in,out] score The current running score for this
+        window. This value is updated using the delta array.
+
+        \param[in,out] minScore The current minimum score. This value
+        is updated after the score is updated to reflect the minimum
+        of score and minScore.
+
+        \param[in] s1Pos The current index position in the 1st strain.
+
+        \param[in] s2Pos The current index position in the 2nd strain.
+        
+        \param[out] s1MinPos The index position (in 1st strain) of the
+        minimum score thus far.  This value is updated only when a new
+        minimum score is encountered.
+
+        \param[out] s1MinPos The index position (in 2nd strain) of the
+        minimum score thus far.  This value is updated only when a new
+        minimum score is encountered.
+    */
+    inline void updateWindow(const int wordIn, const int wordOut,
+                             int& score, int& minScore, int s1Pos, int s2Pos,
+                             int& s1MinPos,
+                             int& s2MinPos) {
+        // Update score and delta for word moving in
+        score -= (delta[wordIn] << 1) - 1;
+        delta[wordIn]--;
+        // Update score and delta for word moving out
+        score += (delta[wordOut] << 1) + 1;
+        delta[wordOut]++;
+        // Track the minimum score.
+        if (score < minScore) {
+            minScore = score;
+            s1MinPos = s1Pos;
+            s2MinPos = s2Pos;
+            std::cout << "D2(" << s1MinPos << ", " << s2MinPos
+                      << ") = " << minScore << std::endl;
+        }
+    }
+    
     /** The core method that run's the D2 algorithm.
 
         This method performs the core D2 analysis. This method is

### 2025-06-29 src/clustering/ESTAnalyzerFactory.cpp

@@ -53,6 +53,9 @@
 #include "D2Cuda.h"
 #endif
 
+#include "AlignmentAnalyzer.h"
+#include "TMAlignAnalyzer.h"
+
 void
 ESTAnalyzerFactory::addCommandLineInfo(ArgParser& argParser) {
     // Create dummy command-line args to make display prettier and
@@ -80,6 +83,8 @@ ESTAnalyzerFactory::addCommandLineInfo(ArgParser& argParser) {
         {"", "d2cuda: CUDA-based D2 distance metric generation algorithm",
          NULL, ArgParser::INFO_MESSAGE},
 #endif
+        {"", "align: Alignment-based distance metric generation",
+         NULL, ArgParser::INFO_MESSAGE},        
         {"", "", NULL, ArgParser::INVALID}
     };
     argParser.addValidArguments(DummyArgs);
@@ -107,6 +112,10 @@ ESTAnalyzerFactory::create(const std::string& name) {
         // return new BatonAnalyzer();
     } else if (name == "primes") {
         return new PrimesESTAnalyzer();
+    } else if (name == "align") {
+        return new AlignmentAnalyzer();
+    } else if (name == "tmalign") {
+        return new TMAlignAnalyzer();
     }
 
 #ifdef HAVE_CUDA

### 2025-06-29 src/clustering/MSTClusterMaker.cpp

@@ -265,7 +265,7 @@ MSTClusterMaker::updateProgress(const int estsAnalyzed, const int totalESTcount,
                                 const int estAdded, const float metric,
                                 const int direction) {
     // Dump the cache for testing purposes
-    cache->print(std::cout);
+    // cache->print(std::cout);
     // Print progress to console if requested.
     if (printProgress) {
         char buffer[128];
@@ -461,10 +461,13 @@ MSTClusterMaker::initialize() {
         // Base class inititalization failed
         return false;
     }
-    if (!analyzer->initialize()) {
-        // Error occured during initialization. Bail out.
-        return false;
-    }
+    // The base class already initializes the analyzer. So we don't
+    // need to do it again here.
+    // if (!analyzer->initialize()) {
+    //     // Error occured during initialization. Bail out.
+    //     return false;
+    // }
+
     // Ensure the cache size is at least 1.
     if (cacheSize < 1) {
         std::cerr << "Invalid cache size (must be greater than zero)\n";

### 2025-06-29 src/clustering/NNMSTClusterMaker.cpp

@@ -30,7 +30,7 @@
 // intellectual property laws, and all other applicable laws of the
 // U.S., and the terms of GNU General Public License (version 3).
 //
-// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+// Authors:   Dhananjai M. Rao          raodm@miamiOH.edu
 //
 //---------------------------------------------------------------------
 
@@ -46,10 +46,24 @@ NNMSTClusterMaker::NNMSTClusterMaker(ESTAnalyzer *analyzer)
     // Initialize other instance variables & command-line arguments to
     // custom default values.
     localESTStartIndex = -1;
+    // The number of times the nn-pool was effective
+    nnPoolHits   = 0;
+    // The number of times the nn-pool was exhausted and had to be refreshed
+    refreshCount = 0;
 }
 
 NNMSTClusterMaker::~NNMSTClusterMaker() {
     // Nothing to be done for now.
+}
+
+void
+NNMSTClusterMaker::displayStats(std::ostream& os) {
+    // First let the base class print statistics
+    MSTClusterMaker::displayStats(os);
+    // Print nn-stats that we have collected
+    os << "Statistics from nn-mst:\n"
+       << "\tPool hits count   : " << nnPoolHits   << "\n"     
+       << "\tPool refresh count: " << refreshCount << std::endl;
 }
 
 bool
@@ -68,7 +82,6 @@ NNMSTClusterMaker::initialize() {
     return true;
 }
 
-
 bool
 NNMSTClusterMaker::computeSMEntry(const int estIdx,
                                   const int otherEstIdx, SMList& smList,
@@ -86,9 +99,8 @@ NNMSTClusterMaker::computeSMEntry(const int estIdx,
     if (chain != NULL) {
         chain->getHint(HeuristicChain::MST_RC, directionData);
     }
-    // Add only the first invalid entry. One is enough to do build
-    // a valid MST. There is no need for multiple vestigial
-    // entries.
+    // Add only the first invalid entry. One is enough to do build a
+    // valid MST. There is no need for multiple invalid entries.
     const float InvalidMetric = analyzer->getInvalidMetric();
     const bool isMetricOK = analyzer->compareMetrics(metric, InvalidMetric);
     if ((isMetricOK) || (addInvalidMetric)) {
@@ -116,8 +128,8 @@ NNMSTClusterMaker::computeSMList(const int estIdx, SMList& smList) {
     // startESTidx for size because the number of ESTs can be in
[…]

### 2025-06-29 src/clustering/NNMSTClusterMaker.h

@@ -30,7 +30,7 @@
 // intellectual property laws, and all other applicable laws of the
 // U.S., and the terms of GNU General Public License (version 3).
 //
-// Authors:   Dhananjai M. Rao          raodm@muohio.edu
+// Authors:   Dhananjai M. Rao          raodm@miamiOH.edu
 //
 //---------------------------------------------------------------------
 
@@ -88,9 +88,20 @@ public:
     */
     virtual ~NNMSTClusterMaker();
 
+    /** Method to display performance statistics.
+
+        This method overrides (but calls) implementation in the base
+        class to display statistics on nn-pool usage.  However, first
+        it calls the base class's method to print its statistics prior
+        to printing nn-pool information.
+
+        \param[out] os The output stream to which the statistics must
+        be written.        
+    */
+    virtual void displayStats(std::ostream& os) override;
+
 protected:
-    /** A method to handle initialization tasks for the
-		MSTClusterMaker.
+    /** A method to handle initialization tasks for the NNMSTClusterMaker.
 		
         This method is called after the ESTs have been loaded and the
         clustering is just about to commence. This method calls the
@@ -98,8 +109,8 @@ protected:
         initialization proceeds successfully (call to base class
         inititalizes the ESTAnalyzer and sets up estList pointer).  It
         then sets up the nearest-neighbor pool.
-
-		\return This method returns \c true if initialization was
+        
+        \return This method returns \c true if initialization was
         successfully completed.  On errors it returns \c false.
     */
     virtual bool initialize();
@@ -165,7 +176,7 @@ protected:
        given output stream.  This is primarily used for
        troubleshooting this class.
 
-       param[out] os The output stream to which the data is to be
+       \param[out] os The output stream to which the data is to be
        written.  Typically, this is just std::cout.
     */
     void printNNPool(std::ostream& os);
@@ -185,40 +196,53 @@ private:
     */
     NNMSTClusterMaker(ESTAnalyzer *analyzer);
 
-	/** Helper method to perform a single analysis and appropriately
-		add an entry to the smList.
-	   
-	*/
-	bool computeSMEntry(const int estIdx,
+    /** Helper method to perform a single analysis and appropriately
+        add an entry to the smList.
+	
+    */
+    bool computeSMEntry(const int estIdx,
                         const int otherEstIdx, SMList& smList,
-						const float InvalidMetric,
-						const bool addInvalidMetric,
+                        const float InvalidMetric,
+                        const bool addInvalidMetric,
                         const bool addToNNPool);
 
[…]

### 2025-06-29 src/clustering/NewUVHeuristic.cpp

@@ -381,14 +381,12 @@ NewUVHeuristic::checkHist(const EST* est1,
                   << (result ? "true" : "false")
                   << ", revCompMatch = "
                   << (revCompMatch ? "true" : "false") << std::endl;
-        if (result) {
-            std::cout << "View histogram (y/n)? " << std::flush;
-            std::string ans;
-            std::cin  >> ans;
-            if (ans == "y") {
-                printHist(est1, est2, revCompMatch, result, nMers);
-            }
-        }
+    }
+    std::cout << "View histogram (y/n)? " << std::flush;
+    std::string ans;
+    std::cin  >> ans;
+    if (ans == "y") {
+        printHist(est1, est2, revCompMatch, result, nMers);
     }
 }
 
@@ -397,7 +395,7 @@ NewUVHeuristic::printHists(const EST* est1,
                            const EST* est2,
                            const bool revCompMatch,
                            const bool result) {
-    checkHist(est1, est2, revCompMatch, result, 6);
+    checkHist(est1, est2, revCompMatch, result, 3);
 }
 
 #endif

### 2025-06-29 src/clustering/TransMSTClusterMaker.h

@@ -66,7 +66,7 @@ typedef std::vector<TransCacheEntry*> MetricCacheMap;
     <ol>
 
     <li>Given three ESTs, say e<sub>1</sub>, e<sub>2</sub>, and
-    e<sub>2</sub>, assume heavy weight metrics m<sub>12</sub> (between
+    e<sub>3</sub>, assume heavy weight metrics m<sub>12</sub> (between
     e<sub>1</sub> &amp; e<sub>2</sub> and m<sub>23</sub> (between
     e<sub>2</sub> &amp; e<sub>3</sub>) have been computed.</li>
 
@@ -205,7 +205,8 @@ protected:
         list of metrics computed.
     */
     virtual void populateCache(const int estIdx,
-							   SMList* UNREFERENCED_PARAMETER(metricList) = NULL);
+                               SMList* UNREFERENCED_PARAMETER(metricList)
+                               = NULL);
     
     /** The default constructor.
         
@@ -222,19 +223,19 @@ protected:
     TransMSTClusterMaker(ESTAnalyzer *analyzer);
 
     /** Helper method to remove invalid metrics from a given list of
-		values.
+        values.
 		
-		This is a helper method that is invoked from populateCache()
-		method to remove invalid/unwanted entries from a given list of
-		metrics. This method iterates over the entries in \c list and
-		removes all entires that have their metric set to a value that
-		is worse than badMetric value.
+        This is a helper method that is invoked from populateCache()
+        method to remove invalid/unwanted entries from a given list of
+        metrics. This method iterates over the entries in \c list and
+        removes all entires that have their metric set to a value that
+        is worse than badMetric value.
 		
-		\param[in] list The list of metrics that need to be
-		pruned.
+        \param[in] list The list of metrics that need to be
+        pruned.
 		
-		\param[out] entries The list of entries that are better than
-		badMetric.
+        \param[out] entries The list of entries that are better than
+        badMetric.
     */
     void pruneMetricEntries(const SMList& list, SMList& entries);
 

### 2025-06-29 src/input/InputFile.cpp

@@ -41,6 +41,7 @@ InputFile::InputFile(const std::string& srcFileName) :
     fileName(srcFileName) {
     randomizeNbases = false;
     maskBases       = false;
+    normalizeNTs    = true;
 }
 
 InputFile::~InputFile() {
@@ -66,8 +67,10 @@ InputFile::getNextEntry(const int estID) {
     }
     // Normalize the sequence by applying makign and ranodmization
     // options via a helper method.
-    sequence = EST::normalizeBases(sequence, maskBases,
-                                   randomizeNbases, (int) position);
+    if (normalizeNTs) {
+        sequence = EST::normalizeBases(sequence, maskBases,
+                                       randomizeNbases, (int) position);
+    }
     // OK, create and return the EST after updating our internal tables
     entryOffset[estID] = position;
     // Normalize the nucleotide sqeuence 
@@ -105,9 +108,10 @@ InputFile::repopulate(EST& est) {
 }
 
 void
-InputFile::setOptions(bool maskBases, bool randomizeNbases) {
+InputFile::setOptions(bool maskBases, bool randomizeNbases, bool normalizeNTs) {
     this->maskBases       = maskBases;
     this->randomizeNbases = randomizeNbases;
+    this->normalizeNTs    = normalizeNTs;
 }
 
 #endif

### 2025-06-29 src/input/InputFile.h

@@ -163,8 +163,12 @@ public:
         'N' characters are \b preserved.  However, if the flag is set
         to \c true then the 'N' characters are randomly converted to a
         regular base character from the list \c ATCG.
+
+        \param[in] normalizeNTs If this value is \c false then the
+        reads are left untouched and the previous 2 flags have no
+        effect.
     */
-    void setOptions(bool maskBases, bool randomizeNbases);
+    void setOptions(bool maskBases, bool randomizeNbases, bool normalizeNTs);
 
     /** Destructor.
 
@@ -279,6 +283,13 @@ private:
         setOptions() method.
     */
     bool randomizeNbases;
+
+    /**
+       If this flag is true then the reads are normalized to be only
+       nucleotide sequences consisting of ATCG.  If this flag is
+       false, then the this normalization is not done.
+     */
+    bool normalizeNTs;
     
     /** A map to maintain the offsets of each EST entry in this
         InputFile.

### 2025-06-29 src/input/InputFileFactory.cpp

@@ -37,6 +37,7 @@
 #include "InputFileFactory.h"
 #include "FASTAFile.h"
 #include "SFFReader.h"
+#include "PDBListFile.h"
 #include "SubSystem.h"
 #include "OnDemandESTList.h"
 #include "RuntimeContext.h"
@@ -46,6 +47,7 @@ InputFileFactory::InputFileFactory() : Component("InputFileFactory") {
     noMaskBases     = false;
     randomizeNbases = false;
     noOnDemand      = false;
+    noNormalizeNTs  = false;
 }
 
 InputFileFactory::~InputFileFactory() {
@@ -83,6 +85,8 @@ InputFileFactory::create(const std::string& fileName,
         inputFile = new FASTAFile(fileName);
     } else if (format == "sff") {
         inputFile = new SFFReader(fileName);
+    } else if (format == "pdblist") {
+        inputFile = new PDBListFile(fileName);
     }
 
     if ((inputFile == NULL) || (!inputFile->good())) {
@@ -96,7 +100,7 @@ InputFileFactory::create(const std::string& fileName,
         }
     } else {
         // File was opened successfully. Setup options
-        inputFile->setOptions(noMaskBases, randomizeNbases);
+        inputFile->setOptions(noMaskBases, randomizeNbases, !noNormalizeNTs);
     }
     // Return file pointer (if any) back to the caller
     return inputFile;
@@ -121,8 +125,12 @@ InputFileFactory::addCommandLineArguments(ArgParser& argParser) {
          &estFileNames, ArgParser::STRING_LIST},
         {"--estFiles", "Space separated data files (format auto-detected) to be processed",
          &estFileNames, ArgParser::STRING_LIST},
+        {"--pdbListFiles", "Set of PDB-list file path(s)",
+         &pdbListNames, ArgParser::STRING_LIST},        
         {"--no-ondemand", "Hold all reads in memory (faster; uses more RAM)",
          &noOnDemand, ArgParser::BOOLEAN},
+        {"--no-normalize", "Use sequences as it and don't normalize them to just nucleotide values (use for protiens or other inputs)",
+         &noNormalizeNTs, ArgParser::BOOLEAN},        
         {"", "", NULL, ArgParser::INVALID}
     };
     argParser.addValidArguments(Arguments);
@@ -137,6 +145,10 @@ InputFileFactory::initialize() {
     }
     // Load all SFF files
     if (!loadFiles(sffFileNames, "sff")) {
+        // Error occured during file loading.
[…]

### 2025-06-29 src/input/InputFileFactory.h

@@ -219,6 +219,19 @@ private:
     */
     ArgParser::StringList sffFileNames;
 
+    /** The set of PDB-List files specified by the user at the command
+        line.
+
+        This instance variable is passed to the command line argument
+        parser and is used to hold the list of pdb-list file names
+        (with optional absolute or relative path) as specified by the
+        user.  The list is set in the addCommandLineArguments() method
+        and is filled-in when PEACE performs the initial round of
+        command line argument processing.  The list is processed in
+        the initialize() method.
+    */    
+    ArgParser::StringList pdbListNames;
+    
     /** The list of data files specified by the user at the command
         line.
 
@@ -267,6 +280,17 @@ private:
     */
     bool randomizeNbases;
 
+    /** Flag to indicate if all nucleotides should be normalized to be
+        ATCG or left untouched.
+
+        If this flag is set to false (default is true), then the reads
+        are left unaltered. Consequently, the randomizeNbases and
+        noMaskBases flags have no effect.  This feature is useful to
+        process protiens or other inputs other than nucleotide
+        sequences as inputs.
+    */    
+    bool noNormalizeNTs;
+    
     /** Flag to indicate if ESTs should be loaded and retained in
         memory instead of loading on-demand.
         

### 2025-06-29 src/shared/ESTAnalyzer.h

@@ -41,6 +41,18 @@
 class HeuristicChain;
 class ESTList;
 class EST;
+class ESTAnalyzer;
+
+/**
+   This is a synonym for an ESTAnalyzer. Earlier, PEACE was primarily
+   used for EST analysis. However, over time this functionality/scope
+   has steadily increased to include other analysis such as
+   clustering, assembly, and clustering using completely different
+   metrics/measures. Accordingly, the source code is continuing to
+   evolve and change with transitions in meanings of the different
+   APIs (includes classes and methods).
+*/
+using Analyzer = ESTAnalyzer;
 
 /** The base class of all EST analyzers.
 

### 2025-06-29 src/shared/MSTCluster.cpp

@@ -99,6 +99,8 @@ MSTCluster::makeClusters(const NodeList& nodeList, const ESTAnalyzer* analyzer,
     // Check if the threshold is less than 0 (treat all negatives as -1)
     if (threshold < 0) {
         threshold = calculateThreshold(nodeList);
+        std::cout << "Clustering threshold computed to be: "
+                  << threshold << std::endl;
     }
     // Create a hash map to track cluster for a given node
     ClusterMap clusterMap;
@@ -126,13 +128,17 @@ MSTCluster::makeClusters(const NodeList& nodeList, const ESTAnalyzer* analyzer,
             // need to traverse backwards in the node list from the
             // current point.
             NodeList::const_iterator parent = node;
-            while (parent != nodeList.begin()) {
+            while (true) {
                 if (parent->estIdx == node->parentIdx) {
                     subCluster->add(*parent);
                     break;
                 }
                 // Check previous node next.
-                parent--;
+                if (parent != nodeList.begin()) {
+                    parent--;
+                } else {
+                    break;  // print a warning about not finding parent?
+                }
             }
         }
         ASSERT ( subCluster != NULL );

### 2025-06-29 tools/ExtractReads.cpp

@@ -44,8 +44,10 @@
 int
 ExtractReads::parseArgs(int argc, char *argv[], std::string& inputPath,
                         std::string& outputPath,
-                        ArgParser::StringList& idxList, int& rndNum) {
+                        ArgParser::StringList& idxList, int& rndNum,
+                        ArgParser::StringList& substrList) {
     bool showOptions = false;
+    subSeqStart = subSeqEnd = -1;
     // Create the list of valid arguments to be used by the arg_parser.
     ArgParser::ArgRecord arg_list[] = {
         {"--input", "The input FASTA file to be randomly shuffled",
@@ -53,18 +55,24 @@ ExtractReads::parseArgs(int argc, char *argv[], std::string& inputPath,
         {"--output", "Output FASTA file to write randomly shuffled reads",
          &outputPath, ArgParser::STRING},
         {"--indexs", "List index values (zero-based) for reads to be extracted",
-         &idxList, ArgParser::STRING_LIST},        
+         &idxList, ArgParser::STRING_LIST},
+        {"--substr", "List of sub-strings for which reads are to be extracted", 
+         &substrList, ArgParser::STRING_LIST},
         {"--options", "Lists options for this tool",
          &showOptions, ArgParser::BOOLEAN},
         {"--random", "Randomly select specified number of reads to extract",
          &rndNum, ArgParser::INTEGER},
+        {"--sub-seq-beg", "Beginning index position for subsequence",
+         &subSeqStart, ArgParser::INTEGER},
+        {"--sub-seq-end", "Ending index position for subsequence",
+         &subSeqEnd, ArgParser::INTEGER},
         {"", "", NULL, ArgParser::INVALID}
     };
     // Get the argument parser to parse and consume the global
     // options.  Based on the options supplied, various variables will
     // be set to appropriate values.
     ArgParser ap;
-    Tool::addCmdLineArgs("ExtractReads", ap);
+    Tool::addCmdLineArgs("extract", ap);
     ap.addValidArguments(arg_list);
     ap.parseArguments(argc, argv, false);
     if (showOptions) {
@@ -79,24 +87,93 @@ ExtractReads::parseArgs(int argc, char *argv[], std::string& inputPath,
                "Use --input option\n");
     CHECK_ARGS(tool, outputPath.empty(), "Output file not specified.\n" \
                "Use --output option\n");
-    CHECK_ARGS(tool, idxList.empty() && rndNum == -1,
[…]

### 2025-06-29 tools/ExtractReads.h

@@ -101,11 +101,70 @@ protected:
 
         \param[out] rndNum An optional random subset of reads to be
         extracted.
+
+        \param[out] substrList An optional list of sub-strings to be
+        used to extract entries.
+
+        \return This method returns zero on success.
     */
-    static int parseArgs(int argc, char *argv[], std::string& inputPath,
-                         std::string& outputPath,
-                         ArgParser::StringList& idxList, int& rndNum);
-    
+    int parseArgs(int argc, char *argv[], std::string& inputPath,
+                  std::string& outputPath,
+                  ArgParser::StringList& idxList, int& rndNum,
+                  ArgParser::StringList& substrList);
+
+    /**
+       Extracts a given list of random reads from the given input list.
+
+       \param[in] rndNums The number of random reads to be extracted.
+
+       \param[in] inputList The list of input EST reads.
+
+       \param[out] output The output to where the data is to be written.
+     */
+    void extractRandomReads(const int rndNums, ESTList& inputList,
+                            std::ostream& output) const;
+
+    /**
+       Extracts a given list of random reads from the given input list.
+
+       \param[in] idxList The list of indexes in the inputList whose
+       entries are to be extracted and written to output.
+
+       \param[in] inputList The list of input EST reads.
+
+       \param[out] output The output to where the data is to be written.
+    */
+    void extractIndex(const ArgParser::StringList& idxList,
+                      ESTList& inputList, std::ostream& output) const;
+
+    /**
+       Extracts reads that contains any of the given set of substrings
+       in their information.
+
+       \param[in] substrList The list of sub-strings that are searched
+       in each EST's info.  If any of the sub-strings are found in the
+       info, then those ESTs are written to the output.
+
+       \param[in] inputList The list of input EST reads.
+
+       \param[out] output The output to where the data is to be written.
+    */
+    void extractSubstr(const ArgParser::StringList& substrList,
+                       ESTList& inputList, std::ostream& output) const;
+
+    /** Helper method to write out EST entry while honoring
+        sub-fragment settings (if any).  This method uses the range in
+        subSeqStart and subSeqEnd instance variables (set via
+        command-ine arguments --sub-seq-beg and --sub-seq-end) to
+        extract and write sub-fragements.
+
+        \param[in] est The read to be written.
+
+        \param[out] out The output stream to where the read is to be
+        written.
+    */
[…]

### 2025-06-29 tools/RandomizeReads.cpp

@@ -66,7 +66,7 @@ RandomizeReads::parseArgs(int argc, char *argv[], std::string& inputPath,
     // options.  Based on the options supplied, various variables will
     // be set to appropriate values.
     ArgParser ap;
-    Tool::addCmdLineArgs("ShowAlignment", ap);
+    Tool::addCmdLineArgs("randomize", ap);
     ap.addValidArguments(arg_list);
     ap.parseArguments(argc, argv, false);
     if (showOptions) {
@@ -107,8 +107,8 @@ RandomizeReads::main(int argc, char *argv[]) {
     for (int i = 0; (i < estList.size()); i++) {
         if (filterNs && estList[i]->getSequenceString().find('N')
             != std::string::npos) {
-            // std::cout << estList[i]->getInfo() << ": Found N at: "
-            //           << estList[i]->getSequenceString().find('N') << '\n';
+            std::cout << estList[i]->getInfo() << ": Found N at: "
+                      << estList[i]->getSequenceString().find('N') << '\n';
             continue;  // This read has 'N'. Ignore it.
         }
         estVector.push_back(estList[i]);

### 2025-06-29 tools/ShowMST.cpp

@@ -56,7 +56,7 @@ ShowMST::main(int argc, char *argv[]) {
          &estFileName, ArgParser::STRING},
         {"--output", "File to which the bracketed MST must be written",
          &outFileName, ArgParser::STRING},
-        {"--clstrFile", "Optional clusters output from PEACE for color coding",
+        {"--clstrFile", "Optional flat clusters output from PEACE for color coding",
          &clstrFileName, ArgParser::STRING},        
         {"--xScale", "X-Scale for emphasizing line lengths",
          &xScale, ArgParser::DOUBLE},
@@ -73,7 +73,7 @@ ShowMST::main(int argc, char *argv[]) {
     // options.  Based on the options supplied, various variables will
     // be set to appropriate values.
     ArgParser ap;
-    Tool::addCmdLineArgs("ShowAlignment", ap);
+    Tool::addCmdLineArgs("ShowMST", ap);
     ap.addValidArguments(arg_list);
     ap.parseArguments(argc, argv, false);
     if (showOptions) {

### 2025-06-29 tools/Tool.cpp

@@ -54,6 +54,48 @@ Tool::loadFastaFile(const std::string& fileName) {
     return peace.loadFile(fileName, "", 0);
 }
 
+// A static helper method to split strings
+std::vector<std::string>
+Tool::splitString(const std::string& str, const std::string& delimiters,
+                  const bool trimTokens) {
+    const std::string s = (trimTokens ? trim(str) : str);
+    std::vector<std::string> tokens;
+    size_t start = 0;
+    size_t end = s.find_first_of(delimiters, start);
+    
+    while (end != std::string::npos) {
+        const std::string token = s.substr(start, end - start);
+        tokens.push_back(trimTokens ? trim(token) : token);
+        start = end + 1;
+        end = s.find_first_of(delimiters, start);
+    }
+    // Add the last token
+    const std::string token = s.substr(start);
+    tokens.push_back(trimTokens ? trim(token) : token);    
+    
+    return tokens;
+}
+
+std::string
+Tool::trim(const std::string& s, const char *whitespace) {
+    // Find the first non-whitespace character
+    size_t first_non_whitespace = s.find_first_not_of(whitespace);
+
+    // If the string contains only whitespace, return an empty string
+    if (first_non_whitespace == std::string::npos) {
+        return "";
+    }
+
+    // Find the last non-whitespace character
+    size_t last_non_whitespace = s.find_last_not_of(whitespace);
+
+    // Calculate the length of the trimmed substring
+    size_t length = last_non_whitespace - first_non_whitespace + 1;
+
+    // Return the trimmed substring
+    return s.substr(first_non_whitespace, length);
+}
+
 bool
 Tool::loadClusterInfo(const std::string& fileName) {
     // Try and open the cluster file
@@ -64,7 +106,6 @@ Tool::loadClusterInfo(const std::string& fileName) {
         return false;
     }
     // Now repeatedly read a line from the file and process it.
-    int clusterNum = -1; // The last seen cluster number
     while (!clstrFile.eof() && clstrFile.good()) {
         // Read a line from the cluster file.
         std::string line;
@@ -74,12 +115,19 @@ Tool::loadClusterInfo(const std::string& fileName) {
             // nothing to process on this line.
             continue;
         }
-        if (line.find("Cluster #") == 0) {
-            // This is a new cluster entry. Update cluster number.
-            sscanf(line.c_str(), "Cluster #%d", &clusterNum);
+        // This is a EST entry. Update color map, from flat format
[…]

### 2025-06-29 tools/Tool.h

@@ -298,6 +298,39 @@ public:
         \return A reference to the EST analyzer setup in PEACE.
     */
     ESTAnalyzer& getAnalyzer();
+
+    /**
+       Helper method to split a string based on one-or-more delimiter
+       characters.
+
+       \param[in] str The string to be split.
+
+       \param[in] delimiters The list of delimiter characters to be
+       used. Note that each delimiter can only be a character.
+
+       \param[in] trimTokens If this flag is true, then each token is
+       trimmed by calling the trim method in this class.
+       
+       \return The list of tokens from the string separated by the
+       specified list of delimiters.
+    */
+    static std::vector<std::string>
+    splitString(const std::string& str, const std::string& delimiters,
+                const bool trimTokens = false);
+
+    /**
+       Convenience method to strip leading and trailing whitespaces
+       from a string.
+
+       \param[in] str The string to be trimmed.
+
+       \param[in] whitespace The list of characters to be deemed as
+       whitespace characters.
+
+       \return Returns the trimmed version of the string.
+     */
+    static std::string trim(const std::string& str,
+                            const char* whitespace = " \t\n\r\f\v");
     
 protected:
     /** The constructor.

### 2025-06-29 tools/main.cpp

@@ -37,7 +37,7 @@
 const char* ToolInfo[] = {
     "ShowAlignment","Show graphical view of EST alignments",
     "DrawMatrix",   "Draw graphical view of D2  matrix of scores",
-    "ShowMST",      "Translate MST to a bracketed form for rendering",
+    "ShowMST",      "Draw xfig diagram of MST generated by peace",
     "Align",        "Simple alignment generation tool using MST alignment data",
     "Matcher",      "Score contigs against source DNA/Transcripts",
     "primes",       "Print n-dimensional features based on prime numbers",
